### 2016-12-15 els2017-radiance/els2017-radiance.tex
Start working on the ELS'17 paper on Radiance.

\documentclass{sig-alternate-05-2015}
\begin{document}

\setcopyright{rightsretained}
\doi{}
\isbn{}
\conferenceinfo{ELS'17}{April 3--4, 2017, Brussel, Belgium}

\begin{CCSXML}
  <ccs2012>
  <concept>
  <concept_id>10011007.10011006.10011066</concept_id>
  <concept_desc>Software and its engineering~Development frameworks and environments</concept_desc>
  <concept_significance>500</concept_significance>
  </concept>
  <concept>
  <concept_id>10002951.10003260.10003282</concept_id>
  <concept_desc>Information systems~Web applications</concept_desc>
  <concept_significance>300</concept_significance>
  </concept>
  </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Development frameworks and environments}
\ccsdesc[300]{Information systems~Web applications}

\title{Radiance - A Web Application Environment}

\numberofauthors{1}
\author{
\alignauthor
Nicolas Hafner\\
       \affaddr{ETH Zürich}\\
       \affaddr{Nürenbergstr. 17B}\\
       \affaddr{8037 Zürich, Switzerland}\\
       \email{shinmera@tymoon.eu}
}
\date{15 November 2016}

\maketitle

\begin{abstract}
  Radiance is a set of libraries that provide an environment for web applications. Unlike traditional web frameworks that focus on a set of tools to support the construction and maintenance of a single application, Radiance attempts to allow you to run as many differing applications as desired in the same environment. This has many advantages, as frequently there are common resources in applications that should be shared between them, but cannot because the applications are separated and developed without regard for sharing. In order to fix this, Radiance provides an extensible interface mechanism to standardise the interaction between users of common resources. To facilitate the separation and encapsulation of the shared URL address space between applications, Radiance also provides a flexible routing mechanism. This mechanism is powerful enough to ensure that any application written with Radiance can be deployed on any server setup, no matter how complicated the constraints, without needing to change the source code of the application.
\end{abstract}

\printccsdesc

\keywords{Common Lisp, web framework, web development, encapsulation, interfacing}

\newpage
\section{Introduction}

\section{Interfaces}

\section{Routing}

\section{Conclusions}

\section{References}
\bibliographystyle{abbrv}
\bibliography{els2017-radiance}
\end{document}

%%% Local Variables:
%%% mode: luatex
%%% TeX-master: t
%%% End:

### 2016-12-16 els2017-radiance/els2017-radiance.tex
Continue writing the article.

@@
 \documentclass{sig-alternate-05-2015}
+\usepackage{float}
+\usepackage{graphicx}
+\usepackage{color}
+\usepackage{minted}
+\usemintedstyle{manni}
+\bibliographystyle{abbrv}
+
 \begin{document}
@@
 \section{Introduction}
+While writing a website or web application is not a very complex or complicated process, doing so still usually involves performing very similar tasks every time.
+This lead to the rise of web frameworks-- collections of tools and libraries to take care of the most common of aspects for you.
+\\
 
-\section{Interfaces}
+A web application framework is comprised of several components.
+This can go from the absolute minimum of an HTTP server and URL dispatch mechanism as seen in microframeworks\cite{microframeworks}, to a vast range of tools including database access, authentication, user management, validation, API endpoints, etc.
+All of the components exist with the aim of reducing the amount of work necessary to write a specific web application.
+\\
 
-\section{Routing}
+Often these frameworks do not address questions about deployment and server setup.
+The assumption is made that the application will be deployed in such a fashion that it is run on a single public-facing domain, which it has complete control over.
+More specifically, it is assumed that the full path of the URL can be controlled.
+This leads to complications when multiple applications should be run on the same server.
+Sharing can be accomplished by using different ports for each application's server, and then using a proxying server that defers requests to the appropriate application depending on the subdomain of the request.
+\\
+
+However, this does not account for all possible setups.
+For example, if you cannot add subdomains and would prefer to run the applications under different path prefixes, it will frequently not be possible to run the application properly.
+While URL rewriting by the proxying server can be used to at least get the application to dispatch, the HTTP documents returned by it will contain links that are not properly resolved and will point to invalid resources.
+\\
+
+This leads to a divide in application design.
+If you set out to write an application that others should be able to use on their own setup, then it needs to be written in a particular way in order to account for the aforementioned setup differences.
+The framework will usually not help you with this in any way.
+\\
+
+The problems for administrators that stem from these single-application focused frameworks do not end there, however.
+Many applications will need access to the same kinds of resources.
+Frequently, information about users, authorisation, sessions, and so forth should be shared, but cannot, because the various applications were designed individually without any concern for a standardised access or even representation of such resources.
+As such, running multiple applications results in duplication of information, which can be especially cumbersome for the users, as they need to track multiple registrations, logins, and so forth despite being on the same service.
+\\
+
+\section{Radiance's Approach}
+Radiance was designed with a different perspective in mind.
+Instead of viewing the application design process from the lens of the developer, it instead looks at it from the view of an administrator that would like to run an application written with Radiance on one of his servers.
+Doing so allowed it to come up with solutions to these problems that can simultaneously account for the needs of an administrator and those of a developer.
+\\
+
+\subsection{Interfaces}
+The first piece is an interface system.
+This system allows the specification of signatures of a package, meaning that functions, macros, variables, etc.
+are defined in their signature only without any actual functionality being implemented that makes them work.
+For example, a simplified interface to a database might look like this:
+
+\begin{minted}[fontsize=\small]{common-lisp}
+(define-interface database
+  (defun select (table where &key limit skip sort))
+  (defun insert (table data))
+  (defun delete (table where &key limit skip sort))
+  (defun update (table data where &key limit skip sort))
+  (defmacro with-transaction ((name) &body body)))
+\end{minted}
+
+While the example here only shows the definition of functions and a macro, it is possible to extend the system to any definition facility desired.
+The description of the general behaviour of each of these definitions cannot be expressed in code however, and must thus be specified in documentation.
+\\
+
+Applications can now make use of this interface and write code against it. As long as the use corresponds with the specified signatures and behaviour, the code is promised to work correctly. \\
+
+The actual behaviour of the individual definitions is left up to an implementation of the respective interface.
+Such an implementation can be a regular system that simply redefines the stubs defined by the interface to do what it is supposed to.
+In the case of our proposed database interface, we could have multiple implementations for it, each of which connects to a particular type of database system.
+The choice of implementation for an interface is left up to the administrator of a Radiance installation.
+Since the applications that make use of an interface and the implementation for it must conform to its specified behaviour, all applications should work regardless of the particular implementation chosen.
+\\
+
+Radiance provides standard interface definitions for most of the common aspects of web application programming.
+This allows the sharing of the previously discussed common resources such as sessions, authentication, users, profile data, and so forth.
+However, it is also sensible to define interfaces for commonly needed tools that may be solved in different manners.
+An example for this would be a caching system; the exact behaviour of the cache can vary --be it database-backed, file-backed, memory-backed, etc.-- but the usage of the system is the same regardless of implementation details.
+\\
+
+In specific, Radiance defines standard interfaces for banning, rate-limiting, an administration panel, caching, authentication, sessions, users, user profiles, the HTTP server, logging, database interaction, and database record abstraction.
+While the interface definitions will always be around, not every interface must load an implementation unless there is an application that specifically needs it.
+Thus, the amount of functionality Radiance provides is also dynamic and without penalty for minimal setups.
+\\
+
+Since the implementation mechanism used is a direct replacement of the interface stubs, rather than relying on some sort of indirection facility, interface functionality incurs zero overhead as well, allowing implementations to be as efficient as possible if the need arises.
+\\
+
+Thus the interface mechanism solves the problem of shared access across applications without constraining the system to a particular implementation, allowing the administrator of the system to choose the appropriate implementation after the fact.
+\\
+
+\subsection{Routing}
+The second piece is a powerful routing system.
+While most frameworks seem to conflate routing and dispatching, Radiance makes a clear distinction.
+Dispatching is the process of determining which function to use to provide the response to a request to a particular URL.
+Routing is the process of transforming a URL into another.
+\\
+
+The routing system lies between the HTTP server and the web applications. This leads to the following request lifecycle for a typical Radiance setup: \\
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=0.42\textwidth]{request.pdf}
+  \caption{Radiance Request Lifecycle}
+\end{figure}
+
+The URL on which the request arrives is transformed into an internal representation by mapping routes.
+This internal representation, called a ``URI'', is composed of a list of subdomains, an optional port, and a path string.
+The top-level domain of the request --the part of the domain we cannot change-- is stripped away.
+Then the request is dispatched to the first handler that matches the URI.
+It can then decide what to do with it, which usually is rendering some kind of template.
+In order to ensure that links and references within the emitted HTML can actually be resolved, the template system must use URIs that are then translated by reversal routes to produce a URL.
+The compelted HTML document is then sent back as a response to the request.
+\\
+
+The routing system is thus responsible for maintaining the separation of the system into two different worlds: an external one that the server and users of the site interact with, and an internal one that applications deal with.
+\\
+
+As an example, let us imagine a server set up on \\ \mintinline{html}{radiance.example.com/test/} that runs a single application.
+The application will be called ``hello'' and has the following code to define a page:
+
+\begin{minted}{common-lisp}
+(define-page index "hello/index" ()
+  (r-clip:process (@template "index.ctml")))
+\end{minted}
+
+If an (untranslated) request to \mintinline{html}{hello/index} is performed, the above code will be executed, and the template system Clip will render the \mintinline{html}{index.ctml} file.
+The template might look as follows:
+
+\begin{minted}{html}
+<link rel="stylesheet"
+      @href="hello/static/hello/index.css" />
+<h1>Hello!</h1>
+\end{minted}
+
+Clip will recognise the \mintinline{html}{@href} attribute, parse its value as a URI, send it through the reversal routes, and set the resulting URL as the value for the \mintinline{html}{href} attribute.
+Thus the cycle illustrated in Figure 1 is followed by our minimal setup here.
+\\
+
+As it stands, Radiance would translate a request to \\ \mintinline{html}{http://radiance.example.com/test/index} into the URI \\\mintinline{html}{radiance/test/index}, which is not what we need if the index page should be invoked.
+Routes need to be set up to handle the conversion to and from the internal representation.
+
+\begin{minted}{common-lisp}
+(define-string-route index :mapping 
+  "radiance/test/(.*)" "hello/\\1")
+(define-string-route index :reversal
+  "hello/(.*)" "radiance/test/\\1")
+\end{minted}
+
+With these routes in place, our previous example URL will be translated into \mintinline{html}{hello/index} as required.
+Furthermore, the requested reversal translation of the URI \\\mintinline{html}{/static/hello/index.css} will be turned into the URL \\\mintinline{html}{http://radiance.example.com/test/static/hello/index.css}, completing the cycle properly.
+\\
+
+While the above routes are fairly simplistic in their behaviour, routes can be written to perform arbitrary operations, including stashing away data in the request object for later use during reversal.
+This allows things like the virtual subdomain route system that is included by default.
+This route recognises when a request is made on the path prefix \mintinline{html}{/!/subdomain/} and translates it into a request with the subdomains set to the subdomain name in the prefix and the path stripped of this prefix.
+It is necessary for the reversal routes to know that the initial request came through this particular route, in order to appropriately translate back to the prefix, as there is no indication in the actual URI that this should happen.
+\\
 
@@
 
-\section{References}
-\bibliographystyle{abbrv}
 \bibliography{els2017-radiance}
@@
 %%% Local Variables:
-%%% mode: luatex
+%%% mode: latex
+%%% TeX-command-extra-options: "-shell-escape"
 %%% TeX-master: t
+%%% TeX-engine: luatex
 %%% End:

### 2016-12-16 els2017-radiance/els2017-radiance.tex
Remove bib for now

@@
 A web application framework is comprised of several components.
-This can go from the absolute minimum of an HTTP server and URL dispatch mechanism as seen in microframeworks\cite{microframeworks}, to a vast range of tools including database access, authentication, user management, validation, API endpoints, etc.
+This can go from the absolute minimum of an HTTP server and URL dispatch mechanism as seen in microframeworks, to a vast range of tools including database access, authentication, user management, validation, API endpoints, etc.
 All of the components exist with the aim of reducing the amount of work necessary to write a specific web application.
@@
 
-\bibliography{els2017-radiance}
+%\bibliography{els2017-radiance}
 \end{document}

### 2016-12-16 els2017-radiance/els2017-radiance.tex
Done writing for now.

@@
 \documentclass{sig-alternate-05-2015}
+\usepackage[english]{babel}
 \usepackage{float}
 \usepackage{graphicx}
-\usepackage{color}
+\usepackage{xcolor}
 \usepackage{minted}
-\usemintedstyle{manni}
+\usemintedstyle{autumn}
 \bibliographystyle{abbrv}
@@
 \begin{abstract}
-  Radiance is a set of libraries that provide an environment for web applications.
+  Radiance is a set of libraries that together provide an environment for web applications.
 Unlike traditional web frameworks that focus on a set of tools to support the construction and maintenance of a single application, Radiance attempts to allow you to run as many differing applications as desired in the same environment.
@@
 
-As an example, let us imagine a server set up on \\ \mintinline{html}{radiance.example.com/test/} that runs a single application.
+As an example, let us imagine a server set up on \\ \texttt{radiance.example.com/test/} that runs a single application.
 The application will be called ``hello'' and has the following code to define a page:
@@
 
-If an (untranslated) request to \mintinline{html}{hello/index} is performed, the above code will be executed, and the template system Clip will render the \mintinline{html}{index.ctml} file.
+If an (untranslated) request to \texttt{hello/index} is performed, the above code will be executed, and the template system Clip will render the \texttt{index.ctml} file.
 The template might look as follows:
@@
 \begin{minted}{html}
-<link rel="stylesheet"
+<link rel="stylesheet" type="text/css"
       @href="hello/static/hello/index.css" />
@@
 
-Clip will recognise the \mintinline{html}{@href} attribute, parse its value as a URI, send it through the reversal routes, and set the resulting URL as the value for the \mintinline{html}{href} attribute.
+Clip will recognise the \texttt{@href} attribute, parse its value as a URI, send it through the reversal routes, and set the resulting URL as the value for the \texttt{href} attribute.
 Thus the cycle illustrated in Figure 1 is followed by our minimal setup here.
@@
 
-As it stands, Radiance would translate a request to \\ \mintinline{html}{http://radiance.example.com/test/index} into the URI \\\mintinline{html}{radiance/test/index}, which is not what we need if the index page should be invoked.
+As it stands, Radiance would translate a request to \\\texttt{http://radiance.example.com/test/index} into the URI \\\texttt{radiance/test/index}, which is not what we need if the index page should be invoked.
 Routes need to be set up to handle the conversion to and from the internal representation.
@@
 
-With these routes in place, our previous example URL will be translated into \mintinline{html}{hello/index} as required.
-Furthermore, the requested reversal translation of the URI \\\mintinline{html}{/static/hello/index.css} will be turned into the URL \\\mintinline{html}{http://radiance.example.com/test/static/hello/index.css}, completing the cycle properly.
+With these routes in place, our previous example URL will be translated into \texttt{hello/index} as required.
+Furthermore, the requested reversal translation of the URI \\\texttt{/static/hello/index.css} will be turned into the URL \\\texttt{http://radiance.example.com/test/static/hello/index.css}, completing the cycle properly.
 \\
@@
 This allows things like the virtual subdomain route system that is included by default.
-This route recognises when a request is made on the path prefix \mintinline{html}{/!/subdomain/} and translates it into a request with the subdomains set to the subdomain name in the prefix and the path stripped of this prefix.
-It is necessary for the reversal routes to know that the initial request came through this particular route, in order to appropriately translate back to the prefix, as there is no indication in the actual URI that this should happen.
+This route recognises when a request is made on the path prefix \texttt{/!/subdomain/} and translates it into a request with the subdomains set to the subdomain name in the prefix and the path stripped of this prefix.
+It is necessary for the corresponding reversal route to know that the initial request came through this particular route in order to appropriately translate back to the prefix, as there is no indication in the actual URI that this should happen.
 \\
 
+In order for everything to fit together and work with routing, applications need to abide by some conventions that Radiance postulates.
+The most important one is that each application gets its own subdomain, including all subdomains on the levels below it.
+This allows applications to assume that they have free reign over the path part of the URL.
+Additionally, the path prefixes \texttt{/static/} and \texttt{/api/} are reserved by Radiance on all domains, and are used to provide static files and REST api endpoints respectively.
+This way, routes can be written in a more generic manner to solve common problems, rather than requiring specifically tailored rules for each application.
+\\
+
+Thus, providing a generic translation mechanism for URLs allows applications to be written in a way that avoids them from clashing with each other, while at the same time gaining the ability for administrators for a setup to configure the URL namespace according to their constraints and needs.
+\\
+
+\subsection{Resources}
+The final piece is a generic resource mechanism that allows the exchanging of information between applications, and particularly between applications and interface implementations.
+This is necessary because applications may need to refer to resources provided by other components in the Radiance environment.
+For example, the login page and mechanism is provided by the authentication interface.
+An application may want to present the users a link to this page, but because the location of this page is not standardised, the application has no direct way of knowing where the page would be.
+Instead of specifying the page location, Radiance provides a system that allows the retrieval of arbitrary resources, and specifies that the implementation must provide the resource as requested.
+\\
+
+Several standard resource types are specified, though more can be defined by an application writer, for example to provide for an extension mechanism.
+The most important standard resource type is the \texttt{page}, which returns a URI to the requested page, if it exists.
+Thus, a code like the following will produce a URL to the appropriate page:
+
+\begin{minted}{common-lisp}
+(uri-to-url (resource :auth :page "login")
+            :representation :external)
+\end{minted}
+
 \section{Conclusions}
+By implementing the components presented in this paper, Radiance allows for an environment that can house multiple applications in the same lisp process and seamlessly handle both the sharing of common resources, and the encapsulation and separation of each application.
+In order to make it feasible for every application to run on any target setup, and to make it adaptable to an administrator's requirements for the public web interface, a powerful routing system is necessary that not only handles the incoming request, but also the outgoing data.
+\\
+
+These components not only present advantages to the administrator of a system, but also to the developer of an application, as they simplify and streamline various parts that would otherwise be implemented ad-hoc.
+This makes the resulting code easier to maintain, read, and adapt for other projects.
 

### 2016-12-16 els2017-radiance/paper.tex
Renamed files.
Don't overflow

--- 2016-12-16 els2017-radiance/els2017-radiance.tex
@@
-\documentclass{sig-alternate-05-2015}
+\documentclass{sig-alternate}
 \usepackage[english]{babel}
@@
 With these routes in place, our previous example URL will be translated into \texttt{hello/index} as required.
-Furthermore, the requested reversal translation of the URI \\\texttt{/static/hello/index.css} will be turned into the URL \\\texttt{http://radiance.example.com/test/static/hello/index.css}, completing the cycle properly.
+Furthermore, the requested reversal translation of the URI \\\texttt{/static/hello/index.css} will be turned into the URL \\\texttt{\small http://radiance.example.com/test/static/hello/index.css},\\ completing the cycle properly.
 \\

### 2016-12-16 els2017-radiance/paper.tex
Some measly references

@@
 \begin{abstract}
-  Radiance is a set of libraries that together provide an environment for web applications.
+  Radiance\cite{radiance} is a set of libraries that together provide an environment for web applications.
 Unlike traditional web frameworks that focus on a set of tools to support the construction and maintenance of a single application, Radiance attempts to allow you to run as many differing applications as desired in the same environment.
@@
 A web application framework is comprised of several components.
-This can go from the absolute minimum of an HTTP server and URL dispatch mechanism as seen in microframeworks, to a vast range of tools including database access, authentication, user management, validation, API endpoints, etc.
+This can go from the absolute minimum of an HTTP server and URL dispatch mechanism as seen in microframeworks\cite{microframeworks}, to a vast range of tools including database access, authentication, user management, validation, API endpoints, etc.
 All of the components exist with the aim of reducing the amount of work necessary to write a specific web application.
@@
 
-If an (untranslated) request to \texttt{hello/index} is performed, the above code will be executed, and the template system Clip will render the \texttt{index.ctml} file.
+If an (untranslated) request to \texttt{hello/index} is performed, the above code will be executed, and the template system Clip\cite{clip} will render the \texttt{index.ctml} file.
 The template might look as follows:
@@
 
-%\bibliography{els2017-radiance}
+\bibliography{paper}
 \end{document}

### 2016-12-16 els2017-radiance/paper.tex
Minor

@@
 \begin{abstract}
-  Radiance\cite{radiance} is a set of libraries that together provide an environment for web applications.
-Unlike traditional web frameworks that focus on a set of tools to support the construction and maintenance of a single application, Radiance attempts to allow you to run as many differing applications as desired in the same environment.
+  Radiance\cite{radiance} is a set of libraries that provides an environment for web applications.
+Unlike traditional web frameworks that focus on a set of tools to support the construction and maintenance of a single application, Radiance attempts to allow one to run as many differing applications as desired in the same environment.
 This has many advantages, as frequently there are common resources in applications that should be shared between them, but cannot because the applications are separated and developed without regard for sharing.

### 2016-12-16 els2017-radiance/paper.tex
Lotsa fixes, mostly stylistic, but also in content.

@@
 Unlike traditional web frameworks that focus on a set of tools to support the construction and maintenance of a single application, Radiance attempts to allow one to run as many differing applications as desired in the same environment.
-This has many advantages, as frequently there are common resources in applications that should be shared between them, but cannot because the applications are separated and developed without regard for sharing.
+This has many advantages, as frequently there are common resources in applications that should be shared between them, but cannot, because the applications are separated and developed without regard for sharing.
 In order to fix this, Radiance provides an extensible interface mechanism to standardise the interaction between users of common resources.
@@
 
-\newpage
 \section{Introduction}
-While writing a website or web application is not a very complex or complicated process, doing so still usually involves performing very similar tasks every time.
-This lead to the rise of web frameworks-- collections of tools and libraries to take care of the most common of aspects for you.
+Writing a website or web application is usually not a very complex or complicated process; doing so involves performing very similar tasks every time.
+This lead to the rise of web frameworks-- collections of tools and libraries to take care of the most common of aspects for the developer.
 \\
@@
 A web application framework is comprised of several components.
-This can go from the absolute minimum of an HTTP server and URL dispatch mechanism as seen in microframeworks\cite{microframeworks}, to a vast range of tools including database access, authentication, user management, validation, API endpoints, etc.
-All of the components exist with the aim of reducing the amount of work necessary to write a specific web application.
+This can go from the absolute minimum of an HTTP server and URL dispatch mechanism as seen in microframeworks\cite{microframeworks}, to a vast range of tools including database access, authentication, user management, validation, API endpoints, and more.
+All of the components exist with the aim of reducing the amount of work necessary to write a single web application.
 \\
 
-Often these frameworks do not address questions about deployment and server setup.
+Often these frameworks do not address many questions about deployment and server setup.
 The assumption is made that the application will be deployed in such a fashion that it is run on a single public-facing domain, which it has complete control over.
@@
 However, this does not account for all possible setups.
-For example, if you cannot add subdomains and would prefer to run the applications under different path prefixes, it will frequently not be possible to run the application properly.
+For example, if one cannot add subdomains and would prefer to run the applications under different path prefixes, it will frequently not be possible to run the application properly.
 While URL rewriting by the proxying server can be used to at least get the application to dispatch, the HTTP documents returned by it will contain links that are not properly resolved and will point to invalid resources.
@@
 This leads to a divide in application design.
-If you set out to write an application that others should be able to use on their own setup, then it needs to be written in a particular way in order to account for the aforementioned setup differences.
-The framework will usually not help you with this in any way.
-\\
+If the task is to write an application that others should be able to use on their own setup, then it needs to be written in a particular way in order to account for the aforementioned setup differences.
+The framework will usually not help with this in any way, and an ad-hoc solution has to be written.
+As an example, Wordpress requires the configuration of a special variable to account for this.\cite{wordpress} \\
 
@@
 Many applications will need access to the same kinds of resources.
-Frequently, information about users, authorisation, sessions, and so forth should be shared, but cannot, because the various applications were designed individually without any concern for a standardised access or even representation of such resources.
-As such, running multiple applications results in duplication of information, which can be especially cumbersome for the users, as they need to track multiple registrations, logins, and so forth despite being on the same service.
+Frequently, information about users, authorisation, sessions, and so forth should be shared, but cannot, because the various applications were designed individually, without any concern for standardised access of such resources.
+As such, running multiple applications on the same server results in duplication of information.
+This can be especially cumbersome for the users, as they need to track multiple registrations, profiles, and so forth, despite being on the same service.
 \\
@@
 Radiance was designed with a different perspective in mind.
-Instead of viewing the application design process from the lens of the developer, it instead looks at it from the view of an administrator that would like to run an application written with Radiance on one of his servers.
-Doing so allowed it to come up with solutions to these problems that can simultaneously account for the needs of an administrator and those of a developer.
+Instead of viewing the application design process from the lens of the developer, it instead looks at it from the view of an administrator that would like to run an application on one of his servers.
+Doing so allowed it to come up with solutions to these problems that can simultaneously account for the needs of an administrator, and those of a developer.
 \\
@@
 \subsection{Interfaces}
-The first piece is an interface system.
-This system allows the specification of signatures of a package, meaning that functions, macros, variables, etc.
-are defined in their signature only without any actual functionality being implemented that makes them work.
+The first piece that came out of this line of thinking is an interface system.
+This system allows the specification of signatures of a package.
+This means that functions, macros, variables, etc.
+are defined in their signature only, without any actual functionality being implemented that makes them work.
 For example, a simplified interface to a database might look like this:
@@
 While the example here only shows the definition of functions and a macro, it is possible to extend the system to any definition facility desired.
-The description of the general behaviour of each of these definitions cannot be expressed in code however, and must thus be specified in documentation.
+The description of the general behaviour of each of these definitions cannot be expressed in code however, and must be specified in documentation.
 \\
@@
 
-The actual behaviour of the individual definitions is left up to an implementation of the respective interface.
-Such an implementation can be a regular system that simply redefines the stubs defined by the interface to do what it is supposed to.
-In the case of our proposed database interface, we could have multiple implementations for it, each of which connects to a particular type of database system.
-The choice of implementation for an interface is left up to the administrator of a Radiance installation.
+The code that actually realises the specific behaviour of the individual definitions is left up to an implementation of the respective interface.
+Such an implementation can be a regular system that simply overwrites the stubs defined by the interface.
+In the case of our proposed database interface, we could have multiple implementations to choose from, each of which provides the glue code to a particular type of database system.
+The choice of the specific implementation for an interface is left up to the administrator of a Radiance installation.
 Since the applications that make use of an interface and the implementation for it must conform to its specified behaviour, all applications should work regardless of the particular implementation chosen.
@@
 Radiance provides standard interface definitions for most of the common aspects of web application programming.
-This allows the sharing of the previously discussed common resources such as sessions, authentication, users, profile data, and so forth.
-However, it is also sensible to define interfaces for commonly needed tools that may be solved in different manners.
+This allows the sharing of the previously discussed common resources.
+However, it is also sensible to define interfaces for commonly needed tools that may have slightly different implementation details.
 An example for this would be a caching system; the exact behaviour of the cache can vary --be it database-backed, file-backed, memory-backed, etc.-- but the usage of the system is the same regardless of implementation details.
@@
 In specific, Radiance defines standard interfaces for banning, rate-limiting, an administration panel, caching, authentication, sessions, users, user profiles, the HTTP server, logging, database interaction, and database record abstraction.
-While the interface definitions will always be around, not every interface must load an implementation unless there is an application that specifically needs it.
+While the interface definitions will always be around, not every interface must load an implementation unless there is an application that specifically requests it.
 Thus, the amount of functionality Radiance provides is also dynamic and without penalty for minimal setups.
@@
 
-Since the implementation mechanism used is a direct replacement of the interface stubs, rather than relying on some sort of indirection facility, interface functionality incurs zero overhead as well, allowing implementations to be as efficient as possible if the need arises.
+Since the implementation mechanism does not rely on some sort of indirection facility, interface functionality incurs zero overhead, allowing implementations to be as efficient as possible.
+This is one of the details of the system that are rather specific to Lisp, since it makes this kind of behaviour trivial to implement.
 \\
 
-Thus the interface mechanism solves the problem of shared access across applications without constraining the system to a particular implementation, allowing the administrator of the system to choose the appropriate implementation after the fact.
+Thus the interface mechanism solves the problem of shared access across applications without constraining the system to a particular implementation.
+This allows the administrator of the system to choose the appropriate implementation for his setup, without having to take care to respect every application's wishes.
 \\
@@
 Dispatching is the process of determining which function to use to provide the response to a request to a particular URL.
-Routing is the process of transforming a URL into another.
+Routing is the process of transforming a URL in some fashion, thus ``routing'' it somewhere else.
 \\
@@
   \centering
-  \includegraphics[width=0.42\textwidth]{request.pdf}
+  \includegraphics[width=0.5\textwidth]{request.pdf}
   \caption{Radiance Request Lifecycle}
@@
 This internal representation, called a ``URI'', is composed of a list of subdomains, an optional port, and a path string.
-The top-level domain of the request --the part of the domain we cannot change-- is stripped away.
+The other information contained in a URL is not necessary for routing or dispatching.
 Then the request is dispatched to the first handler that matches the URI.
-It can then decide what to do with it, which usually is rendering some kind of template.
+It can decide what to do with it, which is usually to render some kind of template.
 In order to ensure that links and references within the emitted HTML can actually be resolved, the template system must use URIs that are then translated by reversal routes to produce a URL.
-The compelted HTML document is then sent back as a response to the request.
+The completed HTML document is then sent back as a response to the request.
 \\
@@
 
-If an (untranslated) request to \texttt{hello/index} is performed, the above code will be executed, and the template system Clip\cite{clip} will render the \texttt{index.ctml} file.
+If an (untranslated) request to \texttt{hello/index} were performed, the above code will be executed, and the template system Clip\cite{clip} will render the \texttt{index.ctml} file.
 The template might look as follows:
@@
 
-As it stands, Radiance would translate a request to \\\texttt{http://radiance.example.com/test/index} into the URI \\\texttt{radiance/test/index}, which is not what we need if the index page should be invoked.
-Routes need to be set up to handle the conversion to and from the internal representation.
+As it stands, Radiance would translate a request to \\\texttt{http://radiance.example.com/test/index} into the URI \\\texttt{radiance.example.com/test/index} and then map it to \\\texttt{radiance/test/index}.
+The top-level domain --the part of the domain we cannot change-- is stripped away by one of the standard routes.
+This is still not a URI that would invoke our page though.
+Therefore, additional routes need to be set up to handle the conversion to and from the internal representation.
 
@@
 
-With these routes in place, our previous example URL will be translated into \texttt{hello/index} as required.
-Furthermore, the requested reversal translation of the URI \\\texttt{/static/hello/index.css} will be turned into the URL \\\texttt{\small http://radiance.example.com/test/static/hello/index.css},\\ completing the cycle properly.
+With these routes in place, our previous example URL will be mapped to \texttt{hello/index} as required.
+Furthermore, the requested reversal of the URI \texttt{/static/hello/index.css} within the template will be turned into the URL \\\texttt{\small http://radiance.example.com/test/static/hello/index.css},\\ completing the cycle properly.
+The top-level domain here is reattached by another standard route.
 \\
 
-While the above routes are fairly simplistic in their behaviour, routes can be written to perform arbitrary operations, including stashing away data in the request object for later use during reversal.
-This allows things like the virtual subdomain route system that is included by default.
-This route recognises when a request is made on the path prefix \texttt{/!/subdomain/} and translates it into a request with the subdomains set to the subdomain name in the prefix and the path stripped of this prefix.
-It is necessary for the corresponding reversal route to know that the initial request came through this particular route in order to appropriately translate back to the prefix, as there is no indication in the actual URI that this should happen.
+While the above routes are fairly simplistic in their behaviour, routes can be written to perform arbitrary operations, including stashing away data for later use during reversal.
+This allows things like the virtual subdomain route that is included by default.
+This route recognises when a request like this \texttt{/!/subdomain/foo} and maps it to \texttt{subdomain/foo}.
+It is necessary for the corresponding reversal route to know that the initial request was mapped by this particular route in order to appropriately translate back to the prefix, as there is no indication in the URI that we want to reverse that this should happen.
 \\
 
-In order for everything to fit together and work with routing, applications need to abide by some conventions that Radiance postulates.
-The most important one is that each application gets its own subdomain, including all subdomains on the levels below it.
+In order for everything to fit together, applications need to abide by some conventions that Radiance postulates.
+The most important one is that each application gets its own subdomain in the internal representation.
 This allows applications to assume that they have free reign over the path part of the URL.
@@
 
-Thus, providing a generic translation mechanism for URLs allows applications to be written in a way that avoids them from clashing with each other, while at the same time gaining the ability for administrators for a setup to configure the URL namespace according to their constraints and needs.
+Thus, providing a generic translation mechanism for URLs allows applications to be written in a way that avoids them from clashing with each other, while at the same time gaining the ability for administrators to configure the URL namespace according to their constraints and needs.
 \\
@@
 \subsection{Resources}
-The final piece is a generic resource mechanism that allows the exchanging of information between applications, and particularly between applications and interface implementations.
+The final piece is a generic resource mechanism that allows the exchange of information between applications, and particularly between applications and interface implementations.
 This is necessary because applications may need to refer to resources provided by other components in the Radiance environment.
 For example, the login page and mechanism is provided by the authentication interface.
-An application may want to present the users a link to this page, but because the location of this page is not standardised, the application has no direct way of knowing where the page would be.
-Instead of specifying the page location, Radiance provides a system that allows the retrieval of arbitrary resources, and specifies that the implementation must provide the resource as requested.
+An application may want to present the users a link to this page.
+However, because the location of this page is not standardised, the application has no direct way of knowing where the page would be.
+Instead of specifying the page location, Radiance provides a system that allows the retrieval of such resources generically.
 \\
 
-Several standard resource types are specified, though more can be defined by an application writer, for example to provide for an extension mechanism.
-The most important standard resource type is the \texttt{page}, which returns a URI to the requested page, if it exists.
-Thus, a code like the following will produce a URL to the appropriate page:
+Several standard resource types are specified, though more can be defined by an application writer; for example to provide for an extension mechanism.
+The most important standard resource type is the \texttt{page}, which returns a URI to a requested page, if it exists.
+Thus, a code like the following will produce a URL to the login screen of the authentication interface:
 
@@
 
+The interface definitions are extended to allow expressing that an implementation must provide for a particular resource, providing the above behaviour as a promise.
+
 \section{Conclusions}
-By implementing the components presented in this paper, Radiance allows for an environment that can house multiple applications in the same lisp process and seamlessly handle both the sharing of common resources, and the encapsulation and separation of each application.
-In order to make it feasible for every application to run on any target setup, and to make it adaptable to an administrator's requirements for the public web interface, a powerful routing system is necessary that not only handles the incoming request, but also the outgoing data.
+By implementing the components presented in this paper, Radiance allows for an environment that can house multiple applications in the same lisp process.
+It seamlessly handles both the sharing of common resources, as well as the encapsulation and separation of each application.
+In order to make it feasible for every application to run on any target setup, and to make it adaptable to an administrator's requirements for the public web interface, a powerful routing system was created that not only handles the incoming request, but also the outgoing data.
 \\
 
-These components not only present advantages to the administrator of a system, but also to the developer of an application, as they simplify and streamline various parts that would otherwise be implemented ad-hoc.
-This makes the resulting code easier to maintain, read, and adapt for other projects.
+These components present clear advantages to the administrator of a system and to the developer of an application, as they simplify and streamline various parts that would otherwise be implemented ad-hoc.
 

### 2016-12-16 els2017-radiance/paper.tex
Some grammar and style fixes are reported by @robert-strandh

@@
+%% FIXME:
+% Use of "this" without being followed by a noun. Eg "this environment sharing", "this kind of repetition"
+% Tweak minted colours
+
 \documentclass{sig-alternate}
@@
 
-The problems for administrators that stem from these single-application focused frameworks do not end there, however.
+For administrators the problems that stem from these single-application focused frameworks do not end there, however.
 Many applications will need access to the same kinds of resources.
@@
 Radiance was designed with a different perspective in mind.
-Instead of viewing the application design process from the lens of the developer, it instead looks at it from the view of an administrator that would like to run an application on one of his servers.
+Instead of viewing the application design process through the lens of the developer, it instead looks at it from the view of an administrator that would like to run an application on one of his servers.
 Doing so allowed it to come up with solutions to these problems that can simultaneously account for the needs of an administrator, and those of a developer.
@@
 
-Applications can now make use of this interface and write code against it. As long as the use corresponds with the specified signatures and behaviour, the code is promised to work correctly. \\
+Applications can now make use of this interface and write code against it. As long as the use corresponds to the specified signatures and behaviour, the code is promised to work correctly. \\
 
@@
 However, it is also sensible to define interfaces for commonly needed tools that may have slightly different implementation details.
-An example for this would be a caching system; the exact behaviour of the cache can vary --be it database-backed, file-backed, memory-backed, etc.-- but the usage of the system is the same regardless of implementation details.
+An example for this would be a caching system; the exact behaviour of the cache can vary --be it database-backed, file-backed, memory-backed, etc.-- but the usage of the system is the same, regardless of implementation details.
 \\
 
-In specific, Radiance defines standard interfaces for banning, rate-limiting, an administration panel, caching, authentication, sessions, users, user profiles, the HTTP server, logging, database interaction, and database record abstraction.
+Specifically, Radiance defines standard interfaces for banning, rate-limiting, an administration panel, caching, authentication, sessions, users, user profiles, the HTTP server, logging, database interaction, and database record abstraction.
 While the interface definitions will always be around, not every interface must load an implementation unless there is an application that specifically requests it.
@@
 
-Thus the interface mechanism solves the problem of shared access across applications without constraining the system to a particular implementation.
-This allows the administrator of the system to choose the appropriate implementation for his setup, without having to take care to respect every application's wishes.
+Thus, the interface mechanism solves the problem of shared access across applications without constraining the system to a particular implementation.
+This allows the administrator of the system to choose the appropriate implementation for his setup, without having to take care to respect the wishes of every application.
 \\
@@
 Clip will recognise the \texttt{@href} attribute, parse its value as a URI, send it through the reversal routes, and set the resulting URL as the value for the \texttt{href} attribute.
-Thus the cycle illustrated in Figure 1 is followed by our minimal setup here.
+Thus, the cycle illustrated in Figure 1 is followed by our minimal setup here.
 \\
@@
 
-Thus, providing a generic translation mechanism for URLs allows applications to be written in a way that avoids them from clashing with each other, while at the same time gaining the ability for administrators to configure the URL namespace according to their constraints and needs.
+Thus, providing a generic translation mechanism for URLs allows applications to be written in a way that prevents them from clashing with each other, while at the same time gaining the ability for administrators to configure the URL namespace according to their constraints and needs.
 \\
@@
 For example, the login page and mechanism is provided by the authentication interface.
-An application may want to present the users a link to this page.
+An application may want to present a link to this page to the users.
 However, because the location of this page is not standardised, the application has no direct way of knowing where the page would be.
@@
 \section{Conclusions}
-By implementing the components presented in this paper, Radiance allows for an environment that can house multiple applications in the same lisp process.
+By implementing the components presented in this paper, Radiance allows for an environment that can house multiple applications in the same Lisp process.
 It seamlessly handles both the sharing of common resources, as well as the encapsulation and separation of each application.

### 2016-12-17 els2017-radiance/paper.tex
Fixed "This .." without subject.

@@
 %% FIXME:
-% Use of "this" without being followed by a noun. Eg "this environment sharing", "this kind of repetition"
 % Tweak minted colours
+% Clarify different perspective (admin vs dev)
 
@@
 Unlike traditional web frameworks that focus on a set of tools to support the construction and maintenance of a single application, Radiance attempts to allow one to run as many differing applications as desired in the same environment.
-This has many advantages, as frequently there are common resources in applications that should be shared between them, but cannot, because the applications are separated and developed without regard for sharing.
+This goal brings many advantages, as frequently there are common resources in applications that should be shared between them, but cannot, because the applications are separated and developed without regard for sharing.
 In order to fix this, Radiance provides an extensible interface mechanism to standardise the interaction between users of common resources.
@@
 Writing a website or web application is usually not a very complex or complicated process; doing so involves performing very similar tasks every time.
-This lead to the rise of web frameworks-- collections of tools and libraries to take care of the most common of aspects for the developer.
+This trend lead to the rise of web frameworks-- collections of tools and libraries to take care of the most common of aspects for the developer.
 \\
@@
 More specifically, it is assumed that the full path of the URL can be controlled.
-This leads to complications when multiple applications should be run on the same server.
+This assumption leads to complications when multiple applications should be run on the same server.
 Sharing can be accomplished by using different ports for each application's server, and then using a proxying server that defers requests to the appropriate application depending on the subdomain of the request.
@@
 
-This leads to a divide in application design.
+This problem leads to a divide in application design.
 If the task is to write an application that others should be able to use on their own setup, then it needs to be written in a particular way in order to account for the aforementioned setup differences.
@@
 As such, running multiple applications on the same server results in duplication of information.
-This can be especially cumbersome for the users, as they need to track multiple registrations, profiles, and so forth, despite being on the same service.
+This duplication can be especially cumbersome for the users, as they need to track multiple registrations, profiles, and so forth, despite being on the same service.
 \\
@@
 This system allows the specification of signatures of a package.
-This means that functions, macros, variables, etc.
+What this means is that functions, macros, variables, etc.
 are defined in their signature only, without any actual functionality being implemented that makes them work.
@@
 
-Radiance provides standard interface definitions for most of the common aspects of web application programming.
-This allows the sharing of the previously discussed common resources.
+Radiance provides standard interface definitions for most of the common aspects of web application programming, allowing the sharing of the previously discussed common resources.
 However, it is also sensible to define interfaces for commonly needed tools that may have slightly different implementation details.
@@
 Since the implementation mechanism does not rely on some sort of indirection facility, interface functionality incurs zero overhead, allowing implementations to be as efficient as possible.
-This is one of the details of the system that are rather specific to Lisp, since it makes this kind of behaviour trivial to implement.
+This advantage is one of the details of the system that are rather specific to Lisp, since it makes this kind of behaviour trivial to implement.
 \\
@@
 Thus, the interface mechanism solves the problem of shared access across applications without constraining the system to a particular implementation.
-This allows the administrator of the system to choose the appropriate implementation for his setup, without having to take care to respect the wishes of every application.
+Thanks to this, the administrator of the system can choose the appropriate implementation for his setup, without having to take care to respect the wishes of every application.
 \\
@@
 
-The routing system lies between the HTTP server and the web applications. This leads to the following request lifecycle for a typical Radiance setup: \\
+The routing system lies between the HTTP server and the web applications, leading to the following request lifecycle for a typical Radiance setup: \\
 
@@
 The top-level domain --the part of the domain we cannot change-- is stripped away by one of the standard routes.
-This is still not a URI that would invoke our page though.
+This URI is still not one that would invoke our page though.
 Therefore, additional routes need to be set up to handle the conversion to and from the internal representation.
@@
 While the above routes are fairly simplistic in their behaviour, routes can be written to perform arbitrary operations, including stashing away data for later use during reversal.
-This allows things like the virtual subdomain route that is included by default.
+This flexibility allows things like the virtual subdomain route that is included by default.
 This route recognises when a request like this \texttt{/!/subdomain/foo} and maps it to \texttt{subdomain/foo}.
@@
 The most important one is that each application gets its own subdomain in the internal representation.
-This allows applications to assume that they have free reign over the path part of the URL.
+This convention allows applications to assume that they have free reign over the path part of the URL.
 Additionally, the path prefixes \texttt{/static/} and \texttt{/api/} are reserved by Radiance on all domains, and are used to provide static files and REST api endpoints respectively.
@@
 The final piece is a generic resource mechanism that allows the exchange of information between applications, and particularly between applications and interface implementations.
-This is necessary because applications may need to refer to resources provided by other components in the Radiance environment.
+This piece is necessary because applications may need to refer to resources provided by other components in the Radiance environment.
 For example, the login page and mechanism is provided by the authentication interface.

### 2016-12-18 els2017-radiance/paper.tex
Tweaked highlighting
Some improvements thanks to @tehrengruber

@@
 %% FIXME:
-% Tweak minted colours
 % Clarify different perspective (admin vs dev)
+% Clarify "on any server setup"
+% Show examples for shitty frameworks
+% Focus more on new perspective rather than making the impossible possible or whatever the fuck I was thinking at the tim
+% Give more code samples
+% In the request cycle image, replace Application with Application/Handler
+% Virual subdomains because localhost no subdomains, what fucking idiots specified this shit
+% Note about lack of configuration for dev vs prod in conclusion
 
@@
 \usepackage{minted}
-\usemintedstyle{autumn}
+\usemintedstyle[html]{borland}
+\usemintedstyle[common-lisp]{pastie}
 \bibliographystyle{abbrv}
@@
        \affaddr{ETH Zürich}\\
-       \affaddr{Nürenbergstr. 17B}\\
-       \affaddr{8037 Zürich, Switzerland}\\
+       \affaddr{Rämistrasse 101}\\
+       \affaddr{8092 Zürich, Switzerland}\\
        \email{shinmera@tymoon.eu}
@@
 \section{Introduction}
-Writing a website or web application is usually not a very complex or complicated process; doing so involves performing very similar tasks every time.
+Most tasks involved in writing a website or web application are not very complex or complicated; they usually involve performing very similar tasks every time.
 This trend lead to the rise of web frameworks-- collections of tools and libraries to take care of the most common of aspects for the developer.
@@
 
-The URL on which the request arrives is transformed into an internal representation by mapping routes.
+The URL on which the request arrives is transformed into an internal representation by so-called ``mapping routes''.
 This internal representation, called a ``URI'', is composed of a list of subdomains, an optional port, and a path string.
@@
 It can decide what to do with it, which is usually to render some kind of template.
-In order to ensure that links and references within the emitted HTML can actually be resolved, the template system must use URIs that are then translated by reversal routes to produce a URL.
+In order to ensure that links and references within the emitted HTML can actually be resolved, the template system must use URIs that are then translated by so-called ``reversal routes'' to produce a URL.
 The completed HTML document is then sent back as a response to the request.

### 2016-12-18 els2017-radiance/paper.tex
Some more minor corrections, notes

@@
 %% FIXME:
-% Clarify different perspective (admin vs dev)
-% Clarify "on any server setup"
-% Show examples for shitty frameworks
-% Focus more on new perspective rather than making the impossible possible or whatever the fuck I was thinking at the tim
-% Give more code samples
-% In the request cycle image, replace Application with Application/Handler
-% Virual subdomains because localhost no subdomains, what fucking idiots specified this shit
-% Note about lack of configuration for dev vs prod in conclusion
+% As it stands the paper's introduction is garbage.
+% Either:
+%   Show examples for shitty frameworks that can't do what Radiance can
+%   Focus more on the new perspective rather than making the impossible possible
 
@@
 Radiance was designed with a different perspective in mind.
-Instead of viewing the application design process through the lens of the developer, it instead looks at it from the view of an administrator that would like to run an application on one of his servers.
+Instead of viewing the application design process through the lens of the developer, it looks at it from the view of an administrator that would like to run an application on one of his servers.
 Doing so allowed it to come up with solutions to these problems that can simultaneously account for the needs of an administrator, and those of a developer.
@@
 It is necessary for the corresponding reversal route to know that the initial request was mapped by this particular route in order to appropriately translate back to the prefix, as there is no indication in the URI that we want to reverse that this should happen.
+The virtual domain routing is especially useful on development setups, where no real domains are directly available.
+It allows you to use \texttt{localhost} or even just the IP address to reach everything.
 \\
@@
 \section{Conclusions}
-By implementing the components presented in this paper, Radiance allows for an environment that can house multiple applications in the same Lisp process.
+By implementing the components presented in this paper, Radiance allows for an environment that can conveniently house multiple applications in the same Lisp process.
 It seamlessly handles both the sharing of common resources, as well as the encapsulation and separation of each application.
@@
 These components present clear advantages to the administrator of a system and to the developer of an application, as they simplify and streamline various parts that would otherwise be implemented ad-hoc.
+Particularly, development and production setups can both be configured with ease.
 

### 2016-12-19 els2017-radiance/paper.tex
Attempt to rewrite the introduction to be less shit.

@@
-%% FIXME:
-% As it stands the paper's introduction is garbage.
-% Either:
-%   Show examples for shitty frameworks that can't do what Radiance can
-%   Focus more on the new perspective rather than making the impossible possible
-
 \documentclass{sig-alternate}
@@
 \keywords{Common Lisp, web framework, web development, encapsulation, interfacing}
+\newpage
 
 \section{Introduction}
-Most tasks involved in writing a website or web application are not very complex or complicated; they usually involve performing very similar tasks every time.
+Most tasks involved in writing a website or web application are not very complex or complicated; they usually involve solving very similar problems every time.
 This trend lead to the rise of web frameworks-- collections of tools and libraries to take care of the most common of aspects for the developer.
@@
 This can go from the absolute minimum of an HTTP server and URL dispatch mechanism as seen in microframeworks\cite{microframeworks}, to a vast range of tools including database access, authentication, user management, validation, API endpoints, and more.
-All of the components exist with the aim of reducing the amount of work necessary to write a single web application.
+All of these components exist with the aim of reducing the amount of work necessary to write a single web application.
 \\
 
-Often these frameworks do not address many questions about deployment and server setup.
-The assumption is made that the application will be deployed in such a fashion that it is run on a single public-facing domain, which it has complete control over.
-More specifically, it is assumed that the full path of the URL can be controlled.
-This assumption leads to complications when multiple applications should be run on the same server.
-Sharing can be accomplished by using different ports for each application's server, and then using a proxying server that defers requests to the appropriate application depending on the subdomain of the request.
+Usually these frameworks have the goal of supporting developers in writing a single, particular application.
+This can lead to issues when the application is supposed to be deployed on wildly different setups than it was developed on, or if it should share the space with other applications.
+Solutions to resource sharing or configuration then have to be designed by the application itself, and are thus often ad-hoc and incompatible with other applications.
+As an example, even if you have two applications like a forum and a blog that were both written with the same framework, you won't feasibly be able to configure your system in such a way that both share the same login and account information.
 \\
 
-However, this does not account for all possible setups.
-For example, if one cannot add subdomains and would prefer to run the applications under different path prefixes, it will frequently not be possible to run the application properly.
-While URL rewriting by the proxying server can be used to at least get the application to dispatch, the HTTP documents returned by it will contain links that are not properly resolved and will point to invalid resources.
-\\
-
-This problem leads to a divide in application design.
-If the task is to write an application that others should be able to use on their own setup, then it needs to be written in a particular way in order to account for the aforementioned setup differences.
-The framework will usually not help with this in any way, and an ad-hoc solution has to be written.
-As an example, Wordpress requires the configuration of a special variable to account for this.\cite{wordpress} \\
-
-For administrators the problems that stem from these single-application focused frameworks do not end there, however.
-Many applications will need access to the same kinds of resources.
-Frequently, information about users, authorisation, sessions, and so forth should be shared, but cannot, because the various applications were designed individually, without any concern for standardised access of such resources.
-As such, running multiple applications on the same server results in duplication of information.
-This duplication can be especially cumbersome for the users, as they need to track multiple registrations, profiles, and so forth, despite being on the same service.
-\\
+This leads to a frustrating experience for the administrator of a server setup, who either has to go through complicated deployment and configuration steps to bend the applications to fit his requirements, or even has to go so far as to change the source code to do what he needs.
 

### 2016-12-21 els2017-radiance/paper.tex
Minor corrections and improvements.

@@
 To facilitate the separation and encapsulation of the shared URL address space between applications, Radiance also provides a flexible routing mechanism.
-This mechanism is powerful enough to ensure that any application written with Radiance can be deployed on any server setup, no matter how complicated the constraints, without needing to change the source code of the application.
+This mechanism is powerful enough to ensure that any application written with Radiance can be deployed on a server setup, no matter how complicated the constraints, without needing to change the source code of the application.
 \end{abstract}
@@
 Solutions to resource sharing or configuration then have to be designed by the application itself, and are thus often ad-hoc and incompatible with other applications.
-As an example, even if you have two applications like a forum and a blog that were both written with the same framework, you won't feasibly be able to configure your system in such a way that both share the same login and account information.
+As an example, even if you have two applications like a forum and a blog that were both written with the same framework, they are usually deployed in such a format, that you won't feasibly be able to configure your system to allow both to share the same login and account information, or other kinds of data.
 \\
@@
 Doing so allowed it to come up with solutions to these problems that can simultaneously account for the needs of an administrator, and those of a developer.
+For example, an administrator may want to configure which database and server to use, confine an application to a particular domain or path, or share a domain across multiple applications.
 \\
@@
 Specifically, Radiance defines standard interfaces for banning, rate-limiting, an administration panel, caching, authentication, sessions, users, user profiles, the HTTP server, logging, database interaction, and database record abstraction.
-While the interface definitions will always be around, not every interface must load an implementation unless there is an application that specifically requests it.
+While the interface definitions will always be around, not every interface needs to load an implementation unless there is an application that specifically requests it.
 Thus, the amount of functionality Radiance provides is also dynamic and without penalty for minimal setups.
@@
 While most frameworks seem to conflate routing and dispatching, Radiance makes a clear distinction.
-Dispatching is the process of determining which function to use to provide the response to a request to a particular URL.
+Dispatching is the process of determining which function should provide the response to a request to a particular URL.
 Routing is the process of transforming a URL in some fashion, thus ``routing'' it somewhere else.
@@
 It can decide what to do with it, which is usually to render some kind of template.
-In order to ensure that links and references within the emitted HTML can actually be resolved, the template system must use URIs that are then translated by so-called ``reversal routes'' to produce a URL.
+In order to ensure that links and references within the emitted HTML can actually be resolved, the template system must use URIs that are translated by so-called ``reversal routes'' to produce a URL.
 The completed HTML document is then sent back as a response to the request.

### 2016-12-21 els2017-radiance/paper.tex
Change address.
Minor crap.

@@
 Nicolas Hafner\\
-       \affaddr{ETH Zürich}\\
-       \affaddr{Rämistrasse 101}\\
-       \affaddr{8092 Zürich, Switzerland}\\
+       \affaddr{Shirakumo.org}\\
+       \affaddr{Zürich, Switzerland}\\
        \email{shinmera@tymoon.eu}
@@
 Most tasks involved in writing a website or web application are not very complex or complicated; they usually involve solving very similar problems every time.
-This trend lead to the rise of web frameworks-- collections of tools and libraries to take care of the most common of aspects for the developer.
+This trend lead to the rise of web frameworks--- collections of tools and libraries to take care of the most common of aspects for the developer.
 \\
@@
 However, it is also sensible to define interfaces for commonly needed tools that may have slightly different implementation details.
-An example for this would be a caching system; the exact behaviour of the cache can vary --be it database-backed, file-backed, memory-backed, etc.-- but the usage of the system is the same, regardless of implementation details.
+An example for this would be a caching system; the exact behaviour of the cache can vary ---be it database-backed, file-backed, memory-backed, etc.--- but the usage of the system is the same, regardless of implementation details.
 \\
@@
 As it stands, Radiance would translate a request to \\\texttt{http://radiance.example.com/test/index} into the URI \\\texttt{radiance.example.com/test/index} and then map it to \\\texttt{radiance/test/index}.
-The top-level domain --the part of the domain we cannot change-- is stripped away by one of the standard routes.
+The top-level domain ---the part of the domain we cannot change--- is stripped away by one of the standard routes.
 This URI is still not one that would invoke our page though.

### 2017-03-13 els2017-radiance/paper.tex
Completely rewrite the paper in hopes that this attempt makes it not suck huge amounts of dicks.

--- 2016-12-19 els2017-radiance/paper.tex
@@
+%% FIXME:
+%% - !! Elaborate on the provided standard interfaces that Radiance's core provides
+%% - Elaborate on mechanisms for required and optional interface dependence
+%% - Elaborate on how the interface mechanism is implemented
+%% - Elaborate on how to actually implement an interface
+%% - Explain what options an adminitrator has to administer the system
+%% - Explain difference in URI to W3C URIs
+%% - Elaborate on the problem of deployment of single-application programs
+%% - Show a real-world example
+
 \documentclass{sig-alternate}
@@
 \usepackage{minted}
+\usepackage{hyperref}
 \usemintedstyle[html]{borland}
 \usemintedstyle[common-lisp]{pastie}
+\newmintinline[code]{common-lisp}{}
 \bibliographystyle{abbrv}
+
+\hypersetup{
+  colorlinks,
+  linkcolor={red!50!black},
+  citecolor={blue!50!black},
+  urlcolor={blue!80!black}
+}
 
@@
 To facilitate the separation and encapsulation of the shared URL address space between applications, Radiance also provides a flexible routing mechanism.
-This mechanism is powerful enough to ensure that any application written with Radiance can be deployed on any server setup, no matter how complicated the constraints, without needing to change the source code of the application.
 \end{abstract}
@@
 \section{Introduction}
-Most tasks involved in writing a website or web application are not very complex or complicated; they usually involve solving very similar problems every time.
-This trend lead to the rise of web frameworks-- collections of tools and libraries to take care of the most common of aspects for the developer.
+As the internet evolved, served websites began to evolve and become more and more dynamic.
+Content is no longer a set of static webpages, but rather automatically updated or even individually composed for the specific user viewing the website.
+Creating such dynamic websites requires a lot of extra work, much of which is usually done in much the same way for every website.
+The requested content needs to be retrieved somehow, transformed as needed, and finally assembled into deliverable HTML.
 \\
 
-A web application framework is comprised of several components.
-This can go from the absolute minimum of an HTTP server and URL dispatch mechanism as seen in microframeworks\cite{microframeworks}, to a vast range of tools including database access, authentication, user management, validation, API endpoints, and more.
-All of these components exist with the aim of reducing the amount of work necessary to write a single web application.
+In order to handle these common aspects, web frameworks have been created.
+These frameworks can come in all kinds of sizes, from the very minimal set of a HTML templating engine and a web server as seen in micro-frameworks\cite{microframeworks}, to extensive sets of utilities to handle user interaction and data.
 \\
 
-Usually these frameworks have the goal of supporting developers in writing a single, particular application.
-This can lead to issues when the application is supposed to be deployed on wildly different setups than it was developed on, or if it should share the space with other applications.
-Solutions to resource sharing or configuration then have to be designed by the application itself, and are thus often ad-hoc and incompatible with other applications.
-As an example, even if you have two applications like a forum and a blog that were both written with the same framework, you won't feasibly be able to configure your system in such a way that both share the same login and account information.
+Typically these frameworks are constructed with the intent of helping you develop a single web application, which is then deployed and run standalone.
+However, this can lead to issues when multiple applications should be deployed side-by-side.
+For example, if the framework does not provide explicit support for a particular feature such as user accounts, the two applications will likely have implemented ad-hoc solutions of their own, which are incompatible with each other and thus can't be trivially merged together.
+Large frameworks may avoid the problems introduced by ad-hoc feature implementation in each application, but instead run the risk of introducing too many features that remain unused.
 \\
 
-This leads to a frustrating experience for the administrator of a server setup, who either has to go through complicated deployment and configuration steps to bend the applications to fit his requirements, or even has to go so far as to change the source code to do what he needs.
-
-\section{Radiance's Approach}
-Radiance was designed with a different perspective in mind.
-Instead of viewing the application design process through the lens of the developer, it looks at it from the view of an administrator that would like to run an application on one of his servers.
-Doing so allowed it to come up with solutions to these problems that can simultaneously account for the needs of an administrator, and those of a developer.
+\section{Example Requirements of an Application}
+In order to illustrate the reasoning behind the individual components present in the Radiance system, we are going to make use of a simple example web application throughout the paper.
+This application should provide a ``paste service'' wherein users can upload text snippets, to which they can then link to.
+If they have a registered account, they should be able to manage pastes they have made before, and configure the look of the site to their liking.
 \\
 
+In order to write this application, we are going to need a database of some sort that stores the pastes, a system to handle user accounts and authentication, a user settings panel, a cache system for performance, an HTTP server, and a template engine.
+\\
+
+For this application to share as many resources as it can with potential third-party applications residing in the same installation, it needs to rely on the framework to provide all of the above, except for the template engine.
+Additionally, it needs to have some system that divides up the URL namespace between applications.
+As the application writer, we cannot have any presumptions about what the final setup will look like-- what kinds of applications will be deployed together, and how the public URLs should resolve.
+
+\section{The Radiance System}
+The core of the Radiance system is rather small in terms of actual code.
+It can provide a plethora of features in the spirit of a macro-framework, but doesn't have to.
+Most of its features are pluggable, which is done through interfaces.
+\\
+
+The division of the URL namespace is provided through the routing system, which allows both a convenient view for developers to work with, and a powerful way for administrators to configure the system to their liking without having to touch application code.
+\\
+
+Finally, the setup of a Radiance configuration is handled through so-called ``environments,'' which specify the setup of the individual components and how they can be reached.
+
 \subsection{Interfaces}
-The first piece that came out of this line of thinking is an interface system.
-This system allows the specification of signatures of a package.
-What this means is that functions, macros, variables, etc.
-are defined in their signature only, without any actual functionality being implemented that makes them work.
-For example, a simplified interface to a database might look like this:
+Interfaces represent a form of contract.
+They allow you to define the signatures of constructs exposed from a package, such as functions, macros, classes, and so forth.
+These function signatures are accompanied by a documentation that specifies their public behaviour.
+Through this, Radiance can provide several ``standard interfaces'' that specify the behaviour of many features that are commonly needed in web applications, without actually having to implement them.
+\\
+
+In order to make this more concrete, let's look at an interface that might provide some form of caching mechanism.
 
 \begin{minted}[fontsize=\small]{common-lisp}
-(define-interface database
-  (defun select (table where &key limit skip sort))
-  (defun insert (table data))
-  (defun delete (table where &key limit skip sort))
-  (defun update (table data where &key limit skip sort))
-  (defmacro with-transaction ((name) &body body)))
+(define-interface cache
+  (defun invalidate (name)
+    "Causes the cached value of NAME to be re-computed.")
+  (defmacro with-caching (name invalidate &body body)
+    "Caches the return value if INVALIDATE is non-NIL."))
 \end{minted}
 
-While the example here only shows the definition of functions and a macro, it is possible to extend the system to any definition facility desired.
-The description of the general behaviour of each of these definitions cannot be expressed in code however, and must be specified in documentation.
+This construct defines a new package called \code{cache}, exports the symbols \code{invalidate} and \code{with-caching} from it, and installs stub definitions for the respective function and macro.
+With this interface specification in place, an application can start writing code against it.
+In our imaginary paste service, we could now cache the page for a specific paste like so:
+
+\begin{minted}[fontsize=\small]{common-lisp}
+(defun paste-page (id)
+  (cache:with-caching id NIL
+    (render-paste (load-paste id))))
+\end{minted}
+
+However, as it currently is, this will obviously not work.
+The \code{with-caching} macro is not implemented with any useful functionality.
+As the writer of the paste application, we don't need to know the specifics of how the caching is implemented, though.
+All we need to do is tell Radiance that we need this interface.
+This can be done by adding the interface as a dependency to our application's ASDF system definition.
+
+\begin{minted}[fontsize=\small]{common-lisp}
+(asdf:defsystem paste-service
+  ...
+  :depends-on (...
+               (:interface :cache)))
+\end{minted}
+
+Radiance extends ASDF's dependency resolution mechanism to allow for this kind of declaration.
+When the \\\code{paste-service} system is now loaded, Radiance will notice the interface dependency and resolve it to an actual system that implements the desired cache functionality.
+This resolution is defined in the currently active environment, and can thus be configured by the administrator of the Radiance installation.
 \\
 
-Applications can now make use of this interface and write code against it. As long as the use corresponds to the specified signatures and behaviour, the code is promised to work correctly. \\
+For completeness, let's look at an implementation of this cache interface.
+The implementation must be loadable through an ASDF system, so that the above dependency resolution can be done.
+The actual implementation mechanism is handed to us by Lisp allowing redefinition.
 
-The code that actually realises the specific behaviour of the individual definitions is left up to an implementation of the respective interface.
-Such an implementation can be a regular system that simply overwrites the stubs defined by the interface.
-In the case of our proposed database interface, we could have multiple implementations to choose from, each of which provides the glue code to a particular type of database system.
-The choice of the specific implementation for an interface is left up to the administrator of a Radiance installation.
-Since the applications that make use of an interface and the implementation for it must conform to its specified behaviour, all applications should work regardless of the particular implementation chosen.
+\begin{minted}[fontsize=\small]{common-lisp}
+(defvar cache::*caches* (make-hash-table))
+
+(defun cache:invalidate (name)
+  (remhash name *caches*))
+
+(defmacro cache:with-caching (name invalidate &body body)
+  (once-only (name)
+    `(or (and (not ,invalidate) (gethash ,name *caches*))
+         (setf (gethash ,name *caches*)
+               (progn ,@body)))))
+\end{minted}
+
+This is a particularly primitive and straight-forward implementation, in order to keep things short.
+Particularly, the above completely avoids creating a new package for the implementation, and instead provides the caching table as an implementation-dependant extension through the cache interface, by making its symbol unexported.
+The function and macro are provided by just overriding the stub definitions that were in place previously.
 \\
 
-Radiance provides standard interface definitions for most of the common aspects of web application programming, allowing the sharing of the previously discussed common resources.
-However, it is also sensible to define interfaces for commonly needed tools that may have slightly different implementation details.
-An example for this would be a caching system; the exact behaviour of the cache can vary --be it database-backed, file-backed, memory-backed, etc.-- but the usage of the system is the same, regardless of implementation details.
+Using direct overwriting of definitions means that all applications must use the same implementation.
+However, usually this is the intended effect, as we want to maximise the sharing between applications.
+If an implementation should have special needs, it can always bypass the interfaces and make direct use of whatever it might depend on.
+This approach does bring for some great benefits however.
+For one, it allows the interfaces to be as efficient as possible, as there is no intermediate layer of redirection.
+It also allows us to expose any kind of definition, including custom, user-defined ones.
 \\
 
-Specifically, Radiance defines standard interfaces for banning, rate-limiting, an administration panel, caching, authentication, sessions, users, user profiles, the HTTP server, logging, database interaction, and database record abstraction.
-While the interface definitions will always be around, not every interface must load an implementation unless there is an application that specifically requests it.
-Thus, the amount of functionality Radiance provides is also dynamic and without penalty for minimal setups.
+Radiance provides standard interface definitions for all of the components we require for the paste application, and more.
+A full list of the interfaces and their descriptions is available in the Radiance documentation\footnote{\url{https://github.com/Shirakumo/radiance/\#interface}}.
+Thus, Radiance can be used like a macro-framework, but does not load on any features unless specifically required.
+Additionally, any of the features can be exchanged by a system administrator for ones that more closely match their requirements, without having to change anything about the application code.\\
+
+Finally, Radiance provides a system by which functionality can be provided if a specific interface should be loaded, without explicitly having to depend on it.
+This is useful to model something like the user settings mentioned in our example application.
+An administrator might not always want to provide an administration or settings panel.
+To make this dependency optional, we can defer the compilation and evaluation of our relevant logic to a later point, after an implementation has been loaded.
+For an imaginary \code{user-settings} interface, this might look like the following:
+
+\begin{minted}[fontsize=\small]{common-lisp}
+(define-implement-trigger user-settings
+  (user-settings:define-option ...))
+\end{minted}
+
+Since all the symbols are already provided by the interface definition, there are no problems when the reader parses the source.
+The forms can thus simply be saved away for evaluation once Radiance notices that an implementation of the interface in question has been loaded.
 \\
 
-Since the implementation mechanism does not rely on some sort of indirection facility, interface functionality incurs zero overhead, allowing implementations to be as efficient as possible.
-This advantage is one of the details of the system that are rather specific to Lisp, since it makes this kind of behaviour trivial to implement.
+Ultimately, interfaces are a form of compromise between providing all possible features at once, and almost no features at all.
+The usefulness of an interface heavily depends on its specification, and in order for implementations to be really exchangeable without modifying the application code, each implementation and application must strictly adhere to the specification.
+Thus interfaces are very restricting, and often cannot expose the full power of an underlying system, to allow a variety of different backing implementations.
+
+\subsection{Routes}
+In order to allow the administrator to change where pages are accessible from, an application cannot hardcode its resource locations and the links in its templates.
+This is doubly important, when it comes to housing multiple applications in one, as the system needs to be set up in such a way that the applications do not clash with each other or potentially confuse pages of one another.
 \\
 
-Thus, the interface mechanism solves the problem of shared access across applications without constraining the system to a particular implementation.
-Thanks to this, the administrator of the system can choose the appropriate implementation for his setup, without having to take care to respect the wishes of every application.
+In many frameworks, like for example Symfony\cite{symfony}, this is solved by naming every resource in the system by a tag, and then allowing the configuration of what each tag resolves to individually.
+Radiance takes a different approach.
+It introduces the idea of two separate namespaces: an external one, which is what a visitor of a website sees and interacts with, and an internal one, which is what the programmer of an application deals with.
+The translation between the two namespaces is the responsibility of the routing system.
 \\
 
-\subsection{Routing}
-The second piece is a powerful routing system.
-While most frameworks seem to conflate routing and dispatching, Radiance makes a clear distinction.
-Dispatching is the process of determining which function to use to provide the response to a request to a particular URL.
-Routing is the process of transforming a URL in some fashion, thus ``routing'' it somewhere else.
+\begin{figure}[H]
+  \includegraphics[width=\columnwidth]{request}
+  \caption{Standard request life-cycle}
+\end{figure}
+
+When a request is dispatched by Radiance, it first parses the request URL into an object presentation that makes it easier to modify.
+It then sends it through the routing system's mapping functions, which turn this external URL into an internal one.
+The request is then dispatched to the according application that ``owns'' the URL in the internal representation.
 \\
 
-The routing system lies between the HTTP server and the web applications, leading to the following request lifecycle for a typical Radiance setup: \\
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=0.5\textwidth]{request.pdf}
-  \caption{Radiance Request Lifecycle}
-\end{figure}
-
-The URL on which the request arrives is transformed into an internal representation by so-called ``mapping routes''.
-This internal representation, called a ``URI'', is composed of a list of subdomains, an optional port, and a path string.
-The other information contained in a URL is not necessary for routing or dispatching.
-Then the request is dispatched to the first handler that matches the URI.
-It can decide what to do with it, which is usually to render some kind of template.
-In order to ensure that links and references within the emitted HTML can actually be resolved, the template system must use URIs that are then translated by so-called ``reversal routes'' to produce a URL.
-The completed HTML document is then sent back as a response to the request.
+Since Radiance has full control over the organisation of the internal representation, it can make strict demands as to how applications need to structure their URLs in order for the translation to work correctly.
+Specifically, it requires each application to put all of its pages on a domain that is named after their application.
 \\
 
-The routing system is thus responsible for maintaining the separation of the system into two different worlds: an external one that the server and users of the site interact with, and an internal one that applications deal with.
+In our sample application, we might write the view page something like this, where \texttt{paste/view} is the internal URL that the view page is reachable on.
+
+\begin{minted}[fontsize=\small]{common-lisp}
+(define-page view "paste/view" ()
+  (render-template (template "view.html")
+                   (get-paste (get-var "paste-id"))))
+\end{minted}
+
+If we wanted to reach this page through the URL \texttt{www.example.com/paste/view}, the mapping route functions would have to strip away the \texttt{www.example.com} domain, recognise the \texttt{paste} folder, and put that as the URL's domain instead.
+We can achieve this relatively easily with the following definition.
+
+\begin{minted}[fontsize=\small]{common-lisp}
+(define-route paste :mapping (uri)
+  (when (begins-with "paste/" (path uri))
+    (setf (domains uri) '("paste")
+          (path uri) (subseq (path uri) 6))))
+\end{minted}
+
+Alternatively, we could also use the even simpler string-based definition.
+
+\begin{minted}[fontsize=\small]{common-lisp}
+(define-string-route paste :mapping
+  "/paste/(.*)" "paste/\\1")
+\end{minted}
+
+Naturally, if you wanted different behaviour depending on which domain the request came from, you'd have to write a more specific translation. \\
+
+With the mapping alone the case is not yet solved, however.
+When a page is emitted that contains links, the links too must be translated, but in the opposite way.
+Since routes can be arbitrarily complex, it is not possible for the system to figure out a reversal route automatically.
+The logic of a reversal route will usually be much the same as it was for the mapping route and should thus not be a problem to figure out, though.
 \\
 
-As an example, let us imagine a server set up on \\ \texttt{radiance.example.com/test/} that runs a single application.
-The application will be called ``hello'' and has the following code to define a page:
+The reversal of URLs should receive special support from the templating system, as it is very frequently needed and should thus be short.
+Radiance does not dictate any specific template system, but offers extensions to some existing systems like Clip\cite{clip} to simplify this process.
 
-\begin{minted}{common-lisp}
-(define-page index "hello/index" ()
-  (r-clip:process (@template "index.ctml")))
+\subsection{Environments}
+Radiance's ``environment'' encapsulates the configuration of a Radiance installation.
+Through it, applications can provide settings that the administrator can set to their liking.
+It also provides the information necessary to figure out which implementation to use for an interface.
+\\
+
+Ideally, an administrator of a system will not have to touch any source code, and instead will be able to configure the system through a set of human-readable configuration files.
+It is, of course, still possible to configure the system through usual Lisp code, should one prefer this approach.
+\\
+
+The environment itself simply dictates a directory that contains all of these configuration files.
+Each application in the system automatically receives its own directory in which it can store both configuration, and temporary data files.
+By simply switching out this environment directory, the system can then be loaded into completely different configurations, for example allowing you to easily keep ``development'' and ``production'' setups.
+\\
+
+In our example application we might want to leave it up to the administrator whether to allow anonymous users to post a paste or not. We can set up a default value like this.
+
+\begin{minted}[fontsize=\small]{common-lisp}
+(define-trigger startup ()
+  (defaulted-config T :anonymous-pastes))
 \end{minted}
 
-If an (untranslated) request to \texttt{hello/index} were performed, the above code will be executed, and the template system Clip\cite{clip} will render the \texttt{index.ctml} file.
-The template might look as follows:
+We need to stick this into a trigger, in order to defer the evaluation to when Radiance is being started up and the environment has been decided.
+During the loading of the system, the environment might not have been set yet, and we would not be able to access the configuration storage.
+\\
 
-\begin{minted}{html}
-<link rel="stylesheet" type="text/css"
-      @href="hello/static/hello/index.css" />
-<h1>Hello!</h1>
+While Radiance does provide default implementations for all of the interfaces, it is likely that some of them are not usable for a production setting.
+In order to change, say, the database interface's implementation, we would then have to modify the Radiance core's configuration file.
+We can either modify the file directly, which can be found through \code{(mconfig-pathname :radiance-core)} like so:
+
+\begin{minted}[fontsize=\small]{common-lisp}
+((:interfaces
+  (:database . "i-postmodern")
+  ...)
+ ...)
 \end{minted}
 
-Clip will recognise the \texttt{@href} attribute, parse its value as a URI, send it through the reversal routes, and set the resulting URL as the value for the \texttt{href} attribute.
-Thus, the cycle illustrated in Figure 1 is followed by our minimal setup here.
+.. or instead use the programmatical way.
+
+\begin{minted}[fontsize=\small]{common-lisp}
+(setf (mconfig :radiance-core :interfaces :database)
+      "i-postmodern")
+\end{minted}
+
+\texttt{i-postmodern} here is the name of a standard implementation of the database interface for PostgreSQL databases. \\
+
+Routes can also be configured through the core configuration file. Our previous example mapping route could be set up in the configuration file like this.
+
+\begin{minted}[fontsize=\small]{common-lisp}
+(...
+ (:routes (paste :mapping "/paste/(.*)" "paste/\\1"))
+ ...)
+\end{minted}
+
+Radiance will take care of converting the configuration data into an actual route definition like we saw above. \\
+
+Every application's individual configuration can be changed in much the same way.
+
+\section{Conclusion}
+Radiance provides a web framework that adjusts itself depending on how many features the application requires.
+By separating the applications from the implementations of these features with an interface, it allows the application programmer to write their software against a well-specified API, and retains the ability for the administrator to decide which implementation is most suitable for their setup.
 \\
 
-As it stands, Radiance would translate a request to \\\texttt{http://radiance.example.com/test/index} into the URI \\\texttt{radiance.example.com/test/index} and then map it to \\\texttt{radiance/test/index}.
-The top-level domain --the part of the domain we cannot change-- is stripped away by one of the standard routes.
-This URI is still not one that would invoke our page though.
-Therefore, additional routes need to be set up to handle the conversion to and from the internal representation.
-
-\begin{minted}{common-lisp}
-(define-string-route index :mapping 
-  "radiance/test/(.*)" "hello/\\1")
-(define-string-route index :reversal
-  "hello/(.*)" "radiance/test/\\1")
-\end{minted}
-
-With these routes in place, our previous example URL will be mapped to \texttt{hello/index} as required.
-Furthermore, the requested reversal of the URI \texttt{/static/hello/index.css} within the template will be turned into the URL \\\texttt{\small http://radiance.example.com/test/static/hello/index.css},\\ completing the cycle properly.
-The top-level domain here is reattached by another standard route.
+By maintaining a strict separation of the URL namespace and providing an automated URL rewriting mechanism, Radiance allows for easy sharing of the namespace between an arbitrary number of applications, while at the same time giving the administrator a convenient way to modify the behaviour of the translation to fit their specific server configuration.
 \\
 
-While the above routes are fairly simplistic in their behaviour, routes can be written to perform arbitrary operations, including stashing away data for later use during reversal.
-This flexibility allows things like the virtual subdomain route that is included by default.
-This route recognises when a request like this \texttt{/!/subdomain/foo} and maps it to \texttt{subdomain/foo}.
-It is necessary for the corresponding reversal route to know that the initial request was mapped by this particular route in order to appropriately translate back to the prefix, as there is no indication in the URI that we want to reverse that this should happen.
-The virtual domain routing is especially useful on development setups, where no real domains are directly available.
-It allows you to use \texttt{localhost} or even just the IP address to reach everything.
+Through the environment system, Radiance standardises the way each application is configured by the administrator and how the system is pieced together when it is loaded.
+As a consequence of that it becomes trivial to switch between different setups of a Radiance installation.
+The simple, human-readable configuration format used allows even users without intimate knowledge of Lisp to set up an instance.
+
+\section{Further Work}
+Currently, while it is possible to dynamically load applications, it is not possible to dynamically unload them.
+Doing so is troublesome, as an application's system might potentially modify any other part of the Lisp process.
+However, if the changes that can be rolled back is constrained in some sense, it should be possible to provide this kind of feature.
 \\
 
-In order for everything to fit together, applications need to abide by some conventions that Radiance postulates.
-The most important one is that each application gets its own subdomain in the internal representation.
-This convention allows applications to assume that they have free reign over the path part of the URL.
-Additionally, the path prefixes \texttt{/static/} and \texttt{/api/} are reserved by Radiance on all domains, and are used to provide static files and REST api endpoints respectively.
-This way, routes can be written in a more generic manner to solve common problems, rather than requiring specifically tailored rules for each application.
+Radiance also does not allow you to change the implementation of an interface on the fly.
+This has much the same problems as the previous issue.
+Implementation switching is even more problematic though, since the system needs to ensure that all dependant applications, including optionally dependant ones, are reloaded to make macro redefinitions take effect.
+Furthermore, since some interfaces expose classes and instances thereof, the system would either have to be able to switch the instances between class definitions, or somehow invalidate the old instances if they are retained in another part of the application.
 \\
 
-Thus, providing a generic translation mechanism for URLs allows applications to be written in a way that prevents them from clashing with each other, while at the same time gaining the ability for administrators to configure the URL namespace according to their constraints and needs.
-\\
-
-\subsection{Resources}
-The final piece is a generic resource mechanism that allows the exchange of information between applications, and particularly between applications and interface implementations.
-This piece is necessary because applications may need to refer to resources provided by other components in the Radiance environment.
-For example, the login page and mechanism is provided by the authentication interface.
-An application may want to present a link to this page to the users.
-However, because the location of this page is not standardised, the application has no direct way of knowing where the page would be.
-Instead of specifying the page location, Radiance provides a system that allows the retrieval of such resources generically.
-\\
-
-Several standard resource types are specified, though more can be defined by an application writer; for example to provide for an extension mechanism.
-The most important standard resource type is the \texttt{page}, which returns a URI to a requested page, if it exists.
-Thus, a code like the following will produce a URL to the login screen of the authentication interface:
-
-\begin{minted}{common-lisp}
-(uri-to-url (resource :auth :page "login")
-            :representation :external)
-\end{minted}
-
-The interface definitions are extended to allow expressing that an implementation must provide for a particular resource, providing the above behaviour as a promise.
-
-\section{Conclusions}
-By implementing the components presented in this paper, Radiance allows for an environment that can conveniently house multiple applications in the same Lisp process.
-It seamlessly handles both the sharing of common resources, as well as the encapsulation and separation of each application.
-In order to make it feasible for every application to run on any target setup, and to make it adaptable to an administrator's requirements for the public web interface, a powerful routing system was created that not only handles the incoming request, but also the outgoing data.
-\\
-
-These components present clear advantages to the administrator of a system and to the developer of an application, as they simplify and streamline various parts that would otherwise be implemented ad-hoc.
-Particularly, development and production setups can both be configured with ease.
+Ultimately it should be made possible to switch out the environment of a Radiance installation on the fly, for example to switch between development and production setups without having to restart completely.
+Doing so could massively improve the time needed to discover differences between different setups.
 

### 2017-03-14 els2017-radiance/paper.tex
Make TeX waste less paper.
Some additions and corrections.

@@
 \usepackage{hyperref}
+\usepackage[all]{hypcap}
 \usemintedstyle[html]{borland}
@@
 In order to handle these common aspects, web frameworks have been created.
-These frameworks can come in all kinds of sizes, from the very minimal set of a HTML templating engine and a web server as seen in micro-frameworks\cite{microframeworks}, to extensive sets of utilities to handle user interaction and data.
+These frameworks can come in all kinds of sizes, from the very minimal set of a HTML templating engine and a web server as seen in ``micro-frameworks''\cite{microframeworks}, to extensive sets of utilities to handle user interaction and data.
 \\
@@
 For example, if the framework does not provide explicit support for a particular feature such as user accounts, the two applications will likely have implemented ad-hoc solutions of their own, which are incompatible with each other and thus can't be trivially merged together.
-Large frameworks may avoid the problems introduced by ad-hoc feature implementation in each application, but instead run the risk of introducing too many features that remain unused.
+Large, ``macro-frameworks'' may avoid the problems introduced by ad-hoc feature implementation in each application, but instead run the risk of introducing too many features that remain unused.
 \\
 
+Radiance initially grew out of the desire to write web applications that could be run together and in such a way that users would not have to create multiple accounts, track multiple logins, and so forth.
+Another primary goal of it was to write it in such a way that it would be possible to exchange various parts of the framework without needing to change application code.
+This would allow flexibility on the side of the administrator, to pick and choose the parts that best fit their environment.
+
 \section{Example Requirements of an Application}
-In order to illustrate the reasoning behind the individual components present in the Radiance system, we are going to make use of a simple example web application throughout the paper.
+In order to illustrate the reasoning behind the individual components present in the Radiance system, we are going to make use of a simple, imaginary example web application throughout the paper.
 This application should provide a ``paste service'' wherein users can upload text snippets, to which they can then link to.
@@
 
-In order to write this application, we are going to need a database of some sort that stores the pastes, a system to handle user accounts and authentication, a user settings panel, a cache system for performance, an HTTP server, and a template engine.
+When talking about an application in the context of Radiance, we think of it as a form of library that is then loaded together with Radiance to form a full webserver.
+This is in contrast to the deployment models of many other frameworks, where the ``library'' is deployed bundled tightly together with the framework as a single application.
+\\
+
+In order to write our paste application, we are going to want a database of some sort that stores the pastes, a system to handle user accounts and authentication, a user settings panel, a cache system for performance, an HTTP server, and a template engine.
 \\
@@
 This is a particularly primitive and straight-forward implementation, in order to keep things short.
-Particularly, the above completely avoids creating a new package for the implementation, and instead provides the caching table as an implementation-dependant extension through the cache interface, by making its symbol unexported.
 The function and macro are provided by just overriding the stub definitions that were in place previously.
+The variable definition is not part of the official interface, and is instead exposed through an unexported symbol, denoting an implementation-dependant extension.
 \\
@@
 It introduces the idea of two separate namespaces: an external one, which is what a visitor of a website sees and interacts with, and an internal one, which is what the programmer of an application deals with.
-The translation between the two namespaces is the responsibility of the routing system.
 \\
 
-\begin{figure}[H]
+This separation allows application programmers to model their pages from a perspective that looks much more like the external view.
+Instead of dealing with named tags, they deal in concrete internal URLs that in a development setup often have a relatively direct one-to-one mapping to the external URLs.
+This makes it easier to visualise and think about the application structure.
+From the administrator's point of view this setup is more convenient, as they too simply need to think in terms of translations, rather than having to adapt potentially hundreds of different URL tags manually.
+\\
+
+The translation between the two namespaces is the responsibility of the routing system, which is illustrated in \hyperref[requestcycle]{Figure 1}. \\
+
+\begin{figure}[h]
   \includegraphics[width=\columnwidth]{request}
   \caption{Standard request life-cycle}
+  \label{requestcycle}
 \end{figure}
@@
 
-If we wanted to reach this page through the URL \texttt{www.example.com/paste/view}, the mapping route functions would have to strip away the \texttt{www.example.com} domain, recognise the \texttt{paste} folder, and put that as the URL's domain instead.
+If, as an administrator of an installation of this paste application, we now wanted to reach this page through the URL \texttt{www.example.com/paste/view}, the mapping route functions would have to strip away the \texttt{www.example.com} domain, recognise the \texttt{paste} folder, and put that as the URL's domain instead.
 We can achieve this relatively easily with the following definition.
@@
 Radiance does not dictate any specific template system, but offers extensions to some existing systems like Clip\cite{clip} to simplify this process.
+As a brief example, in Clip, URLs can be denoted through special attributes like this:
+
+\begin{minted}[fontsize=\small]{html}
+<a @href="paste/list">Public Paste List</a>
+<form method="post">
+  <textarea name="text"></textarea>
+  <input type="submit" @formaction="paste/submit" />
+</form>
+\end{minted}
+
+The templating engine takes care of reading out the \texttt{@} attribute values, running them through the routing system's reversal functions, and putting the resulting URL into the corresponding attribute to produce a valid link.
 
@@
 \begin{minted}[fontsize=\small]{common-lisp}
-((:interfaces
-  (:database . "i-postmodern")
+((:interfaces (:database . "i-postmodern")
               ...)
@@
 \begin{minted}[fontsize=\small]{common-lisp}
-(...
- (:routes (paste :mapping "/paste/(.*)" "paste/\\1"))
+((:routes (paste :mapping "/paste/(.*)" "paste/\\1"))
  ...)

### 2017-03-14 els2017-radiance/paper.tex
Some more fixes.

@@
 All we need to do is tell Radiance that we need this interface.
-This can be done by adding the interface as a dependency to our application's ASDF system definition.
+This can be done by adding the interface as a dependency to our application's ASDF\cite{asdf} system definition.
 
@@
 Implementation switching is even more problematic though, since the system needs to ensure that all dependant applications, including optionally dependant ones, are reloaded to make macro redefinitions take effect.
-Furthermore, since some interfaces expose classes and instances thereof, the system would either have to be able to switch the instances between class definitions, or somehow invalidate the old instances if they are retained in another part of the application.
+Furthermore, since some interfaces expose classes and instances thereof, the system would either have to be able to migrate objects between interface implementations, or somehow invalidate the old instances if they are retained in another part of the application.
 \\

### 2017-03-15 els2017-radiance/paper.tex
Fix some grammar and style problems.

@@
-%% FIXME:
-%% - !! Elaborate on the provided standard interfaces that Radiance's core provides
-%% - Elaborate on mechanisms for required and optional interface dependence
-%% - Elaborate on how the interface mechanism is implemented
-%% - Elaborate on how to actually implement an interface
-%% - Explain what options an adminitrator has to administer the system
-%% - Explain difference in URI to W3C URIs
-%% - Elaborate on the problem of deployment of single-application programs
-%% - Show a real-world example
-
 \documentclass{sig-alternate}
@@
 \section{Introduction}
-As the internet evolved, served websites began to evolve and become more and more dynamic.
-Content is no longer a set of static webpages, but rather automatically updated or even individually composed for the specific user viewing the website.
+As the internet evolved, served websites began to evolve and became more and more dynamic.
+Content is no longer a set of static webpages, but rather automatically updated, or even individually composed for the specific user viewing the website.
 Creating such dynamic websites requires a lot of extra work, much of which is usually done in much the same way for every website.
@@
 In order to handle these common aspects, web frameworks have been created.
-These frameworks can come in all kinds of sizes, from the very minimal set of a HTML templating engine and a web server as seen in ``micro-frameworks''\cite{microframeworks}, to extensive sets of utilities to handle user interaction and data.
+These frameworks can come in all kinds of sizes, from the very minimal set of an HTML templating engine and a web server as seen in ``micro-frameworks''\cite{microframeworks}, to extensive sets of utilities to handle user interaction and data of all kinds.
 \\
@@
 For example, if the framework does not provide explicit support for a particular feature such as user accounts, the two applications will likely have implemented ad-hoc solutions of their own, which are incompatible with each other and thus can't be trivially merged together.
-Large, ``macro-frameworks'' may avoid the problems introduced by ad-hoc feature implementation in each application, but instead run the risk of introducing too many features that remain unused.
+Large, ``macro-frameworks'' may avoid the problems introduced by ad-hoc feature implementation, but instead run the risk of introducing too many features that remain unused.
 \\
 
-Radiance initially grew out of the desire to write web applications that could be run together and in such a way that users would not have to create multiple accounts, track multiple logins, and so forth.
-Another primary goal of it was to write it in such a way that it would be possible to exchange various parts of the framework without needing to change application code.
+Radiance initially grew out of the desire to write web applications that could be run together in such a way, that users would not have to create multiple accounts, track multiple logins, and so forth.
+Another primary goal of it was to write it in such a way, that it would be possible to exchange various parts of the framework without needing to change application code.
 This would allow flexibility on the side of the administrator, to pick and choose the parts that best fit their environment.
@@
 In order to illustrate the reasoning behind the individual components present in the Radiance system, we are going to make use of a simple, imaginary example web application throughout the paper.
-This application should provide a ``paste service'' wherein users can upload text snippets, to which they can then link to.
-If they have a registered account, they should be able to manage pastes they have made before, and configure the look of the site to their liking.
+This application should provide a ``blog platform,'' wherein users can create text posts to which they can link their friends.
+They should also be able to edit the posts again at a later point and customise the look and feel to their preference.
 \\
 
-When talking about an application in the context of Radiance, we think of it as a form of library that is then loaded together with Radiance to form a full webserver.
+When talking about such an application in the context of Radiance, we think of it as a form of library that is then loaded together with Radiance to form a full webserver.
 This is in contrast to the deployment models of many other frameworks, where the ``library'' is deployed bundled tightly together with the framework as a single application.
@@
 
-In order to write our paste application, we are going to want a database of some sort that stores the pastes, a system to handle user accounts and authentication, a user settings panel, a cache system for performance, an HTTP server, and a template engine.
+In order to write our blog application, we are going to want a database that stores the posts, a system to handle user accounts and authentication, a cache system for performance, an HTTP server, and a template engine.
 \\
 
-For this application to share as many resources as it can with potential third-party applications residing in the same installation, it needs to rely on the framework to provide all of the above, except for the template engine.
+Imagine now, if you will, that an administrator also wanted to run a forum on the same host as well, perhaps in order to allow people to discuss and engage with each other about the various posts on the main blog.
+Requiring users to maintain a login for each service would be annoying, to say the least.
+Furthermore, as an administrator, we would like to make sure that as much data is kept in the same place as possible, and arranged in a similar way, to ease maintenance.
+\\
+
+For our blog to share as many resources as it can with potential third-party applications residing in the same installation, it needs to rely on the framework to provide all of our required features, except perhaps for the template engine.
 Additionally, it needs to have some system that divides up the URL namespace between applications.
 As the application writer, we cannot have any presumptions about what the final setup will look like-- what kinds of applications will be deployed together, and how the public URLs should resolve.
+\\
 
 \section{The Radiance System}
-The core of the Radiance system is rather small in terms of actual code.
-It can provide a plethora of features in the spirit of a macro-framework, but doesn't have to.
+The core of the Radiance system is rather small.
+Despite this, it can provide a plethora of features in the spirit of a macro-framework, if needed.
 Most of its features are pluggable, which is done through interfaces.
@@
 
-Finally, the setup of a Radiance configuration is handled through so-called ``environments,'' which specify the setup of the individual components and how they can be reached.
+Finally, the management of a Radiance installation is handled through so-called ``environments,'' which specify the configuration of the individual components.
 
@@
 With this interface specification in place, an application can start writing code against it.
-In our imaginary paste service, we could now cache the page for a specific paste like so:
+In our imaginary blog service, we could now cache the page for a specific post like so:
 
 \begin{minted}[fontsize=\small]{common-lisp}
-(defun paste-page (id)
+(defun post-page (id)
   (cache:with-caching id NIL
-    (render-paste (load-paste id))))
+    (render-post (load-post id))))
 \end{minted}
@@
 The \code{with-caching} macro is not implemented with any useful functionality.
-As the writer of the paste application, we don't need to know the specifics of how the caching is implemented, though.
+As the writer of the blog application, we don't need to know the specifics of how the caching is implemented, though.
 All we need to do is tell Radiance that we need this interface.
@@
 \begin{minted}[fontsize=\small]{common-lisp}
-(asdf:defsystem paste-service
+(asdf:defsystem blog-service
   ...
@@
 Radiance extends ASDF's dependency resolution mechanism to allow for this kind of declaration.
-When the \\\code{paste-service} system is now loaded, Radiance will notice the interface dependency and resolve it to an actual system that implements the desired cache functionality.
+When the \\\code{blog-service} system is now loaded, Radiance will notice the interface dependency and resolve it to an actual system that implements the desired cache functionality.
 This resolution is defined in the currently active environment, and can thus be configured by the administrator of the Radiance installation.
@@
 The implementation must be loadable through an ASDF system, so that the above dependency resolution can be done.
-The actual implementation mechanism is handed to us by Lisp allowing redefinition.
+The actual implementation mechanism is handed to us for free, as Lisp allows redefinition.
 
@@
 Using direct overwriting of definitions means that all applications must use the same implementation.
-However, usually this is the intended effect, as we want to maximise the sharing between applications.
+However, usually this is the intended effect, as we want to maximise the sharing between them.
 If an implementation should have special needs, it can always bypass the interfaces and make direct use of whatever it might depend on.
-This approach does bring for some great benefits however.
-For one, it allows the interfaces to be as efficient as possible, as there is no intermediate layer of redirection.
-It also allows us to expose any kind of definition, including custom, user-defined ones.
+This interfaces approach does bring some benefits, however.
+For one, it allows the implementations to be as efficient as possible, as there is no intermediate layer of redirection.
 \\
 
-Radiance provides standard interface definitions for all of the components we require for the paste application, and more.
+Radiance provides standard interface definitions for all of the components we require for the blog application, and more.
 A full list of the interfaces and their descriptions is available in the Radiance documentation\footnote{\url{https://github.com/Shirakumo/radiance/\#interface}}.
 Thus, Radiance can be used like a macro-framework, but does not load on any features unless specifically required.
-Additionally, any of the features can be exchanged by a system administrator for ones that more closely match their requirements, without having to change anything about the application code.\\
+Additionally, any of the implementations can be exchanged by a system administrator for ones that more closely match their requirements, without having to change anything about the application code.\\
 
-Finally, Radiance provides a system by which functionality can be provided if a specific interface should be loaded, without explicitly having to depend on it.
-This is useful to model something like the user settings mentioned in our example application.
+Finally, Radiance provides a system that allows to optionally depend on an interface.
+This is useful to model something like the user preferences mentioned in our example application.
 An administrator might not always want to provide an administration or settings panel.
@@
 Since all the symbols are already provided by the interface definition, there are no problems when the reader parses the source.
-The forms can thus simply be saved away for evaluation once Radiance notices that an implementation of the interface in question has been loaded.
+The forms can thus simply be saved away to be evaluated once Radiance notices that an implementation of the interface in question has been loaded.
 \\
@@
 Ultimately, interfaces are a form of compromise between providing all possible features at once, and almost no features at all.
-The usefulness of an interface heavily depends on its specification, and in order for implementations to be really exchangeable without modifying the application code, each implementation and application must strictly adhere to the specification.
-Thus interfaces are very restricting, and often cannot expose the full power of an underlying system, to allow a variety of different backing implementations.
+The usefulness of an interface heavily depends on its specification, and for implementations to be really exchangeable without modifying the application code, each implementation and application must strictly adhere to the specification.
 
 \subsection{Routes}
-In order to allow the administrator to change where pages are accessible from, an application cannot hardcode its resource locations and the links in its templates.
-This is doubly important, when it comes to housing multiple applications in one, as the system needs to be set up in such a way that the applications do not clash with each other or potentially confuse pages of one another.
+In order to allow the administrator to change where pages are accessible from, an application cannot hard-code its resource locations and the links in its templates.
+This is doubly important when it comes to housing multiple applications in one, as the system needs to be set up in such a way that the applications do not clash with each other or potentially confuse pages of one another.
 \\
@@
 This separation allows application programmers to model their pages from a perspective that looks much more like the external view.
-Instead of dealing with named tags, they deal in concrete internal URLs that in a development setup often have a relatively direct one-to-one mapping to the external URLs.
+Instead of dealing with named tags, they deal in concrete internal URLs that, in a development setup, often have a relatively direct one-to-one mapping to the external URLs.
 This makes it easier to visualise and think about the application structure.
-From the administrator's point of view this setup is more convenient, as they too simply need to think in terms of translations, rather than having to adapt potentially hundreds of different URL tags manually.
+From the administrator's point of view this setup is more convenient too, as they simply need to think in terms of translations, rather than having to adapt potentially hundreds of different URL tags manually.
 \\
@@
 
-Since Radiance has full control over the organisation of the internal representation, it can make strict demands as to how applications need to structure their URLs in order for the translation to work correctly.
+Since Radiance has full control over the organisation of the internal representation, it can make strict demands as to how applications need to structure their URLs.
 Specifically, it requires each application to put all of its pages on a domain that is named after their application.
@@
 
-In our sample application, we might write the view page something like this, where \texttt{paste/view} is the internal URL that the view page is reachable on.
+In our sample application, we might write the view page something like this, where \texttt{blog/view} is the internal URL that the view page is reachable on.
 
 \begin{minted}[fontsize=\small]{common-lisp}
-(define-page view "paste/view" ()
+(define-page view "blog/view" ()
   (render-template (template "view.html")
-                   (get-paste (get-var "paste-id"))))
+                   (get-post (get-var "post-id"))))
 \end{minted}
 
-If, as an administrator of an installation of this paste application, we now wanted to reach this page through the URL \texttt{www.example.com/paste/view}, the mapping route functions would have to strip away the \texttt{www.example.com} domain, recognise the \texttt{paste} folder, and put that as the URL's domain instead.
+If, as an administrator of an installation of this blog application, we now wanted to reach this page through the URL \texttt{www.example.com/blog/view}, the mapping route functions would have to strip away the \texttt{www.example.com} domain, recognise the \texttt{blog} folder, and put that as the URL's domain instead.
 We can achieve this relatively easily with the following definition.
@@
 \begin{minted}[fontsize=\small]{common-lisp}
-(define-route paste :mapping (uri)
-  (when (begins-with "paste/" (path uri))
-    (setf (domains uri) '("paste")
-          (path uri) (subseq (path uri) 6))))
+(define-route blog :mapping (uri)
+  (when (begins-with "blog/" (path uri))
+    (setf (domains uri) '("blog")
+          (path uri) (subseq (path uri) 5))))
 \end{minted}
@@
 \begin{minted}[fontsize=\small]{common-lisp}
-(define-string-route paste :mapping
-  "/paste/(.*)" "paste/\\1")
+(define-string-route blog :mapping
+  "/blog/(.*)" "blog/\\1")
 \end{minted}
@@
 
-With the mapping alone the case is not yet solved, however.
+With the mapping alone the case is not yet solved, though.
 When a page is emitted that contains links, the links too must be translated, but in the opposite way.
-Since routes can be arbitrarily complex, it is not possible for the system to figure out a reversal route automatically.
+Unfortunately, because routes can be arbitrarily complex, it is not possible for the system to figure out a reversal route automatically.
 The logic of a reversal route will usually be much the same as it was for the mapping route and should thus not be a problem to figure out, though.
@@
 \begin{minted}[fontsize=\small]{html}
-<a @href="paste/list">Public Paste List</a>
+<a @href="blog/list">Latest Blog Posts</a>
 <form method="post">
   <textarea name="text"></textarea>
-  <input type="submit" @formaction="paste/submit" />
+  <input type="submit" @formaction="blog/submit" />
 </form>
@@
 
-Ideally, an administrator of a system will not have to touch any source code, and instead will be able to configure the system through a set of human-readable configuration files.
-It is, of course, still possible to configure the system through usual Lisp code, should one prefer this approach.
-\\
-
 The environment itself simply dictates a directory that contains all of these configuration files.
@@
 
-In our example application we might want to leave it up to the administrator whether to allow anonymous users to post a paste or not. We can set up a default value like this.
+Ideally, an administrator of a system will not have to touch any source code, and instead will be able to configure the system through a set of human-readable configuration files.
+It is, of course, still possible to configure the system through usual Lisp code, should one prefer this approach.
+\\
+
+In our example application we might want to give the administrator the ability to define a global blog title. We could provide a default value for it like this:
 
@@
 (define-trigger startup ()
-  (defaulted-config T :anonymous-pastes))
+  (defaulted-config "Irradiant Blogs" :title))
 \end{minted}
@@
 
-While Radiance does provide default implementations for all of the interfaces, it is likely that some of them are not usable for a production setting.
+While Radiance does provide default implementations for all of its interfaces, it is likely that some of them are not usable for a production setting.
 In order to change, say, the database interface's implementation, we would then have to modify the Radiance core's configuration file.
-We can either modify the file directly, which can be found through \code{(mconfig-pathname :radiance-core)} like so:
+We can either modify the file directly, which can be found through \code{(mconfig-pathname :radiance-core)}, like so:
 
@@
 
-.. or instead use the programmatical way.
+.. or instead use the programmatical way, like so:
 
@@
 \begin{minted}[fontsize=\small]{common-lisp}
-((:routes (paste :mapping "/paste/(.*)" "paste/\\1"))
+((:routes (blog :mapping "/blog/(.*)" "blog/\\1"))
  ...)
@@
 Doing so is troublesome, as an application's system might potentially modify any other part of the Lisp process.
-However, if the changes that can be rolled back is constrained in some sense, it should be possible to provide this kind of feature.
+However, if there is a constraint on what kind of changes can be rolled back, it should be possible to provide a usable form of this kind of feature.
 \\
@@
 This has much the same problems as the previous issue.
-Implementation switching is even more problematic though, since the system needs to ensure that all dependant applications, including optionally dependant ones, are reloaded to make macro redefinitions take effect.
+In addition, though, the system needs to ensure that all dependant applications, including optionally dependant ones, are reloaded to make macro redefinitions take effect.
 Furthermore, since some interfaces expose classes and instances thereof, the system would either have to be able to migrate objects between interface implementations, or somehow invalidate the old instances if they are retained in another part of the application.

### 2017-03-15 els2017-radiance/paper.tex
Minor corrections, fixme notes.

@@
 \begin{abstract}
+  %% FIXME
   Radiance\cite{radiance} is a set of libraries that provides an environment for web applications.
@@
 This makes it easier to visualise and think about the application structure.
-From the administrator's point of view this setup is more convenient too, as they simply need to think in terms of translations, rather than having to adapt potentially hundreds of different URL tags manually.
+From the administrator's point of view this setup is more convenient too, as they simply need to think in terms of translations, rather than specific tag instances.
 \\
@@
 
+%% FIXME
 Ideally, an administrator of a system will not have to touch any source code, and instead will be able to configure the system through a set of human-readable configuration files.
@@
 In order to change, say, the database interface's implementation, we would then have to modify the Radiance core's configuration file.
-We can either modify the file directly, which can be found through \code{(mconfig-pathname :radiance-core)}, like so:
+We can either modify the file directly like so:
 

### 2017-03-15 els2017-radiance/paper.tex
Where the fuck did the DOS line endings come from?
Fix some major issues. Paper is now almost ready.

@@
 \newmintinline[code]{common-lisp}{}
-\bibliographystyle{abbrv}
+\bibliographystyle{unsrt}
 
@@
 \begin{abstract}
-  %% FIXME
-  Radiance\cite{radiance} is a set of libraries that provides an environment for web applications.
-Unlike traditional web frameworks that focus on a set of tools to support the construction and maintenance of a single application, Radiance attempts to allow one to run as many differing applications as desired in the same environment.
-This goal brings many advantages, as frequently there are common resources in applications that should be shared between them, but cannot, because the applications are separated and developed without regard for sharing.
-In order to fix this, Radiance provides an extensible interface mechanism to standardise the interaction between users of common resources.
-To facilitate the separation and encapsulation of the shared URL address space between applications, Radiance also provides a flexible routing mechanism.
+  Radiance\cite{radiance} is a set of libraries that provide an environment for web applications.
+By putting its focus on running multiple web services or applications within the same environment, it has developed a features that set it apart from traditional web frameworks.
+In particular, it shows a new approach to the way different framework features are provided, and how the routing to content is performed.
+These differences allow applications written against Radiance to transparently share common resources such as the HTTP server, user accounts, the authentication, and so forth with each other.
 \end{abstract}
@@
 
-%% FIXME
-Ideally, an administrator of a system will not have to touch any source code, and instead will be able to configure the system through a set of human-readable configuration files.
-It is, of course, still possible to configure the system through usual Lisp code, should one prefer this approach.
+An administrator of a system will not have to touch any source code, and instead can configure the system through a set of human-readable configuration files.
+It is, of course, still possible to configure the system through usual Lisp code as well, should one prefer this approach.
 \\
@@
 
+\section{Acknowledgements}
+Thanks to Joram Shrijver, Till Ehrengruber, and Fernando Borretti for feedback and suggestions on the paper.
+
 \bibliography{paper}

### 2017-03-15 els2017-radiance/paper.tex
Style/grammar corrections.

@@
   Radiance\cite{radiance} is a set of libraries that provide an environment for web applications.
-By putting its focus on running multiple web services or applications within the same environment, it has developed a features that set it apart from traditional web frameworks.
+By putting its focus on running multiple web services or applications within the same environment, it has developed features that set it apart from traditional web frameworks.
 In particular, it shows a new approach to the way different framework features are provided, and how the routing to content is performed.
-These differences allow applications written against Radiance to transparently share common resources such as the HTTP server, user accounts, the authentication, and so forth with each other.
+These differences allow applications written against Radiance to transparently share common resources such as the HTTP server, user accounts, authentication, and so forth with each other.
 \end{abstract}
@@
 \section{Introduction}
-As the internet evolved, served websites began to evolve and became more and more dynamic.
+As the internet evolved, websites began to evolve and became more and more dynamic.
 Content is no longer a set of static webpages, but rather automatically updated, or even individually composed for the specific user viewing the website.
@@
 Additionally, it needs to have some system that divides up the URL namespace between applications.
-As the application writer, we cannot have any presumptions about what the final setup will look like-- what kinds of applications will be deployed together, and how the public URLs should resolve.
+As the application writer, we cannot have any presumptions about what the final setup will look like --- what kinds of applications will be deployed together, and how the public URLs should resolve.
 \\
@@
 They allow you to define the signatures of constructs exposed from a package, such as functions, macros, classes, and so forth.
-These function signatures are accompanied by a documentation that specifies their public behaviour.
+These function signatures are accompanied by documentation that specifies their public behaviour.
 Through this, Radiance can provide several ``standard interfaces'' that specify the behaviour of many features that are commonly needed in web applications, without actually having to implement them.
@@
 
-In order to make this more concrete, let's look at an interface that might provide some form of caching mechanism.
+In order to make this more concrete, let's look at an interface that might provide some sort of caching mechanism.
 
@@
 Radiance extends ASDF's dependency resolution mechanism to allow for this kind of declaration.
-When the \\\code{blog-service} system is now loaded, Radiance will notice the interface dependency and resolve it to an actual system that implements the desired cache functionality.
+When the \\\code{blog-service} system is now loaded, Radiance will notice the interface dependency and resolve it to an actual system that implements the desired caching functionality.
 This resolution is defined in the currently active environment, and can thus be configured by the administrator of the Radiance installation.
@@
 If an implementation should have special needs, it can always bypass the interfaces and make direct use of whatever it might depend on.
-This interfaces approach does bring some benefits, however.
+This approach with interfaces does bring some benefits, however.
 For one, it allows the implementations to be as efficient as possible, as there is no intermediate layer of redirection.
@@
 A full list of the interfaces and their descriptions is available in the Radiance documentation\footnote{\url{https://github.com/Shirakumo/radiance/\#interface}}.
-Thus, Radiance can be used like a macro-framework, but does not load on any features unless specifically required.
+Thus, Radiance can be used like a macro-framework, but does not load any features unless specifically required.
 Additionally, any of the implementations can be exchanged by a system administrator for ones that more closely match their requirements, without having to change anything about the application code.\\
@@
 
-The translation between the two namespaces is the responsibility of the routing system, which is illustrated in \hyperref[requestcycle]{Figure 1}. \\
+The translation between the two namespaces is the responsibility of the routing system. The way it is connected into the life-cycle of a request is illustrated in \hyperref[requestcycle]{Figure 1}. \\
 
@@
 It then sends it through the routing system's mapping functions, which turn this external URL into an internal one.
-The request is then dispatched to the according application that ``owns'' the URL in the internal representation.
+The request is then dispatched to the application that ``owns'' the URL in the internal representation.
 \\
@@
 With the mapping alone the case is not yet solved, though.
-When a page is emitted that contains links, the links too must be translated, but in the opposite way.
+When a page is emitted that contains links, the links must be translated as well, but in the opposite way.
 Unfortunately, because routes can be arbitrarily complex, it is not possible for the system to figure out a reversal route automatically.
@@
 The environment itself simply dictates a directory that contains all of these configuration files.
-Each application in the system automatically receives its own directory in which it can store both configuration, and temporary data files.
+Each application in the system automatically receives its own directory in which it can store both configuration and temporary data files.
 By simply switching out this environment directory, the system can then be loaded into completely different configurations, for example allowing you to easily keep ``development'' and ``production'' setups.
@@
 
-.. or instead use the programmatical way, like so:
+\dots or instead use the programmatical way, like so:
 
@@
 Doing so is troublesome, as an application's system might potentially modify any other part of the Lisp process.
-However, if there is a constraint on what kind of changes can be rolled back, it should be possible to provide a usable form of this kind of feature.
+However, if a constraint on what kind of changes can be rolled back is introduced, it should be possible to provide a usable form of this kind of feature.
 \\
@@
 \section{Acknowledgements}
-Thanks to Joram Shrijver, Till Ehrengruber, and Fernando Borretti for feedback and suggestions on the paper.
+Thanks to Joram Schrijver, Till Ehrengruber, and Fernando Borretti for feedback and suggestions on the paper.
 

### 2017-03-17 els2017-radiance/paper.tex
Incorporated feedback from Robert Strandh

@@
 Typically these frameworks are constructed with the intent of helping you develop a single web application, which is then deployed and run standalone.
-However, this can lead to issues when multiple applications should be deployed side-by-side.
+However, this paradigm can lead to issues when multiple applications should be deployed side-by-side.
 For example, if the framework does not provide explicit support for a particular feature such as user accounts, the two applications will likely have implemented ad-hoc solutions of their own, which are incompatible with each other and thus can't be trivially merged together.
@@
 Another primary goal of it was to write it in such a way, that it would be possible to exchange various parts of the framework without needing to change application code.
-This would allow flexibility on the side of the administrator, to pick and choose the parts that best fit their environment.
+This separation would allow flexibility on the side of the administrator, to pick and choose the parts that best fit their environment.
 
@@
 When talking about such an application in the context of Radiance, we think of it as a form of library that is then loaded together with Radiance to form a full webserver.
-This is in contrast to the deployment models of many other frameworks, where the ``library'' is deployed bundled tightly together with the framework as a single application.
+This approach is in contrast to the deployment models of many other frameworks, where the ``library'' is deployed bundled tightly together with the framework as a single application.
 \\
@@
 These function signatures are accompanied by documentation that specifies their public behaviour.
-Through this, Radiance can provide several ``standard interfaces'' that specify the behaviour of many features that are commonly needed in web applications, without actually having to implement them.
+Through this system, Radiance can provide several ``standard interfaces'' that specify the behaviour of many features that are commonly needed in web applications, without actually having to implement them.
 \\
 
-In order to make this more concrete, let's look at an interface that might provide some sort of caching mechanism.
+In order to make this system more concrete, let's look at an interface that might provide some sort of caching mechanism.
 
@@
 
-However, as it currently is, this will obviously not work.
+However, as it currently is, this function will obviously not work.
 The \code{with-caching} macro is not implemented with any useful functionality.
@@
 
-This is a particularly primitive and straight-forward implementation, in order to keep things short.
+This implementation is a particularly primitive and straight-forward one, in order to keep things short.
 The function and macro are provided by just overriding the stub definitions that were in place previously.
@@
 Using direct overwriting of definitions means that all applications must use the same implementation.
-However, usually this is the intended effect, as we want to maximise the sharing between them.
+However, usually this effect is intended, as we want to maximise the sharing between them.
 If an implementation should have special needs, it can always bypass the interfaces and make direct use of whatever it might depend on.
@@
 
-Finally, Radiance provides a system that allows to optionally depend on an interface.
-This is useful to model something like the user preferences mentioned in our example application.
+Finally, Radiance provides a system that allows the programmer to optionally depend on an interface.
+This mechanism is useful to model something like the user preferences mentioned in our example application.
 An administrator might not always want to provide an administration or settings panel.
 To make this dependency optional, we can defer the compilation and evaluation of our relevant logic to a later point, after an implementation has been loaded.
-For an imaginary \code{user-settings} interface, this might look like the following:
+For an imaginary \code{user-settings} interface, an example might look like the following:
 
@@
 In order to allow the administrator to change where pages are accessible from, an application cannot hard-code its resource locations and the links in its templates.
-This is doubly important when it comes to housing multiple applications in one, as the system needs to be set up in such a way that the applications do not clash with each other or potentially confuse pages of one another.
+It is doubly important to do this when it comes to housing multiple applications in one, as the system needs to be set up in such a way that the applications do not clash with each other or potentially confuse pages of one another.
 \\
 
-In many frameworks, like for example Symfony\cite{symfony}, this is solved by naming every resource in the system by a tag, and then allowing the configuration of what each tag resolves to individually.
+In many frameworks, like for example Symfony\cite{symfony}, this problem is solved by naming every resource in the system by a tag, and then allowing the configuration of what each tag resolves to individually.
 Radiance takes a different approach.
@@
 When a request is dispatched by Radiance, it first parses the request URL into an object presentation that makes it easier to modify.
-It then sends it through the routing system's mapping functions, which turn this external URL into an internal one.
+It then sends it through the mapping function of the routing system, which turn this external URL into an internal one.
 The request is then dispatched to the application that ``owns'' the URL in the internal representation.
@@
 If, as an administrator of an installation of this blog application, we now wanted to reach this page through the URL \texttt{www.example.com/blog/view}, the mapping route functions would have to strip away the \texttt{www.example.com} domain, recognise the \texttt{blog} folder, and put that as the URL's domain instead.
-We can achieve this relatively easily with the following definition.
+We can achieve this behaviour relatively easily with the following definition.
 
@@
 
-The templating engine takes care of reading out the \texttt{@} attribute values, running them through the routing system's reversal functions, and putting the resulting URL into the corresponding attribute to produce a valid link.
+The templating engine takes care of reading out the \texttt{@} attribute values, running them through the reversal functions of the routing system, and putting the resulting URL into the corresponding attribute to produce a valid link.
 
@@
 
-We need to stick this into a trigger, in order to defer the evaluation to when Radiance is being started up and the environment has been decided.
+We need to stick the configuration update into a trigger, in order to defer the evaluation to when Radiance is being started up and the environment has been decided.
 During the loading of the system, the environment might not have been set yet, and we would not be able to access the configuration storage.
@@
 While Radiance does provide default implementations for all of its interfaces, it is likely that some of them are not usable for a production setting.
-In order to change, say, the database interface's implementation, we would then have to modify the Radiance core's configuration file.
+In order to change, say, the implementation of the database interface, we would then have to modify the Radiance core's configuration file.
 We can either modify the file directly like so:
@@
 
-Every application's individual configuration can be changed in much the same way.
+The individual configuration of every application can be changed in much the same way.
 
@@
 Through the environment system, Radiance standardises the way each application is configured by the administrator and how the system is pieced together when it is loaded.
-As a consequence of that it becomes trivial to switch between different setups of a Radiance installation.
+As a consequence of that standardisation it becomes trivial to switch between different setups of a Radiance installation.
 The simple, human-readable configuration format used allows even users without intimate knowledge of Lisp to set up an instance.
@@
 Currently, while it is possible to dynamically load applications, it is not possible to dynamically unload them.
-Doing so is troublesome, as an application's system might potentially modify any other part of the Lisp process.
+Doing so is troublesome, as the system of an application might potentially modify any other part of the Lisp process.
 However, if a constraint on what kind of changes can be rolled back is introduced, it should be possible to provide a usable form of this kind of feature.
@@
 Radiance also does not allow you to change the implementation of an interface on the fly.
-This has much the same problems as the previous issue.
+This feature has much the same problems as the previous issue.
 In addition, though, the system needs to ensure that all dependant applications, including optionally dependant ones, are reloaded to make macro redefinitions take effect.
@@
 \section{Acknowledgements}
-Thanks to Joram Schrijver, Till Ehrengruber, and Fernando Borretti for feedback and suggestions on the paper.
+Thanks to Joram Schrijver, Till Ehrengruber, and Robert Strandh for feedback, corrections, and suggestions on the paper.
 

### 2018-02-21 els2018-glsl-oop/paper.tex
Start working on els2018 paper

\documentclass{sig-alternate}
\usepackage[english]{babel}
\usepackage{float}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{minted}
\usepackage[all]{hypcap}
\usemintedstyle[html]{borland}
\usemintedstyle[common-lisp]{pastie}
\newmintinline[code]{common-lisp}{}
\bibliographystyle{unsrt}

\hypersetup{
  colorlinks,
  linkcolor={red!50!black},
  citecolor={blue!50!black},
  urlcolor={blue!80!black}
}

\begin{document}

\setcopyright{rightsretained}
\doi{}
\isbn{}
\conferenceinfo{ELS'18}{April 16--17, 2018, Marbella, Spain}

\begin{CCSXML}
  <ccs2012>
  <concept>
  <concept_id>10011007.10010940.10010971.10010972.10010979</concept_id>
  <concept_desc>Software and its engineering~Object oriented architectures</concept_desc>
  <concept_significance>500</concept_significance>
  </concept>
  <concept>
  <concept_id>10011007.10010940.10010971.10011682</concept_id>
  <concept_desc>Software and its engineering~Abstraction, modeling and modularity</concept_desc>
  <concept_significance>500</concept_significance>
  </concept>
  <concept>
  <concept_id>10011007.10011006.10011066.10011067</concept_id>
  <concept_desc>Software and its engineering~Object oriented frameworks</concept_desc>
  <concept_significance>300</concept_significance>
  </concept>
  <concept>
  <concept_id>10011007.10011006.10011041.10011688</concept_id>
  <concept_desc>Software and its engineering~Parsers</concept_desc>
  <concept_significance>100</concept_significance>
  </concept>
  <concept>
  <concept_id>10010147.10010371</concept_id>
  <concept_desc>Computing methodologies~Computer graphics</concept_desc>
  <concept_significance>300</concept_significance>
  </concept>
  </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Object oriented architectures}
\ccsdesc[300]{Software and its engineering~Abstraction, modeling and modularity}
\ccsdesc[300]{Software and its engineering~Object oriented frameworks}
\ccsdesc[100]{Software and its engineering~Parsers}
\ccsdesc[300]{Computing methodologies~Computer graphics}

\title{Object Oriented GPU Programming Using CLOS}

\numberofauthors{1}
\author{
\alignauthor
Nicolas Hafner\\
       \affaddr{Shirakumo.org}\\
       \affaddr{Zürich, Switzerland}\\
       \email{shinmera@tymoon.eu}
}
\date{21st February 2018}

\maketitle

\begin{abstract}
  In this we present a new approach to applying object oriented principles to GPU programming via GLSL shaders. Shader code is attached to classes within standard Lisp code and automatically combined according to inheritance rules using the CLOS Meta Object Protocol. This results in a native coupling of both CPU and GPU logic, while preserving the advantages of object oriented inheritance and behaviour.
\end{abstract}

\printccsdesc

\keywords{Common Lisp, GLSL, OpenGL, GPU, CLOS, Object Orientation}
\newpage

\section{Introduction}
In modern real-time computer graphics applications, advanced effects and GPU computations require the use of a programmable graphics pipeline. This pipeline is usually programmable using a specific language that is compiled to GPU instructions by the graphics card drivers. \\

For OpenGL, this language is called GLSL\cite{rost2009opengl} and follows a mostly C-like syntax. A program written in GLSL is called a shader and is used to fill one of several stages in the graphics pipeline. Whenever a primitive is rendered, these shaders are executed on the GPU, each providing a particular stage of processing until finally an image is produced. \\

However, only a single shader can live in a particular stage at a time. This presents an issue for modularity, as effects represented by shaders cannot be easily combined. Instead, it is usually the task of a programmer to craft a single shader for each stage that produces the desired results. \\

In this paper we present a new approach to this problem that is accomplished in two steps. The first step is the parsing and manipulation of native GLSL code. The second step is the integration of shader code into the Common Lisp Object System to allow for automatic combination through inheritance.

\section{Related Work}
Several different approaches to shader combination exist today. \\

Trapp et al\cite{trapp2007automated} use additional constructs introduced to GLSL and a preprocessor to combine effects. This differs from our approach in that we do not extend GLSL syntax in any way and do not present a standard interface to use between shader fragments. \\

McCool et al.\cite{mccool2002shader} present an extension of the C++ language to allow writing GLSL-like code in C++ source. Shader fragments are parsed into an abstract syntax that can be used to perform static analysis and combination of fragments. Combination is however not automatic and needs to be explicitly requested by the programmer. \\

Kuck\cite{kuck2007object} presents an evolution of McCool's approach by allowing the use of C++ method declarations for shader function definitions. In this approach method dispatch is mirrored in the emitted GLSL code using unique identifiers for each class. \\

In VTK\cite{vtk} GLSL's ifdef precompiler macros are used to conditionally exclude or include certain shader functionality. This means that a full shader program of all possible combinations is meticulously hand-crafted, and features are then added or removed as needed by the program by prepending appropriate preprocessor definitions to the shader source. \\

A GLSL extension\cite{arbinclude} for an include preprocessor directive that would splice other shader source files in at the requested position, similar to C's include facility. This would merely allow controlling concatenation of source text within GLSL itself, though.

\section{GLSL Parsing and Manipulation}

\section{Shader Integration with CLOS}

\section{Conclusion}

\section{Further Work}

\section{Acknowledgements}

\bibliography{paper}
\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-command-extra-options: "-shell-escape"
%%% TeX-master: t
%%% TeX-engine: luatex
%%% End:

### 2018-02-21 els2018-glsl-oop/paper.tex
First full draft.

@@
 \usepackage{minted}
+\usepackage{hyperref}
 \usepackage[all]{hypcap}
@@
 }
+
+\def\abovecaptionskip{1pt}
+\def\listingautorefname{listing}
 
@@
 
-\title{Object Oriented GPU Programming Using CLOS}
+\title{Object Oriented Shader Composition Using CLOS}
 
@@
 
-\section{Introduction}
+\section{Introduction}\label{section:1}
 In modern real-time computer graphics applications, advanced effects and GPU computations require the use of a programmable graphics pipeline.
-This pipeline is usually programmable using a specific language that is compiled to GPU instructions by the graphics card drivers.
-\\
+This pipeline is usually programmable using domain-specific languages.
+The programmer passes the programs as textual source code to the graphics driver, which then compiles it down to GPU instructions.\\
 
@@
 
-\section{Related Work}
+\section{Related Work}\label{section:2}
 Several different approaches to shader combination exist today. \\
@@
 
-\section{GLSL Parsing and Manipulation}
+\section{GLSL Parsing and Manipulation}\label{section:3}
+Typically a GLSL program will have the following form:
 
-\section{Shader Integration with CLOS}
+\begin{listing}[h]
+\begin{minted}[fontsize=\small]{glsl}
+uniform mat4 transform_matrix;
+layout (location = 0) in vec3 position;
+out vec4 vertex;
 
-\section{Conclusion}
+void main(){
+  vertex = transform_matrix * vec4(position, 1.0);
+}
+\end{minted}
+\caption{A small example of a vertex shader computing the vertex position based on a transform matrix.}
+\label{lst:vertex}
+\end{listing}
 
-\section{Further Work}
+More specifically, it consists of a set of variable declarations that are either \code{uniform} (exchangeable with the CPU), \code{in} (coming from the previous stage), or \code{out} (going out to the next stage), and a set of function definitions.
+One required function is \code{main}, which is the entry point to the shader.
+\\
 
-\section{Acknowledgements}
+One goal of our approach was to allow the user to keep on using existing GLSL shaders, ideally without having to change anything about them.
+In order to accomplish merging of shader fragments with these constraints, the system needs to be able to automatically rewrite shader code to resolve name conflicts and combine effects.
+\\
+
+In order to do this, a full parser for the GLSL language was implemented in Lisp that turns the textual representation into an AST.
+Based on this AST, code walking and semantic analysis can be performed in order to detect variable definitions, function definitions, and behaviour.
+Using the analysis of two code segments, matching variable declarations can be fused together and their names in the code rewritten as appropriate.
+The main functions can be rewritten and called sequentially in a newly emitted main function.
+\\
+
+For instance, combining the shaders of \autoref{lst:vertex} and \autoref{lst:warp} results in \autoref{lst:combined}. \\
+
+\begin{listing}[H]
+\begin{minted}[fontsize=\small]{glsl}
+layout (location = 0) in vec3 vertex_data;
+out vec4 vertex;
+
+void main(){
+  vertex += sin(vertex_data.x);
+}
+\end{minted}
+\caption{A fragment shader that warps the vertex position.}
+\label{lst:warp}
+\end{listing}
+
+\begin{listing}[h]
+\begin{minted}[fontsize=\small]{glsl}
+uniform mat4 transform_matrix;
+layout (location = 0) in vec3 position;
+out vec4 vertex;
+
+void _GLSLTK_main_1(){
+  vertex = transform_matrix * vec4(position, 1.0);
+}
+
+void _GLSLTK_main_2(){
+  vertex += sin(position.x);
+}
+
+void main(){
+  _GLSLTK_main_1();
+  _GLSLTK_main_2();
+}
+\end{minted}
+\caption{A combination of the shaders from \autoref{lst:vertex} and \autoref{lst:warp}.}
+\label{lst:combined}
+\end{listing}
+
+The system automatically recognises that the \code{vertex} variable is the same in both shaders and omits the second instance.
+It also recognises that the \code{position} and \code{vertex_data} variables denote the same input, omits the second declaration, and renames the variable references to match the first declaration.
+\\
+
+Using this, a wide variety of shaders can be combined, with a minimal amount of awareness of other shaders in the merge being necessary.
+Shortcomings of this technique are elaborated in \autoref{section:5} and \autoref{section:6}.
+
+\section{Shader Integration with CLOS}\label{section:4}
+Usually the shaders do not act on their own.
+They need corresponding CPU-side code in order to provide the proper inputs and parameters for the pipeline to execute properly.
+Thus our system couples the definition of relevant shader code and CPU code together in a single class.
+Combination of features is then automatically provided through the inheritance of classes.
+\\
+
+A new metaclass is implemented that includes a slot for a set of shader fragments, each grouped according to their targeted pipeline stage.
+Shader fragments can then be attached to an instance of this metaclass to form the direct shader fragments.
+When the class hierarchy is finalised, shader fragments from all transitive superclasses are gathered in order of the class-precedence-list to form the list of effective shader fragments.
+This list is then combined into a full shader for each stage using the technique mentioned in \autoref{section:2}.
+\\
+
+Accompanied by this is a set of generic functions that allow specifying the loading and drawing logic of a class.
+These functions encompass the CPU-side code required to run the shaders properly.
+Using the standard method combination, the behaviour can be passed down and combined alongside the shader code.
+An example of such a behaviour is the flat colouring of vertices, as illustrated in \autoref{lst:colored-entity}.\\
+
+\begin{listing}[h]
+\begin{minted}[fontsize=\small]{common-lisp}
+(defclass colored-entity ()
+  ((color :initform (vec 0 0 1 1) :reader color))
+  (:metaclass shader-class))
+
+(defmethod paint :before ((obj colored-entity))
+  (let ((shader (shader-program obj)))
+    (setf (uniform shader "objectcolor") (color obj))))
+
+(define-class-shader (colored-entity :fragment-shader)
+  "uniform vec4 objectcolor;
+out vec4 color;
+
+void main(){
+  color *= objectcolor;
+}")
+\end{minted}
+\caption{A \code{colored-entity} class that encompasses vertex coloring functionality.}
+\label{lst:colored-entity}
+\end{listing}
+
+This \code{colored-entity} class can now simply be specified as a superclass in order to inherit the full behaviour.
+In effect this approach allows us to encapsulate a variety of small behaviours in the form of mixins, which can then be combined to form a full implementation of an object to be drawn.
+
+\section{Conclusion}\label{section:5}
+Using source code analysis allows us to easily re-use shader code written independently of our Lisp program while retaining the ability to merge the effects of multiple shaders together.
+Integrating this into the object system combines the GPU and CPU logic in a single class, allowing the inheritance and re-use of both.
+\\
+
+While the merging works automatically for a large class of shaders, certain ambiguities cannot be automatically rectified and require user input.
+For instance, unless the names of \code{in} or \code{out} declarations match, or their location is explicitly specified, the static analysis cannot determine whether they denote the same thing.
+The shader effects that write to the same \code{out} must also be adapted to be modifying in order for the effects to combine, rather than override.
+Finally, if the declarations do denote the same thing, but differ in qualifiers such as type or allocation, the merger cannot automatically determine how to resolve this conflict.
+
+\section{Further Work}\label{section:6}
+The current merging strategy employed is rather simplistic.
+For instance, no attempt at recognising identical function definitions is made.
+The system could also be extended with a variety of static analysis algorithms to optimise the code ahead of time, or to provide errors and warnings about the code.
+\\
+
+While allowing the use of native GLSL code is handy, providing the user with a system to write more native Lisp-like code for their shaders would improve the system quite a bit.
+To facilitate this, the Varjo\cite{varjo} compiler could be integrated.
+A further development from there could be integration with an IDE to provide the user with automated help information about shader functions.
+
+\section{Acknowledgements}\label{section:7}
+I would like to thank anyone reading this for their time and eventual feedback. I shall include you here by name before submitting the paper if you want to be named.
 

### 2018-02-21 els2018-glsl-oop/paper.tex
Second draft.

@@
 \usepackage[all]{hypcap}
-\usemintedstyle[html]{borland}
-\usemintedstyle[common-lisp]{pastie}
-\newmintinline[code]{common-lisp}{}
+\usemintedstyle[glsl]{default}
+\usemintedstyle[common-lisp]{default}
+\newmintinline[code]{text}{}
 \bibliographystyle{unsrt}
@@
 \begin{abstract}
-  In this we present a new approach to applying object oriented principles to GPU programming via GLSL shaders.
+  We present a new approach to applying object oriented principles to GPU programming via GLSL shaders.
 Shader code is attached to classes within standard Lisp code and automatically combined according to inheritance rules using the CLOS Meta Object Protocol.
 This results in a native coupling of both CPU and GPU logic, while preserving the advantages of object oriented inheritance and behaviour.
+The system allows the use of arbitrary shaders written directly in GLSL, without the need for language extensions.
 \end{abstract}
@@
 
-In this paper we present a new approach to this problem that is accomplished in two steps.
+In this paper we present a new solution to this problem that is accomplished in two steps.
 The first step is the parsing and manipulation of native GLSL code.
@@
 In this approach method dispatch is mirrored in the emitted GLSL code using unique identifiers for each class.
+This system is thus intended as a complete carry-over into shader code, rather than a simple combination of behaviour.
 \\
@@
 
-A GLSL extension\cite{arbinclude} for an include preprocessor directive that would splice other shader source files in at the requested position, similar to C's include facility.
+Khronos proposed a GLSL extension\cite{arbinclude} for an include preprocessor directive that would splice other shader source files in at the requested position, similar to C's include facility.
 This would merely allow controlling concatenation of source text within GLSL itself, though.
@@
 More specifically, it consists of a set of variable declarations that are either \code{uniform} (exchangeable with the CPU), \code{in} (coming from the previous stage), or \code{out} (going out to the next stage), and a set of function definitions.
-One required function is \code{main}, which is the entry point to the shader.
+The \code{main} function presents the entry point for the shader and is required.
 \\
@@
 One goal of our approach was to allow the user to keep on using existing GLSL shaders, ideally without having to change anything about them.
-In order to accomplish merging of shader fragments with these constraints, the system needs to be able to automatically rewrite shader code to resolve name conflicts and combine effects.
+In order to accomplish merging of shader fragments with these constraints, the system needs to be able to automatically rewrite shader code to resolve name conflicts and to combine effects.
 \\
@@
 In order to do this, a full parser for the GLSL language was implemented in Lisp that turns the textual representation into an AST.
-Based on this AST, code walking and semantic analysis can be performed in order to detect variable definitions, function definitions, and behaviour.
-Using the analysis of two code segments, matching variable declarations can be fused together and their names in the code rewritten as appropriate.
+Based on this AST, code walking and semantic analysis can be performed to detect definitions and rewrite behaviour.
+Using the analysis of two code segments, matching variable declarations can be fused together and their references in the code rewritten as appropriate.
 The main functions can be rewritten and called sequentially in a newly emitted main function.
@@
 
-Using this, a wide variety of shaders can be combined, with a minimal amount of awareness of other shaders in the merge being necessary.
-Shortcomings of this technique are elaborated in \autoref{section:5} and \autoref{section:6}.
+Using this, a wide variety of shaders can be combined, with a minimal amount of awareness of other shaders being necessary. Shortcomings of this technique are elaborated in \autoref{section:5}.
 
@@
 Usually the shaders do not act on their own.
-They need corresponding CPU-side code in order to provide the proper inputs and parameters for the pipeline to execute properly.
+They need corresponding CPU-side code in order to provide the inputs and parameters for the pipeline to execute properly.
 Thus our system couples the definition of relevant shader code and CPU code together in a single class.
@@
 
-A new metaclass is implemented that includes a slot for a set of shader fragments, each grouped according to their targeted pipeline stage.
+A new metaclass is implemented that includes a slot for a set of shader fragments, each grouped according to their pipeline stage.
 Shader fragments can then be attached to an instance of this metaclass to form the direct shader fragments.
-When the class hierarchy is finalised, shader fragments from all transitive superclasses are gathered in order of the class-precedence-list to form the list of effective shader fragments.
-This list is then combined into a full shader for each stage using the technique mentioned in \autoref{section:2}.
+When the class hierarchy is finalised, shader fragments from all transitive superclasses are gathered in order of class precedence to form the list of effective shader fragments.
+This list is then combined into a full shader for each stage using the technique from \autoref{section:2}.
 \\
@@
 Using the standard method combination, the behaviour can be passed down and combined alongside the shader code.
-An example of such a behaviour is the flat colouring of vertices, as illustrated in \autoref{lst:colored-entity}.\\
+An example of such a behaviour can be seen in \autoref{lst:colored-entity}.\\
 
@@
 
-This \code{colored-entity} class can now simply be specified as a superclass in order to inherit the full behaviour.
+This \code{colored-entity} class can now be used as a superclass in order to inherit the full behaviour.
 In effect this approach allows us to encapsulate a variety of small behaviours in the form of mixins, which can then be combined to form a full implementation of an object to be drawn.
@@
 \section{Conclusion}\label{section:5}
-Using source code analysis allows us to easily re-use shader code written independently of our Lisp program while retaining the ability to merge the effects of multiple shaders together.
+Source code analysis allows us to easily re-use shader code written independently of our Lisp program while retaining the ability to merge the effects of multiple shaders together.
 Integrating this into the object system combines the GPU and CPU logic in a single class, allowing the inheritance and re-use of both.
@@
 For instance, no attempt at recognising identical function definitions is made.
-The system could also be extended with a variety of static analysis algorithms to optimise the code ahead of time, or to provide errors and warnings about the code.
+The system could also be extended with a variety of static analysis algorithms to optimise the code ahead of time, or to provide errors and warnings about potential user mistakes.
 \\
 
-While allowing the use of native GLSL code is handy, providing the user with a system to write more native Lisp-like code for their shaders would improve the system quite a bit.
+While allowing the use of native GLSL code is handy, providing the user with a system to write shaders in Lisp syntax would improve the system quite a bit.
 To facilitate this, the Varjo\cite{varjo} compiler could be integrated.
-A further development from there could be integration with an IDE to provide the user with automated help information about shader functions.
+A further development from there could be integration with an IDE to provide the user with automated help information about not just Lisp, but also shader functions.
 
+\newpage
 \section{Acknowledgements}\label{section:7}
 I would like to thank anyone reading this for their time and eventual feedback. I shall include you here by name before submitting the paper if you want to be named.
+
+\section{Implementation}\label{section:8}
+An implementation of the proposed system can be found at \href{https://github.com/Shirakumo/trial/blob/e37e0dc73085c401e43da58d5098a9cf02167f8f/shader-entity.lisp}{https://github.com/Shirakumo/trial/blob/\\e37e0dc73085c401e43da58d5098a9cf02167f8f/shader-entity.lisp}.
+A more in-depth discussion can be found at \\\url{https://reader.tymoon.eu/article/362}.
 

### 2018-02-21 els2018-glsl-oop/paper.tex
Incorporate corrections based on Robert Strandh and Philipp Marek's feedback.

@@
   We present a new approach to applying object oriented principles to GPU programming via GLSL shaders.
-Shader code is attached to classes within standard Lisp code and automatically combined according to inheritance rules using the CLOS Meta Object Protocol.
-This results in a native coupling of both CPU and GPU logic, while preserving the advantages of object oriented inheritance and behaviour.
+Using the CLOS Meta Object Protocol, shader code can be attached to classes within standard Lisp code and is automatically combined by the system according to inheritance rules.
+This approach results in a native coupling of both CPU and GPU logic, while preserving the advantages of object oriented inheritance and behaviour.
 The system allows the use of arbitrary shaders written directly in GLSL, without the need for language extensions.
@@
 
-\keywords{Common Lisp, GLSL, OpenGL, GPU, CLOS, Object Orientation}
+\keywords{Common Lisp, GLSL, OpenGL, GPU, CLOS, MOP, Object Orientation}
 \newpage
@@
 In modern real-time computer graphics applications, advanced effects and GPU computations require the use of a programmable graphics pipeline.
-This pipeline is usually programmable using domain-specific languages.
+This is usually done using domain-specific languages.
 The programmer passes the programs as textual source code to the graphics driver, which then compiles it down to GPU instructions.\\
@@
 However, only a single shader can live in a particular stage at a time.
-This presents an issue for modularity, as effects represented by shaders cannot be easily combined.
+This limitation presents an issue for modularity, as effects represented by shaders cannot be easily combined.
 Instead, it is usually the task of a programmer to craft a single shader for each stage that produces the desired results.
@@
 Trapp et al\cite{trapp2007automated} use additional constructs introduced to GLSL and a preprocessor to combine effects.
-This differs from our approach in that we do not extend GLSL syntax in any way and do not present a standard interface to use between shader fragments.
+Their approach differs from ours in that we do not extend GLSL syntax in any way and do not present a standard interface to use between shader fragments.
 \\
@@
 
-In VTK\cite{vtk} GLSL's ifdef precompiler macros are used to conditionally exclude or include certain shader functionality.
-This means that a full shader program of all possible combinations is meticulously hand-crafted, and features are then added or removed as needed by the program by prepending appropriate preprocessor definitions to the shader source.
+In VTK\cite{vtk} GLSL's \code{ifdef} preprocessor macros are used to conditionally exclude or include certain shader functionality.
+This use of preprocessor macros means that a full shader program of all possible combinations is meticulously hand-crafted, and features are then added or removed as needed by the program by prepending appropriate preprocessor definitions to the shader source.
 \\
 
-Khronos proposed a GLSL extension\cite{arbinclude} for an include preprocessor directive that would splice other shader source files in at the requested position, similar to C's include facility.
-This would merely allow controlling concatenation of source text within GLSL itself, though.
+Khronos proposed a GLSL extension\cite{arbinclude} for an \code{include} preprocessor directive that would splice other shader source files in at the requested position, similar to C's include facility.
+This extension would merely allow controlling concatenation of source text within GLSL itself, though.
 
@@
 
-In order to do this, a full parser for the GLSL language was implemented in Lisp that turns the textual representation into an AST.
-Based on this AST, code walking and semantic analysis can be performed to detect definitions and rewrite behaviour.
-Using the analysis of two code segments, matching variable declarations can be fused together and their references in the code rewritten as appropriate.
+In order to do this, we implemented a full parser for the GLSL language in Lisp that turns the textual representation into an AST.
+Based on this AST, code walking and semantic analysis can be performed to detect definitions and to rewrite behaviour.
+By analysing two code segments, matching variable declarations can be fused together and their references in the code rewritten as appropriate.
 The main functions can be rewritten and called sequentially in a newly emitted main function.
@@
 
-Using this, a wide variety of shaders can be combined, with a minimal amount of awareness of other shaders being necessary. Shortcomings of this technique are elaborated in \autoref{section:5}.
+Using this technique, a wide variety of shaders can be combined, with a minimal amount of awareness of other shaders being necessary.
+Shortcomings of this technique are elaborated in \autoref{section:5}.
 
@@
 
-A new metaclass is implemented that includes a slot for a set of shader fragments, each grouped according to their pipeline stage.
+We implement a new metaclass that includes a slot for a set of shader fragments, each grouped according to their pipeline stage.
 Shader fragments can then be attached to an instance of this metaclass to form the direct shader fragments.
@@
 
-Accompanied by this is a set of generic functions that allow specifying the loading and drawing logic of a class.
+Accompanied by this metaclass is a set of generic functions that allow specifying the loading and drawing logic of a class.
 These functions encompass the CPU-side code required to run the shaders properly.
@@
 \end{minted}
-\caption{A \code{colored-entity} class that encompasses vertex coloring functionality.}
+\caption{A \code{colored-entity} class that encompasses vertex colouring functionality.}
 \label{lst:colored-entity}
@@
 In effect this approach allows us to encapsulate a variety of small behaviours in the form of mixins, which can then be combined to form a full implementation of an object to be drawn.
+For instance, we could imagine an entity to emit geometry for a sphere, an entity to apply a wave deformation, an entity to texture the geometry, an entity to apply shading, and so forth.
 
@@
 Source code analysis allows us to easily re-use shader code written independently of our Lisp program while retaining the ability to merge the effects of multiple shaders together.
-Integrating this into the object system combines the GPU and CPU logic in a single class, allowing the inheritance and re-use of both.
+Integrating this analysis into the object system combines the GPU and CPU logic in a single class, allowing the inheritance and re-use of both.
 \\
@@
 For instance, unless the names of \code{in} or \code{out} declarations match, or their location is explicitly specified, the static analysis cannot determine whether they denote the same thing.
-The shader effects that write to the same \code{out} must also be adapted to be modifying in order for the effects to combine, rather than override.
+The shader effects that write to the same \code{out} variable must also be adapted to re-use the existing value in order for the effects to combine.
+For instance, a colour should be combined by multiplying with the previous value, rather than directly setting the new one.
 Finally, if the declarations do denote the same thing, but differ in qualifiers such as type or allocation, the merger cannot automatically determine how to resolve this conflict.
@@
 While allowing the use of native GLSL code is handy, providing the user with a system to write shaders in Lisp syntax would improve the system quite a bit.
-To facilitate this, the Varjo\cite{varjo} compiler could be integrated.
+To facilitate this change, the Varjo\cite{varjo} compiler could be integrated.
 A further development from there could be integration with an IDE to provide the user with automated help information about not just Lisp, but also shader functions.
 
-\newpage
 \section{Acknowledgements}\label{section:7}
-I would like to thank anyone reading this for their time and eventual feedback. I shall include you here by name before submitting the paper if you want to be named.
+I would like to thank Robert Strandh and Philipp Marek for feedback on the paper.
 

### 2018-02-21 els2018-glsl-oop/paper.tex
Forgot to link to glsl-toolkit.

@@
 \end{minted}
-\caption{A \code{colored-entity} class that encompasses vertex colouring functionality.}
+\caption{A \code{colored-entity} class that encompasses object colouring functionality.}
 \label{lst:colored-entity}
@@
 \section{Implementation}\label{section:8}
-An implementation of the proposed system can be found at \href{https://github.com/Shirakumo/trial/blob/e37e0dc73085c401e43da58d5098a9cf02167f8f/shader-entity.lisp}{https://github.com/Shirakumo/trial/blob/\\e37e0dc73085c401e43da58d5098a9cf02167f8f/shader-entity.lisp}.
-A more in-depth discussion can be found at \\\url{https://reader.tymoon.eu/article/362}.
+An implementation of the proposed system can be found at \href{https://github.com/Shirakumo/trial/blob/e37e0dc73085c401e43da58d5098a9cf02167f8f/shader-entity.lisp}{https://github.com/Shirakumo/trial/blob/\\e37e0dc73085c401e43da58d5098a9cf02167f8f/shader-entity.lisp} for the CLOS part, and at \url{https://github.com/Shirakumo/glsl-toolkit} for the merger part.
+\\
+
+A more in-depth discussion of the system can be found at \\\url{https://reader.tymoon.eu/article/362}.
 

### 2018-03-24 els2018-glsl-oop/paper.tex
Minor
Minor corrections.

@@
 \section{GLSL Parsing and Manipulation}\label{section:3}
-Typically a GLSL program will have the following form:
+Typically a GLSL program will have form similar to this example:
 
@@
 void main(){
-  vertex += sin(vertex_data.x);
+  vertex.xyz += vec3(sin(vertex_data.x));
 }
@@
 When the class hierarchy is finalised, shader fragments from all transitive superclasses are gathered in order of class precedence to form the list of effective shader fragments.
-This list is then combined into a full shader for each stage using the technique from \autoref{section:2}.
+This list is then combined into a full shader for each stage using the technique from \autoref{section:3}.
 \\
@@
 For instance, a colour should be combined by multiplying with the previous value, rather than directly setting the new one.
-Finally, if the declarations do denote the same thing, but differ in qualifiers such as type or allocation, the merger cannot automatically determine how to resolve this conflict.
+Typically multiplying the colour will yield the desired combination effect, whereas setting it would simply discard the colour computed by other mixins.
+Finally, if the declarations differ in qualifiers such as type or allocation, the merger cannot automatically determine how to resolve this conflict, even if the authors of the shaders meant it to denote the same conceptual variable.
 

### 2018-04-01 els2018-glsl-oop/paper.tex
Minor
Include an image in section 3.

@@
 \usepackage{float}
+\usepackage[labelfont=bf,textfont=md]{caption}
 \usepackage{graphicx}
@@
 }
-
-\def\abovecaptionskip{1pt}
-\def\listingautorefname{listing}
 
@@
 \newpage
+
+\def\abovecaptionskip{1pt}
+\def\listingautorefname{listing}
+\def\figureautorefname{figure}
 
@@
 void main(){
-  vertex.xyz += vec3(sin(vertex_data.x));
+  vertex.x += sin(vertex_data.y);
 }
@@
 void _GLSLTK_main_2(){
-  vertex += sin(position.x);
+  vertex.x += sin(position.y);
 }
@@
 It also recognises that the \code{position} and \code{vertex_data} variables denote the same input, omits the second declaration, and renames the variable references to match the first declaration.
+An example rendering using \autoref{lst:vertex} and \autoref{lst:combined} can be seen in \autoref{fig:regular}.
 \\
+
+\begin{figure}[h]
+  \begin{center}
+    \begin{minipage}{.2\textwidth}
+      \includegraphics[width=0.9\textwidth]{render-regular.png}
+    \end{minipage}
+    \begin{minipage}{.2\textwidth}
+      \includegraphics[width=0.9\textwidth]{render-warped.png}
+    \end{minipage}
+  \end{center}
+  \caption{Left: rendering of a sphere using \autoref{lst:vertex}. Right: the same with \autoref{lst:warp} added.}
+  \label{fig:regular}
+\end{figure}
 

### 2018-04-04 els2018-glsl-oop/paper.tex
Add some stuff about applications.

@@
     \begin{minipage}{.2\textwidth}
-      \includegraphics[width=0.9\textwidth]{render-regular.png}
+      \includegraphics[width=1.0\textwidth]{render-regular.png}
     \end{minipage}
     \begin{minipage}{.2\textwidth}
-      \includegraphics[width=0.9\textwidth]{render-warped.png}
+      \includegraphics[width=1.0\textwidth]{render-warped.png}
     \end{minipage}
@@
 Using this technique, a wide variety of shaders can be combined, with a minimal amount of awareness of other shaders being necessary.
-Shortcomings of this technique are elaborated in \autoref{section:5}.
+Shortcomings of this technique are elaborated in \autoref{section:6}.
 
@@
 
-\section{Conclusion}\label{section:5}
+\section{Applications}\label{section:5}
+We currently use this technique for two primary applications: the encapsulation of common display functionality into mixins, and the implementation of effects passes.
+The former allows us to separate various concerns such as the read-out of texture data, the transformation of vertex data, and the fragment rendering behaviour into small, self-contained classes, that can then be combined by the user to produce the result they desire.
+\\
+
+The latter allows us to write arbitrary rendering effects as self-contained render pass objects.
+To elaborate on this application, imagine a visual effect that requires you to manipulate some, but not all of the aspects of how an object is rendered.
+A simple example for such an effect is the ``god rays'' that you can see in \autoref{fig:godrays}.
+\\
+
+In order to achieve this effect, all opaque objects in the scene are rendered once normally, and once in black.
+Then the black pass is blurred and added to the normal render.
+Neither the normal render pass nor the combination step require any insight from the pass into how the objects are drawn.
+However, for the black rendering pass, the way the objects are drawn has to be modified in such a way that all fragments are output as black, but everything else, such as vertex transformations, must remain the same.
+\\
+
+With shader combination we can achieve this quite easily: we write a fragment shader that simply outputs the colour black, and then combine this shader with whatever the objects' shaders are.
+In this way we are virtually inserting a superclass for each object in the scene depending on which render pass is currently active.
+Render passes themselves are implemented with the same class shader combination as regular objects, so they too can profit from encapsulating and sharing shader functionality.
+
+\begin{figure}[h]
+  \begin{center}
+    \begin{minipage}{.2\textwidth}
+      \includegraphics[width=1.0\textwidth]{pass-render.png}
+    \end{minipage}
+    \begin{minipage}{.2\textwidth}
+      \includegraphics[width=1.0\textwidth]{pass-black.png}
+    \end{minipage}
+    \vskip0.1cm
+    \begin{minipage}{.4055\textwidth}
+      \includegraphics[width=1.0\textwidth]{pass-combined.png}
+    \end{minipage}
+  \end{center}
+  \caption{Left: phong rendered teapot. Right: black rendered teapot. Bottom: combined god-rays effect.}
+  \label{fig:godrays}
+\end{figure}
+
+\section{Conclusion}\label{section:6}
 Source code analysis allows us to easily re-use shader code written independently of our Lisp program while retaining the ability to merge the effects of multiple shaders together.
@@
 
-\section{Further Work}\label{section:6}
+\section{Further Work}\label{section:7}
 The current merging strategy employed is rather simplistic.
@@
 
-\section{Acknowledgements}\label{section:7}
+\section{Acknowledgements}\label{section:8}
 I would like to thank Robert Strandh and Philipp Marek for feedback on the paper.
 
-\section{Implementation}\label{section:8}
+\section{Implementation}\label{section:9}
 An implementation of the proposed system can be found at \href{https://github.com/Shirakumo/trial/blob/e37e0dc73085c401e43da58d5098a9cf02167f8f/shader-entity.lisp}{https://github.com/Shirakumo/trial/blob/\\e37e0dc73085c401e43da58d5098a9cf02167f8f/shader-entity.lisp} for the CLOS part, and at \url{https://github.com/Shirakumo/glsl-toolkit} for the merger part.

### 2018-04-08 els2018-glsl-oop/paper.tex
Update to new ACM stuff.

@@
-\documentclass{sig-alternate}
+\documentclass[format=sigconf]{acmart}
 \usepackage[english]{babel}
@@
 
+\setcopyright{rightsretained}
+\acmDOI{}
+\acmISBN{978-2-9557474-2-1}
+\acmConference[ELS'18]{the 11th European Lisp Symposium}{April 16--17 2018}{%
+  Marbella, Spain}
+
 \begin{document}
 
-\setcopyright{rightsretained}
-\doi{}
-\isbn{}
-\conferenceinfo{ELS'18}{April 16--17, 2018, Marbella, Spain}
+\title{Object Oriented Shader Composition Using CLOS}
+
+\author{Nicolas Hafner}
+\affiliation{%
+  \institution{Shirakumo.org}
+  \city{Zürich}
+  \country{Switzerland}
+}
+\email{shinmera@tymoon.eu}
+
+\begin{abstract}
+  We present a new approach to applying object oriented principles to GPU programming via GLSL shaders.
+Using the CLOS Meta Object Protocol, shader code can be attached to classes within standard Lisp code and is automatically combined by the system according to inheritance rules.
+This approach results in a native coupling of both CPU and GPU logic, while preserving the advantages of object oriented inheritance and behaviour.
+The system allows the use of arbitrary shaders written directly in GLSL, without the need for language extensions.
+\end{abstract}
 
@@
 
-\title{Object Oriented Shader Composition Using CLOS}
-
-\numberofauthors{1}
-\author{
-  \alignauthor
-  Nicolas Hafner\\
-  \affaddr{Shirakumo.org}\\
-  \affaddr{Zürich, Switzerland}\\
-  \email{shinmera@tymoon.eu}
-}
-\date{21st February 2018}
+\keywords{Common Lisp, GLSL, OpenGL, GPU, CLOS, MOP, Object Orientation}
 
@@
 
-\begin{abstract}
-  We present a new approach to applying object oriented principles to GPU programming via GLSL shaders.
-Using the CLOS Meta Object Protocol, shader code can be attached to classes within standard Lisp code and is automatically combined by the system according to inheritance rules.
-This approach results in a native coupling of both CPU and GPU logic, while preserving the advantages of object oriented inheritance and behaviour.
-The system allows the use of arbitrary shaders written directly in GLSL, without the need for language extensions.
-\end{abstract}
-
-\printccsdesc
-
-\keywords{Common Lisp, GLSL, OpenGL, GPU, CLOS, MOP, Object Orientation}
 \newpage
@@
 \section{Implementation}\label{section:9}
-An implementation of the proposed system can be found at \href{https://github.com/Shirakumo/trial/blob/e37e0dc73085c401e43da58d5098a9cf02167f8f/shader-entity.lisp}{https://github.com/Shirakumo/trial/blob/\\e37e0dc73085c401e43da58d5098a9cf02167f8f/shader-entity.lisp} for the CLOS part, and at \url{https://github.com/Shirakumo/glsl-toolkit} for the merger part.
+An implementation of the proposed system can be found at \\\href{https://github.com/Shirakumo/trial/blob/e37e0dc73085c401e43da58d5098a9cf02167f8f/shader-entity.lisp}{https://github.com/Shirakumo/trial/blob/\\e37e0dc73085c401e43da58d5098a9cf02167f8f/shader-entity.lisp} \\ for the CLOS part, and at \url{https://github.com/Shirakumo/glsl-toolkit} for the merger part.
 \\

### 2019-01-25 els2019-shader-pipeline/paper.tex
Start work on the new paper

\documentclass[format=sigconf]{acmart}
\usepackage[english]{babel}
\usepackage{float}
\usepackage[labelfont=bf,textfont=md]{caption}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{minted}
\usepackage{hyperref}
\usepackage[all]{hypcap}
\usemintedstyle[glsl]{default}
\usemintedstyle[common-lisp]{default}
\newmintinline[code]{text}{}
\bibliographystyle{unsrt}

\hypersetup{
  colorlinks,
  linkcolor={red!50!black},
  citecolor={blue!50!black},
  urlcolor={blue!80!black}
}

\setcopyright{rightsretained}
\acmDOI{}
\acmISBN{}
\acmConference[ELS'19]{the 12th European Lisp Symposium}{April 1--2 2019}{%
  Genova, Italy}

\begin{document}

\title{Shader Pipeline and Effect Encapsulation using CLOS}

\author{Nicolas Hafner}
\affiliation{%
  \institution{Shirakumo.org}
  \city{Zürich}
  \country{Switzerland}
}
\email{shinmera@tymoon.eu}

\begin{abstract}
  %% FIXME: Abstract
\end{abstract}

\begin{CCSXML}
  %% FIXME: CCS
\end{CCSXML}

\keywords{Common Lisp, GLSL, OpenGL, GPU, CLOS, Object Orientation}

\maketitle

\newpage

\def\abovecaptionskip{1pt}
\def\listingautorefname{listing}
\def\figureautorefname{figure}

\section{Introduction}\label{introduction}

\section{Related Work}\label{relatedwork}
Courreges\cite{gtav} presents an in-depth analysis of the rendering procedure employed by the modern, high-production game GTA V. It illustrates the many stages to produce a final image, as well as their data dependencies. \\

Harada et al.'s work on Forward+\cite{forward+}\cite{forward+talk} also clearly illustrates the need for systems that support multi-staged rendering pipelines with complex data interaction schemes. \\

Gyrling\cite{fibers} presents an overview of the techniques used to perform parallel rendering in Naughty Dog's commercial engine. Individual steps within a stage, render stages of a frame, and multiple frame renderings are divided up into many small jobs that can run in parallel and are synchronised using counters on a shared structure. \\

The case study of the Unity game engine by Messaoudi et al\cite{unity} shows the availability of a set of fixed rendering pipelines that can be customised in a very limited extent with custom shaders. However, these shaders must fit into Unity's existing lighting and overall rendering model. While Unity does allow building a custom pipeline via their Scriptable Rendering Pipeline\cite{unitycustom}, they do not seem to offer any specific encapsulation or modularity features. \\

\section{Overview}\label{overview}

\section{Passes}\label{passes}

\section{Pipelines}\label{pipelines}

\section{Allocation}\label{allocation}

\section{Conclusion}\label{conclusion}

\section{Further Work}\label{furtherwork}

\section{Acknowledgements}\label{acknowledgements}

\section{Implementation}\label{implementation}
An implementation of the proposed system can be found at
\\\href{https://github.com/Shirakumo/trial/blob/f34a79f0a6df21d1ed9259e85fbb3c7eed39352b/shader-pass.lisp}{https://github.com/Shirakumo/trial/blob/\\f34a79f0a6df21d1ed9259e85fbb3c7eed39352b/shader-pass.lisp}
\\\href{https://github.com/Shirakumo/trial/blob/f34a79f0a6df21d1ed9259e85fbb3c7eed39352b/pipeline.lisp}{https://github.com/Shirakumo/trial/blob/\\f34a79f0a6df21d1ed9259e85fbb3c7eed39352b/pipeline.lisp}
\\\url{https://github.com/Shinmera/flow} \\

A more in-depth discussion of the system can be found at
\\\url{https://reader.tymoon.eu/article/363}
\\\url{https://reader.tymoon.eu/article/364}

\bibliography{paper}
\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-command-extra-options: "-shell-escape"
%%% TeX-master: t
%%% TeX-engine: luatex
%%% End:

### 2019-01-25 els2019-shader-pipeline/paper.tex
Type some more crap

@@
 
+\setlength{\parskip}{1em}
+
 \begin{document}
@@
 
-\keywords{Common Lisp, GLSL, OpenGL, GPU, CLOS, Object Orientation}
+\keywords{Common Lisp, OpenGL, GPU, CLOS, Object Orientation}
 
@@
 \section{Introduction}\label{introduction}
+Modern graphics systems such as OpenGL Core and DirectX offer a lot of customisation to the programmer.
+Particularly, in order to render an image, they allow the programmer to supply code fragments (shaders) that are run directly on the GPU.
+These code fragments fill in steps of a fixed rendering pipeline that is executed on the GPU in order to transform vertex data into the pixels of an image.
+The pipeline for OpenGL is illustrated in \autoref{fig:openglpipeline}.
+
+\begin{figure}[h]
+  \begin{center}
+    \includegraphics[width=0.3\textwidth]{opengl-pipeline.png}
+  \end{center}
+  \caption{The stages of the OpenGL rendering pipeline. White boxes represent stages that can be customised with shader code.}
+  \label{fig:openglpipeline}
+\end{figure}
+
+For consistency, we refer to a step within the hardware rendering pipeline as a ``stage'', an invocation of the hardware pipeline as a ``pass'', and all invocations of the hardware pipeline to produce an image as a ``frame''.
+
+Each of the customisable stages only accepts a single shader for each pass, making it difficult to separate, encapsulate, and ultimately combine behaviour.
+Furthermore, the steps required in order to change the shaders and shader inputs can be non-trivial and expensive to execute.
+
+The complexity of managing the graphics state and the order of rendering can be massively complicated for modern requirements.
+Rendering a frame often requires a multitude of passes, each with their own parameters and shared data.
+The rendering of each object within a pass can also differ, leading to even more state that needs to be correctly managed.
+
+This results in a difficult challenge for modularity. We attempt to solve this challenge through several systems:
+
+\begin{itemize}
+\item A protocol for communicating information between a pass and the objects rendered within.
+\item A protocol to connect the inputs, outputs, and parameters of different passes.
+\item An algorithm to automatically allocate shared textures.
+\end{itemize}
 
@@
 Courreges\cite{gtav} presents an in-depth analysis of the rendering procedure employed by the modern, high-production game GTA V.
-It illustrates the many stages to produce a final image, as well as their data dependencies.
-\\
+It illustrates the many passes to produce a final image, as well as their data dependencies.
 
-Harada et al.'s work on Forward+\cite{forward+}\cite{forward+talk} also clearly illustrates the need for systems that support multi-staged rendering pipelines with complex data interaction schemes. \\
+Harada et al.'s work on Forward+\cite{forward+}\cite{forward+talk} also clearly illustrates the need for systems that support multi-pass rendering pipelines with complex data interaction schemes.
 
 Gyrling\cite{fibers} presents an overview of the techniques used to perform parallel rendering in Naughty Dog's commercial engine.
-Individual steps within a stage, render stages of a frame, and multiple frame renderings are divided up into many small jobs that can run in parallel and are synchronised using counters on a shared structure.
-\\
+Individual stages within a pass, render passes of a frame, and multiple frame renderings are divided up into many small jobs that can run in parallel and are synchronised using counters on a shared structure.
 
@@
 While Unity does allow building a custom pipeline via their Scriptable Rendering Pipeline\cite{unitycustom}, they do not seem to offer any specific encapsulation or modularity features.
-\\
+
+The work by He et al.\cite{components} introduces a framework for general encapsulation of shaders and their parameters into structures that minimise the overhead of changing GPU state while retaining the ability to dynamically compose shader parts.
+They do however not create a distinction between properties for the rendering of an object, and those for the rendering of an overall pass.
+
+Foley et al.'s work on Spark\cite{spark} presents a high-level graph-based system for defining reusable and composable shader components.
+Their system shows a much more distanced view of the underlying graphics hardware than we attempt.
+Similar to He's work, they do not present a separation between object logic and pass logic.
+
+In our previous work\cite{glsloop} we introduce a system to tie shader code to classes and compose behaviour through inheritance.
+We make use of this system and extend it to allow further control over rendering behaviour in individual passes.
 

### 2019-01-29 els2019-shader-pipeline/paper.tex
Some more stuff
Minor
Complete allocation section

@@
 \documentclass[format=sigconf]{acmart}
+\usepackage{enumitem}
 \usepackage[english]{babel}
@@
 
+\newlist{step}{enumerate}{10}
+\setlist[step]{label*=\arabic*.,leftmargin=2em}
+
 \setcopyright{rightsretained}
@@
   Genova, Italy}
-
-\setlength{\parskip}{1em}
 
@@
 The pipeline for OpenGL is illustrated in \autoref{fig:openglpipeline}.
+\\
 
@@
 For consistency, we refer to a step within the hardware rendering pipeline as a ``stage'', an invocation of the hardware pipeline as a ``pass'', and all invocations of the hardware pipeline to produce an image as a ``frame''.
+\\
 
@@
 Furthermore, the steps required in order to change the shaders and shader inputs can be non-trivial and expensive to execute.
+\\
+
+\begin{figure}[h]
+  \begin{center}
+    \includegraphics[width=0.3\textwidth]{render-pipeline.png}
+  \end{center}
+  \caption{A sample frame pipeline with shadow mapping, deferred rendering, and bloom. Each box represents a pass and each edge a texture buffer.}
+  \label{fig:renderpipeline}
+\end{figure}
 
@@
 The rendering of each object within a pass can also differ, leading to even more state that needs to be correctly managed.
+\\
 
@@
 \item A protocol to connect the inputs, outputs, and parameters of different passes.
-\item An algorithm to automatically allocate shared textures.
+\item An algorithm to automatically allocate shared textures used as buffers between passes.
 \end{itemize}
@@
 It illustrates the many passes to produce a final image, as well as their data dependencies.
+\\
 
-Harada et al.'s work on Forward+\cite{forward+}\cite{forward+talk} also clearly illustrates the need for systems that support multi-pass rendering pipelines with complex data interaction schemes.
+Harada et al.'s work on Forward+\cite{forward+}\cite{forward+talk} also clearly illustrates the need for systems that support multi-pass rendering pipelines with complex data interaction schemes. \\
 
@@
 Individual stages within a pass, render passes of a frame, and multiple frame renderings are divided up into many small jobs that can run in parallel and are synchronised using counters on a shared structure.
+\\
 
@@
 While Unity does allow building a custom pipeline via their Scriptable Rendering Pipeline\cite{unitycustom}, they do not seem to offer any specific encapsulation or modularity features.
+\\
 
@@
 They do however not create a distinction between properties for the rendering of an object, and those for the rendering of an overall pass.
+\\
 
@@
 Similar to He's work, they do not present a separation between object logic and pass logic.
+\\
 
@@
 \section{Allocation}\label{allocation}
+In order to prepare the shader pipeline, several resources need to be allocated.
+Each shader stage needs a ``framebuffer,'' an OpenGL resource that allows one to render to off-screen textures.
+These framebuffers then need to have the required textures to render onto allocated as well.
+As each input and output from a pass can specify constraints on the texture's features, these constraints must be matched up for any connecting edges as well.
+Finally, in order to minimise memory usage we would like to re-use textures where possible.
+
+Thus the allocation proceeds in three phases: reconciling texture constraints on edges between passes, computing how textures are shared between passes, and finally constructing all the necessary resources with the previously gathered information.
+
+\subsection{Constraint Merging}\label{texture-constraints}
+OpenGL textures include a massive amount of information.\cite{texparameter}\cite{teximage2d} When two ports are connected that specify different constraints on the texture properties, a join must be performed.
+Fortunately, a wide range of the texture property values are fundamentally incompatible, meaning that a lot of the logic can simply error.
+For simplicity and brevity, we will focus on the join operator for a single texture property here, the internal format.
+The internal format specifies how many colour channels the texture has, how many bits of precision each channel has, which format each channel has, as well as whether the texture is compressed or has sRGB gamma normalisation applied.
+\\
+
+The list of specified texture formats is quite large.\cite{texinternalformat} Unfortunately OpenGL does not give us an interface to handle these formats in a way that lets us pick the individual features easily.
+Instead, each format is represented by a constant number, with no relation to the features that format includes.
+This means that we need to first destructure each format into a list of its features:
+
+\begin{itemize}
+\item \code{R}, \code{G}, \code{B}, \code{A} How many bits to use for each channel, and the format of the channel (normalised, float, integer, unsigned integer).
+\item \code{depth} How many bits to use for the depth channel.
+\item \code{stencil} How many bits to use for the stencil channel.
+\item \code{shared} Whether bits are shared across the channel, and if so how many.
+\item \code{features} Whether the format has compression, sRGB, RGTC, BPTC, SNORM, or UNORM features.
+\end{itemize}
+
+Once the texture format specifications are destructured, we can perform a join as follows.
+
+\begin{step}
+\item If the \code{features} and sharing are not the same, a join is impossible.
+\item The \code{features} list of the output spec is set to the same as either of the specs.
+\item If both include a \code{depth} component:
+  \begin{step}
+  \item The \code{depth} feature is set to the join of both.
+  \item If either include a \code{stencil} feature, the \code{stencil} feature is set to the join of both.
+  \end{step}
+\item If both include a \code{stencil} component:
+  \begin{step}
+  \item The \code{stencil} feature is set to the join of both.
+  \end{step}
+\item If neither include a \code{depth} component:
+  \begin{step}
+  \item The \code{R} feature is set to the join of both.
+  \item The \code{G} feature is set to the join of both.
+  \item The \code{B} feature is set to the join of both.
+  \item The \code{A} feature is set to the join of both.
+  \end{step}
+\item Otherwise a join is impossible.
+\end{step}
+
+Wherein the ``join of both'' is computed as the join of two channel formats as follows.
+
+\begin{step}
+\item If both include the format:
+  \begin{step}
+  \item If the channel format is not the same, a join is impossible.
+  \item The channel bit depth is set to the maximum of both.
+  \end{step}
+\item If one includes the format, that format is returned.
+\item Otherwise, the absence of the channel is indicated.
+\end{step}
+
+If the join is successful, we then re-encode the texture format specification into OpenGL's constant and use it in the real texture specification.
+Note that this join could texture format specifications that are not legal according to the OpenGL specification.
+However, this can only occur if one of the specs to join is already illegal.
+We thus deem it unnecessary to handle such cases.
+
+\subsection{Port Allocation}\label{port-allocation}
+In compilers, allocation of a graph of variables with use-relations is typically handled with a graph colouring algorithm.
+However, since we represent our graph in a different fashion than usual, with nodes having multiple distinct ports on which edges are connected, we devised a different kind of colouring algorithm to maximise texture sharing.
+\\
+
+Given a set of nodes the allocation algorithm proceeds as follows.
+
+\begin{step}
+\item The set of unique texture specifications is computed by joining each port's texspec with every other and comparing for equality.
+\item For each unique texture specification T:
+  \begin{step}
+  \item The nodes are sorted topologically.
+  \item The number of colours is set to 0.
+  \item For each node N:
+    \begin{step}
+    \item For each output port P of N:
+      \begin{step}
+      \item If P's texspec is joinable with T...
+      \item The number of colours is increased
+      \end{step}
+    \end{step}
+  \item An array is allocated to fit the number of colours. Each index of the array represents a colour and each value at the index whether the colour is currently available or unavailable.
+  \item For each node N in \emph{reverse order}:
+    \begin{step}
+    \item For each input port P of N:
+      \begin{step}
+      \item For each neighbour port O of P:
+        \begin{step}
+        \item If O's texspec is joinable with T...
+        \item and O does not yet have a colour...
+        \item The first available colour is assigned to O.
+        \item This colour is marked as unavailable.
+        \end{step}
+      \end{step}
+    \item For each non-input port P of N:
+      \begin{step}
+      \item If P's texspec is joinable with T...
+      \item and P does not yet have a colour...
+      \item The first available colour is assigned to P.
+      \item This colour is marked as unavailable.
+      \end{step}
+    \item For each port P of N:
+      \begin{step}
+      \item If P's texspec is joinable with T...
+      \item and P has a colour...
+      \item P's colour is marked as available.
+      \end{step}
+    \end{step}
+  \end{step}
+\end{step}
+
+In other words, the algorithm proceeds backwards from the last node in the graph, marking predecessors' output ports with unique colours, then marking unconnected ports with unique colours, and finally marking all colours at its own output ports as available again.
+We repeat this process for each unique texture specification, each time ensuring we only consider ports that share that texture specification.
+\\
+
+This algorithm is by no means efficient, but since pipeline allocation only has to happen during loading phases, we currently do not consider this to be a big problem. \\
+
+Unfortunately print documents cannot yet display animations, so illustrating the algorithm in motion is not possible directly in the document.
+However, a brief animation of the pipeline illustrated in \autoref{fig:renderpipeline} is available online: \\
+
+\href{https://raw.githubusercontent.com/Shinmera/talks/master/els2019-shader-pipeline/pipeline-allocation.gif}{https://raw.githubusercontent.com/Shinmera/talks/master/\\els2019-shader-pipeline/pipeline-allocation.gif}
 
@@
 
-\section{Further Work}\label{furtherwork}
+\section{Further Work}\label{further-work}
 

### 2019-01-29 els2019-shader-pipeline/paper.tex
Add CCS

@@
 \begin{CCSXML}
-  %% FIXME: CCS
+  <ccs2012>
+  <concept>
+  <concept_id>10011007.10010940.10010971.10010972.10010979</concept_id>
+  <concept_desc>Software and its engineering~Object oriented architectures</concept_desc>
+  <concept_significance>500</concept_significance>
+  </concept>
+  <concept>
+  <concept_id>10011007.10010940.10010971.10011682</concept_id>
+  <concept_desc>Software and its engineering~Abstraction, modeling and modularity</concept_desc>
+  <concept_significance>500</concept_significance>
+  </concept>
+  <concept>
+  <concept_id>10011007.10011006.10011066.10011067</concept_id>
+  <concept_desc>Software and its engineering~Object oriented frameworks</concept_desc>
+  <concept_significance>300</concept_significance>
+  </concept>
+  <concept>
+  <concept_id>10010147.10010371</concept_id>
+  <concept_desc>Computing methodologies~Computer graphics</concept_desc>
+  <concept_significance>300</concept_significance>
+  </concept>
+  <concept>
+  <concept_id>10011007.10011006.10011041.10011688</concept_id>
+  <concept_desc>Software and its engineering~Parsers</concept_desc>
+  <concept_significance>100</concept_significance>
+  </concept>
+  </ccs2012>
 \end{CCSXML}
+
+\ccsdesc[500]{Software and its engineering~Object oriented architectures}
+\ccsdesc[300]{Software and its engineering~Abstraction, modeling and modularity}
+\ccsdesc[300]{Software and its engineering~Object oriented frameworks}
+\ccsdesc[300]{Computing methodologies~Computer graphics}
+\ccsdesc[100]{Software and its engineering~Compilers}
 

### 2019-01-30 els2019-shader-pipeline/paper.tex
Finish almost everything

@@
 These code fragments fill in steps of a fixed rendering pipeline that is executed on the GPU in order to transform vertex data into the pixels of an image.
-The pipeline for OpenGL is illustrated in \autoref{fig:openglpipeline}.
+The pipeline for OpenGL is illustrated in \autoref{fig:opengl-pipeline}.
 \\
@@
   \caption{The stages of the OpenGL rendering pipeline. White boxes represent stages that can be customised with shader code.}
-  \label{fig:openglpipeline}
+  \label{fig:opengl-pipeline}
 \end{figure}
@@
   \caption{A sample frame pipeline with shadow mapping, deferred rendering, and bloom. Each box represents a pass and each edge a texture buffer.}
-  \label{fig:renderpipeline}
+  \label{fig:render-pipeline}
 \end{figure}
@@
 \section{Overview}\label{overview}
+The system is composed out of three distinct entities: shader objects, shader passes, and pipelines.
+Both shader objects and shader passes encapsulate rendering logic.
+The pipeline on the other hand represents the assembly of a complete frame and only influences rendering in the sense that it tracks which passes to run in which order.
+\\
+
+Due to the restrictions of OpenGL's rendering pipeline, the combination of the render logic of objects and passes is not trivial.
+We present a protocol to solve this in \hyperref[passes]{section 4}.
+\\
+
+In \hyperref[pipelines]{section 5} we present a protocol to represent input and output information of a pass and how to connect these inputs and outputs between passes together. \\
+
+Due to the complex interactions between passes in a pipeline, allocation of intermediary buffer textures is error-prone and tedious.
+We present algorithms to automate this in \hyperref[allocation]{section 6}.
+\\
+
+Finally, in \hyperref[concept]{section 7} we show the results of an implementation of a medium-sized pipeline using these techniques.
 
 \section{Passes}\label{passes}
+A shader pass should encapsulate the logic for drawing objects in a certain way.
+In order to do this, it needs to be able to partially or fully control the shaders of objects.
+For instance, an object should still be able to have control over how its vertices are constructed and what materials and textures are applied to it.
+The shader pass needs to be able to make use of this information if it needs to, or discard it completely if unneeded.
+\\
+
+In order to allow this we make use of the shader composition capabilities that we presented in our previous work\cite{glsloop} that allows us to combine pieces of GLSL code, and a new protocol centred around shader passes.
+
+\begin{listing}[h]
+\begin{minted}[fontsize=\small]{common-lisp}
+(defclass shader-pass ()
+  ()
+  (:metaclass shader-pass-class))
+
+(defgeneric register-object-for-pass (shader-pass object))
+(defgeneric shader-program-for-pass (shader-pass object))
+(defgeneric make-pass-shader-program (shader-pass object))
+(defgeneric coerce-pass-shader (shader-pass object stage))
+(defgeneric render-with (shader-pass object))
+(defgeneric render (object target))
+\end{minted}
+\caption{The protocol for shader passes.}
+\label{lst:pass}
+\end{listing}
+
+For any object that should be rendered using a given pass, first \code{register-object-for-pass} must be called.
+This allows the pass to prepare the shader program that will be used during rendering for this object.
+A shader program is an OpenGL resource that compiles all shaders for a pass together.
+With \code{shader-program-for-pass} this program can then later be retrieved.
+The shader program is important for objects to access, as it allows them to set values for ``uniform'' variables that are used in the shader code.
+\\
+
+\code{register-object-for-pass} calls \code{make-pass-shader-program} to compute this program and registers it internally in the shader pass so that it can be retrieved later.
+If the shader pass would like to retain complete control over how each object is drawn, it would forego this call and instead generate its own shader program.
+This is especially useful for post-processing effects that don't render any objects at all, and instead simply operate on textures output by previous passes.
+\\
+
+\code{make-pass-shader-program} gathers all the shader sources for a program using \code{coerce-pass-shader} and then generates a representation for an OpenGL shader program.
+This representation also includes additional information such as the data buffers used.
+\\
+
+\code{coerce-pass-shader} computes the effective shader source for a particular shader stage or type.
+Typically this simply involves the combination of the shader sources of both shader pass and object for the given shader type.
+We perform this combination using the same parsing and code walking strategy as we described in our previous work.\cite{glsloop} \\
+
+Objects are typically rendered using the \code{render} function.
+Object classes are encouraged to add methods that specialise the behaviour and perform necessary setup, as well as to perform the final draw call.
+For instance, a very primitive class could look as shown in \autoref{lst:simple-object}.
+
+\begin{listing}[h]
+\begin{minted}[fontsize=\small]{common-lisp}
+(defclass simple-object ()
+  ((vertex-array :accessor vertex-array)
+  (:metaclass shader-class))
+
+(defmethod render ((oject simple-object) target)
+  (let ((vao (vertex-array object)))
+    (gl:bind-vertex-array (gl-name vao))
+    (%gl:draw-elements :triangles (size vao) :unsigned-int 0)))
+
+(defmethod render :before ((object simple-object)
+                           (pass shader-pass))
+  (let ((program (shader-program-for-pass pass object)))
+    (setf (uniform program "projection_matrix")
+          (projection-matrix))))
+\end{minted}
+\caption{A simple object class and its render methods. The first method tells OpenGL to render a list of vertices. The second method sends the projection matrix to the GPU via a uniform variable.}
+\label{lst:simple-object}
+\end{listing}
+
+This function is fine for allowing the object being drawn control over the behaviour.
+However, the pass cannot exert the same amount of control due to the generic function's argument precedence.
+We thus introduce another function with inverted argument order that is called first.
+\\
+
+\code{render-with} thus exists mostly as an entry point for rendering operations to allow shader passes greater control over the rendering.
+Typically it will simply defer to \code{render}.
+This inversion is very important for shader passes that take all control away from the objects.
+Without it, the \code{:before}, \code{:after}, and \code{:around} methods on \code{render} specialised on an object could try to manipulate state before methods specialised on a shader pass could prevent it.
+\\
+
+This protocol thus allows a great amount of control both over the effective shader code used to render an object, as well as the behaviour leading up towards the actual draw call for an object.
 
 \section{Pipelines}\label{pipelines}
+Shader passes not only retain information about how objects are drawn with it, but also about the input and output textures that a pass interacts with.
+For this reason, each shader pass is also a node in a graph, with distinct input and output ports.
+
+\begin{listing}[h]
+\begin{minted}[fontsize=\small]{common-lisp}
+(defclass deferred-render (shader-pass)
+  ((position-map :port-type input)
+   (normal-map :port-type input)
+   (albedo-map :port-type input)
+   (color :port-type output))
+  (:metaclass shader-pass-class)
+\end{minted}
+\caption{An outline of a deferred rendering pass, taking position-, normal-, and albedo-map textures as input, producing a single color texture as output.}
+\label{lst:deferred-render}
+\end{listing}
+
+An example of such a pass is illustrated in \autoref{lst:deferred-render}.
+The ports are modelled as slots on the class, requiring a new metaclass.
+A slot with a port-type will not only hold a texture for its value, but also retain information about how it is connected to other passes.
+Using slots in this manner also allows us to inherit ports from other classes, and thus combine behaviour alongside the shader source code.
+\\
+
+For the sake of brevity and ease of explanation, we have left out the details of texture constraints in these code samples.
+Nevertheless, constraining the features of textures tied to the ports is important, and we discuss the technique for dealing with that in \hyperref[allocation]{section 6}.
+
+\begin{listing}[h]
+\begin{minted}[fontsize=\small]{common-lisp}
+(defclass shadow-render (shader-pass)
+  ((shadow-map :port-type input)
+   (color :port-type output))
+  (:metaclass shader-pass-class))
+\end{minted}
+\caption{An outline of a shadow rendering pass, taking into account the information from a shadow-map to render shadows onto the output color texture.}
+\label{lst:shadow-render}
+\end{listing}
+
+\begin{listing}[h]
+\begin{minted}[fontsize=\small]{common-lisp}
+(defclass high-pass-render (shader-pass)
+  ((high-pass :port-type output))
+  (:metaclass shader-pass-class))
+\end{minted}
+\caption{An outline of a high-pass renderer, which splices off colours of a high intensity into a high-pass output texture.}
+\label{lst:high-pass}
+\end{listing}
+
+For instance, the ``Deferred Lighting'' pass shown in \autoref{fig:render-pipeline} is created by combining the passes illustrated in \autoref{lst:deferred-render}, \autoref{lst:shadow-render}, and \autoref{lst:high-pass} through inheritance.
+\\
+
+When assembling a pipeline, we can then connect these ports together in order to dictate how the various output textures generated by a pass are used as inputs for other passes.
+As an example, the pipeline from \autoref{fig:render-pipeline} is created in \autoref{lst:pipeline}.
+\\
+
+\begin{listing}[h]
+\begin{minted}[fontsize=\small]{common-lisp}
+(let ((pipeline (make-pipeline))
+      (shadow (make-instance 'shadow-map-pass))
+      (geometry (make-instance 'geometry-pass))
+      (deferred (make-instance 'deferred-lighting-pass))
+      (blur (make-instance 'gaussian-blur-pass))
+      (composite (make-instance 'composite-pass)))
+  (connect (port shadow 'shadow-map) 
+           (port deferred 'shadow-map) pipeline)
+  (connect (port geometry 'position-map)
+           (port deferred 'position-map) pipeline)
+  (connect (port geometry 'normal-map)
+           (port deferred 'normal-map) pipeline)
+  (connect (port geometry 'albedo-map)
+           (port deferred 'albedo-map) pipeline)
+  (connect (port deferred 'high-pass)
+           (port blur 'previous-pass) pipeline)
+  (connect (port deferred 'color)
+           (port composite 'color) pipeline)
+  (connect (port blur 'color)
+           (port composite 'bloom) pipeline)
+  (prepare pipeline))
+\end{minted}
+\caption{An assembly of the pipeline shown in \autoref{fig:render-pipeline}.}
+\label{lst:pipeline}
+\end{listing}
+
+The pipeline object itself retains information on the order in which the passes should be executed, and keeps track of the textures that are allocated in order to run the passes.
+How this information is computed is described in \hyperref[allocation]{section 6}.
+\\
+
+The procedure to actually render objects with this pipeline technique is then as follows:
+
+\begin{step}
+\item Create a pipeline object and instances of the desired shader passes.
+\item Connect the ports of the shader passes.
+\item Prepare the pipeline to allocate the needed resources.
+\item Create instances of the desired objects.
+\item Register each object with each pass.
+\item Call \code{render} on the pipeline with a collection of all objects to draw.
+\item The output port's texture of the last shader pass in the pipeline will contain the finished frame.
+\end{step}
+
+The completed frame can then be blitted onto the screen, or be used as the input for another computation.
 
@@
 Finally, in order to minimise memory usage we would like to re-use textures where possible.
+\\
 
@@
 Unfortunately print documents cannot yet display animations, so illustrating the algorithm in motion is not possible directly in the document.
-However, a brief animation of the pipeline illustrated in \autoref{fig:renderpipeline} is available online: \\
+However, a brief animation of the pipeline illustrated in \autoref{fig:render-pipeline} is available online: \\
 
@@
 
+\section{Proof of Concept}\label{concept}
+As a proof of concept we have implemented these protocols and mechanisms, as well as the passes shown in \autoref{fig:render-pipeline}.
+As the code required to do so is quite lengthy and involved, and the graphics techniques used are beyond the scope of this paper, we will omit them.
+The relevant outline of the passes and the pipeline involved has already been shown in \hyperref[pipelines]{section 5}.
+\\
+
+Nevertheless, you can find the complete code online at the following pages: \\
+\href{https://github.com/Shirakumo/trial/blob/b889d50b7906348ad787e773ba57998bb39b16a5/deferred.lisp}{https://github.com/Shirakumo/trial/blob/\\b889d50b7906348ad787e773ba57998bb39b16a5/deferred.lisp} \\
+\href{https://github.com/Shirakumo/trial/blob/b889d50b7906348ad787e773ba57998bb39b16a5/hdr.lisp}{https://github.com/Shirakumo/trial/blob/\\b889d50b7906348ad787e773ba57998bb39b16a5/hdr.lisp} \\
+\href{https://github.com/Shirakumo/trial/blob/b889d50b7906348ad787e773ba57998bb39b16a5/shadow-map.lisp}{https://github.com/Shirakumo/trial/blob/\\b889d50b7906348ad787e773ba57998bb39b16a5/shadow-map.lisp} \\
+\href{https://github.com/Shirakumo/trial/blob/b889d50b7906348ad787e773ba57998bb39b16a5/workbench.lisp}{https://github.com/Shirakumo/trial/blob/\\b889d50b7906348ad787e773ba57998bb39b16a5/workbench.lisp} \\
+
+\autoref{fig:pipeline-gbuffer}, \autoref{fig:pipeline-shadow}, \autoref{fig:pipeline-lighting}, \autoref{fig:pipeline-blur}, and \autoref{fig:pipeline-frame} show the output textures produced by the various stages for a simple scene.
+
+\begin{figure}[H]
+  \begin{center}
+    \begin{minipage}{.2\textwidth}
+      \includegraphics[width=1.0\textwidth]{pipeline-position.png}
+    \end{minipage}
+    \begin{minipage}{.2\textwidth}
+      \includegraphics[width=1.0\textwidth]{pipeline-normal.png}
+    \end{minipage}
+    \vskip0.04cm
+    \begin{minipage}{.2\textwidth}
+      \includegraphics[width=1.0\textwidth]{pipeline-albedo.png}
+    \end{minipage}
+    \begin{minipage}{.2\textwidth}
+      \includegraphics[width=1.0\textwidth]{pipeline-specular.png}
+    \end{minipage}
+  \end{center}
+  \caption{The output of the geometry pass, producing a ``G-buffer'' composed out of a position-, normal-, albedo-, and specular-map.}
+  \label{fig:pipeline-gbuffer}
+\end{figure}
+
+\begin{figure}[H]
+  \begin{center}
+    \includegraphics[width=0.3\textwidth]{pipeline-shadow.png}
+  \end{center}
+  \caption{The depth map produced by the shadow map pass. It shows the distance of surfaces from the point of view of the primary light source.}
+  \label{fig:pipeline-shadow}
+\end{figure}
+
+\begin{figure}[H]
+  \begin{center}
+    \begin{minipage}{.2\textwidth}
+      \includegraphics[width=1.0\textwidth]{pipeline-lighting.png}
+    \end{minipage}
+    \begin{minipage}{.2\textwidth}
+      \includegraphics[width=1.0\textwidth]{pipeline-highpass.png}
+    \end{minipage}
+  \end{center}
+  \caption{The color map (left) and the high-pass map (right) that is output by the deferred render pass.
+It combines the information from the G-buffer from \autoref{fig:pipeline-gbuffer}, the shadow-map from \autoref{fig:pipeline-shadow}, and information for lights to render an image.
+Note that due to gamma adjustments the image appears quite dark.}
+  \label{fig:pipeline-lighting}
+\end{figure}
+
+\begin{figure}[H]
+  \begin{center}
+    \includegraphics[width=0.3\textwidth]{pipeline-blur.png}
+  \end{center}
+  \caption{The blurred high-pass from \autoref{fig:pipeline-lighting} produced by the gaussian blur pass. This is used to produce an effect called bloom in the final composite pass.}
+  \label{fig:pipeline-blur}
+\end{figure}
+
+\begin{figure}[H]
+  \begin{center}
+    \includegraphics[width=0.3\textwidth]{pipeline-frame.png}
+  \end{center}
+  \caption{The final output frame that is produced by adding together the bloom texture from \autoref{fig:pipeline-blur} and the standard colour texture from \autoref{fig:pipeline-lighting}.
+Gamma adjustment and High Dynamic Range reduction are used for colour normalisation.}
+  \label{fig:pipeline-frame}
+\end{figure}
+
 \section{Conclusion}\label{conclusion}
+We have presented a system to handle the separation between objects that should be rendered, and the way in which they should be rendered.
+By using a protocol for the interaction between such objects and passes we are capable of controlling and combining a variety of rendering behaviour.
+With an additional protocol and set of algorithms we also present a high-level view of rendering pipelines, allowing users to conveniently compose effects to produce complicated scenes.
+\\
+
+These protocols and algorithms are largely generic enough that they could also be adapted for use in other languages and systems.
+However, we believe that the Common Lisp Object System gives us a set of very convenient tools to make everything feel more natural and more integrated with the rest of the system than other environments would allow us to.
+\\
+
+Finally, we have demonstrated the capability of this system by constructing a real-time rendering system out of individual, modular pieces of code.
 
 \section{Further Work}\label{further-work}
+Currently there exist several severe restrictions in our framework. \\
+
+Most severe is that there is no standardised protocol to communicate capabilities and data between objects and passes.
+This means that for instance and object that does not possess texturing has no way of communicating this to a shader pass that might need to render it.
+Currently this problem is partially solved by introducing a mandatory superclass that objects need to be a subtype of, if they want to be renderable under a given pass.
+This superclass then loosely defines the interaction.
+However, this does not solve issues of compatibility of shader code used by the object and the shader pass.
+In order to resolve this, we currently see two required features: \\
+
+Shader source merging must be aware of uses-relations.
+This means that the code walker must track where variables, functions, and types are used and reorder the definitions and declarations in such a way that the dependencies are fulfilled.
+In GLSL, variables, functions, and types cannot be referenced before they are declared, requiring this reordering step.
+With this solved, shader code could be written in such a way to be more amiable towards combination and manipulation by external code.
+\\
+
+GLSL's limited expressiveness makes it very difficult to figure out relations between code and to figure out how to properly combine pieces.
+A more high-level language as is used in other frameworks like Spark\cite{spark} or Shader Components\cite{components} should be introduced to allow a more convenient way to declare and write shaders, objects, and passes.
+With a higher-level language a compiler could more easily figure out how to combine features and reconcile differences between pieces of code.
+\\
+
+Another restriction is that there is no way to encapsulate GLSL code other than associating it with a class, which is a rather heavy-weight operation.
+It would be far more useful to introduce a form of library mechanism that allows the definition of standalone shader functions, which can then be imported by shader classes.
 
 \section{Acknowledgements}\label{acknowledgements}
+I would like to thank all the people that will review this paper. Your name could be here!
 
@@
 \\\url{https://github.com/Shinmera/flow} \\
-
-A more in-depth discussion of the system can be found at
-\\\url{https://reader.tymoon.eu/article/363}
-\\\url{https://reader.tymoon.eu/article/364}
 

### 2019-01-30 els2019-shader-pipeline/paper.tex
Doooonnee!

@@
 \begin{abstract}
-  %% FIXME: Abstract
+  Modern real-time graphics make use of a lot of tricks in order to produce stunning visuals.
+Many of these tricks require separate rendering passes, as well as separate rendering logic for each pass.
+These passes are then combined in a variety of ways in order to produce a final image.
+The interaction between such a rendering pass and the objects it draws, as well as the interaction between multiple passes within a pipeline can become quite complex.
+Often times in order to handle this complexity, the passes and objects are generalised, and the render logic is left almost entirely with the object or the pass.
+We present a new method of representing objects, passes, and pipelines, which allows a modular encapsulation of effects and rendering behaviour, as well as object-oriented composition through inheritance.
 \end{abstract}
@@
 A shader pass should encapsulate the logic for drawing objects in a certain way.
-In order to do this, it needs to be able to partially or fully control the shaders of objects.
+In order to do this, the pass needs to be able to partially or fully control the shaders of objects.
 For instance, an object should still be able to have control over how its vertices are constructed and what materials and textures are applied to it.
@@
 This inversion is very important for shader passes that take all control away from the objects.
-Without it, the \code{:before}, \code{:after}, and \code{:around} methods on \code{render} specialised on an object could try to manipulate state before methods specialised on a shader pass could prevent it.
+Without \code{render-with}, the \code{:before}, \code{:after}, and \code{:around} methods on \code{render} specialised on an object could try to manipulate state before methods specialised on a shader pass could prevent it.
 \\
@@
 \section{Pipelines}\label{pipelines}
-Shader passes not only retain information about how objects are drawn with it, but also about the input and output textures that a pass interacts with.
+Shader passes not only retain information about how objects are drawn, but also about the input and output textures that a pass interacts with.
 For this reason, each shader pass is also a node in a graph, with distinct input and output ports.
@@
 Instead, each format is represented by a constant number, with no relation to the features that format includes.
-This means that we need to first destructure each format into a list of its features:
+This means that we need to first destructure each format into a list of features:
 
@@
 
-If the join is successful, we then re-encode the texture format specification into OpenGL's constant and use it in the real texture specification.
+If the join is successful, we then re-encode the texture format specification into OpenGL's constant and use this constant in the real texture specification.
 Note that this join could texture format specifications that are not legal according to the OpenGL specification.
@@
 
-In other words, the algorithm proceeds backwards from the last node in the graph, marking predecessors' output ports with unique colours, then marking unconnected ports with unique colours, and finally marking all colours at its own output ports as available again.
+In other words, the algorithm proceeds backwards from the last node in the graph, marking predecessors' output ports with unique colours, then marking unconnected ports with unique colours, and finally marking all colours at the nodes own output ports as available again.
 We repeat this process for each unique texture specification, each time ensuring we only consider ports that share that texture specification.
@@
 Most severe is that there is no standardised protocol to communicate capabilities and data between objects and passes.
-This means that for instance and object that does not possess texturing has no way of communicating this to a shader pass that might need to render it.
+This means that for instance an object that does not possess texturing has no way of communicating this to a shader pass that might need to render it.
 Currently this problem is partially solved by introducing a mandatory superclass that objects need to be a subtype of, if they want to be renderable under a given pass.

### 2019-01-30 els2019-shader-pipeline/paper.tex
Fix some stuff. Ready for review

@@
 The interaction between such a rendering pass and the objects it draws, as well as the interaction between multiple passes within a pipeline can become quite complex.
-Often times in order to handle this complexity, the passes and objects are generalised, and the render logic is left almost entirely with the object or the pass.
+Often times, in order to handle this complexity, the passes and objects are generalised, and the render logic is left almost entirely with either the object or the pass.
 We present a new method of representing objects, passes, and pipelines, which allows a modular encapsulation of effects and rendering behaviour, as well as object-oriented composition through inheritance.
@@
 
-Each of the customisable stages only accepts a single shader for each pass, making it difficult to separate, encapsulate, and ultimately combine behaviour.
+Each of the customisable stages only accept a single shader for each pass, making it difficult to separate, encapsulate, and ultimately combine behaviour.
 Furthermore, the steps required in order to change the shaders and shader inputs can be non-trivial and expensive to execute.
@@
 
-The complexity of managing the graphics state and the order of rendering can be massively complicated for modern requirements.
+The complexity of managing this graphics state and the order of rendering can be massively complicated for modern requirements.
 Rendering a frame often requires a multitude of passes, each with their own parameters and shared data.
@@
 
-Gyrling\cite{fibers} presents an overview of the techniques used to perform parallel rendering in Naughty Dog's commercial engine.
+Gyrling\cite{fibers} presents an overview of the techniques used to perform parallel rendering in Naughty Dog's private game engine.
 Individual stages within a pass, render passes of a frame, and multiple frame renderings are divided up into many small jobs that can run in parallel and are synchronised using counters on a shared structure.
+How rendering logic requirements are communicated is however not explained.
 \\
@@
 The case study of the Unity game engine by Messaoudi et al\cite{unity} shows the availability of a set of fixed rendering pipelines that can be customised in a very limited extent with custom shaders.
-However, these shaders must fit into Unity's existing lighting and overall rendering model.
+These shaders must fit into Unity's existing lighting and overall rendering model.
 While Unity does allow building a custom pipeline via their Scriptable Rendering Pipeline\cite{unitycustom}, they do not seem to offer any specific encapsulation or modularity features.
@@
 Both shader objects and shader passes encapsulate rendering logic.
-The pipeline on the other hand represents the assembly of a complete frame and only influences rendering in the sense that it tracks which passes to run in which order.
+The pipeline on the other hand represents the assembly of a complete frame and only influences rendering in the sense that it tracks which passes to run in what order.
 \\
@@
 
-Finally, in \hyperref[concept]{section 7} we show the results of an implementation of a medium-sized pipeline using these techniques.
+Finally, in \hyperref[concept]{section 7} we show the results of implementing a medium-sized pipeline using these techniques.
 
@@
 For instance, an object should still be able to have control over how its vertices are constructed and what materials and textures are applied to it.
-The shader pass needs to be able to make use of this information if it needs to, or discard it completely if unneeded.
+The shader pass should be able to make use of this information if it needs to, or discard it completely if unneeded.
 \\
 
-In order to allow this we make use of the shader composition capabilities that we presented in our previous work\cite{glsloop} that allows us to combine pieces of GLSL code, and a new protocol centred around shader passes.
+In order to allow this we make use of the shader composition capabilities that we presented in our previous work\cite{glsloop}.
+This technique allows us to combine pieces of GLSL code.
+On top of this shader combination sits a new protocol to control the interaction between shader passes and objects.
+\\
 
@@
 This allows the pass to prepare the shader program that will be used during rendering for this object.
-A shader program is an OpenGL resource that compiles all shaders for a pass together.
-With \code{shader-program-for-pass} this program can then later be retrieved.
+A shader program is an OpenGL resource that compiles the shaders for the stages of a pass together.
+Using \\ \code{shader-program-for-pass} this program can then later be retrieved.
 The shader program is important for objects to access, as it allows them to set values for ``uniform'' variables that are used in the shader code.
@@
 Objects are typically rendered using the \code{render} function.
-Object classes are encouraged to add methods that specialise the behaviour and perform necessary setup, as well as to perform the final draw call.
+Users are encouraged to add methods that specialise the behaviour and perform necessary setup, as well as to perform the final draw call for their custom object classes.
 For instance, a very primitive class could look as shown in \autoref{lst:simple-object}.
+\\
 
@@
 
-This function is fine for allowing the object being drawn control over the behaviour.
+This function is fine for allowing the objec control over the behaviour.
 However, the pass cannot exert the same amount of control due to the generic function's argument precedence.
@@
 
-\code{render-with} thus exists mostly as an entry point for rendering operations to allow shader passes greater control over the rendering.
+\code{render-with} thus exists mostly as an entry point to allow shader passes greater control over the rendering.
 Typically it will simply defer to \code{render}.
@@
 
-For instance, the ``Deferred Lighting'' pass shown in \autoref{fig:render-pipeline} is created by combining the passes illustrated in \autoref{lst:deferred-render}, \autoref{lst:shadow-render}, and \autoref{lst:high-pass} through inheritance.
+For instance, the ``Deferred Render'' pass shown in \autoref{fig:render-pipeline} is created by combining the passes illustrated in \autoref{lst:deferred-render}, \autoref{lst:shadow-render}, and \autoref{lst:high-pass} through inheritance.
 \\
@@
       (geometry (make-instance 'geometry-pass))
-      (deferred (make-instance 'deferred-lighting-pass))
+      (deferred (make-instance 'deferred+shadow-pass))
       (blur (make-instance 'gaussian-blur-pass))
@@
 For simplicity and brevity, we will focus on the join operator for a single texture property here, the internal format.
-The internal format specifies how many colour channels the texture has, how many bits of precision each channel has, which format each channel has, as well as whether the texture is compressed or has sRGB gamma normalisation applied.
+The internal format is arguably the most important property.
+This property specifies how many colour channels the texture has, how many bits of precision each channel has, which format each channel has, as well as whether the texture is compressed or has sRGB gamma normalisation applied.
 \\
@@
 The list of specified texture formats is quite large.\cite{texinternalformat} Unfortunately OpenGL does not give us an interface to handle these formats in a way that lets us pick the individual features easily.
-Instead, each format is represented by a constant number, with no relation to the features that format includes.
-This means that we need to first destructure each format into a list of features:
+Instead, each format is represented by a constant, whose value has no relation to the features that format includes.
+This means that we need to first destructure each format name into a list of features: \\
 
@@
 If the join is successful, we then re-encode the texture format specification into OpenGL's constant and use this constant in the real texture specification.
-Note that this join could texture format specifications that are not legal according to the OpenGL specification.
-However, this can only occur if one of the specs to join is already illegal.
+Note that this join could produce texture format specifications that are not legal according to the OpenGL specification.
+However, this can only occur if one of the specifications to join is already illegal.
 We thus deem it unnecessary to handle such cases.
@@
 \begin{step}
-\item The set of unique texture specifications is computed by joining each port's texspec with every other and comparing for equality.
+\item The set of unique texture specifications is computed by joining each port's texture specification with every other and comparing for equality.
 \item For each unique texture specification T:
@@
       \begin{step}
-      \item If P's texspec is joinable with T...
+      \item If P's texture specification is joinable with T...
       \item The number of colours is increased
@@
         \begin{step}
-        \item If O's texspec is joinable with T...
+        \item If O's texture specification is joinable with T...
         \item and O does not yet have a colour...
@@
       \begin{step}
-      \item If P's texspec is joinable with T...
+      \item If P's texture specification is joinable with T...
       \item and P does not yet have a colour...
@@
       \begin{step}
-      \item If P's texspec is joinable with T...
+      \item If P's texture specification is joinable with T...
       \item and P has a colour...
@@
 
-This algorithm is by no means efficient, but since pipeline allocation only has to happen during loading phases, we currently do not consider this to be a big problem. \\
+This algorithm is by no means efficient, but since pipeline allocation only has to happen during loading phases, we currently do not consider this to be a big problem.
+We also have not performed any analysis as to whether the algorithm produces optimal allocation results in every case.\\
 
@@
 As a proof of concept we have implemented these protocols and mechanisms, as well as the passes shown in \autoref{fig:render-pipeline}.
-As the code required to do so is quite lengthy and involved, and the graphics techniques used are beyond the scope of this paper, we will omit them.
+As the code required to do so is quite lengthy and involved, and the graphics techniques used are beyond the scope of this paper, we will omit the code.
 The relevant outline of the passes and the pipeline involved has already been shown in \hyperref[pipelines]{section 5}.
@@
 
-\begin{figure}[H]
+\begin{figure}[h]
   \begin{center}
@@
   \end{center}
-  \caption{The output of the geometry pass, producing a ``G-buffer'' composed out of a position-, normal-, albedo-, and specular-map.}
+  \caption{The output of the geometry pass, producing a ``G-buffer'' composed out of a position- (top left), normal- (top right), albedo- (bottom left), and specular-map (bottom right).}
   \label{fig:pipeline-gbuffer}
@@
   \begin{center}
-    \includegraphics[width=0.3\textwidth]{pipeline-frame.png}
+    \includegraphics[width=0.45\textwidth]{pipeline-frame.png}
   \end{center}
@@
 These protocols and algorithms are largely generic enough that they could also be adapted for use in other languages and systems.
-However, we believe that the Common Lisp Object System gives us a set of very convenient tools to make everything feel more natural and more integrated with the rest of the system than other environments would allow us to.
+However, we believe that the Common Lisp Object System gives us a set of very convenient tools to make everything feel more natural and more integrated with the rest of the system than other environments would allow us to do.
 \\
@@
 \section{Further Work}\label{further-work}
-Currently there exist several severe restrictions in our framework. \\
+Currently there exist several restrictions in our framework. \\
 
@@
 
-Another restriction is that there is no way to encapsulate GLSL code other than associating it with a class, which is a rather heavy-weight operation.
-It would be far more useful to introduce a form of library mechanism that allows the definition of standalone shader functions, which can then be imported by shader classes.
+Another restriction is that there is currently no way to encapsulate GLSL code other than associating it with a class, which is a rather heavy-weight operation.
+It would be far more useful to introduce a library mechanism that allows the definition of standalone shader functions, which can then be required by shader classes.
 
@@
 An implementation of the proposed system can be found at
-\\\href{https://github.com/Shirakumo/trial/blob/f34a79f0a6df21d1ed9259e85fbb3c7eed39352b/shader-pass.lisp}{https://github.com/Shirakumo/trial/blob/\\f34a79f0a6df21d1ed9259e85fbb3c7eed39352b/shader-pass.lisp}
-\\\href{https://github.com/Shirakumo/trial/blob/f34a79f0a6df21d1ed9259e85fbb3c7eed39352b/pipeline.lisp}{https://github.com/Shirakumo/trial/blob/\\f34a79f0a6df21d1ed9259e85fbb3c7eed39352b/pipeline.lisp}
+\\\href{https://github.com/Shirakumo/trial/blob/b889d50b7906348ad787e773ba57998bb39b16a5/shader-pass.lisp}{https://github.com/Shirakumo/trial/blob/\\b889d50b7906348ad787e773ba57998bb39b16a5/shader-pass.lisp}
+\\\href{https://github.com/Shirakumo/trial/blob/b889d50b7906348ad787e773ba57998bb39b16a5/pipeline.lisp}{https://github.com/Shirakumo/trial/blob/\\b889d50b7906348ad787e773ba57998bb39b16a5/pipeline.lisp}
 \\\url{https://github.com/Shinmera/flow} \\

### 2019-01-30 els2019-shader-pipeline/paper.tex
Incorporate @phoe's feedback.
Unfortunately ACM's style is fucked up and does not permit non-latin characters. For some reason, even with inputenc and bable and all that jazz Michał is not displayed correctly.

@@
 \documentclass[format=sigconf]{acmart}
+\usepackage[utf8]{inputenc}
 \usepackage{enumitem}
-\usepackage[english]{babel}
 \usepackage{float}
@@
 
-This function is fine for allowing the objec control over the behaviour.
+This function is fine for allowing the object control over the behaviour.
 However, the pass cannot exert the same amount of control due to the generic function's argument precedence.
+For example, if an \code{:around} method specialised on the object exists, a pass would not have any way of preventing it from firing, as its own \code{:around} method would be executed afterwards.
 We thus introduce another function with inverted argument order that is called first.
@@
 This inversion is very important for shader passes that take all control away from the objects.
-Without \code{render-with}, the \code{:before}, \code{:after}, and \code{:around} methods on \code{render} specialised on an object could try to manipulate state before methods specialised on a shader pass could prevent it.
 \\
@@
 Nevertheless, you can find the complete code online at the following pages: \\
-\href{https://github.com/Shirakumo/trial/blob/b889d50b7906348ad787e773ba57998bb39b16a5/deferred.lisp}{https://github.com/Shirakumo/trial/blob/\\b889d50b7906348ad787e773ba57998bb39b16a5/deferred.lisp} \\
-\href{https://github.com/Shirakumo/trial/blob/b889d50b7906348ad787e773ba57998bb39b16a5/hdr.lisp}{https://github.com/Shirakumo/trial/blob/\\b889d50b7906348ad787e773ba57998bb39b16a5/hdr.lisp} \\
-\href{https://github.com/Shirakumo/trial/blob/b889d50b7906348ad787e773ba57998bb39b16a5/shadow-map.lisp}{https://github.com/Shirakumo/trial/blob/\\b889d50b7906348ad787e773ba57998bb39b16a5/shadow-map.lisp} \\
-\href{https://github.com/Shirakumo/trial/blob/b889d50b7906348ad787e773ba57998bb39b16a5/workbench.lisp}{https://github.com/Shirakumo/trial/blob/\\b889d50b7906348ad787e773ba57998bb39b16a5/workbench.lisp} \\
+\url{https://github.com/Shirakumo/trial/blob/b889d50/deferred.lisp} \\
+\url{https://github.com/Shirakumo/trial/blob/b889d50/hdr.lisp} \\
+\url{https://github.com/Shirakumo/trial/blob/b889d50/shadow-map.lisp} \\
+\url{https://github.com/Shirakumo/trial/blob/b889d50/workbench.lisp} \\
 
@@
 \section{Acknowledgements}\label{acknowledgements}
-I would like to thank all the people that will review this paper. Your name could be here!
+I would like to thank Michal ``phoe'' Herda, and all the people that will review this paper. Your name could be here!
 
 \section{Implementation}\label{implementation}
-An implementation of the proposed system can be found at
-\\\href{https://github.com/Shirakumo/trial/blob/b889d50b7906348ad787e773ba57998bb39b16a5/shader-pass.lisp}{https://github.com/Shirakumo/trial/blob/\\b889d50b7906348ad787e773ba57998bb39b16a5/shader-pass.lisp}
-\\\href{https://github.com/Shirakumo/trial/blob/b889d50b7906348ad787e773ba57998bb39b16a5/pipeline.lisp}{https://github.com/Shirakumo/trial/blob/\\b889d50b7906348ad787e773ba57998bb39b16a5/pipeline.lisp}
-\\\url{https://github.com/Shinmera/flow} \\
+An implementation of the proposed system can be found at \\
+\url{https://github.com/Shirakumo/trial/blob/b889d50/shader-pass.lisp} \\
+\url{https://github.com/Shirakumo/trial/blob/b889d50/pipeline.lisp} \\
+\url{https://github.com/Shinmera/flow} \\
 

### 2019-01-30 els2019-shader-pipeline/paper.tex
Fix the figures slightly, !@#/# TEX!!!
Incorporate Robert Strandh's feedback

@@
 We present a new method of representing objects, passes, and pipelines, which allows a modular encapsulation of effects and rendering behaviour, as well as object-oriented composition through inheritance.
+We make use of CLOS' multiple inheritance, multimethods, and standard method combination to form an extensible protocol.
+We also make use of the MOP, in order to introduce additional metadata to classes.
 \end{abstract}
@@
 \def\abovecaptionskip{1pt}
-\def\listingautorefname{listing}
-\def\figureautorefname{figure}
+\def\listingautorefname{Listing}
+\def\figureautorefname{Figure}
 
@@
 
-Each of the customisable stages only accept a single shader for each pass, making it difficult to separate, encapsulate, and ultimately combine behaviour.
+Each of the customisable stages accepts only a single shader for each pass, making it difficult to separate, encapsulate, and ultimately combine behaviour.
 Furthermore, the steps required in order to change the shaders and shader inputs can be non-trivial and expensive to execute.
@@
 
-This results in a difficult challenge for modularity. We attempt to solve this challenge through several systems:
+This complexity results in a difficult challenge for modularity. We attempt to solve this challenge through several systems:
 
@@
 
-The case study of the Unity game engine by Messaoudi et al\cite{unity} shows the availability of a set of fixed rendering pipelines that can be customised in a very limited extent with custom shaders.
+The case study of the Unity game engine by Messaoudi et al\cite{unity} shows the availability of a set of fixed rendering pipelines that can be customised to a very limited extent with custom shaders.
 These shaders must fit into Unity's existing lighting and overall rendering model.
@@
 The work by He et al.\cite{components} introduces a framework for general encapsulation of shaders and their parameters into structures that minimise the overhead of changing GPU state while retaining the ability to dynamically compose shader parts.
-They do however not create a distinction between properties for the rendering of an object, and those for the rendering of an overall pass.
+They do however not create a distinction between properties for rendering an object, and those for rendering an overall pass.
 \\
@@
 
-In our previous work\cite{glsloop} we introduce a system to tie shader code to classes and compose behaviour through inheritance.
+In our previous work\cite{glsloop} we introduce a system to tie shader code to classes and to compose behaviour through inheritance.
 We make use of this system and extend it to allow further control over rendering behaviour in individual passes.
@@
 \section{Overview}\label{overview}
-The system is composed out of three distinct entities: shader objects, shader passes, and pipelines.
+The system is composed of three distinct entities: shader objects, shader passes, and pipelines.
 Both shader objects and shader passes encapsulate rendering logic.
-The pipeline on the other hand represents the assembly of a complete frame and only influences rendering in the sense that it tracks which passes to run in what order.
+The pipeline, on the other hand, represents the assembly of a complete frame, and only influences rendering in the sense that it tracks which passes to run in what order.
 \\
 
-Due to the restrictions of OpenGL's rendering pipeline, the combination of the render logic of objects and passes is not trivial.
-We present a protocol to solve this in \hyperref[passes]{section 4}.
+Due to the restrictions of the rendering pipeline of OpenGL, the combination of the render logic of objects and of passes is not trivial.
+We present a protocol to solve this difficulty in \hyperref[passes]{section 4}.
 \\
 
-In \hyperref[pipelines]{section 5} we present a protocol to represent input and output information of a pass and how to connect these inputs and outputs between passes together. \\
+In \hyperref[pipelines]{section 5} we present a protocol for representing input and output information of a pass and for connecting these inputs and outputs between passes together. \\
 
@@
 A shader pass should encapsulate the logic for drawing objects in a certain way.
-In order to do this, the pass needs to be able to partially or fully control the shaders of objects.
+In order to accomplish its task, the pass needs to be able to partially or fully control the shaders of objects.
 For instance, an object should still be able to have control over how its vertices are constructed and what materials and textures are applied to it.
@@
 
-In order to allow this we make use of the shader composition capabilities that we presented in our previous work\cite{glsloop}.
+In order to permit this amount of control, we make use of the shader composition capabilities that we presented in our previous work\cite{glsloop}.
 This technique allows us to combine pieces of GLSL code.
@@
 For any object that should be rendered using a given pass, first \code{register-object-for-pass} must be called.
-This allows the pass to prepare the shader program that will be used during rendering for this object.
+This call allows the pass to prepare the shader program that will be used during rendering for this object.
 A shader program is an OpenGL resource that compiles the shaders for the stages of a pass together.
@@
 
-\code{register-object-for-pass} calls \code{make-pass-shader-program} to compute this program and registers it internally in the shader pass so that it can be retrieved later.
+\code{register-object-for-pass} calls \code{make-pass-shader-program} to compute this program, and registers it internally in the shader pass so that it can be retrieved later.
 If the shader pass would like to retain complete control over how each object is drawn, it would forego this call and instead generate its own shader program.
-This is especially useful for post-processing effects that don't render any objects at all, and instead simply operate on textures output by previous passes.
+Doing so is especially useful for post-processing effects that don't render any objects at all, and instead simply operate on textures that are output by previous passes.
 \\
@@
 \code{coerce-pass-shader} computes the effective shader source for a particular shader stage or type.
-Typically this simply involves the combination of the shader sources of both shader pass and object for the given shader type.
+Typically this computation simply involves the combination of the shader sources of both the shader pass and of the object, for the given shader type.
 We perform this combination using the same parsing and code walking strategy as we described in our previous work.\cite{glsloop} \\
@@
 This function is fine for allowing the object control over the behaviour.
-However, the pass cannot exert the same amount of control due to the generic function's argument precedence.
+However, the pass cannot exert the same amount of control, due to the generic function's argument precedence.
 For example, if an \code{:around} method specialised on the object exists, a pass would not have any way of preventing it from firing, as its own \code{:around} method would be executed afterwards.
@@
 An example of such a pass is illustrated in \autoref{lst:deferred-render}.
-The ports are modelled as slots on the class, requiring a new metaclass.
+The ports are modelled as slots of the class that carry additional metadata, requiring a new metaclass.
 A slot with a port-type will not only hold a texture for its value, but also retain information about how it is connected to other passes.
@@
 
-When assembling a pipeline, we can then connect these ports together in order to dictate how the various output textures generated by a pass are used as inputs for other passes.
+When assembling a pipeline, we can then connect these ports together, in order to dictate how the various output textures generated by a pass are used as inputs for other passes.
 As an example, the pipeline from \autoref{fig:render-pipeline} is created in \autoref{lst:pipeline}.
@@
 \item Call \code{render} on the pipeline with a collection of all objects to draw.
-\item The output port's texture of the last shader pass in the pipeline will contain the finished frame.
+\item The output texture of the output port of the last shader pass in the pipeline will contain the finished frame.
 \end{step}
@@
 These framebuffers then need to have the required textures to render onto allocated as well.
-As each input and output from a pass can specify constraints on the texture's features, these constraints must be matched up for any connecting edges as well.
-Finally, in order to minimise memory usage we would like to re-use textures where possible.
+As each input and output from a pass can specify constraints on the features of the texture, these constraints must be matched up for any connecting edges as well.
+Finally, in order to minimise memory usage, we would like to re-use textures where possible.
 \\
 
-Thus the allocation proceeds in three phases: reconciling texture constraints on edges between passes, computing how textures are shared between passes, and finally constructing all the necessary resources with the previously gathered information.
+Thus, the allocation proceeds in three phases: reconciling texture constraints on edges between passes, computing how textures are shared between passes, and finally constructing all the necessary resources with the previously gathered information.
 
 \subsection{Constraint Merging}\label{texture-constraints}
-OpenGL textures include a massive amount of information.\cite{texparameter}\cite{teximage2d} When two ports are connected that specify different constraints on the texture properties, a join must be performed.
+OpenGL textures include a massive amount of information\cite{texparameter}\cite{teximage2d}.
+When two ports are connected that specify different constraints on the texture properties, a join must be performed.
 Fortunately, a wide range of the texture property values are fundamentally incompatible, meaning that a lot of the logic can simply error.
-For simplicity and brevity, we will focus on the join operator for a single texture property here, the internal format.
+For simplicity and brevity, we will focus on the join operator for a single texture property here: the internal format.
 The internal format is arguably the most important property.
@@
 
-The list of specified texture formats is quite large.\cite{texinternalformat} Unfortunately OpenGL does not give us an interface to handle these formats in a way that lets us pick the individual features easily.
-Instead, each format is represented by a constant, whose value has no relation to the features that format includes.
-This means that we need to first destructure each format name into a list of features: \\
+The list of specified texture formats is quite large\cite{texinternalformat}.
+Unfortunately OpenGL does not give us an interface to handle these formats in a way that lets us pick the individual features easily.
+Instead, each format is represented by a constant whose value has no relation to the features that format includes.
+This representation means that we first need to destructure each format name into a list of features: \\
 
@@
 
-In other words, the algorithm proceeds backwards from the last node in the graph, marking predecessors' output ports with unique colours, then marking unconnected ports with unique colours, and finally marking all colours at the nodes own output ports as available again.
+In other words, the algorithm proceeds backwards from the last node in the graph, marking output ports of predecessors with unique colours, then marking unconnected ports with unique colours, and finally marking all colours at the node's own output ports as available again.
 We repeat this process for each unique texture specification, each time ensuring we only consider ports that share that texture specification.
@@
 
-Unfortunately print documents cannot yet display animations, so illustrating the algorithm in motion is not possible directly in the document.
+Unfortunately printed documents cannot yet display animations, so illustrating the algorithm in motion is not possible directly in the document.
 However, a brief animation of the pipeline illustrated in \autoref{fig:render-pipeline} is available online: \\
@@
   \begin{center}
-    \includegraphics[width=0.3\textwidth]{pipeline-shadow.png}
+    \includegraphics[width=0.2\textwidth]{pipeline-shadow.png}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=0.3\textwidth]{pipeline-blur.png}
+    \includegraphics[width=0.2\textwidth]{pipeline-blur.png}
   \end{center}
@@
 
-\begin{figure}[H]
+\begin{figure}[h]
   \begin{center}
-    \includegraphics[width=0.45\textwidth]{pipeline-frame.png}
+    \includegraphics[width=0.4\textwidth]{pipeline-frame.png}
   \end{center}
@@
 We have presented a system to handle the separation between objects that should be rendered, and the way in which they should be rendered.
-By using a protocol for the interaction between such objects and passes we are capable of controlling and combining a variety of rendering behaviour.
+By using a protocol for the interaction between such objects and passes, we are capable of controlling and combining a variety of rendering behaviour.
 With an additional protocol and set of algorithms we also present a high-level view of rendering pipelines, allowing users to conveniently compose effects to produce complicated scenes.
@@
 Most severe is that there is no standardised protocol to communicate capabilities and data between objects and passes.
-This means that for instance an object that does not possess texturing has no way of communicating this to a shader pass that might need to render it.
+This restriction means that for instance an object that does not possess texturing has no way of communicating this to a shader pass that might need to render it.
 Currently this problem is partially solved by introducing a mandatory superclass that objects need to be a subtype of, if they want to be renderable under a given pass.
 This superclass then loosely defines the interaction.
-However, this does not solve issues of compatibility of shader code used by the object and the shader pass.
-In order to resolve this, we currently see two required features: \\
+However, this additional class does not solve issues of compatibility of shader code used by the object and by the shader pass.
+In order to properly resolve this restriction, we currently see two required features: \\
 
-Shader source merging must be aware of uses-relations.
-This means that the code walker must track where variables, functions, and types are used and reorder the definitions and declarations in such a way that the dependencies are fulfilled.
+Shader source merging must be aware of uses-relations, meaning that the code walker must track where variables, functions, and types are used and reorder the definitions and declarations in such a way that the dependencies are fulfilled.
 In GLSL, variables, functions, and types cannot be referenced before they are declared, requiring this reordering step.
-With this solved, shader code could be written in such a way to be more amiable towards combination and manipulation by external code.
+With this problem solved, shader code could be written in such a way to be more amiable towards combination and manipulation by external code.
 \\
@@
 GLSL's limited expressiveness makes it very difficult to figure out relations between code and to figure out how to properly combine pieces.
-A more high-level language as is used in other frameworks like Spark\cite{spark} or Shader Components\cite{components} should be introduced to allow a more convenient way to declare and write shaders, objects, and passes.
-With a higher-level language a compiler could more easily figure out how to combine features and reconcile differences between pieces of code.
+A more high-level language, as is used in other frameworks like Spark\cite{spark} or Shader Components\cite{components}, should be introduced to allow a more convenient way to declare and write shaders, objects, and passes.
+With a higher-level language, a compiler could more easily figure out how to combine features and reconcile differences between pieces of code.
 \\
@@
 \section{Acknowledgements}\label{acknowledgements}
-I would like to thank Michal ``phoe'' Herda, and all the people that will review this paper. Your name could be here!
+I would like to thank Michal ``phoe'' Herda, Robert Strandh, and all the people that will review this paper. Your name could be here!
 

### 2019-01-30 els2019-shader-pipeline/paper.tex
Incorporate feedback by Michael Reis

@@
 The interaction between such a rendering pass and the objects it draws, as well as the interaction between multiple passes within a pipeline can become quite complex.
-Often times, in order to handle this complexity, the passes and objects are generalised, and the render logic is left almost entirely with either the object or the pass.
+Often times, in order to handle this complexity, the passes and objects are generalised, and the render logic is left to be executed almost entirely with either the object or the pass.
 We present a new method of representing objects, passes, and pipelines, which allows a modular encapsulation of effects and rendering behaviour, as well as object-oriented composition through inheritance.
@@
 
-The complexity of managing this graphics state and the order of rendering can be massively complicated for modern requirements.
+Managing this graphics state and the order of rendering can be very complicated for modern requirements.
 Rendering a frame often requires a multitude of passes, each with their own parameters and shared data.
@@
 The work by He et al.\cite{components} introduces a framework for general encapsulation of shaders and their parameters into structures that minimise the overhead of changing GPU state while retaining the ability to dynamically compose shader parts.
-They do however not create a distinction between properties for rendering an object, and those for rendering an overall pass.
+They do however not create a distinction between properties for rendering an object and for those for rendering an overall pass.
 \\
@@
 Nevertheless, constraining the features of textures tied to the ports is important, and we discuss the technique for dealing with that in \hyperref[allocation]{section 6}.
+\\
 
@@
 Each shader stage needs a ``framebuffer,'' an OpenGL resource that allows one to render to off-screen textures.
-These framebuffers then need to have the required textures to render onto allocated as well.
+These framebuffers then need to have the required textures to render allocated as well.
 As each input and output from a pass can specify constraints on the features of the texture, these constraints must be matched up for any connecting edges as well.
@@
 When two ports are connected that specify different constraints on the texture properties, a join must be performed.
-Fortunately, a wide range of the texture property values are fundamentally incompatible, meaning that a lot of the logic can simply error.
+A wide range of the texture property values are fundamentally incompatible, meaning that a lot of the logic can simply error.
+However, other options require more complicated joining logic.
 For simplicity and brevity, we will focus on the join operator for a single texture property here: the internal format.
@@
 We have presented a system to handle the separation between objects that should be rendered, and the way in which they should be rendered.
-By using a protocol for the interaction between such objects and passes, we are capable of controlling and combining a variety of rendering behaviour.
+By using a protocol for the interaction between such objects and passes, we are capable of controlling and combining a variety of rendering behaviours.
 With an additional protocol and set of algorithms we also present a high-level view of rendering pipelines, allowing users to conveniently compose effects to produce complicated scenes.
@@
 Most severe is that there is no standardised protocol to communicate capabilities and data between objects and passes.
-This restriction means that for instance an object that does not possess texturing has no way of communicating this to a shader pass that might need to render it.
+This restriction means that, for instance, an object that does not possess texturing has no way of communicating this to a shader pass that might need to render it.
 Currently this problem is partially solved by introducing a mandatory superclass that objects need to be a subtype of, if they want to be renderable under a given pass.
@@
 
-GLSL's limited expressiveness makes it very difficult to figure out relations between code and to figure out how to properly combine pieces.
+GLSL's limited expressiveness makes it very difficult to figure out relations between code and how to properly combine pieces.
 A more high-level language, as is used in other frameworks like Spark\cite{spark} or Shader Components\cite{components}, should be introduced to allow a more convenient way to declare and write shaders, objects, and passes.
@@
 \section{Acknowledgements}\label{acknowledgements}
-I would like to thank Michal ``phoe'' Herda, Robert Strandh, and all the people that will review this paper. Your name could be here!
+I would like to thank Michal ``phoe'' Herda, Robert Strandh, Michael Reis, and all the people that will review this paper. Your name could be here!
 

### 2019-01-31 els2019-shader-pipeline/paper.tex
Minor

@@
 We present a new method of representing objects, passes, and pipelines, which allows a modular encapsulation of effects and rendering behaviour, as well as object-oriented composition through inheritance.
-We make use of CLOS' multiple inheritance, multimethods, and standard method combination to form an extensible protocol.
+We make use of CLOS' multiple inheritance, multimethods, and standard method combination to form extensible protocols that allow this new method.
 We also make use of the MOP, in order to introduce additional metadata to classes.
@@
 \item \code{stencil} How many bits to use for the stencil channel.
-\item \code{shared} Whether bits are shared across the channel, and if so how many.
+\item \code{shared} Whether bits are shared across the channels, and if so how many.
 \item \code{features} Whether the format has compression, sRGB, RGTC, BPTC, SNORM, or UNORM features.

### 2019-02-27 els2019-shader-pipeline/paper.tex
Incorporate some peer review feedback

@@
 The interaction between such a rendering pass and the objects it draws, as well as the interaction between multiple passes within a pipeline can become quite complex.
-Often times, in order to handle this complexity, the passes and objects are generalised, and the render logic is left to be executed almost entirely with either the object or the pass.
+Often times, in order to handle this complexity, the passes and objects are generalised, and the render logic is controlled almost entirely by either the object or the pass.
 We present a new method of representing objects, passes, and pipelines, which allows a modular encapsulation of effects and rendering behaviour, as well as object-oriented composition through inheritance.
@@
 In order to permit this amount of control, we make use of the shader composition capabilities that we presented in our previous work\cite{glsloop}.
-This technique allows us to combine pieces of GLSL code.
+This technique allows us to combine pieces of OpenGL Shader Language (GLSL) code.
 On top of this shader combination sits a new protocol to control the interaction between shader passes and objects.
@@
 
-This protocol thus allows a great amount of control both over the effective shader code used to render an object, as well as the behaviour leading up towards the actual draw call for an object.
+This protocol thus allows a great amount of control both over the effective shader code used to render an object, as well as the behaviour leading up towards the actual draw call for an object. \\
+
+The overall protocol thus only invokes overhead for the dispatch of the \code{render-with}, \code{render}, and \code{shader-program-for-pass} generic functions for each rendered object per frame.
+All other functions are only invoked during loading, or in exceptional situations.
+In our use of the system so far we have not found this to impact performance significantly.
 
@@
 
-\href{https://raw.githubusercontent.com/Shinmera/talks/master/els2019-shader-pipeline/pipeline-allocation.gif}{https://raw.githubusercontent.com/Shinmera/talks/master/\\els2019-shader-pipeline/pipeline-allocation.gif}
+\href{https://raw.githubusercontent.com/Shinmera/talks/master/els2019-shader-pipeline/pipeline-allocation.gif}{https://raw.githubusercontent.com/Shinmera/talks/master/\\els2019-shader-pipeline/pipeline-allocation.gif} \\
+
+With this algorithm we reduce the number of necessary textures to allocate down to five.
+With a primitive allocation of one texture per output, we would instead end up with seven textures.
+Note that the shadow map output is not assigned, as it requires a different type of texture from the rest -- a depth texture.
+This texture would be allocated in a second pass.
 
@@
 
+\newpage ~ % Fucking latex, give me a new page.
+
 \section{Conclusion}\label{conclusion}

### 2021-10-08 gic2021-highly-dynamic/paper.tex
Start work on GIC2021 paper.

\documentclass[a4paper]{paper}
\usepackage[utf8]{inputenc}
\usepackage[margin=2.5cm]{geometry}
\usepackage{enumitem}
\usepackage{float}
\usepackage[labelfont=bf,textfont=md]{caption}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{minted}
\usepackage{hyperref}
\usepackage[all]{hypcap}
\usemintedstyle[common-lisp]{default}
\newmintinline[code]{text}{}
\bibliographystyle{plainnat}

\hypersetup{
  colorlinks,
  linkcolor={red!50!black},
  citecolor={blue!50!black},
  urlcolor={blue!80!black}
}

\newlist{step}{enumerate}{10}
\setlist[step]{label*=\arabic*.,leftmargin=2em}

\title{Using a Highly Dynamic Language for Development}
\subtitle{Advantages of and lessons learned from using Common Lisp in games}
\author{Nicolas Hafner}
\institution{Shirakumo Games}

\begin{document}
\twocolumn[\maketitle]

\begin{abstract}
  Games face an interesting challenge. They require rapid development, are highly interactive, and pose hard real-time performance constraints. While smaller games these days have also been developed in dynamic languages such as Python or Lua, traditionally engines are still written in static languages like C++ and C, with an additional scripting languages on top to handle gameplay mechanics. In this paper we explore using Common Lisp for full stack game development and outline the advantages this can bring to the development process, as well as the pitfalls and problem areas that are unique to this kind of approach.
\end{abstract}
\begin{keywords}
  Common Lisp, game development, dynamic languages, object orientation
\end{keywords}

\def\abovecaptionskip{1pt}
\def\listingautorefname{Listing}
\def\figureautorefname{Figure}

\section{Introduction}
Video games pose an interesting engineering challenge. They are highly dynamic in their nature, as users can perform various, sometimes far-reaching changes to the program at any time, and yet they must remain responsive under hard real-time constraints. Additionally, the development of games itself is highly dynamic, as changes to the game require constant testing and refinement. Long pauses between making a change and being able to properly evaluate its effects can gravely discourage testing, which leads to a much worse product.

A typical approach to solve this set of constraints is to use multiple languages in combination. A rather low-level language like C++ or C to handle the ``core engine'', and an integrated scripting language like Lua to handle gameplay logic. However, this approach has multiple issues of its own: it can be hard to distinguish which parts should be a part of the core engine, and which should not. The scripting cannot integrate with everything the engine offers, as an explicit interface has to be designed that can deal with the scripting language's own data types and routines. For performance reasons a highly dynamic part may also need to be lowered down into the static language, making iteration much slower and harder to deal with.

Finally, the lack of runtime debugging means that any problems appearing in the core engine often lead to a crash of the entire program, which makes diagnosing and fixing the issue much harder. This difficulty often leads to defensive programming strategies, where errors are simply ignored or otherwise coerced, leading them to cause issues further down the line, complicating debugging of the final product even more.

In this paper we instead take a holistic approach, using Common Lisp for the full stack of both the core engine and the gameplay tools and mechanics. Common Lisp is a highly dynamic language, allowing runtime redefinition of functions, variables, and classes, even to the point of completely reloading or changing an underlying library or system while the program is being executed. Yet, despite this dynamism, Common Lisp is a compiled language that takes great care to support the writing of efficient code. Highly optimising compilers like SBCL allow you to write fast code without having to drop down into another language.

We explore some of the aspects of Common Lisp that make it particularly suited for games in detail, and also discuss some of the pitfalls we encountered and how to combat them.

\section{Related Works}

\section{Modularity Through Mixins}
The Common Lisp Object System (CLOS) has a couple of traits that remain rare in programming languages in use today, but make for excellent tools to support game development. Relevant to this section are serialised multiple inheritance and the standard generic function method combination.

In CLOS methods are not attached to classes, but are instead parts of a generic function. Methods ``specialise'' on one or several of the required arguments of the function to a class. When the generic function is called, the system must first determine the set of ``applicable methods''. This set depends on the method combination used by the generic function. For brevity's sake, we will only look at the standard method combination here.

The standard method combination offers methods in four flavours: \code{primary}, \code{:before}, \code{:after}, and \code{:around}. These methods are grouped together, then filtered for applicable methods by checking whether the arguments passed to the generic function match their specialised classes, and finally sorted by \textit{how} specific the specialisations are to the arguments passed. With this completed set, the methods are invoked as illustrated in \autoref{fig:method combination}.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.4\textwidth]{method combination.png}
  \caption{The standard method combination behaviour illustrated}
  \label{fig:method combination}
\end{figure}

As a brief example, let us consider \autoref{lst:method definition}. Here we define a generic function of two arguments called \code{handle}, as well as three methods. All of the methods require the first argument to be a subclass of \code{tick}. Method 1 and 2 also require the second argument to be a subclass of \code{player}, and method 3 a subclass of \code{enemy}.

\begin{listing}[h]
\begin{minted}[fontsize=\scriptsize]{common-lisp}
(defgeneric handle (event object))

(defmethod handle         ((ev tick) (object player))) ; 1
(defmethod handle :before ((ev tick) (object player))) ; 2
(defmethod handle         ((ev tick) (object enemy)))  ; 3
\end{minted}
\caption{A brief example of method definition}
\label{lst:method definition}
\end{listing}

\hypertarget{first-call}{When \code{handle} is now called with a \code{tick} and a \code{player} instance, first the method 2 is executed, followed by the method 1. Method 3 is ignored, as it does not match the arguments.}

This method combination mechanism only really shines once we consider inheritance and especially multiple inheritance and the arising ``mixin classes''. Let us now define the classes we used for the previous listing. \autoref{lst:inheritance} shows five classes being defined, with \code{tick} being a subclass of \code{event}, and \code{player} and \code{enemy} being subclasses of \\\code{physics-object}.

\begin{listing}[H]
\begin{minted}[fontsize=\scriptsize]{common-lisp}
(defclass event () ())
(defclass tick (event) ())

(defclass physics-object () ())
(defclass player (physics-object) ())
(defclass enemy (physics-object) ())
\end{minted}
\caption{A brief example of method definition}
\label{lst:inheritance}
\end{listing}

We'll now also change the second method to specialise on the \code{physics-object} instead of \code{player}. For example, since both \code{player} and \code{enemy} are moving, we could be handling the collision resolution there. If we now perform the \hyperlink{first-call}{same call from before}, Methods 2 and 1 will still be invoked. However, unlike before, if we call the function with \code{tick} and \code{enemy}, now methods 2 and 3 will be invoked, instead of only method 3.

\begin{listing}[h]
\begin{minted}[fontsize=\scriptsize]{common-lisp}
(defmethod handle :before ((ev tick) (object moving-entity)))
\end{minted}
\caption{The updated second method definition}
\label{lst:updated method}
\end{listing}

In other OOP paradigms a similar behaviour can usually be achieved by calling the superclass' method with \code{super}, but notice here that the behaviour of method 2 does not require changing anything about the other methods or subclasses. It thus remains wholly encapsulated in its own behaviour.

So far so good. Now imagine that we decide enemies, should emit a light so that they're always visible. To implement this, we're going to define a new class, \code{emitter}, and define the light flicker behaviour in a new \code{handle} method as shown in \autoref{lst:emitter}.

\begin{listing}[h]
\begin{minted}[fontsize=\scriptsize]{common-lisp}
(defclass emitter () ())

(defmethod handle :before ((ev tick) (object emitter)))
\end{minted}
\caption{The new class and method to handle light flickering}
\label{lst:emitter}
\end{listing}

To make the enemy adopt this new behaviour, all we need to do is add \code{emitter} to the \code{enemy} class' superclass list. Thanks to the method combination, calls to \code{handle} will now include the \code{emitter}'s method, and we have achieved the combination of the two behaviours.

\begin{listing}[H]
\begin{minted}[fontsize=\scriptsize]{common-lisp}
(defclass enemy (physics-object emitter) ())
\end{minted}
\caption{The updated enemy class definition}
\label{lst:new-enemy}
\end{listing}

The order of the superclasses here is important. Classes that appear earlier in the list have ``precedence'' and are thus considered to be ``more specific'' when determining the set of applicable methods. By having \code{emitter} appear after \code{physics-object}, we ensure that the \code{emitter}'s \code{:before} method runs \textit{after} that of \code{physics-object}, ensuring the collisions have already resolved properly once we consider the lighting update.

Not having the \code{emitter} be a subclass of \\\code{physics-object} ensures that we can also use it as a superclass for other things such as completely static lanterns that have no business having any interactivity.

As an example from our actual code base, as of writing, the \code{player} class has 8 direct superclasses whose behaviours are combined together with the \code{player}'s own. This combination of behaviours allows encapsulating different parts and re-using them in many cases. Keep in mind, too, that all of these classes and methods can be redefined at runtime to change, add, and remove behaviours from an object while the game is still running.

However, one issue that crops up when segregating behaviours into such small compartments is that you might not have enough control over the combination of them. The standard method combination offers no fine-grained control to exclude or reorder methods for a specific specialisation of an argument. One could devise their own method combination strategy to allow this kind of specialised behaviour, however we are not convinced that this would not lead to an ultimately even more confusing design.

While the standard method combination and mixins offer a great deal of flexibility out of the gate, great care must still be taken when designing the overall class hierarchy and function protocol. Otherwise behaviours will not combine cleanly and lead to strange bugs, or might not combine properly at all.

\section{Restarts}

\section{Optimisation}

\section{Garbage Collection}

\section{Conclusion}

\section{Acknowledgements}

\bibliography{paper}
\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-command-extra-options: "-shell-escape"
%%% TeX-master: t
%%% TeX-engine: luatex
%%% End:

### 2021-10-09 gic2021-highly-dynamic/paper.tex
First draft complete

@@
 \usepackage{minted}
+\usepackage[square,numbers]{natbib}
 \usepackage{hyperref}
@@
 While smaller games these days have also been developed in dynamic languages such as Python or Lua, traditionally engines are still written in static languages like C++ and C, with an additional scripting languages on top to handle gameplay mechanics.
-In this paper we explore using Common Lisp for full stack game development and outline the advantages this can bring to the development process, as well as the pitfalls and problem areas that are unique to this kind of approach.
+Common Lisp offers an environment that's both dynamic and performant enough to allow for a full stack game development system that is highly favourable to fast iteration and modular design.
 \end{abstract}
@@
 \section{Related Works}
+Please see our prior work on using Common Lisp for game development and real-time computer graphics\cite{hafner2018shader}\cite{hafner2019shader}.
+As this is otherwise primarily an overview of Common Lisp facilities and our experiences, we do not compare this paper to other work.
 
@@
 
-However, one issue that crops up when segregating behaviours into such small compartments is that you might not have enough control over the combination of them.
+We've also developed a system to tie shaders into the CLOS inheritance model, allowing us to attach GPU rendering behaviour to classes, and combine it in a similar fashion\cite{hafner2019shader}.
+
+One issue that crops up when segregating behaviours into such small compartments is that you might not have enough control over the combination of them.
 The standard method combination offers no fine-grained control to exclude or reorder methods for a specific specialisation of an argument.
@@
 
-\section{Restarts}
+\section{Conditions, Handlers, and Restarts}
+Debugging is of course an important aspect of programming, but this is even more so the case when the system can be redefined and changed during runtime.
+Fortunately, Common Lisp offers an incredibly capable system to deal with runtime bugs.
+This system comes in three parts, only one of which is commonly found in other languages.
+
+The first part is what's called ``conditions'', or Exceptions in other languages.
+Conditions are ``signalled'' (thrown) and can then be ``handled'' (caught) by a piece of code lower down in the stack.
+When handled, the stack unwinds to the point of the handler, and the matching handler function is invoked to resolve the issue.
+For our purposes here, Common Lisp offers two types of conditions: warnings and errors.
+When a warning is signalled that is not handled further up the stack, it simply vanishes and execution continues from the signalling point unimpeded.
+If an error remains unhandled, a dynamic debugger is invoked instead and the signalling thread is paused.
+From there one of the unique features comes to shine.
+
+Restarts are on first look similar to conditions and handlers.
+When a restart is established, it sets an unwind point for a particularly named restart with a function to be called when that restart is triggered.
+Restarts themselves can be triggered via a function called \code{invoke-restart}, which can also pass along arguments for the restart function.
+The idea with restarts is to provide one or more ways for execution to continue or recover safely from an error.
+When the debugger is invoked from an unhandled error, it now has access to a list of active restarts on the stack, and can invoke one of them dynamically as the user sees fit.
+
+This means that unlike traditional languages where an error causes a stack trace and then tries to continue or simply crashes, in Common Lisp execution is halted, waiting for you to analyse and potentially fix the bug.
+Once you're confident you know what to do or have solved the issue, you can pick from one of several ways to continue the execution from the error.
+This leaves the program running at all times and means you don't need to worry about long recompilation times or setup times to reproduce the error, you can immediately continue where you left off.
+
+The debugger also allows you to evaluate expressions at runtime while in a particular stack frame, allowing access to local variables in the process.
+Since the Common Lisp runtime includes the full compiler suite at all times, these expressions can be arbitrary Lisp code, allowing for far more invasive exploration of the bug and changes to the environment than easily possible with traditional out-of-process debuggers.
+
+The final piece is the existence of \code{handler-bind}.
+This is a type of handler for conditions, but unlike the regular handlers that cause the stack to unwind, these handlers are invoked \textit{on top of the stack} of where the condition was signalled.
+As such they have full access to the dynamic environment at error point.
+In fact, this is also how the top level debugger is invoked.
+However, with \code{handler-bind} you can automate the error resolution and avoid requiring user input.
+
+A system with well-designed recovery points through restarts can then be set up to automatically recover from most errors in a way much more refined than traditional Exceptions, as there is an inversion of control.
+
+\begin{listing}[H]
+\begin{minted}[fontsize=\scriptsize]{common-lisp}
+(defmethod render ((scene scene))
+  (for ((object in scene))
+    (with-simple-restart (continue)
+      (render object))))
+
+(defmethod render ((main main))
+  (handler-bind ((error (lambda (condition)
+                          (invoke-restart 'continue))))
+    (render (scene main))))
+\end{minted}
+\caption{Simplified code illustrating the control inversion of restarts}
+\label{lst:restarts}
+\end{listing}
+
+In \autoref{lst:restarts} we show two simple methods, the first to render a \code{scene}, which simply iterates over each object in the scene and establishes a \code{continue} restart around the call to \code{render} the object.
+When this restart is invoked, the stack unwinds to within the \code{for} and then simply continues with the next object.
+
+If we now call \code{render} on a \code{scene} directly and an error occurs, the debugger gets invoked as usual.
+But this time, we can invoke the \code{continue} restart from the debugger to skip rendering the object.
+We could also define a ``retry'' restart to simply retry rendering the object as well, if we wanted to.
+
+However, importantly, for the \code{main} \code{render}\\ method, this restart is automatically invoked, ensuring that any potential crashes at runtime of the game are gracefully ignored.
+This handler could also be defined elsewhere even higher up the stack, or only invoke the restart under special conditions.
+
+Restarts and handlers are fantastic tools to create a more error friendly environment and to recover from unfavourable situations.
+Combined with the dynamic debugger and runtime compilation environment, these tools make debugging problems and evolving a system over time a lot easier.
 
 \section{Optimisation}
+A frequent problem with highly dynamic language is optimisation, as the dynamic nature of the language prohibits making assumptions at compile time, forcing repeated runtime dispatching.
+The focus on safety will also insert bounds checks and similar assertions into the code, which can have a big impact on runtime in the hot loop.
+
+To help with these issues, Common Lisp includes several tools in the form of sorts of ``compiler promises''.
+You can use a declaration to promise to the compiler that a variable's value assumes a certain type.
+The compiler can then use this promise to perform inference and eliminate runtime dispatching.
+SBCL is often times good enough at automatically inferring the types to eliminate much of runtime dispatch on its own, but for highly optimised functions, declaring the types manually can be very beneficial still.
+
+Additionally, Common Lisp offers several compiler tuning declarations.
+For instance, increasing the \code{speed} switch to its maximum will tell the compiler to try and optimise the code as much as it can.
+SBCL will then also emit warnings about cases where it doesn't have sufficient information to eliminate dispatch, or where it must allocate on the heap.
+Using this extra information the programmer can then refactor the code or add additional promises to ensure the compiler can generate efficient code.
+
+Another switch is the \code{safety} switch, which when set to its minimum will tell the compiler to avoid inserting safety checks such as runtime type checks, bounds checks, or other features that typically help catch bugs.
+Naturally this is a dangerous option, as the generated code now becomes as unsafe as C or assembly code, and misuse of functions declared with this option can lead to memory corruptions.
+
+You can also declare the type of a function, to ensure that the compiler can infer return values and arguments across function boundaries.
+SBCL's block compilation feature, when activated, also allows it to reason across function boundaries within the same compilation unit on its own.
+We have not yet properly explored block compilation in our code base, as block compilation has a few provisions that we do not fulfil yet.
+
+Finally you can also declare functions to be \code{inline}, allowing the compiler to inline the definitions at call sites.
+This comes with the usual advantage of aiding inference and removing call overhead, but does hamper dynamism, as a change to an inlined function now requires recompiling call sites as well.
+
+When focusing on the SBCL implementation, there are several other optimisations that can be done.
+SBCL offers full access to its code optimisation facilities and even its assembler routines.
+With enough effort, code can be optimised on the assembly level, and one could even emit special assembly instructions where needed.
+Some libraries already make use of these systems to speed up computations, but we have not touched them much ourselves.
+We do however make use of the \code{disassemble} function to check the generated assembly of functions, as another aid in optimisation.
+We also make use of the integrated statistical profiler to observe the runtime of the game and determine choke points.
+
+These techniques mostly focus on type dispatch within a function, with special focus on arithmetic functions.
+Dispatch for generic functions cannot be eliminated like this, and especially generic functions with a lot of methods attached or complicated method signatures can become a significant choke point.
+
+This can be largely combated with traditional techniques of simplifying method signatures, splitting up generic functions, or rewriting them into statically dispatching functions that can be inlined to take advantage of type inference.
+We are also investigating an alternate dispatch technique outlined by Robert Strandh\cite{strandh2014fast} that should significantly improve generic function dispatch time, potentially even outperforming traditional vtable based approaches as used in C++ or Java.
+
+While all these opportunities mean that fast code can be written in SBCL, the default is still on the side of safety over performance, something we consider an ultimately good thing.
+Fully optimising code on the level of C, C++, or assembly requires significant, but not insurmountable work.
 
 \section{Garbage Collection}
+Being such a highly dynamic language, it's unavoidable that a garbage collector be involved.
+This can be problematic for games where a GC cycle can easily eclipse the time budget of a single frame.
+There already exists a breadth of literature on the topic of GC types, GC tuning, etc.
+out there so we will not go into a debate on the nature of GC or its drawbacks and advantages here.
+Instead we will focus on the specific issues we have encountered and what we've done to mitigate them.
+
+The GC we use is SBCL's default GC, a generational, compacting, single-threaded stop-the-world collector.
+We do hope that SBCL will receive a better GC at some point, but for now we've been focusing on minimising garbage production in our code to avoid frequent and extensive GCs.
+
+Garbage can be minimised in a number of ways, the most general being object pooling, or in other words manual garbage collection.
+By allocating\\known needed objects ahead of time and simply recycling them, we can avoid generating a lot of intermittent garbage.
+Common Lisp gives us some very useful tools here that make this process a lot easier.
+
+For instance, we can ask the compiler to allocate a number of objects on the stack by using a declaration (\code{(declare (dynamic-extent x))}).
+However, this only works if the compiler can know ahead of time what size the object is, meaning only fixed size arrays and known structs can be stack allocated.
+Furthermore, stack space is much more limited, so bigger allocations still have to happen on the heap.
+
+For cases where the object is a class instance, simply too big for the stack, or has to escape the stack sometimes, we can still easily cache the instance locally by using \code{load-time-value}.\\\code{load-time-value} is a special operator that causes the form it wraps to be executed when the code is first loaded into the system, and then puts the result of that executed form in the place in the code the \code{load-time-value} was at.
+This effectively creates an anonymous global variable where the object is stored, which is then put in place of the original \code{load-time-value}.
+
+This trick is very useful when we know the code path is only ever accessed from one thread at a time. Multi-threaded schemes would need to employ more complex methods.
+
+Another point where \code{load-time-value} becomes very useful is when objects are said to be immutable, and their construction arguments are known at compile time.
+In this case, we can define a ``compiler macro'' which is a macro that is executed to replace a call to a particular function.
+This compiler macro can then analyse the arguments to the function at compile-time and choose to substitute the call with a \code{load-time-value} form if all arguments can be determined statically.
+This then avoids the construction and allocation of the object, or even the entire computation of the function at runtime in a, to the user, completely transparent fashion.
+
+In our UI toolkit Alloy we use this trick extensively to avoid the allocation of colour values, sizes, and dimensional units.
+
+Even with all of these tricks, it remains very hard to completely eliminate allocations, as other parts such as boxing of values or generation of bignums can cause allocations in spots where it isn't immediately obvious.
+So far we have managed to keep the game running smoothly enough by offloading big GC cycles to load screens and transitions, but it is clear that bigger games do need additional work put into them to optimise them for garbage management.
 
 \section{Conclusion}
+We believe that Common Lisp offers a set of unique features that makes it a great candidate for the kind of highly iterative development processes often required for games.
+
+The expressiveness of the Common Lisp Object System allows for rapid changes in game object behaviour, as well as the creation of reusable pieces of behaviour that can often be seamlessly combined together.
+With extensions to the Meta Object Protocol, this can even be extended to graphics rendering routines and shaders.
+
+The presence of the entire compiler environment at runtime allows the developer to change any part of the game while it is running.
+This is especially feasible thanks to the presence of the restarts and interactive debugger, which allow the game to recover from a bug without having to crash and restart the entire system.
+
+While these features offer great amounts of flexibility, there is a performance cost to be paid.
+Many of these performance penalties can be avoided or circumvented with careful planning and design, but often by sacrificing some of the dynamism.
+These optimisations also come at a greater development cost.
+
+We believe that, since these optimisations can be performed gradually over time and because no optimisation opportunities are fundamentally excluded, Common Lisp still makes for a great candidate language for game development.
+
+While there are many more intricacies to the Common Lisp ecosystem, we hope that this paper gives a brief insight into some of the advantages and challenges present when working in a full stack Common Lisp environment.
 
 \section{Acknowledgements}
+We would like to thank the very handsome and attractive people that helped proofread this draft ;)
 

### 2021-10-09 gic2021-highly-dynamic/paper.tex
Minor

@@
 They require rapid development, are highly interactive, and pose hard real-time performance constraints.
-While smaller games these days have also been developed in dynamic languages such as Python or Lua, traditionally engines are still written in static languages like C++ and C, with an additional scripting languages on top to handle gameplay mechanics.
+While smaller games these days have also been developed in dynamic languages such as Python or Lua, traditionally engines are still written in static languages like C++ and C, with an additional scripting language on top to handle gameplay mechanics.
 Common Lisp offers an environment that's both dynamic and performant enough to allow for a full stack game development system that is highly favourable to fast iteration and modular design.

### 2021-10-09 gic2021-highly-dynamic/paper.tex
Paper updates

@@
   \centering
-  \includegraphics[width=0.4\textwidth]{method combination.png}
+  \includegraphics[width=0.45\textwidth]{method combination.png}
   \caption{The standard method combination behaviour illustrated}
@@
 SBCL's block compilation feature, when activated, also allows it to reason across function boundaries within the same compilation unit on its own.
-We have not yet properly explored block compilation in our code base, as block compilation has a few provisions that we do not fulfil yet.
 
@@
 With enough effort, code can be optimised on the assembly level, and one could even emit special assembly instructions where needed.
-Some libraries already make use of these systems to speed up computations, but we have not touched them much ourselves.
+Some libraries already make use of these systems to speed up computations, but we have not touched them much ourselves due to a lack of time so far.
 We do however make use of the \code{disassemble} function to check the generated assembly of functions, as another aid in optimisation.
@@
 The GC we use is SBCL's default GC, a generational, compacting, single-threaded stop-the-world collector.
-We do hope that SBCL will receive a better GC at some point, but for now we've been focusing on minimising garbage production in our code to avoid frequent and extensive GCs.
+We do hope that SBCL will receive a better GC such as the Memory Pool System\cite{brooksby2002memory} at some point, but for now we've been focusing on minimising garbage production in our code to avoid frequent and extensive GCs.
 
@@
 \section{Conclusion}
-We believe that Common Lisp offers a set of unique features that makes it a great candidate for the kind of highly iterative development processes often required for games.
-
 The expressiveness of the Common Lisp Object System allows for rapid changes in game object behaviour, as well as the creation of reusable pieces of behaviour that can often be seamlessly combined together.

### 2021-10-10 gic2021-highly-dynamic/paper.tex
Minor revisions

@@
   Games face an interesting challenge.
-They require rapid development, are highly interactive, and pose hard real-time performance constraints.
+They require rapid development, are highly interactive, and pose soft real-time performance constraints.
 While smaller games these days have also been developed in dynamic languages such as Python or Lua, traditionally engines are still written in static languages like C++ and C, with an additional scripting language on top to handle gameplay mechanics.
@@
 Video games pose an interesting engineering challenge.
-They are highly dynamic in their nature, as users can perform various, sometimes far-reaching changes to the program at any time, and yet they must remain responsive under hard real-time constraints.
+They are highly dynamic in their nature, as users can perform various, sometimes far-reaching changes to the program at any time, and yet they must remain responsive under soft real-time constraints.
 Additionally, the development of games itself is highly dynamic, as changes to the game require constant testing and refinement.
@@
 \begin{minted}[fontsize=\scriptsize]{common-lisp}
-(defmethod handle :before ((ev tick) (object moving-entity)))
+(defmethod handle :before ((ev tick) (object physics-object)))
 \end{minted}
@@
 
-In our UI toolkit Alloy we use this trick extensively to avoid the allocation of colour values, sizes, and dimensional units.
+In Alloy, our UI toolkit, we use this trick extensively to avoid the allocation of colour values, sizes, and dimensional units.
 
@@
 
-While these features offer great amounts of flexibility, there is a performance cost to be paid.
+While these features offer great amounts of flexibility, there is a performance cost.
 Many of these performance penalties can be avoided or circumvented with careful planning and design, but often by sacrificing some of the dynamism.

### 2021-10-11 gic2021-highly-dynamic/paper.tex
minor

@@
 Relevant to this section are serialised multiple inheritance and the standard generic function method combination.
+These features can be used to implement a system resembling ECS, though also offer a few advantages that make combination of behaviours more natural.
 
@@
 
-While the standard method combination and mixins offer a great deal of flexibility out of the gate, great care must still be taken when designing the overall class hierarchy and function protocol.
+While the standard method combination and mixins offer a great deal of flexibility out of the box, great care must still be taken when designing the overall class hierarchy and function protocol.
 Otherwise behaviours will not combine cleanly and lead to strange bugs, or might not combine properly at all.
@@
 Restarts are on first look similar to conditions and handlers.
-When a restart is established, it sets an unwind point for a particularly named restart with a function to be called when that restart is triggered.
+When a restart is established through \code{restart-case}, it sets an unwind point for a particularly named restart with a function to be called when that restart is triggered.
 Restarts themselves can be triggered via a function called \code{invoke-restart}, which can also pass along arguments for the restart function.
@@
   (for ((object in scene))
-    (with-simple-restart (continue)
-      (render object))))
+    (restart-case
+        (render object)
+      (continue ())))
 
@@
 This handler could also be defined elsewhere even higher up the stack, or only invoke the restart under special conditions.
+This works as the \code{lambda} of the \code{handler-bind} is invoked at the point of error on top of the stack, where the \code{continue} restart is visible.
 

### 2021-10-16 gic2021-highly-dynamic/paper.tex
Update stuff with feedback from Robert

@@
 
-We explore some of the aspects of Common Lisp that make it particularly suited for games in detail, and also discuss some of the pitfalls we encountered and how to combat them.
+We explore in detail some of the aspects of Common Lisp that make it particularly suited for games, and also discuss some of the pitfalls we encountered and how to combat them.
 
@@
 
-In CLOS methods are not attached to classes, but are instead parts of a generic function.
-Methods ``specialise'' on one or several of the required arguments of the function to a class.
+In CLOS, methods are not attached to classes, but are instead parts of a generic function.
+Methods ``specialise'' to one or several of the required arguments of the function to a class.
 When the generic function is called, the system must first determine the set of ``applicable methods''.
@@
 Here we define a generic function of two arguments called \code{handle}, as well as three methods.
-All of the methods require the first argument to be a subclass of \code{tick}.
+Each of the methods requires the first argument to be a subclass of \code{tick}.
 Method 1 and 2 also require the second argument to be a subclass of \code{player}, and method 3 a subclass of \code{enemy}.
@@
 
-This method combination mechanism only really shines once we consider inheritance and especially multiple inheritance and the arising ``mixin classes''.
+This method combination mechanism really shines only once we consider inheritance and especially multiple inheritance and the arising ``mixin classes''.
 Let us now define the classes we used for the previous listing.
@@
 
-In other OOP paradigms a similar behaviour can usually be achieved by calling the superclass' method with \code{super}, but notice here that the behaviour of method 2 does not require changing anything about the other methods or subclasses.
+In other OOP paradigms a similar behaviour can usually be achieved by calling the method of the superclass with \code{super}, but notice here that the behaviour of method 2 does not require changing anything about the other methods or subclasses.
 It thus remains wholly encapsulated in its own behaviour.
@@
 So far so good.
-Now imagine that we decide enemies, should emit a light so that they're always visible.
-To implement this, we're going to define a new class, \code{emitter}, and define the light flicker behaviour in a new \code{handle} method as shown in \autoref{lst:emitter}.
+Now imagine that we decide enemies should emit a light so that they're always visible.
+To implement this behaviour, we're going to define a new class, \code{emitter}, and define the light flicker behaviour in a new \code{handle} method as shown in \autoref{lst:emitter}.
 
@@
 
-To make the enemy adopt this new behaviour, all we need to do is add \code{emitter} to the \code{enemy} class' superclass list.
+To make the enemy adopt this new behaviour, all we need to do is add \code{emitter} to the \code{enemy} list of superclasses of the enemy class.
 Thanks to the method combination, calls to \code{handle} will now include the \code{emitter}'s method, and we have achieved the combination of the two behaviours.
@@
 The order of the superclasses here is important.
-Classes that appear earlier in the list have ``precedence'' and are thus considered to be ``more specific'' when determining the set of applicable methods.
+Classes that appear earlier in the list have ``precedence'' and are thus considered to be ``more specific'' when the set of applicable methods is determined.
 By having \code{emitter} appear after \code{physics-object}, we ensure that the \code{emitter}'s \code{:before} method runs \textit{after} that of \code{physics-object}, ensuring the collisions have already resolved properly once we consider the lighting update.
@@
 
-One issue that crops up when segregating behaviours into such small compartments is that you might not have enough control over the combination of them.
+One issue that crops up when behaviours are segregated into such small compartments is that you might not have enough control over the combination of them.
 The standard method combination offers no fine-grained control to exclude or reorder methods for a specific specialisation of an argument.
@@
 
-While the standard method combination and mixins offer a great deal of flexibility out of the box, great care must still be taken when designing the overall class hierarchy and function protocol.
+While the standard method combination and mixins offer a great deal of flexibility out of the box, great care must still be taken when the overall class hierarchy and function protocol is designed.
 Otherwise behaviours will not combine cleanly and lead to strange bugs, or might not combine properly at all.
@@
 Conditions are ``signalled'' (thrown) and can then be ``handled'' (caught) by a piece of code lower down in the stack.
-When handled, the stack unwinds to the point of the handler, and the matching handler function is invoked to resolve the issue.
+When handled by a standard condition handler, the stack unwinds to the point of the handler, and the matching handler function is invoked to resolve the issue.
 For our purposes here, Common Lisp offers two types of conditions: warnings and errors.
@@
 Once you're confident you know what to do or have solved the issue, you can pick from one of several ways to continue the execution from the error.
-This leaves the program running at all times and means you don't need to worry about long recompilation times or setup times to reproduce the error, you can immediately continue where you left off.
+Doing so leaves the program running at all times and means you don't need to worry about long recompilation times or setup times to reproduce the error.
+You can immediately continue where you left off.
 
@@
 This is a type of handler for conditions, but unlike the regular handlers that cause the stack to unwind, these handlers are invoked \textit{on top of the stack} of where the condition was signalled.
-As such they have full access to the dynamic environment at error point.
+As such, they have full access to the dynamic environment at the point of the error.
 In fact, this is also how the top level debugger is invoked.
@@
 
-A system with well-designed recovery points through restarts can then be set up to automatically recover from most errors in a way much more refined than traditional Exceptions, as there is an inversion of control.
+A system with well designed recovery points through restarts can then be set up to automatically recover from most errors in a way much more refined than traditional Exceptions, as there is an inversion of control.
 
@@
 However, importantly, for the \code{main} \code{render}\\ method, this restart is automatically invoked, ensuring that any potential crashes at runtime of the game are gracefully ignored.
-This handler could also be defined elsewhere even higher up the stack, or only invoke the restart under special conditions.
+This handler could also be defined elsewhere even higher up the stack, or invoke the restart only under special conditions.
 This works as the \code{lambda} of the \code{handler-bind} is invoked at the point of error on top of the stack, where the \code{continue} restart is visible.
 
-Restarts and handlers are fantastic tools to create a more error friendly environment and to recover from unfavourable situations.
+Restarts and handlers are fantastic tools to create a more error-friendly environment and to recover from unfavourable situations.
 Combined with the dynamic debugger and runtime compilation environment, these tools make debugging problems and evolving a system over time a lot easier.
@@
 SBCL will then also emit warnings about cases where it doesn't have sufficient information to eliminate dispatch, or where it must allocate on the heap.
-Using this extra information the programmer can then refactor the code or add additional promises to ensure the compiler can generate efficient code.
+Using this extra information, the programmer can then refactor the code or add additional promises to ensure the compiler can generate efficient code.
 
-Another switch is the \code{safety} switch, which when set to its minimum will tell the compiler to avoid inserting safety checks such as runtime type checks, bounds checks, or other features that typically help catch bugs.
+Another switch is the \code{safety} switch, which, when set to its minimum, will tell the compiler to avoid inserting safety checks such as runtime type checks, bounds checks, or other features that typically help catch bugs.
 Naturally this is a dangerous option, as the generated code now becomes as unsafe as C or assembly code, and misuse of functions declared with this option can lead to memory corruptions.
@@
 Finally you can also declare functions to be \code{inline}, allowing the compiler to inline the definitions at call sites.
-This comes with the usual advantage of aiding inference and removing call overhead, but does hamper dynamism, as a change to an inlined function now requires recompiling call sites as well.
+This declaration comes with the usual advantage of aiding inference and removing call overhead, but does hamper dynamism, as a change to an inlined function now requires recompiling call sites as well.
 
@@
 
-These techniques mostly focus on type dispatch within a function, with special focus on arithmetic functions.
+These techniques focus mostly on type dispatch within a function, with special focus on arithmetic functions.
 Dispatch for generic functions cannot be eliminated like this, and especially generic functions with a lot of methods attached or complicated method signatures can become a significant choke point.
 
-This can be largely combated with traditional techniques of simplifying method signatures, splitting up generic functions, or rewriting them into statically dispatching functions that can be inlined to take advantage of type inference.
+This can be combated largely with traditional techniques of simplifying method signatures, splitting up generic functions, or rewriting them into statically dispatching functions that can be inlined to take advantage of type inference.
 We are also investigating an alternate dispatch technique outlined by Robert Strandh\cite{strandh2014fast} that should significantly improve generic function dispatch time, potentially even outperforming traditional vtable based approaches as used in C++ or Java.
@@
 For instance, we can ask the compiler to allocate a number of objects on the stack by using a declaration (\code{(declare (dynamic-extent x))}).
-However, this only works if the compiler can know ahead of time what size the object is, meaning only fixed size arrays and known structs can be stack allocated.
+However, this works only if the compiler can know ahead of time what size the object is, meaning only fixed size arrays and known structs can be stack allocated.
 Furthermore, stack space is much more limited, so bigger allocations still have to happen on the heap.
@@
 For cases where the object is a class instance, simply too big for the stack, or has to escape the stack sometimes, we can still easily cache the instance locally by using \code{load-time-value}.\\\code{load-time-value} is a special operator that causes the form it wraps to be executed when the code is first loaded into the system, and then puts the result of that executed form in the place in the code the \code{load-time-value} was at.
-This effectively creates an anonymous global variable where the object is stored, which is then put in place of the original \code{load-time-value}.
+Another way to understand it is that an anonymous global variable is created where the object is stored, which is then put in place of the original \code{load-time-value}.
 
@@
 In this case, we can define a ``compiler macro'' which is a macro that is executed to replace a call to a particular function.
-This compiler macro can then analyse the arguments to the function at compile-time and choose to substitute the call with a \code{load-time-value} form if all arguments can be determined statically.
-This then avoids the construction and allocation of the object, or even the entire computation of the function at runtime in a, to the user, completely transparent fashion.
+This compiler macro can then analyse the arguments to the function at compile-time and choose to replace the call with a \code{load-time-value} form if all arguments can be determined statically.
+This substitution then avoids the construction and allocation of the object, or even the entire computation of the function at runtime in a, to the user, completely transparent fashion.
 
@@
 The expressiveness of the Common Lisp Object System allows for rapid changes in game object behaviour, as well as the creation of reusable pieces of behaviour that can often be seamlessly combined together.
-With extensions to the Meta Object Protocol, this can even be extended to graphics rendering routines and shaders.
+With extensions to the Meta Object Protocol, this flexibility can even be extended to graphics rendering routines and shaders.
 
 The presence of the entire compiler environment at runtime allows the developer to change any part of the game while it is running.
-This is especially feasible thanks to the presence of the restarts and interactive debugger, which allow the game to recover from a bug without having to crash and restart the entire system.
+Doing so is especially feasible thanks to the presence of the restarts and interactive debugger, which allow the game to recover from a bug without having to crash and restart the entire system.
 
@@
 \section{Acknowledgements}
-We would like to thank the very handsome and attractive people that helped proofread this draft ;)
+We would like to thank Robert Strandh, Florian Hanke, Tobias Mansfield-Williams, and Selwyn Simsek for their feedback on the paper.
 

### 2022-01-19 els2022-lichat/paper.tex
stub

--- 2019-01-30 els2019-shader-pipeline/paper.tex
@@
 \newmintinline[code]{text}{}
-\bibliographystyle{unsrt}
+\bibliographystyle{plainnat}
 
@@
 
+\acmConference[ELS'22]{the 15th European Lisp Symposium}{April 01--02 2022}{%
+  Porto, Portugal}
+\acmISBN{978-2-9557474-3-8}
+\acmDOI{10.5281/zenodo.2636508}
 \setcopyright{rightsretained}
-\acmDOI{}
-\acmISBN{}
-\acmConference[ELS'19]{the 12th European Lisp Symposium}{April 1--2 2019}{%
-  Genova, Italy}
+\copyrightyear{2022}
 
@@
 
-\title{Shader Pipeline and Effect Encapsulation using CLOS}
+\title{}
 
@@
 \begin{abstract}
-  Modern real-time graphics make use of a lot of tricks in order to produce stunning visuals.
-Many of these tricks require separate rendering passes, as well as separate rendering logic for each pass.
-These passes are then combined in a variety of ways in order to produce a final image.
-The interaction between such a rendering pass and the objects it draws, as well as the interaction between multiple passes within a pipeline can become quite complex.
-Often times, in order to handle this complexity, the passes and objects are generalised, and the render logic is left to be executed almost entirely with either the object or the pass.
-We present a new method of representing objects, passes, and pipelines, which allows a modular encapsulation of effects and rendering behaviour, as well as object-oriented composition through inheritance.
-We make use of CLOS' multiple inheritance, multimethods, and standard method combination to form an extensible protocol.
-We also make use of the MOP, in order to introduce additional metadata to classes.
+
 \end{abstract}
@@
 \begin{CCSXML}
-  <ccs2012>
-  <concept>
-  <concept_id>10011007.10010940.10010971.10010972.10010979</concept_id>
-  <concept_desc>Software and its engineering~Object oriented architectures</concept_desc>
-  <concept_significance>500</concept_significance>
-  </concept>
-  <concept>
-  <concept_id>10011007.10010940.10010971.10011682</concept_id>
-  <concept_desc>Software and its engineering~Abstraction, modeling and modularity</concept_desc>
-  <concept_significance>500</concept_significance>
-  </concept>
-  <concept>
-  <concept_id>10011007.10011006.10011066.10011067</concept_id>
-  <concept_desc>Software and its engineering~Object oriented frameworks</concept_desc>
-  <concept_significance>300</concept_significance>
-  </concept>
-  <concept>
-  <concept_id>10010147.10010371</concept_id>
-  <concept_desc>Computing methodologies~Computer graphics</concept_desc>
-  <concept_significance>300</concept_significance>
-  </concept>
-  <concept>
-  <concept_id>10011007.10011006.10011041.10011688</concept_id>
-  <concept_desc>Software and its engineering~Parsers</concept_desc>
-  <concept_significance>100</concept_significance>
-  </concept>
-  </ccs2012>
+
 \end{CCSXML}
 
-\ccsdesc[500]{Software and its engineering~Object oriented architectures}
-\ccsdesc[300]{Software and its engineering~Abstraction, modeling and modularity}
-\ccsdesc[300]{Software and its engineering~Object oriented frameworks}
-\ccsdesc[300]{Computing methodologies~Computer graphics}
-\ccsdesc[100]{Software and its engineering~Compilers}
-
-\keywords{Common Lisp, OpenGL, GPU, CLOS, Object Orientation}
+\keywords{Common Lisp}
 
@@
 \section{Introduction}\label{introduction}
-Modern graphics systems such as OpenGL Core and DirectX offer a lot of customisation to the programmer.
-Particularly, in order to render an image, they allow the programmer to supply code fragments (shaders) that are run directly on the GPU.
-These code fragments fill in steps of a fixed rendering pipeline that is executed on the GPU in order to transform vertex data into the pixels of an image.
-The pipeline for OpenGL is illustrated in \autoref{fig:opengl-pipeline}.
-\\
-
-\begin{figure}[h]
-  \begin{center}
-    \includegraphics[width=0.3\textwidth]{opengl-pipeline.png}
-  \end{center}
-  \caption{The stages of the OpenGL rendering pipeline. White boxes represent stages that can be customised with shader code.}
-  \label{fig:opengl-pipeline}
-\end{figure}
-
-For consistency, we refer to a step within the hardware rendering pipeline as a ``stage'', an invocation of the hardware pipeline as a ``pass'', and all invocations of the hardware pipeline to produce an image as a ``frame''.
-\\
-
-Each of the customisable stages accepts only a single shader for each pass, making it difficult to separate, encapsulate, and ultimately combine behaviour.
-Furthermore, the steps required in order to change the shaders and shader inputs can be non-trivial and expensive to execute.
-\\
-
-\begin{figure}[h]
-  \begin{center}
-    \includegraphics[width=0.3\textwidth]{render-pipeline.png}
-  \end{center}
-  \caption{A sample frame pipeline with shadow mapping, deferred rendering, and bloom. Each box represents a pass and each edge a texture buffer.}
-  \label{fig:render-pipeline}
-\end{figure}
-
-Managing this graphics state and the order of rendering can be very complicated for modern requirements.
-Rendering a frame often requires a multitude of passes, each with their own parameters and shared data.
-The rendering of each object within a pass can also differ, leading to even more state that needs to be correctly managed.
-\\
-
-This complexity results in a difficult challenge for modularity. We attempt to solve this challenge through several systems:
-
-\begin{itemize}
-\item A protocol for communicating information between a pass and the objects rendered within.
-\item A protocol to connect the inputs, outputs, and parameters of different passes.
-\item An algorithm to automatically allocate shared textures used as buffers between passes.
-\end{itemize}
 
 \section{Related Work}\label{relatedwork}
-Courreges\cite{gtav} presents an in-depth analysis of the rendering procedure employed by the modern, high-production game GTA V.
-It illustrates the many passes to produce a final image, as well as their data dependencies.
-\\
-
-Harada et al.'s work on Forward+\cite{forward+}\cite{forward+talk} also clearly illustrates the need for systems that support multi-pass rendering pipelines with complex data interaction schemes. \\
-
-Gyrling\cite{fibers} presents an overview of the techniques used to perform parallel rendering in Naughty Dog's private game engine.
-Individual stages within a pass, render passes of a frame, and multiple frame renderings are divided up into many small jobs that can run in parallel and are synchronised using counters on a shared structure.
-How rendering logic requirements are communicated is however not explained.
-\\
-
-The case study of the Unity game engine by Messaoudi et al\cite{unity} shows the availability of a set of fixed rendering pipelines that can be customised to a very limited extent with custom shaders.
-These shaders must fit into Unity's existing lighting and overall rendering model.
-While Unity does allow building a custom pipeline via their Scriptable Rendering Pipeline\cite{unitycustom}, they do not seem to offer any specific encapsulation or modularity features.
-\\
-
-The work by He et al.\cite{components} introduces a framework for general encapsulation of shaders and their parameters into structures that minimise the overhead of changing GPU state while retaining the ability to dynamically compose shader parts.
-They do however not create a distinction between properties for rendering an object and for those for rendering an overall pass.
-\\
-
-Foley et al.'s work on Spark\cite{spark} presents a high-level graph-based system for defining reusable and composable shader components.
-Their system shows a much more distanced view of the underlying graphics hardware than we attempt.
-Similar to He's work, they do not present a separation between object logic and pass logic.
-\\
-
-In our previous work\cite{glsloop} we introduce a system to tie shader code to classes and to compose behaviour through inheritance.
-We make use of this system and extend it to allow further control over rendering behaviour in individual passes.
 
 \section{Overview}\label{overview}
-The system is composed of three distinct entities: shader objects, shader passes, and pipelines.
-Both shader objects and shader passes encapsulate rendering logic.
-The pipeline, on the other hand, represents the assembly of a complete frame, and only influences rendering in the sense that it tracks which passes to run in what order.
-\\
-
-Due to the restrictions of the rendering pipeline of OpenGL, the combination of the render logic of objects and of passes is not trivial.
-We present a protocol to solve this difficulty in \hyperref[passes]{section 4}.
-\\
-
-In \hyperref[pipelines]{section 5} we present a protocol for representing input and output information of a pass and for connecting these inputs and outputs between passes together. \\
-
-Due to the complex interactions between passes in a pipeline, allocation of intermediary buffer textures is error-prone and tedious.
-We present algorithms to automate this in \hyperref[allocation]{section 6}.
-\\
-
-Finally, in \hyperref[concept]{section 7} we show the results of implementing a medium-sized pipeline using these techniques.
-
-\section{Passes}\label{passes}
-A shader pass should encapsulate the logic for drawing objects in a certain way.
-In order to accomplish its task, the pass needs to be able to partially or fully control the shaders of objects.
-For instance, an object should still be able to have control over how its vertices are constructed and what materials and textures are applied to it.
-The shader pass should be able to make use of this information if it needs to, or discard it completely if unneeded.
-\\
-
-In order to permit this amount of control, we make use of the shader composition capabilities that we presented in our previous work\cite{glsloop}.
-This technique allows us to combine pieces of GLSL code.
-On top of this shader combination sits a new protocol to control the interaction between shader passes and objects.
-\\
-
-\begin{listing}[h]
-\begin{minted}[fontsize=\small]{common-lisp}
-(defclass shader-pass ()
-  ()
-  (:metaclass shader-pass-class))
-
-(defgeneric register-object-for-pass (shader-pass object))
-(defgeneric shader-program-for-pass (shader-pass object))
-(defgeneric make-pass-shader-program (shader-pass object))
-(defgeneric coerce-pass-shader (shader-pass object stage))
-(defgeneric render-with (shader-pass object))
-(defgeneric render (object target))
-\end{minted}
-\caption{The protocol for shader passes.}
-\label{lst:pass}
-\end{listing}
-
-For any object that should be rendered using a given pass, first \code{register-object-for-pass} must be called.
-This call allows the pass to prepare the shader program that will be used during rendering for this object.
-A shader program is an OpenGL resource that compiles the shaders for the stages of a pass together.
-Using \\ \code{shader-program-for-pass} this program can then later be retrieved.
-The shader program is important for objects to access, as it allows them to set values for ``uniform'' variables that are used in the shader code.
-\\
-
-\code{register-object-for-pass} calls \code{make-pass-shader-program} to compute this program, and registers it internally in the shader pass so that it can be retrieved later.
-If the shader pass would like to retain complete control over how each object is drawn, it would forego this call and instead generate its own shader program.
-Doing so is especially useful for post-processing effects that don't render any objects at all, and instead simply operate on textures that are output by previous passes.
-\\
-
-\code{make-pass-shader-program} gathers all the shader sources for a program using \code{coerce-pass-shader} and then generates a representation for an OpenGL shader program.
-This representation also includes additional information such as the data buffers used.
-\\
-
-\code{coerce-pass-shader} computes the effective shader source for a particular shader stage or type.
-Typically this computation simply involves the combination of the shader sources of both the shader pass and of the object, for the given shader type.
-We perform this combination using the same parsing and code walking strategy as we described in our previous work.\cite{glsloop} \\
-
-Objects are typically rendered using the \code{render} function.
-Users are encouraged to add methods that specialise the behaviour and perform necessary setup, as well as to perform the final draw call for their custom object classes.
-For instance, a very primitive class could look as shown in \autoref{lst:simple-object}.
-\\
-
-\begin{listing}[h]
-\begin{minted}[fontsize=\small]{common-lisp}
-(defclass simple-object ()
-  ((vertex-array :accessor vertex-array)
-  (:metaclass shader-class))
-
-(defmethod render ((oject simple-object) target)
-  (let ((vao (vertex-array object)))
-    (gl:bind-vertex-array (gl-name vao))
-    (%gl:draw-elements :triangles (size vao) :unsigned-int 0)))
-
-(defmethod render :before ((object simple-object)
-                           (pass shader-pass))
-  (let ((program (shader-program-for-pass pass object)))
-    (setf (uniform program "projection_matrix")
-          (projection-matrix))))
-\end{minted}
-\caption{A simple object class and its render methods. The first method tells OpenGL to render a list of vertices. The second method sends the projection matrix to the GPU via a uniform variable.}
-\label{lst:simple-object}
-\end{listing}
-
-This function is fine for allowing the object control over the behaviour.
-However, the pass cannot exert the same amount of control, due to the generic function's argument precedence.
-For example, if an \code{:around} method specialised on the object exists, a pass would not have any way of preventing it from firing, as its own \code{:around} method would be executed afterwards.
-We thus introduce another function with inverted argument order that is called first.
-\\
-
-\code{render-with} thus exists mostly as an entry point to allow shader passes greater control over the rendering.
-Typically it will simply defer to \code{render}.
-This inversion is very important for shader passes that take all control away from the objects.
-\\
-
-This protocol thus allows a great amount of control both over the effective shader code used to render an object, as well as the behaviour leading up towards the actual draw call for an object.
-
-\section{Pipelines}\label{pipelines}
-Shader passes not only retain information about how objects are drawn, but also about the input and output textures that a pass interacts with.
-For this reason, each shader pass is also a node in a graph, with distinct input and output ports.
-
-\begin{listing}[h]
-\begin{minted}[fontsize=\small]{common-lisp}
-(defclass deferred-render (shader-pass)
-  ((position-map :port-type input)
-   (normal-map :port-type input)
-   (albedo-map :port-type input)
-   (color :port-type output))
-  (:metaclass shader-pass-class)
-\end{minted}
-\caption{An outline of a deferred rendering pass, taking position-, normal-, and albedo-map textures as input, producing a single color texture as output.}
-\label{lst:deferred-render}
-\end{listing}
-
-An example of such a pass is illustrated in \autoref{lst:deferred-render}.
-The ports are modelled as slots of the class that carry additional metadata, requiring a new metaclass.
-A slot with a port-type will not only hold a texture for its value, but also retain information about how it is connected to other passes.
-Using slots in this manner also allows us to inherit ports from other classes, and thus combine behaviour alongside the shader source code.
-\\
-
-For the sake of brevity and ease of explanation, we have left out the details of texture constraints in these code samples.
-Nevertheless, constraining the features of textures tied to the ports is important, and we discuss the technique for dealing with that in \hyperref[allocation]{section 6}.
-\\
-
-\begin{listing}[h]
-\begin{minted}[fontsize=\small]{common-lisp}
-(defclass shadow-render (shader-pass)
-  ((shadow-map :port-type input)
-   (color :port-type output))
-  (:metaclass shader-pass-class))
-\end{minted}
-\caption{An outline of a shadow rendering pass, taking into account the information from a shadow-map to render shadows onto the output color texture.}
-\label{lst:shadow-render}
-\end{listing}
-
-\begin{listing}[h]
-\begin{minted}[fontsize=\small]{common-lisp}
-(defclass high-pass-render (shader-pass)
-  ((high-pass :port-type output))
-  (:metaclass shader-pass-class))
-\end{minted}
-\caption{An outline of a high-pass renderer, which splices off colours of a high intensity into a high-pass output texture.}
-\label{lst:high-pass}
-\end{listing}
-
-For instance, the ``Deferred Render'' pass shown in \autoref{fig:render-pipeline} is created by combining the passes illustrated in \autoref{lst:deferred-render}, \autoref{lst:shadow-render}, and \autoref{lst:high-pass} through inheritance.
-\\
-
-When assembling a pipeline, we can then connect these ports together, in order to dictate how the various output textures generated by a pass are used as inputs for other passes.
-As an example, the pipeline from \autoref{fig:render-pipeline} is created in \autoref{lst:pipeline}.
-\\
-
-\begin{listing}[h]
-\begin{minted}[fontsize=\small]{common-lisp}
-(let ((pipeline (make-pipeline))
-      (shadow (make-instance 'shadow-map-pass))
-      (geometry (make-instance 'geometry-pass))
-      (deferred (make-instance 'deferred+shadow-pass))
-      (blur (make-instance 'gaussian-blur-pass))
-      (composite (make-instance 'composite-pass)))
-  (connect (port shadow 'shadow-map) 
-           (port deferred 'shadow-map) pipeline)
-  (connect (port geometry 'position-map)
-           (port deferred 'position-map) pipeline)
-  (connect (port geometry 'normal-map)
-           (port deferred 'normal-map) pipeline)
-  (connect (port geometry 'albedo-map)
-           (port deferred 'albedo-map) pipeline)
-  (connect (port deferred 'high-pass)
-           (port blur 'previous-pass) pipeline)
-  (connect (port deferred 'color)
-           (port composite 'color) pipeline)
-  (connect (port blur 'color)
-           (port composite 'bloom) pipeline)
-  (prepare pipeline))
-\end{minted}
-\caption{An assembly of the pipeline shown in \autoref{fig:render-pipeline}.}
-\label{lst:pipeline}
-\end{listing}
-
-The pipeline object itself retains information on the order in which the passes should be executed, and keeps track of the textures that are allocated in order to run the passes.
-How this information is computed is described in \hyperref[allocation]{section 6}.
-\\
-
-The procedure to actually render objects with this pipeline technique is then as follows:
-
-\begin{step}
-\item Create a pipeline object and instances of the desired shader passes.
-\item Connect the ports of the shader passes.
-\item Prepare the pipeline to allocate the needed resources.
-\item Create instances of the desired objects.
-\item Register each object with each pass.
-\item Call \code{render} on the pipeline with a collection of all objects to draw.
-\item The output texture of the output port of the last shader pass in the pipeline will contain the finished frame.
-\end{step}
-
-The completed frame can then be blitted onto the screen, or be used as the input for another computation.
-
-\section{Allocation}\label{allocation}
-In order to prepare the shader pipeline, several resources need to be allocated.
-Each shader stage needs a ``framebuffer,'' an OpenGL resource that allows one to render to off-screen textures.
-These framebuffers then need to have the required textures to render allocated as well.
-As each input and output from a pass can specify constraints on the features of the texture, these constraints must be matched up for any connecting edges as well.
-Finally, in order to minimise memory usage, we would like to re-use textures where possible.
-\\
-
-Thus, the allocation proceeds in three phases: reconciling texture constraints on edges between passes, computing how textures are shared between passes, and finally constructing all the necessary resources with the previously gathered information.
-
-\subsection{Constraint Merging}\label{texture-constraints}
-OpenGL textures include a massive amount of information\cite{texparameter}\cite{teximage2d}.
-When two ports are connected that specify different constraints on the texture properties, a join must be performed.
-A wide range of the texture property values are fundamentally incompatible, meaning that a lot of the logic can simply error.
-However, other options require more complicated joining logic.
-For simplicity and brevity, we will focus on the join operator for a single texture property here: the internal format.
-The internal format is arguably the most important property.
-This property specifies how many colour channels the texture has, how many bits of precision each channel has, which format each channel has, as well as whether the texture is compressed or has sRGB gamma normalisation applied.
-\\
-
-The list of specified texture formats is quite large\cite{texinternalformat}.
-Unfortunately OpenGL does not give us an interface to handle these formats in a way that lets us pick the individual features easily.
-Instead, each format is represented by a constant whose value has no relation to the features that format includes.
-This representation means that we first need to destructure each format name into a list of features: \\
-
-\begin{itemize}
-\item \code{R}, \code{G}, \code{B}, \code{A} How many bits to use for each channel, and the format of the channel (normalised, float, integer, unsigned integer).
-\item \code{depth} How many bits to use for the depth channel.
-\item \code{stencil} How many bits to use for the stencil channel.
-\item \code{shared} Whether bits are shared across the channel, and if so how many.
-\item \code{features} Whether the format has compression, sRGB, RGTC, BPTC, SNORM, or UNORM features.
-\end{itemize}
-
-Once the texture format specifications are destructured, we can perform a join as follows.
-
-\begin{step}
-\item If the \code{features} and sharing are not the same, a join is impossible.
-\item The \code{features} list of the output spec is set to the same as either of the specs.
-\item If both include a \code{depth} component:
-  \begin{step}
-  \item The \code{depth} feature is set to the join of both.
-  \item If either include a \code{stencil} feature, the \code{stencil} feature is set to the join of both.
-  \end{step}
-\item If both include a \code{stencil} component:
-  \begin{step}
-  \item The \code{stencil} feature is set to the join of both.
-  \end{step}
-\item If neither include a \code{depth} component:
-  \begin{step}
-  \item The \code{R} feature is set to the join of both.
-  \item The \code{G} feature is set to the join of both.
-  \item The \code{B} feature is set to the join of both.
-  \item The \code{A} feature is set to the join of both.
-  \end{step}
-\item Otherwise a join is impossible.
-\end{step}
-
-Wherein the ``join of both'' is computed as the join of two channel formats as follows.
-
-\begin{step}
-\item If both include the format:
-  \begin{step}
-  \item If the channel format is not the same, a join is impossible.
-  \item The channel bit depth is set to the maximum of both.
-  \end{step}
-\item If one includes the format, that format is returned.
-\item Otherwise, the absence of the channel is indicated.
-\end{step}
-
-If the join is successful, we then re-encode the texture format specification into OpenGL's constant and use this constant in the real texture specification.
-Note that this join could produce texture format specifications that are not legal according to the OpenGL specification.
-However, this can only occur if one of the specifications to join is already illegal.
-We thus deem it unnecessary to handle such cases.
-
-\subsection{Port Allocation}\label{port-allocation}
-In compilers, allocation of a graph of variables with use-relations is typically handled with a graph colouring algorithm.
-However, since we represent our graph in a different fashion than usual, with nodes having multiple distinct ports on which edges are connected, we devised a different kind of colouring algorithm to maximise texture sharing.
-\\
-
-Given a set of nodes the allocation algorithm proceeds as follows.
-
-\begin{step}
-\item The set of unique texture specifications is computed by joining each port's texture specification with every other and comparing for equality.
-\item For each unique texture specification T:
-  \begin{step}
-  \item The nodes are sorted topologically.
-  \item The number of colours is set to 0.
-  \item For each node N:
-    \begin{step}
-    \item For each output port P of N:
-      \begin{step}
-      \item If P's texture specification is joinable with T...
-      \item The number of colours is increased
-      \end{step}
-    \end{step}
-  \item An array is allocated to fit the number of colours. Each index of the array represents a colour and each value at the index whether the colour is currently available or unavailable.
-  \item For each node N in \emph{reverse order}:
-    \begin{step}
-    \item For each input port P of N:
-      \begin{step}
-      \item For each neighbour port O of P:
-        \begin{step}
-        \item If O's texture specification is joinable with T...
-        \item and O does not yet have a colour...
-        \item The first available colour is assigned to O.
-        \item This colour is marked as unavailable.
-        \end{step}
-      \end{step}
-    \item For each non-input port P of N:
-      \begin{step}
-      \item If P's texture specification is joinable with T...
-      \item and P does not yet have a colour...
-      \item The first available colour is assigned to P.
-      \item This colour is marked as unavailable.
-      \end{step}
-    \item For each port P of N:
-      \begin{step}
-      \item If P's texture specification is joinable with T...
-      \item and P has a colour...
-      \item P's colour is marked as available.
-      \end{step}
-    \end{step}
-  \end{step}
-\end{step}
-
-In other words, the algorithm proceeds backwards from the last node in the graph, marking output ports of predecessors with unique colours, then marking unconnected ports with unique colours, and finally marking all colours at the node's own output ports as available again.
-We repeat this process for each unique texture specification, each time ensuring we only consider ports that share that texture specification.
-\\
-
-This algorithm is by no means efficient, but since pipeline allocation only has to happen during loading phases, we currently do not consider this to be a big problem.
-We also have not performed any analysis as to whether the algorithm produces optimal allocation results in every case.\\
-
-Unfortunately printed documents cannot yet display animations, so illustrating the algorithm in motion is not possible directly in the document.
-However, a brief animation of the pipeline illustrated in \autoref{fig:render-pipeline} is available online: \\
-
-\href{https://raw.githubusercontent.com/Shinmera/talks/master/els2019-shader-pipeline/pipeline-allocation.gif}{https://raw.githubusercontent.com/Shinmera/talks/master/\\els2019-shader-pipeline/pipeline-allocation.gif}
-
-\section{Proof of Concept}\label{concept}
-As a proof of concept we have implemented these protocols and mechanisms, as well as the passes shown in \autoref{fig:render-pipeline}.
-As the code required to do so is quite lengthy and involved, and the graphics techniques used are beyond the scope of this paper, we will omit the code.
-The relevant outline of the passes and the pipeline involved has already been shown in \hyperref[pipelines]{section 5}.
-\\
-
-Nevertheless, you can find the complete code online at the following pages: \\
-\url{https://github.com/Shirakumo/trial/blob/b889d50/deferred.lisp} \\
-\url{https://github.com/Shirakumo/trial/blob/b889d50/hdr.lisp} \\
-\url{https://github.com/Shirakumo/trial/blob/b889d50/shadow-map.lisp} \\
-\url{https://github.com/Shirakumo/trial/blob/b889d50/workbench.lisp} \\
-
-\autoref{fig:pipeline-gbuffer}, \autoref{fig:pipeline-shadow}, \autoref{fig:pipeline-lighting}, \autoref{fig:pipeline-blur}, and \autoref{fig:pipeline-frame} show the output textures produced by the various stages for a simple scene.
-
-\begin{figure}[h]
-  \begin{center}
-    \begin{minipage}{.2\textwidth}
-      \includegraphics[width=1.0\textwidth]{pipeline-position.png}
-    \end{minipage}
-    \begin{minipage}{.2\textwidth}
-      \includegraphics[width=1.0\textwidth]{pipeline-normal.png}
-    \end{minipage}
-    \vskip0.04cm
-    \begin{minipage}{.2\textwidth}
-      \includegraphics[width=1.0\textwidth]{pipeline-albedo.png}
-    \end{minipage}
-    \begin{minipage}{.2\textwidth}
-      \includegraphics[width=1.0\textwidth]{pipeline-specular.png}
-    \end{minipage}
-  \end{center}
-  \caption{The output of the geometry pass, producing a ``G-buffer'' composed out of a position- (top left), normal- (top right), albedo- (bottom left), and specular-map (bottom right).}
-  \label{fig:pipeline-gbuffer}
-\end{figure}
-
-\begin{figure}[h]
-  \begin{center}
-    \includegraphics[width=0.2\textwidth]{pipeline-shadow.png}
-  \end{center}
-  \caption{The depth map produced by the shadow map pass. It shows the distance of surfaces from the point of view of the primary light source.}
-  \label{fig:pipeline-shadow}
-\end{figure}
-
-\begin{figure}[h]
-  \begin{center}
-    \begin{minipage}{.2\textwidth}
-      \includegraphics[width=1.0\textwidth]{pipeline-lighting.png}
-    \end{minipage}
-    \begin{minipage}{.2\textwidth}
-      \includegraphics[width=1.0\textwidth]{pipeline-highpass.png}
-    \end{minipage}
-  \end{center}
-  \caption{The color map (left) and the high-pass map (right) that is output by the deferred render pass.
-It combines the information from the G-buffer from \autoref{fig:pipeline-gbuffer}, the shadow-map from \autoref{fig:pipeline-shadow}, and information for lights to render an image.
-Note that due to gamma adjustments the image appears quite dark.}
-  \label{fig:pipeline-lighting}
-\end{figure}
-
-\begin{figure}[h]
-  \begin{center}
-    \includegraphics[width=0.2\textwidth]{pipeline-blur.png}
-  \end{center}
-  \caption{The blurred high-pass from \autoref{fig:pipeline-lighting} produced by the gaussian blur pass. This is used to produce an effect called bloom in the final composite pass.}
-  \label{fig:pipeline-blur}
-\end{figure}
-
-\begin{figure}[h]
-  \begin{center}
-    \includegraphics[width=0.4\textwidth]{pipeline-frame.png}
-  \end{center}
-  \caption{The final output frame that is produced by adding together the bloom texture from \autoref{fig:pipeline-blur} and the standard colour texture from \autoref{fig:pipeline-lighting}.
-Gamma adjustment and High Dynamic Range reduction are used for colour normalisation.}
-  \label{fig:pipeline-frame}
-\end{figure}
 
 \section{Conclusion}\label{conclusion}
-We have presented a system to handle the separation between objects that should be rendered, and the way in which they should be rendered.
-By using a protocol for the interaction between such objects and passes, we are capable of controlling and combining a variety of rendering behaviours.
-With an additional protocol and set of algorithms we also present a high-level view of rendering pipelines, allowing users to conveniently compose effects to produce complicated scenes.
-\\
-
-These protocols and algorithms are largely generic enough that they could also be adapted for use in other languages and systems.
-However, we believe that the Common Lisp Object System gives us a set of very convenient tools to make everything feel more natural and more integrated with the rest of the system than other environments would allow us to do.
-\\
-
-Finally, we have demonstrated the capability of this system by constructing a real-time rendering system out of individual, modular pieces of code.
 
 \section{Further Work}\label{further-work}
-Currently there exist several restrictions in our framework. \\
-
-Most severe is that there is no standardised protocol to communicate capabilities and data between objects and passes.
-This restriction means that, for instance, an object that does not possess texturing has no way of communicating this to a shader pass that might need to render it.
-Currently this problem is partially solved by introducing a mandatory superclass that objects need to be a subtype of, if they want to be renderable under a given pass.
-This superclass then loosely defines the interaction.
-However, this additional class does not solve issues of compatibility of shader code used by the object and by the shader pass.
-In order to properly resolve this restriction, we currently see two required features: \\
-
-Shader source merging must be aware of uses-relations, meaning that the code walker must track where variables, functions, and types are used and reorder the definitions and declarations in such a way that the dependencies are fulfilled.
-In GLSL, variables, functions, and types cannot be referenced before they are declared, requiring this reordering step.
-With this problem solved, shader code could be written in such a way to be more amiable towards combination and manipulation by external code.
-\\
-
-GLSL's limited expressiveness makes it very difficult to figure out relations between code and how to properly combine pieces.
-A more high-level language, as is used in other frameworks like Spark\cite{spark} or Shader Components\cite{components}, should be introduced to allow a more convenient way to declare and write shaders, objects, and passes.
-With a higher-level language, a compiler could more easily figure out how to combine features and reconcile differences between pieces of code.
-\\
-
-Another restriction is that there is currently no way to encapsulate GLSL code other than associating it with a class, which is a rather heavy-weight operation.
-It would be far more useful to introduce a library mechanism that allows the definition of standalone shader functions, which can then be required by shader classes.
 
 \section{Acknowledgements}\label{acknowledgements}
-I would like to thank Michal ``phoe'' Herda, Robert Strandh, Michael Reis, and all the people that will review this paper. Your name could be here!
-
-\section{Implementation}\label{implementation}
-An implementation of the proposed system can be found at \\
-\url{https://github.com/Shirakumo/trial/blob/b889d50/shader-pass.lisp} \\
-\url{https://github.com/Shirakumo/trial/blob/b889d50/pipeline.lisp} \\
-\url{https://github.com/Shinmera/flow} \\
 

### 2022-01-19 els2022-lichat/paper.tex
Start writing the paper.

@@
 \usepackage{hyperref}
+\usepackage[parfill]{parskip}
 \usepackage[all]{hypcap}
-\usemintedstyle[glsl]{default}
 \usemintedstyle[common-lisp]{default}
@@
 
-\acmConference[ELS'22]{the 15th European Lisp Symposium}{April 01--02 2022}{%
+\acmConference[ELS'22]{the 15th European Lisp Symposium}{March 21--22 2022}{%
   Porto, Portugal}
@@
 
-\title{}
+\title{Lichat, A Lightweight Chat Protocol}
 
@@
 \begin{abstract}
-
+  In this paper we present Lichat, a simple and lightweight chat protocol.
+We take particular interest in how relying on Common Lisp as a base language led to a fast prototyping of the protocol and rapid evolution of its design, and then evaluate the consequences of these decisions when porting the protocol to other languages such as Java, JavaScript, Python, and Elixir.
 \end{abstract}
@@
 \begin{CCSXML}
-
+  <ccs2012>
+  <concept>
+  <concept_id>10003033.10003039.10003051</concept_id>
+  <concept_desc>Networks~Application layer protocols</concept_desc>
+  <concept_significance>500</concept_significance>
+  </concept>
+  <concept>
+  <concept_id>10002951.10003260.10003304</concept_id>
+  <concept_desc>Information systems~Web services</concept_desc>
+  <concept_significance>500</concept_significance>
+  </concept>
+  <concept>
+  <concept_id>10011007.10011074.10011092.10010876</concept_id>
+  <concept_desc>Software and its engineering~Software prototyping</concept_desc>
+  <concept_significance>300</concept_significance>
+  </concept>
+  </ccs2012>
 \end{CCSXML}
 
-\keywords{Common Lisp}
+\ccsdesc[500]{Networks~Application layer protocols}
+\ccsdesc[500]{Information systems~Web services}
+\ccsdesc[300]{Software and its engineering~Software prototyping}
+
+\keywords{Common Lisp, Chat, Protocol, Design, Networking}
 
@@
 \section{Introduction}\label{introduction}
+The IRC (Internet Relay Chat) protocol has long stood the test of time as a reliable protocol for text chat communication.
+Large networks and a plethora of mature clients still exist and are in use today.
+However, The IRC protocol is also plagued by a number of issues that stem from its historical design and largely distributed evolution.
+In particular, the protocol has many different, incompatible extensions, and features many idiosyncrasies that make it difficult to write new clients or servers.
+
+IRC has also failed to catch up with several changes in user expectations for chat protocols today.
+It lacks support for custom emoticons, search, formatting, does not permit multiple connections to the same user, and does not even offer an account registration or authorisation process.
+Instead the onus is on the servers to offer extensions, and on the clients to implement such features completely locally (forgoing synchronisation).
+
+Over time many other chat systems such as AIM, ICQ, Skype, have come and largely gone, though all of these focus on a different model of operation than IRC.
+They focus on a friends-list based approach where people talk to each other directly, rather than relying on sometimes huge public chat rooms.
+
+More recently, proprietary systems such as Slack, Discord, and Microsoft Teams have sprung up that follow the IRC model more closely, though all focus on a more sheltered experience where communities are gated off via separate ``servers'' in which the channels reside.
+People cannot see channels from other servers without being invited.
+All of these solutions use proprietary, closed-source protocols and often even prohibit the usage of third-party clients, severely restricting user-freedom and endangering the longevity of the platform.
+
+On the open source side, the Matrix (formerly Riot) protocol has been trying to offer a modernised, federated approach to the IRC model.
+Being federated and open it has a much higher chance at staying around for decades to come than many of the aforementioned proprietary solutions.
+However, just like the proprietary counterparts, the Matrix protocol is complex and far from trivial to implement.
+
+With Lichat we instead focused on developing a protocol that should be trivial to implement a client for, and not too hard to implement a server for.
+We also focused on a design that would avoid many of the pitfalls we found in IRC, and in general strives for a few basic properties that ensure reliable and easy to understand communication between the client and server.
+The core protocol of Lichat is very small, only covering the most basic elements, instead relying on protocol extensions to provide additional functionality that bring it up to par with competing systems such as Discord or Matrix.
+
+In \autoref{initial-design} we present our initial prototyping for the protocol and illustrate the advantages of picking Lisp as a starting point.
+In \autoref{ratified-design} we discuss how the design evolved after the initial prototyping phase and discuss the porting of the protocol implementations to other languages.
+Finally, in \autoref{extensions} we discuss the use and development of extensions to provide additional functionality.
 
 \section{Related Work}\label{relatedwork}
+IRC
 
-\section{Overview}\label{overview}
+Matrix
+
+XMPP
+
+Whatever?
+
+\section{Initial Design}\label{initial-design}
+As a basis for the protocol we chose to use UTF-8 character streams for the transport.
+This ensures that protocol messages can be read and written by a human, which eases development and debugging.
+Specifying the character encoding also avoids ambiguity when interpreting the underlying bytes.
+For the actual data format, we relied on Common Lisp s-expressions, which could simply be \code{read} and \code{print}ed using the native Lisp functions.
+
+Doing so both eliminated the need to write encoders for the prototype, and ensured that we had a data structure basis that could support more complex data payloads further down the line.
+The actual message data types communicated with were also directly represented using Common Lisp standard classes, allowing for tight encapsulation of data attributes, and the natural combination thereof through multiple inheritance.
+
+Each message to be sent was trivially encoded by printing a list consisting of the message's type and its construction arguments.
+The decoding then followed just as trivially.
+Doing so meant we could set up a server-client communication with as close to no effort as possible while remaining extensible for future changes.
+Particularly, the choice of symbols for message type names and argument names allowed us to use packages to isolate additions made by extensions into their own namespaces, avoiding future conflicts.
+
+Another of the issues with IRC's design is that it is not strictly possible to verify whether messages arrived at the server at all, and if a response is given, which request that response belongs to.
+To rectify this issue, we require the inclusion of an \code{:id} field in every message sent, which the server will either re-use on a mirror-reply, or include as a reference in case of a failure.
+We also require the server to always reply to a message with some form of a response, even if it is just sending the original message back to the client.
+
+The client should also include a \code{:clock} field of the local Universal-Time at the time of sending.
+This ensures that, if distributed to other users, they can get an accurate ordering of events that reflects the sending user's point of view, without potential network delays changing the sequence of events.
+It also allows the server and clients to estimate lag in the network and issue warnings on unstable connections, or terminate them early, ensuring an ultimately more stable communication.
+
+We also require the inclusion of a \code{:version} field on connection to allow the server to reject the connection or switch to a compatibility mode.
+We also include a \code{:extensions} field to allow the listing of supported protocol extensions from the client, which the server then modifies with its own set of supported extensions when confirming the connection.
+This way both server and client are fully aware of supported extensions that they should deal with.
+
+On the client to client communication side, we decided to enforce all communication between clients to occur over channels.
+This simplifies the design on both sides, as there don't need to be any special rules to deal with direct communication.
+With all of this groundwork laid, we could implement a basic chat server that allowed for communication over multiple channels, as shown in \autoref{lst:basic-exchange}.
+
+\begin{listing}[h]
+\begin{minted}[fontsize=\scriptsize]{common-lisp}
+> (connect :id 0 :clock 5243 :from "tester" :version "1.0" :extensions ())
+< (connect :id 0 :clock 5243 :from "tester" :version "1.0" :extensions ())
+> (join :id 0 :clock 5243 :from "tester" :channel "lichatters")
+< (join :id 0 :clock 5243 :from "tester" :channel "lichatters")
+< (message :id 0 :clock 5244 :from "shinmera" :channel "lichatters" :text "Welcome!")
+> (message :id 0 :clock 5246 :from "tester" :channel "lichatters" :text "Hiya")
+< (message :id 0 :clock 5246 :from "tester" :channel "lichatters" :text "Hiya")
+\end{minted}
+\caption{A basic protocol exchange.}
+\label{lst:basic-exchange}
+\end{listing}
+
+After this most of the protocol design revolved around the implementation of a permissions system to allow channel operators better control over what users are allowed to do, and the ratification of distinct error types for potential failures.
+
+The permissions system in general is very simple: each channel holds a map where each update type is associated with with either a list of permitted users, or a list of denied users.
+As all communication between users must occur through channels, this provides a simple, but expressive model to handle moderation.
+For instance, banning a user simply requires removing their ability to send a \code{join} message to the channel.
+
+For failures, a simple hierarchy was designed:
+
+\begin{step}
+\item \code{failure} Any kind of issue
+  \begin{step}
+  \item \code{malformed-update} The message could not be read at all
+  \item \code{update-too-long} The message was too long and was skipped wholesale
+  \item \code{update-failure} An issue with a request the user sent
+    \begin{step}
+    \item \code{insufficient-permissions} The message was denied
+    \item \code{no-such-channel} The requested channel does not exist
+    \item ...
+    \end{step}
+  \end{step}
+\end{step}
+
+Particularly, \code{update-failure}s carry a field called \code{:update-id} which refers back to the ID of the original request, allowing the client to correlate the failure with the previous request, avoiding the issue of confusing responses should multiple requests and responses be in flight at the same time.
+
+Finally, in order to ensure that permissions tables actually have any grip at all, and to allow users to use multiple connections at once, we introduced an accounts system against which the user is authenticated when connecting.
+
+\section{Ratified Design}\label{ratified-design}
+After the initial proof of concept, we decided to simplify the protocol in order to ease the porting process.
+The first part was to reduce the s-expression syntax to make the writing of a compliant parser easier, and safer.
+To this end, we restricted the possible data types to the following:
+
+\begin{step}
+\item \code{number}
+  \begin{step}
+  \item \code{integer}
+  \item \code{float}
+  \end{step}
+\item \code{symbol} (names are case-insensitive, no gensyms)
+  \begin{step}
+  \item \code{keyword}
+  \item \code{boolean}
+  \end{step}
+\item \code{list}
+\item \code{string}
+\end{step}
+
+Further, we required the use of a null byte as an end of message marker.
+This is useful so that a server or client can restore the stream of messages should a parser failure occur, or should the update be rejected due to excessive length.
+
+To ease the initiation and maintenance of the connection, the \code{:from} and \code{:clock} fields were made optional, instead requiring the server to automatically fill them in should they be missing.
+This makes it even more trivial to write a simple, first client.
+
+We decided to specify two distinctions in channel use from regular channels: the ``primary'' channel, and ``anonymous'' channels.
+The primary channel is used to keep track of all users on the server, as all users are automatically joined to this channel after connecting.
+Anonymous channels are used for private ``direct'' communications between users.
+They are named such as the name is picked automatically by the server, and they won't appear in channel listings.
+Otherwise they operate exactly the same as any other channel, making it trivial for clients to implement private user to user communication, which can also be extended to private group chats by inviting other users.
+
+The primary channel is particularly useful to manage permissions, as all updates that do not target a specific channel instead default to having the access checked against the primary channel.
+
+We then implemented a client in JavaScript, Java, and Python, as well as a server in Elixir.
+For each of those the porting process for the basic syntax of the protocol was fairly simple, only requiring a recursive descent parser that is trivial enough to write by hand.
+Most of the issues arose when porting the message class hierarchy.
+
+\section{Extensions}\label{extensions}
 

### 2022-01-20 els2022-lichat/paper.tex
More work.

@@
 
+The initial client and server implementations were fully in Common Lisp, with a JavaScript client intended for public use.
+However, we ran into scaling and concurrency issues with the Common Lisp server that proved extremely difficult to debug and understand, as they often took days on the public server to be triggered.
+The lack of primitives for efficient asynchronous programming and lightweight processes pushed us to attempt a rewrite of the server in Elixir instead, once the core protocol was ratified fully.
+
 \section{Ratified Design}\label{ratified-design}
@@
 
-The primary channel is particularly useful to manage permissions, as all updates that do not target a specific channel instead default to having the access checked against the primary channel.
+The primary channel is particularly useful to manage permissions, as all messages that do not target a specific channel instead default to having the access checked against the primary channel.
 
-We then implemented a client in JavaScript, Java, and Python, as well as a server in Elixir.
+Finally, we ratified the definition of all message types in a machine-readable format (\autoref{lst:spec}) that can be parsed using the same parser of the wire format.
+This allows protocol implementations to easily parse up-to-date representations of the entire message type hierarchy.
+
+\begin{listing}
+\begin{minted}[fontsize=\footnotesize]{common-lisp}
+(define-package "lichat")
+
+(define-object lichat:update ()
+  (:id id)
+  (:clock integer :optional)
+  (:from string :optional))
+
+(define-object lichat:channel-update (lichat:update)
+  (:channel string))
+
+(define-object lichat:text-update (lichat:update)
+  (:text string))
+
+(define-object lichat:message (lichat:channel-update lichat:text-update))
+\end{minted}
+\caption{A section of the machine-readable protocol specification}
+\label{lst:spec}
+\end{listing}
+
+The core protocol specification has been incredibly stable since these changes, only requiring minor amendments to clarify and disambiguate certain points.
+Almost all of the active development since happened within extensions and their specifications, instead.
+
+\section{Porting}
+We implemented client software in JavaScript, Java, and Python, as well as a server in Elixir.
 For each of those the porting process for the basic syntax of the protocol was fairly simple, only requiring a recursive descent parser that is trivial enough to write by hand.
 Most of the issues arose when porting the message class hierarchy.
+
+In Python we were able to make use of multiple superclasses as we would in Common Lisp, keeping the implementation quite straight-forward.
+
+JavaScript does not support multiple inheritance out of the box, so we implemented our own scheme following the serialised inheritance model of CLOS.
+This does however still not integrate neatly with the native JavaScript type system, meaning we also require the use of a specialised type test for dispatch.
+
+Similarly to JavaScript, Java does not support multiple inheritance.
+Instead of defining an ad-hoc system however, we opt for a stubbing approach where we create a single inheritance chain, instead transforming extraneous types into additional class fields.
+The Java implementation is further restricted by its static typing requirements, forcing additional logic when parsing concrete class instances from the message representation.
+Finally, Java does not support runtime creation of classes, requiring us to write a code generator that emits static Java code, instead of creating the classes at runtime like the Python and JavaScript implementations.
+
+Our Elixir server implementation eschews inheritance altogether, as the language does not share an object-oriented system at all.
+We instead manually flatten the inheritance by specifying all inherited fields on each message type.
+Each message type is represented through a module, with a function to handle the update on the server-side.
+Thanks to Elixir's macro system we were able to avoid most of the boilerplate overhead of doing this, but we have so far not implemented automated creation of the message representations from the machine-readable specification.
 
@@
 \section{Further Work}\label{further-work}
+On the protocol side, we would like to develop further extensions, primarily to allow for VOIP, federation, and encrypted messaging.
+
+We would also like to develop further client implementations, particularly for C/C++ for use in the multi-protocol client Pidgin.
 

### 2022-01-21 els2022-lichat/paper.tex
Minor

@@
 \section{Extensions}\label{extensions}
+While the core protocol only provides text-only chat, we have specified and implemented a number of extensions that provide additional functionality commonly found in more modern systems such as Discord.
+
+We've specified over twenty extensions at the time of writing. Instead of outlining them all, we will instead name a few of particular import.
+
+A particular lament with IRC is the lack of easy file sharing.
+Users have come to expect being able to share images, videos, and other small files directly through the chat platform.
+IRC does have the DCC (Direct Client to Client) protocol extension, but it only allows exchanging files between two users directly, rather than broadcasting it to a channel.
+The ``shirakumo-data'' extension introduces a new message type that allows sending a base64-encoded payload, with an associated mime-type.
+The server can restrict permitted mime-types, and the general message length checking (with the null-byte skipping mechanism) allows dropping data payloads that are too large.
+The ``shirakumo-link'' extension further optimises this by letting the server save data-payloads and distribute them via an internet link instead, reducing the amount of data needing to be sent out to the channel drastically.
+
+The ``shirakumo-channel-info'' and ``shirakumo-user-info'' extensions in turn allow associating metadata with channels and user accounts.
+Like with the data extension, the server controls what metadata can be stored and how much of it.
+These extensions allow storing common things such as channel rules, a topic, icons, or other user contact data.
+
+With the ``shirakumo-emotes'' and ``shirakumo-reactions'' extensions, users can upload custom emoticons for their channels and use them in messages, or as ``reactions'' to other user's messages.
+These features have become especially popular on platforms such as Discord and Slack.
+
+Finally, the ``shirakumo-history'' extension allows the server (if requested) to store the chat logs of a channel, and then allows users to search through the history of a channel, and also in combination with the ``shirakumo-backfill'' extension request a replay of the most recent messages after joining a channel, to get some context for the conversation.
 
 \section{Conclusion}\label{conclusion}
+We have demonstrated a protocol design that offers a solid base for a chat system, which is both simple to implement, and easily extensible to provide advanced features that elevate it beyond simple text chat.
+
+With our extension in place, user experience is very close to that offered by Discord, Slack, or Matrix, with the primary difference being that, as the protocol currently stands, a user account is always local to a specific server, rather than being a part of the global network of servers.
+
+Our base protocol design has proven very robust, even in the face of these many extensions and the addition of multiple new clients across different languages.
 
@@
 \section{Acknowledgements}\label{acknowledgements}
+We would like to thank all the very sexy people that read this paper ;)
 

### 2022-01-23 els2022-lichat/paper.tex
Minor section

@@
 Large networks and a plethora of mature clients still exist and are in use today.
-However, The IRC protocol is also plagued by a number of issues that stem from its historical design and largely distributed evolution.
+However, the IRC protocol is also plagued by a number of issues that stem from its historical design and largely distributed evolution.
 In particular, the protocol has many different, incompatible extensions, and features many idiosyncrasies that make it difficult to write new clients or servers.
@@
 
-Over time many other chat systems such as AIM, ICQ, Skype, have come and largely gone, though all of these focus on a different model of operation than IRC.
+Over time many other chat systems such as AIM, ICQ, and Skype have come and largely gone, though all of these focus on a different model of operation than IRC.
 They focus on a friends-list based approach where people talk to each other directly, rather than relying on sometimes huge public chat rooms.
@@
 < (connect :id 0 :clock 5243 :from "tester" :version "1.0" :extensions ())
-> (join :id 0 :clock 5243 :from "tester" :channel "lichatters")
-< (join :id 0 :clock 5243 :from "tester" :channel "lichatters")
-< (message :id 0 :clock 5244 :from "shinmera" :channel "lichatters" :text "Welcome!")
-> (message :id 0 :clock 5246 :from "tester" :channel "lichatters" :text "Hiya")
-< (message :id 0 :clock 5246 :from "tester" :channel "lichatters" :text "Hiya")
+> (join :id 1 :clock 5243 :from "tester" :channel "lichatters")
+< (join :id 1 :clock 5243 :from "tester" :channel "lichatters")
+< (message :id 37 :clock 5244 :from "shinmera" :channel "lichatters" :text "Welcome!")
+> (message :id 2 :clock 5246 :from "tester" :channel "lichatters" :text "Hiya")
+< (message :id 2 :clock 5246 :from "tester" :channel "lichatters" :text "Hiya")
 \end{minted}
@@
 
-The permissions system in general is very simple: each channel holds a map where each update type is associated with with either a list of permitted users, or a list of denied users.
+The permissions system in general is very simple: each channel holds a map where each update type is associated with either a list of permitted users, or a list of denied users.
 As all communication between users must occur through channels, this provides a simple, but expressive model to handle moderation.
@@
 This is useful so that a server or client can restore the stream of messages should a parser failure occur, or should the update be rejected due to excessive length.
+We also specified that an optional, missing field is equivalent to specifying the field to be \code{nil}, and vice-versa.
+This simplifies initialising and dealing with optional values a lot.
 
@@
 Most of the issues arose when porting the message class hierarchy.
+
+Aside from the wire protocol and message type hierarchy, each client follows a pretty simple scheme as outlined in \autoref{lst:client}.
+Particularly simplifying is the fact that protocol-level pings are sent by the server, removing the need for complex timeout tracking in the client.
+
+\begin{listing}
+\begin{minted}[fontsize=\footnotesize]{common-lisp}
+(defun connect ()
+  (socket-connect)
+  (send 'connect :from username)
+  (let ((reply (make-update (read-list))))
+    (etypecase reply
+      (connect
+       (setf username (from reply))
+       (setf supported-extensions (extensions reply))))))
+
+(defun skip ()
+  (loop until (= 0 (read-byte)))
+  (continue))
+
+(defun reconnect ()
+  (socket-disconnect)
+  (connect)
+  (continue))
+
+(defun message-loop ()
+  (loop (with-simple-restart (continue)
+          (handle (make-update (read-list))))))
+
+(defun main ()
+  (connect)
+  (handler-bind ((syntax-error #'skip)
+                 (update-malformed #'continue)
+                 (timeout #'reconnect))
+    (message-loop)))
+\end{minted}
+\caption{A pseudo-code example of a client's main loop}
+\label{lst:client}
+\end{listing}
 

### 2022-01-24 els2022-lichat/paper.tex
Sigh

@@
 \section{Introduction}\label{introduction}
-The IRC (Internet Relay Chat) protocol has long stood the test of time as a reliable protocol for text chat communication.
+The IRC\cite{oikarinen1993internet} (Internet Relay Chat) protocol has long stood the test of time as a reliable protocol for text chat communication.
 Large networks and a plethora of mature clients still exist and are in use today.
@@
 \section{Related Work}\label{relatedwork}
-IRC
+Penpoppichanan et al.\cite{penpoppichanan2001security} identify a number of security and privacy related issues with the IRC protocol.
+While Lichat shares the basic concerns about privacy, being an unencrypted protocol, it does fix several of the identity and authentication related issues identified in the thesis.
 
-Matrix
+The XMPP core specification\cite{saint2004extensible} alone is over 200 pages long, and does not include the already non-trivial question of parsing XML, which is used for its wire format.
+Saint-Andre et al.\cite{saint2009xmpp} also point out the lack of a proper end-to-end encryption scheme for the protocol.
 
-XMPP
-
-Whatever?
+Jennings et al.\cite{jennings2006study} compare a variety of Instant Messenger systems such as AIM and MSN, which are all now largely defunct.
+The ppaper focuses primarily on the distribution of sessions and the scalability of the networks.
+Lichat is explicitly focused on small to mid-sized communities and is thus not concerned with the scalability to tens of thousands of simultaneous users.
 
@@
 On the protocol side, we would like to develop further extensions, primarily to allow for VOIP, federation, and encrypted messaging.
+Encryption and privacy are of especially big interest, though also pose quite challenging as end-to-end encrypted one-to-many communication is difficult to achieve efficiently\cite{nabeel2017many}\cite{whisper2014}.
 
@@
 \section{Acknowledgements}\label{acknowledgements}
-We would like to thank all the very sexy people that read this paper ;)
+We would like to thank Georgiy Tugai, Jan Moringe, and Selwyn Simsek for their feedback on this paper and contributions to Lichat's development.
 

### 2022-01-24 els2022-lichat/paper.tex
Minor

@@
 Jennings et al.\cite{jennings2006study} compare a variety of Instant Messenger systems such as AIM and MSN, which are all now largely defunct.
-The ppaper focuses primarily on the distribution of sessions and the scalability of the networks.
+The paper focuses primarily on the distribution of sessions and the scalability of the networks.
 Lichat is explicitly focused on small to mid-sized communities and is thus not concerned with the scalability to tens of thousands of simultaneous users.
@@
 \section{Acknowledgements}\label{acknowledgements}
-We would like to thank Georgiy Tugai, Jan Moringe, and Selwyn Simsek for their feedback on this paper and contributions to Lichat's development.
+We would like to thank Georgiy Tugai, Jan Moringe, Bart Botta, and Selwyn Simsek for their feedback on this paper and contributions to Lichat's development.
 

### 2022-01-25 els2022-lichat/paper.tex
Cite for matrix.

@@
 Lichat is explicitly focused on small to mid-sized communities and is thus not concerned with the scalability to tens of thousands of simultaneous users.
+
+The Matrix core specification\cite{matrix} is extremely extensive.
+It covers a lot of ground including encryption, federation, and other advanced features supported by the protocol.
+Especially notable is that the protocol is quite complex and features a lot of out-of-stream APIs, whereas Lichat handles everything within the single stream between the user and the server.
 
@@
 On the protocol side, we would like to develop further extensions, primarily to allow for VOIP, federation, and encrypted messaging.
-Encryption and privacy are of especially big interest, though also pose quite challenging as end-to-end encrypted one-to-many communication is difficult to achieve efficiently\cite{nabeel2017many}\cite{whisper2014}.
+Encryption and privacy are of especially big interest, though are quite challenging, as end-to-end encrypted one-to-many communication is difficult to achieve efficiently\cite{nabeel2017many}\cite{whisper2014}.
 

### 2022-01-25 els2022-lichat/paper.tex
More fixes

@@
 \section{Related Work}\label{relatedwork}
-Penpoppichanan et al.\cite{penpoppichanan2001security} identify a number of security and privacy related issues with the IRC protocol.
+Penpoppichanan\cite{penpoppichanan2001security} identifies a number of security and privacy related issues with the IRC protocol.
 While Lichat shares the basic concerns about privacy, being an unencrypted protocol, it does fix several of the identity and authentication related issues identified in the thesis.
@@
 The XMPP core specification\cite{saint2004extensible} alone is over 200 pages long, and does not include the already non-trivial question of parsing XML, which is used for its wire format.
-Saint-Andre et al.\cite{saint2009xmpp} also point out the lack of a proper end-to-end encryption scheme for the protocol.
+Saint-Andre\cite{saint2009xmpp} also points out the lack of a proper end-to-end encryption scheme for the protocol.
 
@@
 As a basis for the protocol we chose to use UTF-8 character streams for the transport.
-This ensures that protocol messages can be read and written by a human, which eases development and debugging.
+This choice ensures that protocol messages can be read and written by a human, which eases development and debugging.
 Specifying the character encoding also avoids ambiguity when interpreting the underlying bytes.
@@
 The client should also include a \code{:clock} field of the local Universal-Time at the time of sending.
-This ensures that, if distributed to other users, they can get an accurate ordering of events that reflects the sending user's point of view, without potential network delays changing the sequence of events.
+This field ensures that, if distributed to other users, they can get an accurate ordering of events that reflects the sending user's point of view, without potential network delays changing the sequence of events.
 It also allows the server and clients to estimate lag in the network and issue warnings on unstable connections, or terminate them early, ensuring an ultimately more stable communication.
@@
 On the client to client communication side, we decided to enforce all communication between clients to occur over channels.
-This simplifies the design on both sides, as there don't need to be any special rules to deal with direct communication.
+This decision simplifies the design on both sides, as there don't need to be any special rules to deal with direct communication.
 With all of this groundwork laid, we could implement a basic chat server that allowed for communication over multiple channels, as shown in \autoref{lst:basic-exchange}.
@@
 
-After this most of the protocol design revolved around the implementation of a permissions system to allow channel operators better control over what users are allowed to do, and the ratification of distinct error types for potential failures.
+After these changes most of the protocol design revolved around the implementation of a permissions system to allow channel operators better control over what users are allowed to do, and the ratification of distinct error types for potential failures.
 
 The permissions system in general is very simple: each channel holds a map where each update type is associated with either a list of permitted users, or a list of denied users.
-As all communication between users must occur through channels, this provides a simple, but expressive model to handle moderation.
+As all communication between users must occur through channels, this system provides a simple, but expressive model to handle moderation.
 For instance, banning a user simply requires removing their ability to send a \code{join} message to the channel.
@@
 Further, we required the use of a null byte as an end of message marker.
-This is useful so that a server or client can restore the stream of messages should a parser failure occur, or should the update be rejected due to excessive length.
+The null byte is useful so that a server or client can restore the stream of messages should a parser failure occur, or should the update be rejected due to excessive length.
 We also specified that an optional, missing field is equivalent to specifying the field to be \code{nil}, and vice-versa.
-This simplifies initialising and dealing with optional values a lot.
+This aliasing simplifies initialising and dealing with optional values a lot.
 
 To ease the initiation and maintenance of the connection, the \code{:from} and \code{:clock} fields were made optional, instead requiring the server to automatically fill them in should they be missing.
-This makes it even more trivial to write a simple, first client.
+This behaviour makes it even more trivial to write a simple, first client.
 
@@
 Finally, we ratified the definition of all message types in a machine-readable format (\autoref{lst:spec}) that can be parsed using the same parser of the wire format.
-This allows protocol implementations to easily parse up-to-date representations of the entire message type hierarchy.
+These machine-readable specs allow protocol implementations to easily parse up-to-date representations of the entire message type hierarchy.
 
@@
 JavaScript does not support multiple inheritance out of the box, so we implemented our own scheme following the serialised inheritance model of CLOS.
-This does however still not integrate neatly with the native JavaScript type system, meaning we also require the use of a specialised type test for dispatch.
+This scheme does however still not integrate neatly with the native JavaScript type system, meaning we also require the use of a specialised type test for dispatch.
 
@@
 Each message type is represented through a module, with a function to handle the update on the server-side.
-Thanks to Elixir's macro system we were able to avoid most of the boilerplate overhead of doing this, but we have so far not implemented automated creation of the message representations from the machine-readable specification.
+Thanks to Elixir's macro system we were able to avoid most of the boilerplate overhead of writing each message type, but we have so far not implemented automated creation of the message representations from the machine-readable specification.
 
@@
 The server can restrict permitted mime-types, and the general message length checking (with the null-byte skipping mechanism) allows dropping data payloads that are too large.
-The ``shirakumo-link'' extension further optimises this by letting the server save data-payloads and distribute them via an internet link instead, reducing the amount of data needing to be sent out to the channel drastically.
+The ``shirakumo-link'' extension further optimises sending media by letting the server save data-payloads and distribute them via an internet link instead, reducing the amount of data needing to be sent out to the channel drastically.
 
@@
 
+\section{Source Code}\label{source}
+The lichat specification and implementations can be found freely on the web:
+\begin{itemize}
+\item Lichat \url{https://shirakumo.org/projects/lichat}
+\item JavaScript \url{https://shirakumo.org/projects/lichat-js}
+\item Elixir \url{https://shirakumo.org/projects/ex-lichat}
+\item Java \url{https://shirakumo.org/projects/jlichat}
+\item Python \url{https://shirakumo.org/projects/py-lichat}
+\end{itemize}
+
 \bibliography{paper}

### 2022-01-25 els2022-lichat/paper.tex
Minor

@@
 
+Another topic of interest is the implementation of a full bridge for the Matrix protocol, which would allow Lichat servers to interact with the Matrix federated ecosystem.
+
 We would also like to develop further client implementations, particularly for C/C++ for use in the multi-protocol client Pidgin.
+
+Finally we would like to extend the existing clients to support all of the drafted extensions, and further improve the Elixir server to support statistics gathering and allow for better spam and abuse control.
 

### 2023-01-21 els2023-kandria/paper.tex
init

\documentclass[format=sigconf]{acmart}
\usepackage[utf8]{inputenc}
\usepackage{enumitem}
\usepackage{float}
\usepackage[labelfont=bf,textfont=md]{caption}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{minted}
\usepackage{hyperref}
\usepackage[parfill]{parskip}
\usepackage[all]{hypcap}
\usemintedstyle[common-lisp]{default}
\newmintinline[code]{text}{}
\bibliographystyle{plainnat}

\hypersetup{
  colorlinks,
  linkcolor={red!50!black},
  citecolor={blue!50!black},
  urlcolor={blue!80!black}
}

\newlist{step}{enumerate}{10}
\setlist[step]{label*=\arabic*.,leftmargin=2em}

\acmConference[ELS'23]{the 15th European Lisp Symposium}{March 21--22 2023}{%
  }
\acmISBN{}
\acmDOI{}
\setcopyright{rightsretained}
\copyrightyear{2023}

\begin{document}

\title{Experience Report: Kandria - A Game in Common Lisp}

\author{Nicolas Hafner}
\affiliation{%
  \institution{Shirakumo.org}
  \city{Zürich}
  \country{Switzerland}
}
\email{shinmera@tymoon.eu}

\begin{abstract}

\end{abstract}

\begin{CCSXML}

\end{CCSXML}

\keywords{Common Lisp, Games, Video Games, Computer Graphics, Experience Report}

\maketitle

\def\abovecaptionskip{1pt}
\def\listingautorefname{Listing}
\def\figureautorefname{Figure}

\section{Introduction}\label{introduction}

\section{Related Work}\label{relatedwork}

\section{Missing Components}\label{yaks}

\section{Garbage Collection}\label{gc}

\section{Performance}\label{performance}

\section{Deployment}\label{deployment}

\section{Conclusion}\label{conclusion}

\section{Further Work}\label{further-work}
Currently a sizable amount of the work in Kandria has not been backported into Trial for more general purpose use. We would like to extract a few of the systems and generalise them to make them available for other users.

We are also working on implementing several new subsystems in Trial to allow creating 3D games, as well. A skeletal animation system has been completed, and we're currently working on a physics subsystem. Also needed will be several spatial query structures to speed up collision testing.

Finally we are also exploring the possibility of porting the engine to work on closed platforms such as the Nintendo Switch. This presents several challenges that we unfortunately cannot elaborate on here due to non-disclosure agreements.

At some point we would like to get rid of the dependency on GLFW to reduce further dependence on C libraries and the issues arising from them. However, GLFW has proven extremely stable in this regard, and as such this is of very low priority for us.

\section{Acknowledgements}\label{acknowledgements}
We would like to thank \textit{you} for being beautiful and nice :)
\bibliography{paper}
\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-command-extra-options: "-shell-escape"
%%% TeX-master: t
%%% TeX-engine: luatex
%%% End:

### 2023-02-02 els2023-kandria/paper.tex
Actually write some stuff.

@@
 \section{Introduction}\label{introduction}
+Kandria is a video game in the ``action RPG'' genre that was released worldwide on the \href{https://kandria.com/steam}{Valve Steam} platform for Windows and Linux in January of 2023, receiving very positive reviews.
+The game was developed using the Trial engine and thus relies almost entirely upon code and libraries written in Common Lisp -- to our knowledge the first commercial game like this to be released.
+
+\begin{figure}[h]
+  \begin{centering}
+    \includegraphics[width=0.4\textwidth]{screenshot 8.png}
+  \end{centering}
+  \caption{A screenshot of Kandria}
+\end{figure}
+
+Games lie at an intersection of many different computer science disciplines such as audio, graphics, interfaces, soft real-time, artificial intelligence, and more.
+As such games provide a unique challenge in combining all of these disciplines together into one product, and ultimately shipping this product to paying customers.
+
+In this paper we will outline the challenges we faced in realising Kandria as related to Common Lisp, discuss advantages of our approach, and take a look at the work still ahead of us in expanding the capabilities of our engine to better address the requirements of even more complex games we would like to develop in the future.
+
+In particular we will note our own experiences with a few common Boogeymen, such as the overhead of CLOS dispatch, the pause times of GC, the maturity of the library ecosystem, and the stability and size of deployed binaries.
+
+All of our work, including Trial, and even Kandria, is open source and available on our \href{https://github.com/shirakumo}{GitHub}, in the hopes that it will inspire others to create new projects based upon them.
 
 \section{Related Work}\label{relatedwork}
+In \cite{hafner2021} we discussed many similar points that this paper touches upon in a format more suitable for people unfamiliar with the intricacies of lisp and especially Common Lisp.
 
-\section{Missing Components}\label{yaks}
+\section{Library Ecosystem}\label{yaks}
+In this section we will outline our general experience working in the Common Lisp ecosystem, and particularly the contributions we've made to it in order to implement Kandria.
+
+Out of the 110 libraries Kandria depends upon, 54 were written by us.
+This does mean that we've spent a rather significant amount of time ``yak shaving'' and creating libraries to fulfil a variety of needs that were, prior to their creation, either completely unfulfilled, or unsatisfactorily so.
+We will go into detail on some of the more important ones in the following subsections.
+
+While this is undeniably a lot of work, it has been done now and is available to others.
+Furthermore, of the other half of libraries that we were not authors of, the vast majority have been extremely stable.
+We only needed to supply extremely minimal patches to a select few libraries, most of which were reviewed and accepted relatively swiftly.
+
+Among this we also count the SBCL implementation itself.
+While we strive to write implementation independent code in our separate libraries, for Kandria itself we decided to only focus on SBCL in order to reduce the development overhead.
+SBCL offers great native code performance and is available for all platform configurations that we require.
+And, perhaps most importantly, it is very actively maintained.
+Most of the issues we've encountered in releases were usually fixed within a few weeks, if not days.
+
+We are also actively investigating the possibility of porting SBCL and Kandria to the proprietary Nintendo Switch platform, though due to Non-Disclosure Agreements we are unfortunately not at liberty to speak of the specifics involved in that at this time.
+
+We'll now touch on a few specific areas that we developed libraries for. This is by no means exhaustive, but we consider these to be the most relevant to the general community.
+
+\subsection{Math Libraries}\label{math}
+While there is no shortage of math libraries available to Common Lisp, especially linear algebra implementations, most of those libraries focus on large scale scientific computing, often by integrating with foreign libraries such as BLAS and LAPACK.
+For basic computer graphics and especially games, this is overkill.
+Most of the linear algebra stays within the confines of 2x2, 3x3, 4x4 matrices, and 2, 3, and 4 element vectors.
+
+It is more important to provide a very convenient interface that allows us to perform computations on these elements with adequate speed.
+Back when Trial started out, no linear algebra libraries with such a limited focus existed, and as such the 3d-Vectors and 3d-Matrices libraries were born.
+We have since extended this set of libraries to include 3d-Quaternions for common operations with rotations, and 3d-Transforms, for the convenient encapsulation of a ``transform gizmo'' that can represent rotation, translation, and scaling without Gimbal Lock.
+
+All of these libraries rely very heavily on macros to reduce code duplication and automatically generate code for loop unrolling and other common tactics in linear algebra code.
+The current versions of these libraries all work by emitting an \code{etypecase} for every operation, which then handles dispatch based on the provided argument types.
+The operation functions are inlined, such that the compiler can eliminate the dispatch altogether if the argument types are locally known.
+This allows us to provide a generic interface to the user that's quite convenient, while still staying competitive in performance critical sections.
+
+Unfortunately this approach, while portable, is also riddled with issues: since every operation is inlined, this leads to explosive code growth for the compiler before it can reduce the code back down again by eliminating superfluous dispatch \code{etypecase}s.
+Type inference is also much more complicated, and stack allocation is usually only possible via careful manual rewriting of the operations involved.
+The libraries are also limited to a single float type, meaning you can by default only create vectors with single-floats as elements.
+
+In this case the lack of static typing facilities and lack of portable compiler hooks for integrating with type inference really hurts the compile speed, implementation clarity, and ultimate performance of the resulting code.
+
+We have started work on a full rewrite of all libraries that take a fundamentally different approach: instead of emitting etypecases and relying on inlining, we create a sort of ``template mechanism'' by which we can generate all possible permutations of a singular function for all involved argument types.
+This gives us very tiny, but perfectly optimised base functions for all required operations.
+We then create a dispatcher function on top which falls back to emitting an \code{etypecase} on other implementations, but will hook into SBCL's \code{deftransform} and similar facilities to better handle expansion and type propagation.
+Finally we create variadic functions on top of the dispatchers which transform any possible variadic call into calls to dispatched 2-argument operation functions, while retaining proper type propagation.
+
+Ultimately this results in much more easily understandable and performing code.
+However, it still comes at a cost: because we cannot know ahead of time which type combinations and operations the user will actually need, we must generate all possible combinations ahead of time.
+This potentially results in thousands of functions being generated that are never used.
+We would like a compiler facility that allows us to hook into the expansion process of a function call to generate the required permutations on-demand.
+
+While such a facility is conceivable, and making it work for unknown runtime types by delayed compilation is, too, we currently have no plans to implement such an advanced strategy.
+More importantly, we hope that at some point in the future implementors can come to some kind of consensus that would allow a portability library to expose a similar mechanism to SBCL's \code{deftransform} for faster, more convenient, type-inference-informed call expansion.
+
+\subsection{User Interfaces}\label{ui}
+When Trial initially began its development in 2016 it was directly integrated with the Qt4 UI toolkit (via CommonQt / Qtools).
+Since Qt4 is a rather large dependency that not only increases the deployed package size, but also invites a lot of C and C++ interop that can cause hard to debug issues, and has not been maintained in many years, it was quickly abandoned, however.
+These days Trial does not depend on a UI toolkit directly, or even a specific backend for its OpenGL use.
+
+Aside from CommonQt, the options for a user interface in Common Lisp were, and remain, limited: GTK runs the same issues as Qt, though with even worse deployment and support aspects, LTK is far too limited and cannot integrate with OpenGL at all, McCLIM similarly does not possess an OpenGL backend or method of integration and is still very limited in its theming capabilities, and the newly established CLOG requires a browser to be shipped, which is far too heavy-weight.
+
+As a result we decided to implement our own toolkit, called Alloy.
+Alloy is separated into different protocols, with the core being completely independent of any rendering or input method, instead only handling the layouting decisions, the input handling via a generic event protocol, and a system to dynamically react to changes in the represented data.
+
+How visual elements can be rendered is then offloaded to several other protocols: the ``Simple Rendering Protocol'' which provides a basic text and shape rendering API, the ``Presentations Protocol'' which allows users to describe how to compose the look of a visual element via the basic shapes from the Simple protocol, and the ``Animations Protocol'' which allows users to describe how shapes change over time as properties of a visual element are changed.
+
+Finally, the Simple protocol needs to be implemented by a backend, such as the ``OpenGL Renderer'' to actually provide a way to draw the shapes in some way, the Core protocol needs to receive input events from the surrounding context to actually react to user input, and some method of rendering and layouting text needs to be provided.
+Text in particular is separated out as it is a very complex topic in its own right.
+
+This separation of concerns via protocols implemented in CLOS allows us to make Alloy far more amenable to being ported to different backends.
+In particular, it allows us to use Alloy in contexts that are otherwise rather unusual for UI toolkits, such as within a game where the actual operating system interaction and rendering logic cannot be directly controlled by the toolkit itself, but is instead handled by the game engine.
+
+Usual desktop UI toolkits are rather cumbersome to style effectively, as most desktop applications are expected to present themselves in a ``native look and feel'', while games are expected to have rather elaborately customised and animated interfaces.
+Alloy's presentations protocol allows us to style the UI quite extensively, with relative ease (\autoref{lst:presentation}).
+
+\begin{listing}
+\begin{minted}[fontsize=\small]{common-lisp}
+;; Define the shapes that compose the element
+(presentations:define-realization (ui alloy:switch)
+  ((:border simple:rectangle)
+   (alloy:margins)
+   :pattern colors:white
+   :line-width (alloy:un 2))
+  ((:background simple:rectangle)
+   (alloy:margins))
+  ((:switch simple:rectangle)
+   (alloy:extent 0 0 (alloy:pw 0.3) (alloy:ph))))
+
+;; Define how shape properties change
+(presentations:define-update (ui alloy:switch)
+  (:border
+   :hidden-p NIL
+   :pattern (if alloy:focus
+                (colored:color 0.9 0.9 0.9)
+                colors:gray))
+  (:background
+   :pattern (if (alloy:active-p alloy:renderable)
+                colors:dark-accent
+                (colored:color 0 0 0 0.5)))
+  (:switch
+   :pattern (if (alloy:active-p alloy:renderable)
+                colors:accent
+                colors:black)))
+
+;; Define how to animate the properties
+(presentations:define-animated-shapes alloy:switch
+  (:background (simple:pattern :duration 0.2))
+  (:switch (simple:pattern :duration 0.3)
+           (simple:bounds :duration 0.5)))
+\end{minted}
+\caption{An example of the presentations mechanism in Alloy to define the look and visual behaviour of a switch.}
+\label{lst:presentation}
+\end{listing}
+
+This protocol separation would not be possible to implement without the high degree of flexibility CLOS offers us in combining behaviours, and even without advanced techniques such as shadow mixins and metaclasses.
+
+\subsection{Input Handling}\label{input}
+While keyboard, text, and mouse input is provided by the windowing system on each platform, and thus by whatever OpenGL context backend Trial may use, the same does not go for other input devices, most notably gamepads.
+
+Gamepads or game controllers are special purpose devices composed out of a few analog directional sticks, buttons, and analog pressure triggers.
+We implemented a library that directly interfaces with the operating system on each of the platforms to discover these devices and read out their state.
+Thanks to CFFI it is usually unnecessary to actually rely on a C library to interface with operating system APIs, and we can instead write the code directly in Lisp, retaining superior interactive debugging capabilities.
+
+Another challenge, besides the direct system interaction, is that gamepads do not follow a universally defined scheme, and many devices will have partially broken drivers, or simply report their inputs incorrectly.
+Our library must adapt to this and present as uniform of a scheme to the engine as possible.
+
+To deal with this we have started building a database that maps the device's input IDs onto a standardised set of names and ranges, and have provided a simple setup wizard to configure a new device for which the auto-detection fails.
+
+\subsection{Audio Processing}\label{audio}
+
+\subsection{Operating System Interaction}\label{os}
+
+\subsection{Service Integration}\label{steam}
 

### 2023-02-03 els2023-kandria/paper.tex
Initial draft

@@
 \begin{abstract}
-
+  In this paper we outline the experience we've gathered while developing Kandria using Common Lisp.
+Kandria is a video game in the ``action RPG'' genre and has been developed for Windows and Linux with the SBCL implementation.
+Being a video game, the project touches a unique combination of disciplines within computer science, and as such provides an in-depth and comprehensive view of the development process of a large-scale project in Lisp, and the general language ecosystem.
 \end{abstract}
@@
 
-All of our work, including Trial, and even Kandria, is open source and available on our \href{https://github.com/shirakumo}{GitHub}, in the hopes that it will inspire others to create new projects based upon them.
+Since this is an experience report, we cannot present any specific figures on performance characteristics, statistics on used projects, or some sort of unified thesis.
+Instead we hope that this insight will be valuable for readers to gain a better understanding of the complexities involved, the benefits we have identified with Lisp over other ecosystems, and, most importantly, the areas in which work still needs to be done.
+
+All of the work we've done, including Trial, and even Kandria, is open source and available on our \href{https://github.com/shirakumo}{GitHub}, in the hopes that it will inspire others to create new projects based upon them.
 
@@
 We are also actively investigating the possibility of porting SBCL and Kandria to the proprietary Nintendo Switch platform, though due to Non-Disclosure Agreements we are unfortunately not at liberty to speak of the specifics involved in that at this time.
+
+Overall while we certainly had to create a lot of libraries to fulfil our needs, and those libraries presented a significant amount of effort to implement, we remain convinced that we were only able to implement these systems in a respectable amount of time due to the convenience factors that Lisp offers us.
 
@@
 
-\subsection{Input Handling}\label{input}
-While keyboard, text, and mouse input is provided by the windowing system on each platform, and thus by whatever OpenGL context backend Trial may use, the same does not go for other input devices, most notably gamepads.
+\subsection{Audio Processing}\label{audio}
+Of the entire system, audio processing is where the toughest constraints apply, as audio is very sensitive to latency.
+We cannot process audio in large buffers to smooth over processing hiccups, as then sound effects would be desynchronised from their visual counterparts.
+Audio processing can also often be very computation intensive, and benefits greatly from vectorisation.
 
-Gamepads or game controllers are special purpose devices composed out of a few analog directional sticks, buttons, and analog pressure triggers.
-We implemented a library that directly interfaces with the operating system on each of the platforms to discover these devices and read out their state.
-Thanks to CFFI it is usually unnecessary to actually rely on a C library to interface with operating system APIs, and we can instead write the code directly in Lisp, retaining superior interactive debugging capabilities.
+For these reasons we have decided to write the main bulk of our processing as a C library, rather than Common Lisp, as at the time it was the fastest way for us to get the kinds of performance constraints met and the kinds of capabilities we needed.
+Now with the advent of the sb-simd contrib, writing a competitive alternative (albeit constrained to SBCL) may be feasible.
 
-Another challenge, besides the direct system interaction, is that gamepads do not follow a universally defined scheme, and many devices will have partially broken drivers, or simply report their inputs incorrectly.
-Our library must adapt to this and present as uniform of a scheme to the engine as possible.
+In any case, our library, libmixed, follows a very strict C style and is extensible and interoperable from Lisp.
+You can write ``segments'' that process audio from Lisp as well, and integrate with the rest of the processing system neatly.
 
-To deal with this we have started building a database that maps the device's input IDs onto a standardised set of names and ranges, and have provided a simple setup wizard to configure a new device for which the auto-detection fails.
+This allows us to write the more hairy parts such as audio format decoding, and audio playback in Lisp, instead, without sacrificing the performance gains in the bulk of the processing.
+Libmixed then takes care of unpacking and decoding audio streams, applying a variety of effects, mixing multiple audio streams together, and even resampling everything to fit into the expected, and often differing, sample rates at the input and output points.
 
-\subsection{Audio Processing}\label{audio}
+Libmixed also includes a full introspection API, allowing us to not only put the audio processing pipeline together from Lisp, but also to interactively inspect the state of the processing segments and modify them at runtime.
+We can also use the extensibility to prototype new effects from Lisp before lowering them down to C should the performance requirements be strict enough.
+
+On the Lisp side we have also crafted a more high-level system called Harmony, which makes it trivial to put together these processing pipelines, and interactively play music and sound effects back.
+In Kandria we use these pipelines for a variety of effects such as cross fading multiple music layers (called ``horizontal mixing''), ducking the audio with a low pass filter when the player is under water, and slowing down audio playback when the player enters a slow motion segment.
 
 \subsection{Operating System Interaction}\label{os}
+In order to output sound and graphics, process input, and receive user information we need to interact with the surrounding operating system.
+On all three targets we care about, this is done via calls to a C API of some sort.
+Thanks to CFFI it is usually unnecessary to actually rely on an external C library to do so, and we can instead code the interaction directly in Lisp, retaining the interactive implementation and debugging.
+
+Despite this though, the interaction still involves C and the usual memory safety perils, and depending on the interface the documentation often leaves things to be desired, especially on MacOS.
+So while our implementation definitely benefits from a much faster retry cycle, it is still a very arduous process to write these OS interop libraries.
+
+Particularly we had to write libraries to do the following:
+\begin{enumerate}
+\item Input handling for game controllers.
+This is rather involved, especially on Windows, where multiple APIs need to be supported simultaneously.
+Game controller devices are also prone to broken drivers or bad input mapping, which our library needs to deal with.
+\item Audio output. We've implemented several different output APIs on both Linux and Windows, as both systems can differ on their supported output interfaces depending on version and setup.
+\item Native dialog boxes. In order to show emergency error boxes or prompt the user for files.
+\item Font discovery. To search the available fonts for a matching set we need to query operating system APIs.
+\item Language querying. To ensure the game launches with the user's language (if supported of course), we again need to query the environment for the preferred localisation.
+\end{enumerate}
+
+For graphics output and general window interaction we currently rely on the GLFW C library, as it has proven extremely stable and portable.
+It is conceivable that we will replace this at some point to reduce C dependencies, but given that we've had zero issues arising from its use, this is rather low priority for us.
 
 \subsection{Service Integration}\label{steam}
+In order to publish a game on the Steam platform, you must integrate with their SteamWorks SDK.
+The actually required amount of integration is very minimal, and merely involves loading their shared library and calling a single function.
+However, the SteamWorks SDK offers a lot of other functionality that games can make use of, like social networking features, user generated content sharing, multiplayer systems, and more.
 
-\section{Garbage Collection}\label{gc}
+Unlike the prior OS interfaces, the default expected interaction mode with the SteamWorks SDK is via C++, which is quite difficult to talk with directly via CFFI.
+Fortunately, they also offer a raw C API, but this API is not fully documented and fraught with strange gotchas.
+The SDK does ship a ``machine readable'' description of the endpoints, structures, and types in the form of a JSON file, but as we have found this file is both incomplete and partially incorrect.
 
-\section{Performance}\label{performance}
+In our implementation we analyse the JSON file along with a manually supplied file with the lacking data, and a couple of the shipped static headers to automatically generate the CFFI wrapper data structures, types, constants, and functions.
+Since the analysis of these files is rather involved however, we have chosen not to use macroexpansion, but rather provide a separate system which emits Lisp code to a new file.
+
+In addition to this generated interface, we also had to supply a minimal ``shim'' C library that handles a few select API calls that rely on structure-by-value passing, a feature which is not natively supported on most Lisp implementations at this point.
+Another workaround would be to use libffi, but libffi comes with its own issue, and would be another C library to depend on anyhow, so we opted for the much simpler and easier to understand alternative of writing our own minimal shim library.
+
+In Kandria we make use of the extended services to present the user with an on-screen keyboard when they are using a controller, to read out the username for default save file naming, and to integrate with the Steam platform's ``achievements'' system, which players have come to expect.
+
+\section{Common Lisp Object System}\label{clos}
+In both Trial and Kandria we make rather extensive use of CLOS, both its basic features of classes, multiple dispatch, and deep hierarchies via mixins, and the advanced features of metaclasses.
+For instance, our entire event handling system is simply a singular function \code{(handle event receiver)} which we define methods on to receive events.
+A central ``event loop'' object then just calls the handle function for every event it receives on every receiver it has in its internal list.
+
+As described in our prior paper \cite{hafner2018} we use metaclasses to attach ``shader fragments'' (code that is executed on the GPU during rendering) to classes and inherit the behaviour of these fragments together, allowing composition of rendering behaviour through inheritance as well.
+
+Overall the use of classes as an organisational structure lends itself extremely well to games, especially in the presence of multiple inheritance and mixins.
+We also often use this to introduce ``marker classes'' that contain no behaviour or data of their own, but act as type information that other parts of the system use, such as whether a class can be instantiated by the editor, resized, should be treated as solid during collision, should not be deleted when loading a prior save state, etc.
+
+We make use of structures over standard objects only in very select cases where the impact of the increased verbosity can be contained, and the performance requirements actually indicated that we needed to optimise.
+Specifically this is for the hit information during collision resolution, and for the node graph that powers the pathing AI.
+In those cases the initial class approach produced too much overhead, so we switched to a few specific structures.
+
+The dispatch overhead of CLOS has never been a real problem for us.
+Even with thousands of objects that receive events through our \code{handle} function, hundreds of methods attached to it, and tens of events every frame, not to mention all the other parts in the game that use CLOS, we still manage to easily hit a steady 120 frames per second on a ten year old machine.
+
+\section{Performance \& Garbage Collection}\label{performance}
+Overall we have needed to do surprisingly little actual performance analysis and optimisation work to make Kandria run well.
+This is definitely in large part thanks to SBCL's quite good native code compiler and type inference systems, and the prior work we've done to design critical libraries to not be completely obscene in terms of their performance characteristics.
+
+We have done larger scale rewrites of several subsystems in Trial, which have provided us with general performance increases, though these rewrites were primarily done to increase the code clarity and usability, with the performance being a nice bonus.
+
+As usual, far more important than constant factor improvements like those has been reducing the asymptotic behaviour.
+Introducing a Bounding Volume Hierarchy to speed up spatial queries and especially collision detection from a primitive quadratic search down to a logarithmic one has definitely provided the most significant performance improvement.
+
+While asymptotic behaviour improvements will be applicable for projects in any language, with SBCL we've had the definite advantage of being able to rely on the statistical profiler to determine hot spots in the code.
+This is especially advantageous in Lisp as we can turn the profiler on and off at any time, to capture exact segments that we are interested in, rather than having to capture the full run of a program and then massaging the data manually.
+
+Newly developed tools to better visualise the data gathered by the statistical profiler, such as the ``flamegraph'' visualiser developed by Jan Moringe have also been invaluable in gaining a better understanding of the performance characteristics of the code.
+
+As outlined in \autoref{math} above, there are still areas in which Lisp struggles to be truly competitive, largely in part due to its very dynamic typing, an aspect that is otherwise very advantageous for development.
+We've also encountered issues in eliminating superfluous consing, as many convenient styles of writing code will prevent stack allocation and other garbage elimination.
+
+We have managed to keep our garbage production down largely with two tricks:
+
+\begin{itemize}
+\item Pooling and preallocation.
+By storing objects in a pool and manually allocating and freeing them we can avoid allocating them at runtime, leaving less work for the GC to deal with.
+Since the objects are long lived, they will also be quickly promoted to later generations, lessening the work to scan other generations.
+The obvious downside is that we lose automatic freeing and may introduce double-use cases, but for many cases the lifetimes of the objects can be predetermined and managed relatively painlessly.
+\item Using \code{load-time-value} in lieu of stack allocation.
+With this trick we can allocate a ``local object'' that will be modified at runtime rather than allocated fresh.
+The downside of this approach is that whichever function provides the load-time-value object cannot be called concurrently, and the programmer must be vigilant in tracking the lifetime of the load-time-value object should it escape the dynamic extent of the function.
+\end{itemize}
+
+We are also very interested in a recent proposal to add memory arenas to SBCL, which would allow us to capture all of the objects allocated within a dynamic extent and free them immediately on exit, further lessening the burden of the global GC in cases where we know the exact lifetimes.
+
+Aside from memory arenas, we're also very interested in recent work by Hayley Patton to parallelise SBCL's garbage collector, though it is currently unknown when this feature will become mature and be merged into mainline SBCL.
+
+In general we have not felt that we've had to do a lot of work to keep garbage production down.
+On a usual system the GC seems to trigger every ten second or so, with no noticeable stutters caused by the pause time.
+We have been made aware of some rare systems on which the GC pause \textit{does} cause visible stutters, but have not been able to identify why that should happen or what to do about it.
+Completely eliminating all runtime garbage production does not seem like an effective way to spend our time, however.
 
 \section{Deployment}\label{deployment}
+Finally we would like to talk about our experience actually delivering Kandria to users and the process we have developed for doing so.
+
+The first step in the deployment pipeline is the build of the actual game binary, for which we use ASDF and the Deploy library.
+With these all game code is compiled fresh, dumped into an executable suitable for the local platform, and bundled together with the depended-upon shared libraries, producing an executable ready for deployment to target machines.
+
+However, since the build does not happen from scratch via a platform independent compiler, we have to ensure that we use an SBCL version that has been compiled on our minimal target platform version.
+For instance, we build the SBCL we use for deployment in a virtual machine on an old Linux version to ensure that internal symbol version dependencies in glibc can be satisfied on target machines.
+We have to take similar care for all required shared libraries that are shipped.
+Once this version has been compiled however, we can simply invoke it on any Linux version we like to use, and don't have to build the entire system inside a VM.
+
+We can also directly build Windows version of the game on Linux by using WINE.
+This has worked flawlessly for us, and we imagine that it would also be possible to do the inverse on Windows by using the Windows Subsystem for Linux to build Linux versions of the games.
+We've also investigated the possibility of using Darling to build MacOS versions.
+Unfortunately Darling is not able to run SBCL as of the writing of this paper.
+
+For all generated binaries we also make use of SBCL's core compression to reduce the binary size to about 35MB per platform.
+The newly introduced support for zstd also brings improved compression rates and startup speeds over the older zlib.
+This binary size is more than acceptable to us and is completely dwarfed by the resource files for sound and graphics.
+
+We also make use of hooks in the Deploy library to extract these resource files from their source directory and bundle them alongside the binary when building.
+Paths to resource files are resolved at runtime in the engine, and are thus trivial to redirect to binary-relative paths when deployed.
+
+Finally we have created a release system that invokes the necessary SBCL subprocesses to kick off a build, prunes out unnecessary files from the generated release, then bundles it all into a ZIP for easy delivery to testers.
+The system can also upload the release directly to a variety of distribution platforms including Steam, Itch.io, Keygen, and generic HTTP or FTP servers.
+Ultimately this allows us to compile, bundle, package, and upload new builds with a single command, drastically reducing the time and complexity involved in supplying testers and users with bugfixes.
+
+Another aspect that is advantageous to Lisp's dynamic nature is that when an unhandled condition occurs on a target system we can, in almost all cases, still use the system well enough to gather telemetry and submit an automated crash report.
+This has been invaluable for us to detect rare bugs, especially ones related to uncommon system configurations that we would otherwise never have been able to catch.
 
 \section{Conclusion}\label{conclusion}
+Overall the biggest hurdle we have had in our development of Kandria has nothing to do with the language itself, but rather with the general lack of manpower behind the community.
+We have had to spend large amounts of time implementing, testing, and documenting auxiliary systems that have often close to nothing to do with the core process of implementing a game.
+This is also why we have been extremely light on topics that concern the actual game implementation process in this paper.
+
+On the other hand, it is very clear to us that interactive development is an immense boon to the development of a game.
+Being able to redefine game behaviour at runtime to quickly iterate on the game content and feel is invaluable.
+So much so that almost every engine in use will have some form of scripting language available for game logic specifically, such that designers can iterate quickly.
+However, having the full stack of code available, debuggable, and performing just as well as any other part of the system is undeniably far more convenient.
+
+We also fully recognise that while Kandria is a full game project, it is by its nature rather limited in the required processing capabilities.
+A lot more work is needed to support more complex games, which we intend on focusing on in the near future.
+However, we do not at present see any deal breakers that would make it unfeasible to create such games using Common Lisp and the base ecosystem we have helped establish so far.
+
+If anything the recent advances in SBCL's capabilities show us a very promising future and we are excited to make use of them to further improve the situation, and with much of the ``grunt work'' now done we should be able to focus our efforts onto problems more directly associated with game development instead.
 
 \section{Further Work}\label{further-work}
-Currently a sizable amount of the work in Kandria has not been backported into Trial for more general purpose use.
+Currently a sizable amount of work in Kandria has not been backported into Trial for more general purpose use.
 We would like to extract a few of the systems and generalise them to make them available for other users.
@@
 A skeletal animation system has been completed, and we're currently working on a physics subsystem.
-Also needed will be several spatial query structures to speed up collision testing.
+Also needed will be several spatial query structures to speed up collision testing, along with a more unified rendering subsystem to support Physics Based Rendering pipelines.
 
@@
 
-At some point we would like to get rid of the dependency on GLFW to reduce further dependence on C libraries and the issues arising from them.
-However, GLFW has proven extremely stable in this regard, and as such this is of very low priority for us.
-
 \section{Acknowledgements}\label{acknowledgements}
 We would like to thank \textit{you} for being beautiful and nice :)
+
+Kandria was funded in part by the Pro Helvetia Interactive Media Grant and the KPT Poland Prize Digital Dragons Accelerator.
 \bibliography{paper}

### 2023-02-11 els2023-kandria/paper.tex
Revision

@@
 
-\author{Nicolas Hafner}
+\author{Nicolas ``Shinmera'' Hafner}
 \affiliation{%
@@
 }
-\email{shinmera@tymoon.eu}
 
@@
 \begin{CCSXML}
-
+  <ccs2012>
+  <concept>
+  <concept_id>10010147.10010371</concept_id>
+  <concept_desc>Computing methodologies~Computer graphics</concept_desc>
+  <concept_significance>300</concept_significance>
+  </concept>
+  <concept>
+  <concept_id>10010520.10010570.10010574</concept_id>
+  <concept_desc>Computer systems organization~Real-time system architecture</concept_desc>
+  <concept_significance>300</concept_significance>
+  </concept>
+  <concept>
+  <concept_id>10011007.10011006.10011066.10011070</concept_id>
+  <concept_desc>Software and its engineering~Application specific development environments</concept_desc>
+  <concept_significance>500</concept_significance>
+  </concept>
+  <concept>
+  <concept_id>10011007.10011074.10011092.10011093</concept_id>
+  <concept_desc>Software and its engineering~Object oriented development</concept_desc>
+  <concept_significance>100</concept_significance>
+  </concept>
+  <concept>
+  <concept_id>10011007.10011074.10011092.10011691</concept_id>
+  <concept_desc>Software and its engineering~Error handling and recovery</concept_desc>
+  <concept_significance>100</concept_significance>
+  </concept>
+  <concept>
+  <concept_id>10011007.10011074.10011081.10011082</concept_id>
+  <concept_desc>Software and its engineering~Software development methods</concept_desc>
+  <concept_significance>100</concept_significance>
+  </concept>
+  </ccs2012>
 \end{CCSXML}
 
+\ccsdesc[300]{Computing methodologies~Computer graphics}
+\ccsdesc[300]{Computer systems organization~Real-time system architecture}
+\ccsdesc[500]{Software and its engineering~Application specific development environments}
+\ccsdesc[100]{Software and its engineering~Object oriented development}
+\ccsdesc[100]{Software and its engineering~Error handling and recovery}
+\ccsdesc[100]{Software and its engineering~Software development methods}
 \keywords{Common Lisp, Games, Video Games, Computer Graphics, Experience Report}
@@
 
-All of the work we've done, including Trial, and even Kandria, is open source and available on our \href{https://github.com/shirakumo}{GitHub}, in the hopes that it will inspire others to create new projects based upon them.
+All of the work we've done, including Trial, and even Kandria, is open source and available on our GitHub\cite{github}, in the hopes that it will inspire others to create new projects based upon them.
 
@@
 In \cite{hafner2021} we discussed many similar points that this paper touches upon in a format more suitable for people unfamiliar with the intricacies of lisp and especially Common Lisp.
+
+Strandh\cite{strandh2014fast} proposes a new approach to improve the performance characteristics of generic function dispatch.
+Since we make heavy use of CLOS in our game, such improvements are very relevant.
+
+Patton\cite{swcl-gc} are working on a new parallelisation of the SBCL garbage collector to allow for faster collection and thus reduced pause times.
+GC pause times are a frequently reported issue in games, as they can cause unexpected latency, and thus lead to drops in the framerate.
+
+Mäkelä\cite{makela2021development} describe their experience in developing a game using the open source game engine Godot, with C\#.
+Godot is currently the leading open source game engine, and is striving to be a viable alternative to commercial products like Unity and Unreal.
+
+Craighead et al.\cite{craighead2008using} cover a case study of using Unity, a proprietary game engine, to build a small game. Unity is currently most-used game engine for small to mid-sized games.
+
+Nurminen\cite{nurminen1990rft} describe their experience developing and deploying a Common Lisp application to users.
 
@@
 Usual desktop UI toolkits are rather cumbersome to style effectively, as most desktop applications are expected to present themselves in a ``native look and feel'', while games are expected to have rather elaborately customised and animated interfaces.
-Alloy's presentations protocol allows us to style the UI quite extensively, with relative ease (\autoref{lst:presentation}).
-
-\begin{listing}
-\begin{minted}[fontsize=\small]{common-lisp}
-;; Define the shapes that compose the element
-(presentations:define-realization (ui alloy:switch)
-  ((:border simple:rectangle)
-   (alloy:margins)
-   :pattern colors:white
-   :line-width (alloy:un 2))
-  ((:background simple:rectangle)
-   (alloy:margins))
-  ((:switch simple:rectangle)
-   (alloy:extent 0 0 (alloy:pw 0.3) (alloy:ph))))
-
-;; Define how shape properties change
-(presentations:define-update (ui alloy:switch)
-  (:border
-   :hidden-p NIL
-   :pattern (if alloy:focus
-                (colored:color 0.9 0.9 0.9)
-                colors:gray))
-  (:background
-   :pattern (if (alloy:active-p alloy:renderable)
-                colors:dark-accent
-                (colored:color 0 0 0 0.5)))
-  (:switch
-   :pattern (if (alloy:active-p alloy:renderable)
-                colors:accent
-                colors:black)))
-
-;; Define how to animate the properties
-(presentations:define-animated-shapes alloy:switch
-  (:background (simple:pattern :duration 0.2))
-  (:switch (simple:pattern :duration 0.3)
-           (simple:bounds :duration 0.5)))
-\end{minted}
-\caption{An example of the presentations mechanism in Alloy to define the look and visual behaviour of a switch.}
-\label{lst:presentation}
-\end{listing}
+Alloy's presentations protocol allows us to style the UI quite extensively.
+Thanks to macros we can present the user with a declarative style interface to define these behaviours with relative ease.
 
@@
 \begin{enumerate}
-\item Input handling for game controllers.
-This is rather involved, especially on Windows, where multiple APIs need to be supported simultaneously.
-Game controller devices are also prone to broken drivers or bad input mapping, which our library needs to deal with.
+\item Input handling for game controllers. This is rather involved, especially on Windows, where multiple APIs need to be supported simultaneously.
 \item Audio output. We've implemented several different output APIs on both Linux and Windows, as both systems can differ on their supported output interfaces depending on version and setup.
@@
 We make use of structures over standard objects only in very select cases where the impact of the increased verbosity can be contained, and the performance requirements actually indicated that we needed to optimise.
-Specifically this is for the hit information during collision resolution, and for the node graph that powers the pathing AI.
-In those cases the initial class approach produced too much overhead, so we switched to a few specific structures.
+Specifically this is for the hit information during collision resolution, for the node graph that powers the pathing AI, and the glyph information used for text layouting and rendering.
+In those cases the initial class approach produced too much overhead with repeated slot access dispatch, so we switched to a few specific structures.
 
-The dispatch overhead of CLOS has never been a real problem for us.
+Especially with regards to propagating type information CLOS can be a hindrance, as you are prevented from declaring the argument and return types.
+Structures give us the advantage of fixing a return type on their accessors, propagating type information more easily and thus leading to better performing code at their call sites.
+There are projects that allow inlining method dispatch as well, eliminating some of the type ambiguity, but we have not seriously investigated these approaches yet.
+
+Overall the dispatch overhead of CLOS has never been a real problem for us.
 Even with thousands of objects that receive events through our \code{handle} function, hundreds of methods attached to it, and tens of events every frame, not to mention all the other parts in the game that use CLOS, we still manage to easily hit a steady 120 frames per second on a ten year old machine.
@@
 
-The first step in the deployment pipeline is the build of the actual game binary, for which we use ASDF and the Deploy library.
+The first step in the deployment pipeline is the build of the actual game binary, for which we use the ASDF build system and the Deploy library.
 With these all game code is compiled fresh, dumped into an executable suitable for the local platform, and bundled together with the depended-upon shared libraries, producing an executable ready for deployment to target machines.

### 2023-02-11 els2023-kandria/paper.tex
Minor

@@
 
-Newly developed tools to better visualise the data gathered by the statistical profiler, such as the ``flamegraph'' visualiser developed by Jan Moringe have also been invaluable in gaining a better understanding of the performance characteristics of the code.
+Newly developed tools to better visualise the data gathered by the statistical profiler, such as the ``flamegraph'' visualiser developed by Jan Moringen have also been invaluable in gaining a better understanding of the performance characteristics of the code.
 
@@
 \section{Acknowledgements}\label{acknowledgements}
-We would like to thank \textit{you} for being beautiful and nice :)
+We would like to thank the various contributors to all of the projects that have made it possible to make Kandria in the first place, and we would like to thank \textit{you} for being beautiful and nice.
 

### 2023-04-01 els2023-kandria/paper.tex
Adjustments

@@
 \author{Nicolas ``Shinmera'' Hafner}
+\email{shinmera@tymoon.eu}
 \affiliation{%
@@
 
-All of the work we've done, including Trial, and even Kandria, is open source and available on our GitHub\cite{github}, in the hopes that it will inspire others to create new projects based upon them.
+All of the work we've done, including Trial\cite{trial}, and even Kandria\cite{kandria}, is open source and available on our GitHub\cite{github}, in the hopes that it will inspire others to create new projects based upon them.
 
@@
 
-Patton\cite{swcl-gc} are working on a new parallelisation of the SBCL garbage collector to allow for faster collection and thus reduced pause times.
+Patton\cite{swcl-gc} are working on a new parallelisation of the SBCL\cite{sbcl} garbage collector to allow for faster collection and thus reduced pause times.
 GC pause times are a frequently reported issue in games, as they can cause unexpected latency, and thus lead to drops in the framerate.
 
-Mäkelä\cite{makela2021development} describe their experience in developing a game using the open source game engine Godot, with C\#.
+Mäkelä\cite{makela2021development} describe their experience in developing a game using the open source game engine Godot\cite{godot} with C\# in their BSc.
+thesis.
 Godot is currently the leading open source game engine, and is striving to be a viable alternative to commercial products like Unity and Unreal.
 
-Craighead et al.\cite{craighead2008using} cover a case study of using Unity, a proprietary game engine, to build a small game. Unity is currently most-used game engine for small to mid-sized games.
+Craighead et al.\cite{craighead2008using} cover a case study of using Unity\cite{unity}, a proprietary game engine, to build a small game.
+Unity is currently most-used game engine for small to mid-sized games.
 
@@
 
-We are also actively investigating the possibility of porting SBCL and Kandria to the proprietary Nintendo Switch platform, though due to Non-Disclosure Agreements we are unfortunately not at liberty to speak of the specifics involved in that at this time.
+We are also actively investigating the possibility of porting SBCL and Kandria to the proprietary Nintendo Switch platform, though due to non-disclosure agreements we are unfortunately not at liberty to speak of the specifics involved in that at this time.
 
@@
 For basic computer graphics and especially games, this is overkill.
-Most of the linear algebra stays within the confines of 2x2, 3x3, 4x4 matrices, and 2, 3, and 4 element vectors.
+Most of the linear algebra stays within the confines of $2\times 2$, $3\times 3$, $4\times 4$ matrices, and 2, 3, and 4 element vectors.
 
@@
 Back when Trial started out, no linear algebra libraries with such a limited focus existed, and as such the 3d-Vectors and 3d-Matrices libraries were born.
-We have since extended this set of libraries to include 3d-Quaternions for common operations with rotations, and 3d-Transforms, for the convenient encapsulation of a ``transform gizmo'' that can represent rotation, translation, and scaling without Gimbal Lock.
+We have since extended this set of libraries to include 3d-Quaternions for common operations with rotations, and 3d-Transforms, for the convenient encapsulation of a ``transform gizmo'' that can represent rotation, translation, and scaling without gimbal Lock.
 
@@
 We then create a dispatcher function on top which falls back to emitting an \code{etypecase} on other implementations, but will hook into SBCL's \code{deftransform} and similar facilities to better handle expansion and type propagation.
-Finally we create variadic functions on top of the dispatchers which transform any possible variadic call into calls to dispatched 2-argument operation functions, while retaining proper type propagation.
+Finally we create variadic functions on top of the dispatchers which transform any possible variadic call into calls to dispatched two-argument operation functions, while retaining proper type propagation.
 
@@
 When Trial initially began its development in 2016 it was directly integrated with the Qt4 UI toolkit (via CommonQt / Qtools).
-Since Qt4 is a rather large dependency that not only increases the deployed package size, but also invites a lot of C and C++ interop that can cause hard to debug issues, and has not been maintained in many years, it was quickly abandoned, however.
+Since Qt4 is a rather large dependency that not only increases the deployed package size, but also invites a lot of C and C++ interoperation that can cause hard to debug issues, and has not been maintained in many years, it was quickly abandoned, however.
 These days Trial does not depend on a UI toolkit directly, or even a specific backend for its OpenGL use.
 
-Aside from CommonQt, the options for a user interface in Common Lisp were, and remain, limited: GTK runs the same issues as Qt, though with even worse deployment and support aspects, LTK is far too limited and cannot integrate with OpenGL at all, McCLIM similarly does not possess an OpenGL backend or method of integration and is still very limited in its theming capabilities, and the newly established CLOG requires a browser to be shipped, which is far too heavy-weight.
+Aside from CommonQt, the options for a user interface in Common Lisp were, and remain, limited: GTK runs the same issues as Qt, though with even worse deployment and support aspects, LTK is far too limited and cannot integrate with OpenGL at all, McCLIM similarly does not possess an OpenGL backend or method of integration and is still very limited in its theming capabilities, and the newly established CLOG requires a browser to be shipped, which is far too heavyweight.
 
 As a result we decided to implement our own toolkit, called Alloy.
-Alloy is separated into different protocols, with the core being completely independent of any rendering or input method, instead only handling the layouting decisions, the input handling via a generic event protocol, and a system to dynamically react to changes in the represented data.
+Alloy is separated into different protocols, with the core being completely independent of any rendering or input method, instead only handling the layout decisions, the input handling via a generic event protocol, and a system to dynamically react to changes in the represented data.
 
@@
 
-On the Lisp side we have also crafted a more high-level system called Harmony, which makes it trivial to put together these processing pipelines, and interactively play music and sound effects back.
-In Kandria we use these pipelines for a variety of effects such as cross fading multiple music layers (called ``horizontal mixing''), ducking the audio with a low pass filter when the player is under water, and slowing down audio playback when the player enters a slow motion segment.
+On the Lisp side we have also crafted a higher-level system called Harmony, which makes it trivial to put together these processing pipelines, and interactively play music and sound effects back.
+In Kandria we use these pipelines for a variety of effects such as cross fading multiple music layers (called ``horizontal mixing''), ducking the audio with a low pass filter when the player is underwater, and slowing down audio playback when the player enters a slow motion segment.
 
@@
 Despite this though, the interaction still involves C and the usual memory safety perils, and depending on the interface the documentation often leaves things to be desired, especially on MacOS.
-So while our implementation definitely benefits from a much faster retry cycle, it is still a very arduous process to write these OS interop libraries.
+So while our implementation definitely benefits from a much faster retry cycle, it is still a very arduous process to write these OS interoperation libraries.
 
@@
 \begin{enumerate}
-\item Input handling for game controllers. This is rather involved, especially on Windows, where multiple APIs need to be supported simultaneously.
-\item Audio output. We've implemented several different output APIs on both Linux and Windows, as both systems can differ on their supported output interfaces depending on version and setup.
-\item Native dialog boxes. In order to show emergency error boxes or prompt the user for files.
-\item Font discovery. To search the available fonts for a matching set we need to query operating system APIs.
-\item Language querying. To ensure the game launches with the user's language (if supported of course), we again need to query the environment for the preferred localisation.
+\item Input handling for game controllers. This is rather involved, especially on Windows, where multiple APIs need to be supported simultaneously.\cite{cl-gamepad}
+\item Audio output.
+We've implemented several different output APIs on both Linux and Windows, as both systems can differ on their supported output interfaces depending on version and setup.\cite{libmixed}\cite{cl-mixed}\cite{harmony}
+\item Native dialog boxes. In order to show emergency error boxes or prompt the user for files.\cite{messagebox}\cite{file-select}
+\item Font discovery. To search the available fonts for a matching set we need to query operating system APIs.\cite{font-discovery}
+\item Language querying. To ensure the game launches with the user's language (if supported of course), we again need to query the environment for the preferred localisation.\cite{system-locale}
 \end{enumerate}
 
-For graphics output and general window interaction we currently rely on the GLFW C library, as it has proven extremely stable and portable.
+For graphics output and general window interaction we currently rely on the GLFW\cite{glfw} C library, as it has proven extremely stable and portable.
 It is conceivable that we will replace this at some point to reduce C dependencies, but given that we've had zero issues arising from its use, this is rather low priority for us.
@@
 As usual, far more important than constant factor improvements like those has been reducing the asymptotic behaviour.
-Introducing a Bounding Volume Hierarchy to speed up spatial queries and especially collision detection from a primitive quadratic search down to a logarithmic one has definitely provided the most significant performance improvement.
+Introducing a Bounding Volume Hierarchy to speed up spatial queries and especially collision detection from a primitive linear search per object down to a logarithmic one has definitely provided the most significant performance improvement.
 
@@
 In general we have not felt that we've had to do a lot of work to keep garbage production down.
-On a usual system the GC seems to trigger every ten second or so, with no noticeable stutters caused by the pause time.
+On a usual system the GC seems to trigger every ten seconds or so, with no noticeable stutters caused by the pause time.
 We have been made aware of some rare systems on which the GC pause \textit{does} cause visible stutters, but have not been able to identify why that should happen or what to do about it.
@@
 
-The first step in the deployment pipeline is the build of the actual game binary, for which we use the ASDF build system and the Deploy library.
+The first step in the deployment pipeline is the build of the actual game binary, for which we use the ASDF build system and the Deploy\cite{deploy} library.
 With these all game code is compiled fresh, dumped into an executable suitable for the local platform, and bundled together with the depended-upon shared libraries, producing an executable ready for deployment to target machines.
@@
 
-We can also directly build Windows version of the game on Linux by using WINE.
+We can also directly build Windows version of the game on Linux by using WINE\cite{wine}.
 This has worked flawlessly for us, and we imagine that it would also be possible to do the inverse on Windows by using the Windows Subsystem for Linux to build Linux versions of the games.
@@
 A lot more work is needed to support more complex games, which we intend on focusing on in the near future.
-However, we do not at present see any deal breakers that would make it unfeasible to create such games using Common Lisp and the base ecosystem we have helped establish so far.
+However, we do not at present see any deal-breakers that would make it unfeasible to create such games using Common Lisp and the base ecosystem we have helped establish so far.
 

### 2023-04-03 els2023-kandria/paper.tex
Update

@@
 \usepackage[utf8]{inputenc}
+\usepackage[showframe]{geometry}
 \usepackage{enumitem}
@@
 Kandria was funded in part by the Pro Helvetia Interactive Media Grant and the KPT Poland Prize Digital Dragons Accelerator.
+
+\begin{figure*}
+  \begin{centering}
+    \includegraphics[width=1.0\textwidth]{flamegraph.png}
+  \end{centering}
+  \caption{A flamegraph of the game's main thread.
+Due to the high nesting not much is visible at this scale.
+We've identified two specific blocks belonging to collision resolution and rendering.
+The graph of a release build would look slightly different, as several things are optimised away for a release build.}
+\end{figure*}
+
 \bibliography{paper}
+
 \end{document}

### 2023-12-28 els2024-convex-covering/paper.tex
Scaffolding

--- 2023-02-11 els2023-kandria/paper.tex
@@
 \usepackage[utf8]{inputenc}
+\usepackage{geometry}
 \usepackage{enumitem}
@@
 
-\acmConference[ELS'23]{the 15th European Lisp Symposium}{March 21--22 2023}{%
-  }
-\acmISBN{}
+\acmConference[ELS’24]{the 17th European Lisp Symposium}
+{May 6--7 2024}{Vienna, Austria}
 \acmDOI{}
 \setcopyright{rightsretained}
-\copyrightyear{2023}
+\copyrightyear{2024}
 
@@
 
-\title{Experience Report: Kandria - A Game in Common Lisp}
+\title{Paper: Convex Covering}
 
-\author{Nicolas ``Shinmera'' Hafner}
+\author{Jan ``scymtym'' Moringen}
+\email{jmoringe@techfak.uni-bielefeld.de}
+\author{Yukari ``Shinmera'' Hafner}
+\email{shinmera@tymoon.eu}
 \affiliation{%
@@
 
-\begin{abstract}
-  In this paper we outline the experience we've gathered while developing Kandria using Common Lisp.
-Kandria is a video game in the ``action RPG'' genre and has been developed for Windows and Linux with the SBCL implementation.
-Being a video game, the project touches a unique combination of disciplines within computer science, and as such provides an in-depth and comprehensive view of the development process of a large-scale project in Lisp, and the general language ecosystem.
-\end{abstract}
-
 \begin{CCSXML}
@@
   <concept>
-  <concept_id>10010147.10010371</concept_id>
-  <concept_desc>Computing methodologies~Computer graphics</concept_desc>
-  <concept_significance>300</concept_significance>
-  </concept>
-  <concept>
-  <concept_id>10010520.10010570.10010574</concept_id>
-  <concept_desc>Computer systems organization~Real-time system architecture</concept_desc>
-  <concept_significance>300</concept_significance>
-  </concept>
-  <concept>
-  <concept_id>10011007.10011006.10011066.10011070</concept_id>
-  <concept_desc>Software and its engineering~Application specific development environments</concept_desc>
+  <concept_id>10010147.10010371.10010396.10010398</concept_id>
+  <concept_desc>Computing methodologies~Mesh geometry models</concept_desc>
   <concept_significance>500</concept_significance>
@@
   <concept>
-  <concept_id>10011007.10011074.10011092.10011093</concept_id>
-  <concept_desc>Software and its engineering~Object oriented development</concept_desc>
-  <concept_significance>100</concept_significance>
+  <concept_id>10010147.10010371</concept_id>
+  <concept_desc>Computing methodologies~Computer graphics</concept_desc>
+  <concept_significance>500</concept_significance>
   </concept>
   <concept>
-  <concept_id>10011007.10011074.10011092.10011691</concept_id>
-  <concept_desc>Software and its engineering~Error handling and recovery</concept_desc>
-  <concept_significance>100</concept_significance>
-  </concept>
-  <concept>
-  <concept_id>10011007.10011074.10011081.10011082</concept_id>
-  <concept_desc>Software and its engineering~Software development methods</concept_desc>
+  <concept_id>10010405.10010469.10010474</concept_id>
+  <concept_desc>Applied computing~Media arts</concept_desc>
   <concept_significance>100</concept_significance>
@@
 
-\ccsdesc[300]{Computing methodologies~Computer graphics}
-\ccsdesc[300]{Computer systems organization~Real-time system architecture}
-\ccsdesc[500]{Software and its engineering~Application specific development environments}
-\ccsdesc[100]{Software and its engineering~Object oriented development}
-\ccsdesc[100]{Software and its engineering~Error handling and recovery}
-\ccsdesc[100]{Software and its engineering~Software development methods}
-\keywords{Common Lisp, Games, Video Games, Computer Graphics, Experience Report}
+\ccsdesc[500]{Computing methodologies~Mesh geometry models}
+\ccsdesc[500]{Computing methodologies~Computer graphics}
+\ccsdesc[100]{Applied computing~Media arts}
+
+\begin{abstract}
+  A
+\end{abstract}
+
+\keywords{Common Lisp, Convex Decomposition, Games, Video Games, Computer Graphics, Experience Report}
 
@@
 \section{Introduction}\label{introduction}
-Kandria is a video game in the ``action RPG'' genre that was released worldwide on the \href{https://kandria.com/steam}{Valve Steam} platform for Windows and Linux in January of 2023, receiving very positive reviews.
-The game was developed using the Trial engine and thus relies almost entirely upon code and libraries written in Common Lisp -- to our knowledge the first commercial game like this to be released.
-
-\begin{figure}[h]
-  \begin{centering}
-    \includegraphics[width=0.4\textwidth]{screenshot 8.png}
-  \end{centering}
-  \caption{A screenshot of Kandria}
-\end{figure}
-
-Games lie at an intersection of many different computer science disciplines such as audio, graphics, interfaces, soft real-time, artificial intelligence, and more.
-As such games provide a unique challenge in combining all of these disciplines together into one product, and ultimately shipping this product to paying customers.
-
-In this paper we will outline the challenges we faced in realising Kandria as related to Common Lisp, discuss advantages of our approach, and take a look at the work still ahead of us in expanding the capabilities of our engine to better address the requirements of even more complex games we would like to develop in the future.
-
-In particular we will note our own experiences with a few common Boogeymen, such as the overhead of CLOS dispatch, the pause times of GC, the maturity of the library ecosystem, and the stability and size of deployed binaries.
-
-Since this is an experience report, we cannot present any specific figures on performance characteristics, statistics on used projects, or some sort of unified thesis.
-Instead we hope that this insight will be valuable for readers to gain a better understanding of the complexities involved, the benefits we have identified with Lisp over other ecosystems, and, most importantly, the areas in which work still needs to be done.
-
-All of the work we've done, including Trial, and even Kandria, is open source and available on our GitHub\cite{github}, in the hopes that it will inspire others to create new projects based upon them.
 
 \section{Related Work}\label{relatedwork}
-In \cite{hafner2021} we discussed many similar points that this paper touches upon in a format more suitable for people unfamiliar with the intricacies of lisp and especially Common Lisp.
 
-Strandh\cite{strandh2014fast} proposes a new approach to improve the performance characteristics of generic function dispatch.
-Since we make heavy use of CLOS in our game, such improvements are very relevant.
+Liu et al.\cite{liu2008convex}
 
-Patton\cite{swcl-gc} are working on a new parallelisation of the SBCL garbage collector to allow for faster collection and thus reduced pause times.
-GC pause times are a frequently reported issue in games, as they can cause unexpected latency, and thus lead to drops in the framerate.
+\section{Algorithm}\label{algorithm}
 
-Mäkelä\cite{makela2021development} describe their experience in developing a game using the open source game engine Godot, with C\#.
-Godot is currently the leading open source game engine, and is striving to be a viable alternative to commercial products like Unity and Unreal.
-
-Craighead et al.\cite{craighead2008using} cover a case study of using Unity, a proprietary game engine, to build a small game. Unity is currently most-used game engine for small to mid-sized games.
-
-Nurminen\cite{nurminen1990rft} describe their experience developing and deploying a Common Lisp application to users.
-
-\section{Library Ecosystem}\label{yaks}
-In this section we will outline our general experience working in the Common Lisp ecosystem, and particularly the contributions we've made to it in order to implement Kandria.
-
-Out of the 110 libraries Kandria depends upon, 54 were written by us.
-This does mean that we've spent a rather significant amount of time ``yak shaving'' and creating libraries to fulfil a variety of needs that were, prior to their creation, either completely unfulfilled, or unsatisfactorily so.
-We will go into detail on some of the more important ones in the following subsections.
-
-While this is undeniably a lot of work, it has been done now and is available to others.
-Furthermore, of the other half of libraries that we were not authors of, the vast majority have been extremely stable.
-We only needed to supply extremely minimal patches to a select few libraries, most of which were reviewed and accepted relatively swiftly.
-
-Among this we also count the SBCL implementation itself.
-While we strive to write implementation independent code in our separate libraries, for Kandria itself we decided to only focus on SBCL in order to reduce the development overhead.
-SBCL offers great native code performance and is available for all platform configurations that we require.
-And, perhaps most importantly, it is very actively maintained.
-Most of the issues we've encountered in releases were usually fixed within a few weeks, if not days.
-
-We are also actively investigating the possibility of porting SBCL and Kandria to the proprietary Nintendo Switch platform, though due to Non-Disclosure Agreements we are unfortunately not at liberty to speak of the specifics involved in that at this time.
-
-Overall while we certainly had to create a lot of libraries to fulfil our needs, and those libraries presented a significant amount of effort to implement, we remain convinced that we were only able to implement these systems in a respectable amount of time due to the convenience factors that Lisp offers us.
-
-We'll now touch on a few specific areas that we developed libraries for. This is by no means exhaustive, but we consider these to be the most relevant to the general community.
-
-\subsection{Math Libraries}\label{math}
-While there is no shortage of math libraries available to Common Lisp, especially linear algebra implementations, most of those libraries focus on large scale scientific computing, often by integrating with foreign libraries such as BLAS and LAPACK.
-For basic computer graphics and especially games, this is overkill.
-Most of the linear algebra stays within the confines of 2x2, 3x3, 4x4 matrices, and 2, 3, and 4 element vectors.
-
-It is more important to provide a very convenient interface that allows us to perform computations on these elements with adequate speed.
-Back when Trial started out, no linear algebra libraries with such a limited focus existed, and as such the 3d-Vectors and 3d-Matrices libraries were born.
-We have since extended this set of libraries to include 3d-Quaternions for common operations with rotations, and 3d-Transforms, for the convenient encapsulation of a ``transform gizmo'' that can represent rotation, translation, and scaling without Gimbal Lock.
-
-All of these libraries rely very heavily on macros to reduce code duplication and automatically generate code for loop unrolling and other common tactics in linear algebra code.
-The current versions of these libraries all work by emitting an \code{etypecase} for every operation, which then handles dispatch based on the provided argument types.
-The operation functions are inlined, such that the compiler can eliminate the dispatch altogether if the argument types are locally known.
-This allows us to provide a generic interface to the user that's quite convenient, while still staying competitive in performance critical sections.
-
-Unfortunately this approach, while portable, is also riddled with issues: since every operation is inlined, this leads to explosive code growth for the compiler before it can reduce the code back down again by eliminating superfluous dispatch \code{etypecase}s.
-Type inference is also much more complicated, and stack allocation is usually only possible via careful manual rewriting of the operations involved.
-The libraries are also limited to a single float type, meaning you can by default only create vectors with single-floats as elements.
-
-In this case the lack of static typing facilities and lack of portable compiler hooks for integrating with type inference really hurts the compile speed, implementation clarity, and ultimate performance of the resulting code.
-
-We have started work on a full rewrite of all libraries that take a fundamentally different approach: instead of emitting etypecases and relying on inlining, we create a sort of ``template mechanism'' by which we can generate all possible permutations of a singular function for all involved argument types.
-This gives us very tiny, but perfectly optimised base functions for all required operations.
-We then create a dispatcher function on top which falls back to emitting an \code{etypecase} on other implementations, but will hook into SBCL's \code{deftransform} and similar facilities to better handle expansion and type propagation.
-Finally we create variadic functions on top of the dispatchers which transform any possible variadic call into calls to dispatched 2-argument operation functions, while retaining proper type propagation.
-
-Ultimately this results in much more easily understandable and performing code.
-However, it still comes at a cost: because we cannot know ahead of time which type combinations and operations the user will actually need, we must generate all possible combinations ahead of time.
-This potentially results in thousands of functions being generated that are never used.
-We would like a compiler facility that allows us to hook into the expansion process of a function call to generate the required permutations on-demand.
-
-While such a facility is conceivable, and making it work for unknown runtime types by delayed compilation is, too, we currently have no plans to implement such an advanced strategy.
-More importantly, we hope that at some point in the future implementors can come to some kind of consensus that would allow a portability library to expose a similar mechanism to SBCL's \code{deftransform} for faster, more convenient, type-inference-informed call expansion.
-
-\subsection{User Interfaces}\label{ui}
-When Trial initially began its development in 2016 it was directly integrated with the Qt4 UI toolkit (via CommonQt / Qtools).
-Since Qt4 is a rather large dependency that not only increases the deployed package size, but also invites a lot of C and C++ interop that can cause hard to debug issues, and has not been maintained in many years, it was quickly abandoned, however.
-These days Trial does not depend on a UI toolkit directly, or even a specific backend for its OpenGL use.
-
-Aside from CommonQt, the options for a user interface in Common Lisp were, and remain, limited: GTK runs the same issues as Qt, though with even worse deployment and support aspects, LTK is far too limited and cannot integrate with OpenGL at all, McCLIM similarly does not possess an OpenGL backend or method of integration and is still very limited in its theming capabilities, and the newly established CLOG requires a browser to be shipped, which is far too heavy-weight.
-
-As a result we decided to implement our own toolkit, called Alloy.
-Alloy is separated into different protocols, with the core being completely independent of any rendering or input method, instead only handling the layouting decisions, the input handling via a generic event protocol, and a system to dynamically react to changes in the represented data.
-
-How visual elements can be rendered is then offloaded to several other protocols: the ``Simple Rendering Protocol'' which provides a basic text and shape rendering API, the ``Presentations Protocol'' which allows users to describe how to compose the look of a visual element via the basic shapes from the Simple protocol, and the ``Animations Protocol'' which allows users to describe how shapes change over time as properties of a visual element are changed.
-
-Finally, the Simple protocol needs to be implemented by a backend, such as the ``OpenGL Renderer'' to actually provide a way to draw the shapes in some way, the Core protocol needs to receive input events from the surrounding context to actually react to user input, and some method of rendering and layouting text needs to be provided.
-Text in particular is separated out as it is a very complex topic in its own right.
-
-This separation of concerns via protocols implemented in CLOS allows us to make Alloy far more amenable to being ported to different backends.
-In particular, it allows us to use Alloy in contexts that are otherwise rather unusual for UI toolkits, such as within a game where the actual operating system interaction and rendering logic cannot be directly controlled by the toolkit itself, but is instead handled by the game engine.
-
-Usual desktop UI toolkits are rather cumbersome to style effectively, as most desktop applications are expected to present themselves in a ``native look and feel'', while games are expected to have rather elaborately customised and animated interfaces.
-Alloy's presentations protocol allows us to style the UI quite extensively.
-Thanks to macros we can present the user with a declarative style interface to define these behaviours with relative ease.
-
-This protocol separation would not be possible to implement without the high degree of flexibility CLOS offers us in combining behaviours, and even without advanced techniques such as shadow mixins and metaclasses.
-
-\subsection{Audio Processing}\label{audio}
-Of the entire system, audio processing is where the toughest constraints apply, as audio is very sensitive to latency.
-We cannot process audio in large buffers to smooth over processing hiccups, as then sound effects would be desynchronised from their visual counterparts.
-Audio processing can also often be very computation intensive, and benefits greatly from vectorisation.
-
-For these reasons we have decided to write the main bulk of our processing as a C library, rather than Common Lisp, as at the time it was the fastest way for us to get the kinds of performance constraints met and the kinds of capabilities we needed.
-Now with the advent of the sb-simd contrib, writing a competitive alternative (albeit constrained to SBCL) may be feasible.
-
-In any case, our library, libmixed, follows a very strict C style and is extensible and interoperable from Lisp.
-You can write ``segments'' that process audio from Lisp as well, and integrate with the rest of the processing system neatly.
-
-This allows us to write the more hairy parts such as audio format decoding, and audio playback in Lisp, instead, without sacrificing the performance gains in the bulk of the processing.
-Libmixed then takes care of unpacking and decoding audio streams, applying a variety of effects, mixing multiple audio streams together, and even resampling everything to fit into the expected, and often differing, sample rates at the input and output points.
-
-Libmixed also includes a full introspection API, allowing us to not only put the audio processing pipeline together from Lisp, but also to interactively inspect the state of the processing segments and modify them at runtime.
-We can also use the extensibility to prototype new effects from Lisp before lowering them down to C should the performance requirements be strict enough.
-
-On the Lisp side we have also crafted a more high-level system called Harmony, which makes it trivial to put together these processing pipelines, and interactively play music and sound effects back.
-In Kandria we use these pipelines for a variety of effects such as cross fading multiple music layers (called ``horizontal mixing''), ducking the audio with a low pass filter when the player is under water, and slowing down audio playback when the player enters a slow motion segment.
-
-\subsection{Operating System Interaction}\label{os}
-In order to output sound and graphics, process input, and receive user information we need to interact with the surrounding operating system.
-On all three targets we care about, this is done via calls to a C API of some sort.
-Thanks to CFFI it is usually unnecessary to actually rely on an external C library to do so, and we can instead code the interaction directly in Lisp, retaining the interactive implementation and debugging.
-
-Despite this though, the interaction still involves C and the usual memory safety perils, and depending on the interface the documentation often leaves things to be desired, especially on MacOS.
-So while our implementation definitely benefits from a much faster retry cycle, it is still a very arduous process to write these OS interop libraries.
-
-Particularly we had to write libraries to do the following:
-\begin{enumerate}
-\item Input handling for game controllers. This is rather involved, especially on Windows, where multiple APIs need to be supported simultaneously.
-\item Audio output. We've implemented several different output APIs on both Linux and Windows, as both systems can differ on their supported output interfaces depending on version and setup.
-\item Native dialog boxes. In order to show emergency error boxes or prompt the user for files.
-\item Font discovery. To search the available fonts for a matching set we need to query operating system APIs.
-\item Language querying. To ensure the game launches with the user's language (if supported of course), we again need to query the environment for the preferred localisation.
-\end{enumerate}
-
-For graphics output and general window interaction we currently rely on the GLFW C library, as it has proven extremely stable and portable.
-It is conceivable that we will replace this at some point to reduce C dependencies, but given that we've had zero issues arising from its use, this is rather low priority for us.
-
-\subsection{Service Integration}\label{steam}
-In order to publish a game on the Steam platform, you must integrate with their SteamWorks SDK.
-The actually required amount of integration is very minimal, and merely involves loading their shared library and calling a single function.
-However, the SteamWorks SDK offers a lot of other functionality that games can make use of, like social networking features, user generated content sharing, multiplayer systems, and more.
-
-Unlike the prior OS interfaces, the default expected interaction mode with the SteamWorks SDK is via C++, which is quite difficult to talk with directly via CFFI.
-Fortunately, they also offer a raw C API, but this API is not fully documented and fraught with strange gotchas.
-The SDK does ship a ``machine readable'' description of the endpoints, structures, and types in the form of a JSON file, but as we have found this file is both incomplete and partially incorrect.
-
-In our implementation we analyse the JSON file along with a manually supplied file with the lacking data, and a couple of the shipped static headers to automatically generate the CFFI wrapper data structures, types, constants, and functions.
-Since the analysis of these files is rather involved however, we have chosen not to use macroexpansion, but rather provide a separate system which emits Lisp code to a new file.
-
-In addition to this generated interface, we also had to supply a minimal ``shim'' C library that handles a few select API calls that rely on structure-by-value passing, a feature which is not natively supported on most Lisp implementations at this point.
-Another workaround would be to use libffi, but libffi comes with its own issue, and would be another C library to depend on anyhow, so we opted for the much simpler and easier to understand alternative of writing our own minimal shim library.
-
-In Kandria we make use of the extended services to present the user with an on-screen keyboard when they are using a controller, to read out the username for default save file naming, and to integrate with the Steam platform's ``achievements'' system, which players have come to expect.
-
-\section{Common Lisp Object System}\label{clos}
-In both Trial and Kandria we make rather extensive use of CLOS, both its basic features of classes, multiple dispatch, and deep hierarchies via mixins, and the advanced features of metaclasses.
-For instance, our entire event handling system is simply a singular function \code{(handle event receiver)} which we define methods on to receive events.
-A central ``event loop'' object then just calls the handle function for every event it receives on every receiver it has in its internal list.
-
-As described in our prior paper \cite{hafner2018} we use metaclasses to attach ``shader fragments'' (code that is executed on the GPU during rendering) to classes and inherit the behaviour of these fragments together, allowing composition of rendering behaviour through inheritance as well.
-
-Overall the use of classes as an organisational structure lends itself extremely well to games, especially in the presence of multiple inheritance and mixins.
-We also often use this to introduce ``marker classes'' that contain no behaviour or data of their own, but act as type information that other parts of the system use, such as whether a class can be instantiated by the editor, resized, should be treated as solid during collision, should not be deleted when loading a prior save state, etc.
-
-We make use of structures over standard objects only in very select cases where the impact of the increased verbosity can be contained, and the performance requirements actually indicated that we needed to optimise.
-Specifically this is for the hit information during collision resolution, for the node graph that powers the pathing AI, and the glyph information used for text layouting and rendering.
-In those cases the initial class approach produced too much overhead with repeated slot access dispatch, so we switched to a few specific structures.
-
-Especially with regards to propagating type information CLOS can be a hindrance, as you are prevented from declaring the argument and return types.
-Structures give us the advantage of fixing a return type on their accessors, propagating type information more easily and thus leading to better performing code at their call sites.
-There are projects that allow inlining method dispatch as well, eliminating some of the type ambiguity, but we have not seriously investigated these approaches yet.
-
-Overall the dispatch overhead of CLOS has never been a real problem for us.
-Even with thousands of objects that receive events through our \code{handle} function, hundreds of methods attached to it, and tens of events every frame, not to mention all the other parts in the game that use CLOS, we still manage to easily hit a steady 120 frames per second on a ten year old machine.
-
-\section{Performance \& Garbage Collection}\label{performance}
-Overall we have needed to do surprisingly little actual performance analysis and optimisation work to make Kandria run well.
-This is definitely in large part thanks to SBCL's quite good native code compiler and type inference systems, and the prior work we've done to design critical libraries to not be completely obscene in terms of their performance characteristics.
-
-We have done larger scale rewrites of several subsystems in Trial, which have provided us with general performance increases, though these rewrites were primarily done to increase the code clarity and usability, with the performance being a nice bonus.
-
-As usual, far more important than constant factor improvements like those has been reducing the asymptotic behaviour.
-Introducing a Bounding Volume Hierarchy to speed up spatial queries and especially collision detection from a primitive quadratic search down to a logarithmic one has definitely provided the most significant performance improvement.
-
-While asymptotic behaviour improvements will be applicable for projects in any language, with SBCL we've had the definite advantage of being able to rely on the statistical profiler to determine hot spots in the code.
-This is especially advantageous in Lisp as we can turn the profiler on and off at any time, to capture exact segments that we are interested in, rather than having to capture the full run of a program and then massaging the data manually.
-
-Newly developed tools to better visualise the data gathered by the statistical profiler, such as the ``flamegraph'' visualiser developed by Jan Moringen have also been invaluable in gaining a better understanding of the performance characteristics of the code.
-
-As outlined in \autoref{math} above, there are still areas in which Lisp struggles to be truly competitive, largely in part due to its very dynamic typing, an aspect that is otherwise very advantageous for development.
-We've also encountered issues in eliminating superfluous consing, as many convenient styles of writing code will prevent stack allocation and other garbage elimination.
-
-We have managed to keep our garbage production down largely with two tricks:
-
-\begin{itemize}
-\item Pooling and preallocation.
-By storing objects in a pool and manually allocating and freeing them we can avoid allocating them at runtime, leaving less work for the GC to deal with.
-Since the objects are long lived, they will also be quickly promoted to later generations, lessening the work to scan other generations.
-The obvious downside is that we lose automatic freeing and may introduce double-use cases, but for many cases the lifetimes of the objects can be predetermined and managed relatively painlessly.
-\item Using \code{load-time-value} in lieu of stack allocation.
-With this trick we can allocate a ``local object'' that will be modified at runtime rather than allocated fresh.
-The downside of this approach is that whichever function provides the load-time-value object cannot be called concurrently, and the programmer must be vigilant in tracking the lifetime of the load-time-value object should it escape the dynamic extent of the function.
-\end{itemize}
-
-We are also very interested in a recent proposal to add memory arenas to SBCL, which would allow us to capture all of the objects allocated within a dynamic extent and free them immediately on exit, further lessening the burden of the global GC in cases where we know the exact lifetimes.
-
-Aside from memory arenas, we're also very interested in recent work by Hayley Patton to parallelise SBCL's garbage collector, though it is currently unknown when this feature will become mature and be merged into mainline SBCL.
-
-In general we have not felt that we've had to do a lot of work to keep garbage production down.
-On a usual system the GC seems to trigger every ten second or so, with no noticeable stutters caused by the pause time.
-We have been made aware of some rare systems on which the GC pause \textit{does} cause visible stutters, but have not been able to identify why that should happen or what to do about it.
-Completely eliminating all runtime garbage production does not seem like an effective way to spend our time, however.
-
-\section{Deployment}\label{deployment}
-Finally we would like to talk about our experience actually delivering Kandria to users and the process we have developed for doing so.
-
-The first step in the deployment pipeline is the build of the actual game binary, for which we use the ASDF build system and the Deploy library.
-With these all game code is compiled fresh, dumped into an executable suitable for the local platform, and bundled together with the depended-upon shared libraries, producing an executable ready for deployment to target machines.
-
-However, since the build does not happen from scratch via a platform independent compiler, we have to ensure that we use an SBCL version that has been compiled on our minimal target platform version.
-For instance, we build the SBCL we use for deployment in a virtual machine on an old Linux version to ensure that internal symbol version dependencies in glibc can be satisfied on target machines.
-We have to take similar care for all required shared libraries that are shipped.
-Once this version has been compiled however, we can simply invoke it on any Linux version we like to use, and don't have to build the entire system inside a VM.
-
-We can also directly build Windows version of the game on Linux by using WINE.
-This has worked flawlessly for us, and we imagine that it would also be possible to do the inverse on Windows by using the Windows Subsystem for Linux to build Linux versions of the games.
-We've also investigated the possibility of using Darling to build MacOS versions.
-Unfortunately Darling is not able to run SBCL as of the writing of this paper.
-
-For all generated binaries we also make use of SBCL's core compression to reduce the binary size to about 35MB per platform.
-The newly introduced support for zstd also brings improved compression rates and startup speeds over the older zlib.
-This binary size is more than acceptable to us and is completely dwarfed by the resource files for sound and graphics.
-
-We also make use of hooks in the Deploy library to extract these resource files from their source directory and bundle them alongside the binary when building.
-Paths to resource files are resolved at runtime in the engine, and are thus trivial to redirect to binary-relative paths when deployed.
-
-Finally we have created a release system that invokes the necessary SBCL subprocesses to kick off a build, prunes out unnecessary files from the generated release, then bundles it all into a ZIP for easy delivery to testers.
-The system can also upload the release directly to a variety of distribution platforms including Steam, Itch.io, Keygen, and generic HTTP or FTP servers.
-Ultimately this allows us to compile, bundle, package, and upload new builds with a single command, drastically reducing the time and complexity involved in supplying testers and users with bugfixes.
-
-Another aspect that is advantageous to Lisp's dynamic nature is that when an unhandled condition occurs on a target system we can, in almost all cases, still use the system well enough to gather telemetry and submit an automated crash report.
-This has been invaluable for us to detect rare bugs, especially ones related to uncommon system configurations that we would otherwise never have been able to catch.
+\section{Extensions}\label{extensions}
 
 \section{Conclusion}\label{conclusion}
-Overall the biggest hurdle we have had in our development of Kandria has nothing to do with the language itself, but rather with the general lack of manpower behind the community.
-We have had to spend large amounts of time implementing, testing, and documenting auxiliary systems that have often close to nothing to do with the core process of implementing a game.
-This is also why we have been extremely light on topics that concern the actual game implementation process in this paper.
-
-On the other hand, it is very clear to us that interactive development is an immense boon to the development of a game.
-Being able to redefine game behaviour at runtime to quickly iterate on the game content and feel is invaluable.
-So much so that almost every engine in use will have some form of scripting language available for game logic specifically, such that designers can iterate quickly.
-However, having the full stack of code available, debuggable, and performing just as well as any other part of the system is undeniably far more convenient.
-
-We also fully recognise that while Kandria is a full game project, it is by its nature rather limited in the required processing capabilities.
-A lot more work is needed to support more complex games, which we intend on focusing on in the near future.
-However, we do not at present see any deal breakers that would make it unfeasible to create such games using Common Lisp and the base ecosystem we have helped establish so far.
-
-If anything the recent advances in SBCL's capabilities show us a very promising future and we are excited to make use of them to further improve the situation, and with much of the ``grunt work'' now done we should be able to focus our efforts onto problems more directly associated with game development instead.
 
 \section{Further Work}\label{further-work}
-Currently a sizable amount of work in Kandria has not been backported into Trial for more general purpose use.
-We would like to extract a few of the systems and generalise them to make them available for other users.
-
-We are also working on implementing several new subsystems in Trial to allow creating 3D games, as well.
-A skeletal animation system has been completed, and we're currently working on a physics subsystem.
-Also needed will be several spatial query structures to speed up collision testing, along with a more unified rendering subsystem to support Physics Based Rendering pipelines.
-
-Finally we are also exploring the possibility of porting the engine to work on closed platforms such as the Nintendo Switch.
-This presents several challenges that we unfortunately cannot elaborate on here due to non-disclosure agreements.
 
 \section{Acknowledgements}\label{acknowledgements}
-We would like to thank the various contributors to all of the projects that have made it possible to make Kandria in the first place, and we would like to thank \textit{you} for being beautiful and nice.
 
-Kandria was funded in part by the Pro Helvetia Interactive Media Grant and the KPT Poland Prize Digital Dragons Accelerator.
 \bibliography{paper}
+
 \end{document}

### 2023-12-28 els2024-convex-covering/paper.tex
Start working on related works

@@
 \section{Related Work}\label{relatedwork}
+Liu et al.\cite{liu2008convex}'s work serves as the baseline for our implementation.
+Unfortunately their descriptions of the algorithm's details aren't entirely precise, making it difficult to reproduce their results exactly.
+We were also unable to find any publication of source code at all, let alone a working implementation.
 
-Liu et al.\cite{liu2008convex}
+Mamout et al.\cite{mamou2016volumetric}'s work and their open implementation, \href{https://github.com/kmammou/v-hacd}{``V-HACD''}, provide a high-quality \textit{approximate} convex decomposition algorithm.
+Their algorithm relies on a voxelisation step, which forces the source mesh into a watertight 2-manifold representation, and introduces deviations from the source mesh's vertices.
+This can lead to noticeably different collision behaviour for terrain than the source mesh would produce.
+It can also easily drastically increase the total number of vertices compared to the source mesh, degrading collision performance.
+Their algorithm does allow tuning the complexity and number of produced hulls, but doing so requires human evaluation and there is no good general case behaviour.
 

### 2023-12-28 els2024-convex-covering/paper.tex
Two more citations

@@
 
+Wei et al.\cite{wei2022coacd} present a much improved method for approximate convex decomposition that is especially tuned for collision handling.
+However, their approach is extremely complex and difficult to implement, relying on many other algorithm implementations, such as 2D triangulation and monte-carlo tree search, which we would have had to reproduce in Lisp.
+Liu et al.'s original work only requires on an implementation of the Quickhull\cite{barber1996quickhull} algorithm.
+
 \section{Algorithm}\label{algorithm}

### 2024-01-01 els2024-convex-covering/paper.tex
Eh.

@@
 
-\author{Jan ``scymtym'' Moringen}
+\author{Jan ``Scymtym'' Moringen}
 \email{jmoringe@techfak.uni-bielefeld.de}
@@
 \section{Introduction}\label{introduction}
+Many geometrical applications such as games require efficient intersection tests between various geometrical shapes.
+Efficient algorithms for such intersection tests such as GJK\cite{gilbert1988fast} require the involved shapes to be convex, however.
+This is a difficult constraint, as most shapes encountered in nature and in environments designed for games are non-convex.
+A simple solution to this problem is to compute a convex-hull over the concave shape.
+However, this creates imprecision in the collision result, which may not always be permissible.
+Especially for highly non-convex objects such as terrain or building interiors, merely computing a convex-hull is not feasible.
+Instead, we need to decompose the shape into multiple convex shapes.
+
+Since computing an optimal convex decomposition is NP-hard\cite{1056648} all algorithms to perform such decompositions operate on heuristics and various relaxations of the constraints.
+A frequent relaxation is to allow the resulting shapes to be ``approximately convex''.
+This is usually permissible as the intersection test algorithms still deliver results that are accurate enough even in the presence of very small concavities.
+Another relaxation is to allow the convex hulls to not have to match the input surface exactly.
+However, for the case of environment geometry in games, this approximation means that the resulting shapes can produce geometry that leads to collisions that would not be present if a test were performed against the input mesh instead.
+Such protrusions and inaccuracies can be significant enough to be noticeable to the player.
+
+For efficiency reasons we also desire a couple of additional, competing qualities:
+\begin{itemize}
+\item The number of vertices in a convex hull should be small. GJK degrades in performance as the number of vertices on a convex hull increases.
+\item The number of convex hulls should be small.
+If we keep trading vertices per hull for more convex hulls, we instead trivialise the problem and create too many potential candidates for which the costly fine intersection test needs to be performed.
+\item The size of convex hulls should be small.
+If the resulting convex hull has a large extent, the broad-phase intersection test will include it for many cases, leading to more candidates on average.
+\end{itemize}
+
+In this paper we present an algorithm that produces a convex decomposition that does not produce additional collisions and can be tuned to optimise the number and complexity of convex hulls.
+It is also usable as an automatic off-the-shelf component that will generate an acceptable result without manual tuning of its parameters.
 
@@
 \section{Algorithm}\label{algorithm}
+\subsection{Overview}
+The basic idea of the algorithm is quite simple as illustrated by the pseudo-code in \autoref{lst:algorithm}
+
+\begin{listing}[h]
+\begin{minted}[fontsize=\small]{lisp}
+(defun decompose (faces)
+  (let ((patches (mapcar #'make-patch faces)))
+    (loop for candidates = (find-merge-candidates patches)
+          while candidates
+          do (sort candidates #'< :key #'merge-cost)
+             (merge-patches (first candidates) patches))
+    patches))
+
+(defun find-merge-candidates (patches)
+  (loop for patch in patches
+        append (loop for neighbour in (patch-neighbours patch)
+                     when (merge-possible-p patch neighbour)
+                     collect (cons patch neighbour))))
+\end{minted}
+\caption{A pseudo-code illustration of the basic decomposition algorithm}
+\label{lst:algorithm}
+\end{listing}
+
+We consider a set of ``patches'', each of which forms a convex hull.
+Initially each face is simply turned into its own patch, after which we try to iteratively merge patches where possible.
+A merge is only valid if the resulting patch is still convex and, most importantly, does not introduce new faces that would cause collision behaviour to differ.
+In order to merge the actual patch geometry, we simply run the Quickhull algorithm, which gives us a minimal convex hull of the two input patches.
+
+Since the result of the algorithm depends on the order in which patches are merged, we also perform a heuristic merge cost estimation to select the best merge candidate in each iteration.
+The exact behaviour of this merge cost estimation has great consequences on the quality of the resulting mesh and the speed at which the algorithm converges.
+
+\subsection{Merge Criteria}
+\subsection{Merge Cost}
+\subsection{Optimisations}
 
 \section{Extensions}\label{extensions}
+\subsection{Parallelisation}
+\subsection{Merge Tolerance}
+
+\section{Results}\label{results}
 

### 2024-01-05 els2024-convex-covering/paper.tex
Cite MPR

@@
 Many geometrical applications such as games require efficient intersection tests between various geometrical shapes.
-Efficient algorithms for such intersection tests such as GJK\cite{gilbert1988fast} require the involved shapes to be convex, however.
+Efficient algorithms for such intersection tests such as GJK\cite{gilbert1988fast} or MPR\cite{snethen2008xenocollide} require the involved shapes to be convex, however.
 This is a difficult constraint, as most shapes encountered in nature and in environments designed for games are non-convex.
@@
 
-Since computing an optimal convex decomposition is NP-hard\cite{1056648} all algorithms to perform such decompositions operate on heuristics and various relaxations of the constraints.
+Since computing an optimal convex decomposition is NP-hard\cite{1056648} all used algorithms to perform such decompositions operate on heuristics and various relaxations of the constraints.
 A frequent relaxation is to allow the resulting shapes to be ``approximately convex''.

### 2024-02-05 els2024-convex-covering/paper.tex
Add some notes to ELS2024 paper

@@
           while candidates
-          do (sort candidates #'< :key #'merge-cost)
-             (merge-patches (first candidates) patches))
+          do (let ((best (sort candidates #'<
+                               :key #'merge-cost)))
+               (merge-patches best patches)))
     patches))
@@
 \subsection{Merge Criteria}
+
+\subsubsection{Vertex and Edge Criterion}
+
+Differences from paper:
+\begin{itemize}
+\item Vertices and edges handled in the same test
+\item Heuristic for ``touching'' constellation gives rise to tolerance (see below)
+\end{itemize}
+
+\subsubsection{Matching Normals Criterion}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=.5\linewidth]{normal-criterion.png}
+  \caption{Illustration of the \emph{Matching Normals Criterion}: The orange patch hull (drawn at an offset) is being tested for validity.
+For two hull facets and two mesh faces, the normals of the hull do not match the mesh normals.
+The mismatches are indicated in red.}
+  \label{fig:matching-normals}
+\end{figure}
+
+See figure~\ref{fig:matching-normals}.
+
+\subsubsection{Boundary Constraint Bars Criterion}
+
+\subsubsection{Touching Triangles on Negative Side Criterion}
+
 \subsection{Merge Cost}
+
+\subsection{Flat Patches}
+
 \subsection{Optimisations}
 
+\subsubsection{Spatial Indices}
+
+As mentioned in Section~\ref{}, the merge criteria within the patch validity computation involve predicates over all vertices, edges and faces of the input mesh.
+However, only predicate results computed on mesh features that spatially close to the patch feature in question actually contribute to the result.
+For this reason, the computation can be restricted to such mesh features from the beginning by querying spatial indices which support efficient spatial queries for mesh features.
+We use a modified kd-tree~\cite{} which supports AABB intersection queries TODO explain.
+
+\subsubsection{Priority Queue for Merging}
+
+Cite implementation? Describe cost to integer conversion?
+
 \section{Extensions}\label{extensions}
+
 \subsection{Parallelisation}
+
+\begin{itemize}
+\item Initial computation of index structures is parallelized using \texttt{laparallel:plet}: vertex, edge and face indices as well as boundary edge index
+\item When two patches are merged, new patch links with the new patch at one end of the link are generated which requires expensive computations for evaluating the validity of the merged patch as well as computing the convex hull of the merged.
+\end{itemize}
+
 \subsection{Merge Tolerance}
@@
 \section{Results}\label{results}
+
+Illustration of some generic results
+
+Illustration of ``tolerance''?
+
+Run times for some input meshes?
 

### 2025-01-16 els2025-sbcl-nx/paper.tex
Start.

\documentclass[format=sigconf]{acmart}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\usepackage{enumitem}
\usepackage{float}
\usepackage[labelfont=bf,textfont=md]{caption}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{minted}
\usepackage{hyperref}
\usepackage[parfill]{parskip}
\usepackage[all]{hypcap}
\usemintedstyle[common-lisp]{default}
\newmintinline[code]{text}{}
\bibliographystyle{plainnat}

\hypersetup{
  colorlinks,
  linkcolor={red!50!black},
  citecolor={blue!50!black},
  urlcolor={blue!80!black}
}

\newlist{step}{enumerate}{10}
\setlist[step]{label*=\arabic*.,leftmargin=2em}

\acmConference[ELS’25]{the 18th European Lisp Symposium}
{May 19--20 2025}{Zürich, Switzerland}
\acmDOI{}
\setcopyright{rightsretained}
\copyrightyear{2025}

\begin{document}

\title{Porting the Steel Bank Common Lisp Compiler and Runtime to the Nintendo Switch NX Platform}

\author{Charles Zhang}
\email{charleszhang99@yahoo.com}
\author{Yukari ``Shinmera'' Hafner}
\email{shinmera@tymoon.eu}
\affiliation{%
  \institution{Shirakumo.org}
  \city{Zürich}
  \country{Switzerland}
}

\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10011007.10011006.10011041.10011048</concept_id>
       <concept_desc>Software and its engineering~Runtime environments</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10011007.10011006.10011041.10011045</concept_id>
       <concept_desc>Software and its engineering~Dynamic compilers</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10011007.10010940.10010941.10010949.10010950.10010954</concept_id>
       <concept_desc>Software and its engineering~Garbage collection</concept_desc>
       <concept_significance>300</concept_significance>
       </concept>
   <concept>
       <concept_id>10011007.10011074</concept_id>
       <concept_desc>Software and its engineering~Software creation and management</concept_desc>
       <concept_significance>100</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Runtime environments}
\ccsdesc[500]{Software and its engineering~Dynamic compilers}
\ccsdesc[300]{Software and its engineering~Garbage collection}
\ccsdesc[100]{Software and its engineering~Software creation and management}

\begin{abstract}
  The Nintendo Switch (NX) is a 64-bit ARM-based platform for video games with a proprietary micro-kernel operating system. Notably this system does not give programs the ability to mark pages as executable or give access to thread signal handlers, both of which present a significant hurdle to SBCL's intended bootstrap process and runtime operation. We present our efforts to adapt the SBCL runtime and compiler to deploy applications onto the NX platform.
\end{abstract}

\keywords{Common Lisp, SBCL, porting, ARM, aarch64, NX, Experience Report}

\maketitle

\def\abovecaptionskip{1pt}
\def\listingautorefname{Listing}
\def\figureautorefname{Figure}

\section{Introduction}\label{introduction}

\section{Related Work}\label{relatedwork}
Rhodes\cite{rhodes2008sbcl} outlines the methodology behind the SBCL bootstrapping process.

A pure Common Lisp bootstrapping process as Durand et al.\cite{durand2019bootstrapping} outline would however not notably improve our process, as all our challenges arise from not being able to bootstrap on the desired target platform in the first place, and needing to handle the low-level system construction processes to be amenable for the NX' restrictions.

Citing information on the NX' operating system is difficult as it is a closed-source platform with all usual information placed under NDA. All publicly available information is from security research such as by Roussel-Tarbouriech et al.\cite{roussel2019methodically}.

Particularly, we are unaware of any publication about the porting of other runtime environments to the NX, such as C\#, JavaScript, Lua, etc.

\section{Build System}\label{build}

\section{Relocation}\label{relocation}

\section{Garbage Collection}\label{gc}

\section{Conclusion}\label{conclusion}

\section{Further Work}\label{further-work}

\section{Acknowledgements}\label{acknowledgements}
We would like to thank Douglas Katzman for his help and advice for various parts of the porting effort, as well as the rest of the SBCL maintenance team for their continuous improvements to the SBCL platform.

\bibliography{paper}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-command-extra-options: "-shell-escape"
%%% TeX-master: t
%%% TeX-engine: luatex
%%% End:

### 2025-01-16 els2025-sbcl-nx/paper.tex
Intro

@@
 \section{Introduction}\label{introduction}
+The Nintendo Switch (codename NX) is a handheld video game console based on the ARM 4 Cortex-A57 64-bit architecture\cite{morgan2020}.
+It features a proprietary micro-kernel operating system called the ``Nintendo Switch system software''\cite{roussel2019methodically} (NX OS), and normally only runs software that has been licensed, approved, and digitally signed and encrypted by Nintendo.
+
+Developing licensed software for the NX has to be done via Nintendo's own proprietary SDK, which they only distribute under a non-disclosure agreement (NDA).
+An open-source third-party alternative to the SDK is available\cite{switchbrew}, but this cannot be used for licensed software.
+
+The OS-provided runtime environment is only directly suitable for C and C++ software.
+Other game engines such as Unity, Unreal, and Godot do provide exporting functionality to the NX, meaning that ports for the runtime environments they rely on such as C\#, Lua, and GDScript have been developed, but remain closed-source.
+
+A port of a C-based runtime for Common Lisp such as ECL\cite{attardi1994embeddable} is conceivable.
+However, due to the high performance requirements of video games and the relatively low-power platform of the NX, we deemed it unlikely for such a port to lead to a useful platform.
+Besides, relying on a C runtime would not eliminate all the difficulties we've encountered in porting SBCL anyway.
+
+Particularly, every known Common Lisp implementation compiles a part of its runtime from Lisp source code incrementally into itself, and all user-code is also incrementally compiled into the Lisp image, as opposed to being batch-compiled and linked like other languages such as C.
+We cannot incrementally compile code on the NX due to platform security restrictions.
+The consequence is that all implementation and all user code needs to be compiled ahead of time.
+We discuss the intricacies of this in \autoref{build} and \autoref{relocation}.
+
+The NX OS further does not provide user signal handlers.
+This is a problem for the Garbage Collector (GC), as SBCL relies on inter-thread signalling to park threads during garbage collection.
+While SBCL does provide a safepoints mechanism for GC that is used on Windows, this mechanism had not been tested on any other platform, and as-is still did not meet the requirements of this locked-down platform.
+We discuss the GC in detail in \autoref{gc}.
+
+While we can now compile and deploy complex Common Lisp applications to the NX, some parts of the runtime remain unsupported, and we discuss our future efforts in this regard in \autoref{further-work}.
 
@@
 Citing information on the NX' operating system is difficult as it is a closed-source platform with all usual information placed under NDA.
-All publicly available information is from security research such as by Roussel-Tarbouriech et al.\cite{roussel2019methodically}.
+All publicly available information is from security research such as by Roussel-Tarbouriech et al.\cite{roussel2019methodically} and reverse-engineering\cite{switchbrew}.
 

### 2025-01-16 els2025-sbcl-nx/paper.tex
Paper wooork.

@@
 \section{Build System}\label{build}
+The usual SBCL build process proceeds in several distinct phases, some of which need to be run on a ``host system'' and others on the ``target system''.
+This does allow for limited cross-compilation, wherein the ``host system'' steps can be run on an operating system and platform that is not that of the target we are trying to compile for.
+However, the ``target system'' steps cannot currently be performed on a host system, and full incremental code execution is required on the target.
+As mentioned in the Introduction, we cannot perform incremental code execution on the NX, as the NX OS does not allow us to map new executable pages.
+
+The basic idea of our solution to this issue is to replace the NX as \textit{initial} target with an ARM64-based Linux.
+Once everything, including user code, has been compiled on this Linux target, we extract all Lisp code and data out using a process called \textit{shrinkwrapping} and link it together with the SBCL C runtime as compiled for the NX.
+This process ultimately results in a fully static ELF executable that does not perform any dynamic memory mapping or compilation.
+
+Our build still functions largely the same as the one outlined by Rhodes\cite{rhodes2008sbcl}, though being run roughly twice with special configurations to accomplish the hybrid build.
+
+\begin{enumerate}
+\item \code{build-config} (NX) \\
+  This step gathers whatever build configuration options for the target and spits them out into a readable format for the rest of the build process.
+We run this on some host system (which may not be our ARM64 Linux intermediary), using a special flag to indicate that we're building for the NX.
+We also enable the fasteval contrib, which we need to step in for any place where we would usually invoke the compiler at runtime.
+\item \code{make-host-1} (NX) \\
+  Next we build the cross-compiler with the host Lisp compiler, and at the same time emit C header files describing Lisp object layouts in memory as C structs for the next step.
+\item \code{make-target-1} (NX) \\
+  Now we use the C compiler the Nintendo SDK provides for us, which can cross-compile the SBCL C runtime for the NX.
+We had to make adjustments to the C runtime bits, as the NX OS is not POSIX compliant and lacks a few features the SBCL C runtime usually takes advantage of.
+The SBCL C runtime contains the GC and OS glue bits that the Lisp code needs.
+\item \code{build-config} (Linux$\sim$NX) \\
+  We now create an almost-normal ARM64 Linux system with the same feature set as for the NX.
+This involves the usual steps as before, though with a special flag to inform some parts of the Lisp process that we're going to ultimately target the NX.
+\item \code{make-host-1} (Linux$\sim$NX)
+\item \code{make-target-1} (Linux$\sim$NX)
+\item \code{make-host-2} (Linux$\sim$NX)
+  With the target runtime built, we build the target Lisp system (compiler and the standard library) using the Lisp cross-compiler built by the Lisp host compiler in \code{make-host-1} (Linux).
+This step produces a "cold core" that the runtime can jump into, and can be done purely on the host machine.
+This cold core is not complete, and needs to be executed on the target machine with the target runtime to finish bootstrapping, notably to initialise the object system, which requires runtime compilation.
+\item \code{make-target-2} (Linux$\sim$NX) \\
+  The cold core produced in the last step is loaded into the target runtime, and finishes the bootstrapping procedure to compile and load the rest of the Lisp system.
+After the Lisp system is loaded into memory, the memory is dumped out into a "warm core", which can be loaded back into memory in a new process with the target runtime.
+From this point on, new code can be loaded and images can be dumped at will.
+\item \code{user} \\
+  For user code we now perform some tricks to make it think it's running on the NX, rather than on Linux.
+In particular we modify \code{*features*} to include \code{:nx} and not \code{:linux}, \code{:unix}, or \code{:posix}.
+Once that is set up and ASDF has been neutered, we can compile our program "as usual" and at the end dump out a new core.
+\item \code{shrinkwrap} \\
+  Once all our code has been loaded and a final core has been produced, we ``shrinkwrap'' the image to produce assembly files.
+The details of this are outlined in \autoref{relocation}.
+These assembly files can then be linked together with the SBCL C runtime compiled for the NX in step (3) with the Nintendo SDK toolchain, producing a final ELF executable.
+\item \code{package} \\
+  The final step is to run all the other SDK toolchain bits to produce a signed application package, which can be deployed to a Nintendo Switch development kit to be run.
+\end{enumerate}
+
+A notable extra wrinkle in this procedure is that the SDK is only available for Windows.
+To accommodate this, the custom build system we developed can be run either directly from Windows using another ARM machine remotely to run the Linux bits, or it can be run from a Linux system with the ARM bits being run either locally or remotely and the Windows bits being run either remotely or through Wine\cite{amstadt1994wine}.
 

### 2025-01-17 els2025-sbcl-nx/paper.tex
Some more paper work

@@
 
+Patton\cite{patton2023parallel} and Schwartz\cite{schwartz2018dynamic} describe some details about SBCL's garbage collector which we modify for our port.
+
 \section{Build System}\label{build}
@@
 
+\colorbox{red}{
+  \textbf{TODO CHARLES}: Add some details here.
+}
+
 \section{Garbage Collection}\label{gc}
+SBCL' default garbage collector, ``gencgc'', is a stop-the-world collector.
+In order to start collecting, it needs to ``park'' all threads in a safe location where they won't access any memory that the collector might move or change during collection.
+
+On Unix systems signals are used to park the threads.
+By sending a signal to every other thread, each thread will enter a signal handler, which can then park the thread.
+This is advantageous, since it leverages an operating system mechanism to interrupt thread execution and frees the thread from checking whether it should park.
+
+On Windows no equivalent mechanism is available, so a technique called ``safepoints'' is employed instead.
+The compiler injects code at strategic locations, such as at function call boundaries and loop returns, which checks whether the thread should park explicitly.
+This strategy is notably slower than using signals, since the safepoints are not free, and is finicky, as the compiler could inject too many safepoints (slowing execution of user code significantly) or too few safepoints (making threads take too long to park).
+
+In order to make the safepoint test as small and cheap as possible SBCL does employ another trick on Windows: each safepoint simply writes to a memory region.
+When GC is triggered, this region is marked as protected, causing any thread that tries to write to it to trap, which SBCL can intercept to park the thread instead.
+
+On the NX we cannot install trap handlers, so the current safepoint strategy fails.
+
+\colorbox{red}{
+  \textbf{TODO CHARLES}: Add some details here.
+}
 
@@
 \section{Further Work}\label{further-work}
+Currently we rely on the fasteval system to circumvent runtime compilation restrictions.
+This is especially vital for CLOS, since the discriminating function is only compiled on first call of a generic function.
+However, since we know that we don't dynamically add or remove methods, we should be able to pre-compile all of these functions as well.
+We'd like to add such a CLOS freeze step to the build pipeline.
+
+Callbacks from C to Lisp are currently left unimplemented.
+Callbacks in SBCL are implemented in part as hand-written assembly stubs, and are very hard to maintain.
+The current stubs for ARM64 do not work correctly in the presence of the shrinkwrapping steps that we make use of.
+
+Finally there are a number of improvements we've made that we would like to upstream to lessen the maintenance burden.
+We've already contributed a bunch of the changes back upstream, but a lot of it is also tied to the proprietary SDK and cannot be open-sourced due to the NDA, and some other changes are too contentious for the rest of the SBCL team to want to upstream.
 

### 2025-01-17 els2025-sbcl-nx/paper.tex
Add a photo

@@
 \section{Conclusion}\label{conclusion}
+We have demonstrated that it is possible to port SBCL even to a very restrictive platform that is hostile towards dynamic runtimes such as used for Lisp.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{photo.png}
+  \caption{A photo of a demo running in the Trial Common Lisp game engine on the Nintendo Switch hardware.}
+\end{figure}
+
+Using a substitute build host that is similar enough to the target platform we can compile all code ahead of time, then substitute the base runtime and rewrite the resulting code in order to create an executable suitable for the target platform.
+
+We have also managed to update the garbage collector to work on operating systems with more restrictive feature sets than are found on a typical desktop operating system.
 

### 2025-01-31 els2025-sbcl-nx/paper.tex
Edit build section and add section on relocation.

@@
   The Nintendo Switch (NX) is a 64-bit ARM-based platform for video games with a proprietary micro-kernel operating system.
-Notably this system does not give programs the ability to mark pages as executable or give access to thread signal handlers, both of which present a significant hurdle to SBCL's intended bootstrap process and runtime operation.
+Notably this system does not give programs the ability to mark pages as executable at load time or give access to thread signal handlers, both of which present a significant hurdle to SBCL's intended bootstrap process and runtime operation.
 We present our efforts to adapt the SBCL runtime and compiler to deploy applications onto the NX platform.
@@
 
-A port of a C-based runtime for Common Lisp such as ECL\cite{attardi1994embeddable} is conceivable.
-However, due to the high performance requirements of video games and the relatively low-power platform of the NX, we deemed it unlikely for such a port to lead to a useful platform.
-Besides, relying on a C runtime would not eliminate all the difficulties we've encountered in porting SBCL anyway.
+Many high-performance native-code Common Lisp implementations operate by compiling and loading user code at runtime into the same process (the state of which is traditionally termed an `image' in the Lisp world).
+These implementations typically provide a way to dump an image in a way where the runtime of a new process can load it.
+This contrasts with other languages such as C where code is compiled and linked `offline' into an executable whose entire code and runtime gets loaded into its own process by the operating system.
+We cannot compile and load new code at runtime after an executable is loaded by the operating system on the NX due to platform security restrictions.
+Because large portions of the system such as the compiler of such image-based Common Lisp implementations are also written in Common Lisp, this means that the implementation must be bootstrapped and all user code compiled ahead of time off of the NX, before the compiled Lisp code in the image can then somehow be linked into a runtime built for the NX by the operating system, rather than relying on the runtime to load the image.
+We discuss the intricacies of this for an image-based implementation such as SBCL in \autoref{build} and \autoref{relocation}.
 
-Particularly, every known Common Lisp implementation compiles a part of its runtime from Lisp source code incrementally into itself, and all user-code is also incrementally compiled into the Lisp image, as opposed to being batch-compiled and linked like other languages such as C.
-We cannot incrementally compile code on the NX due to platform security restrictions.
-The consequence is that all implementation and all user code needs to be compiled ahead of time.
-We discuss the intricacies of this in \autoref{build} and \autoref{relocation}.
+A port of a Common Lisp implementation such as ECL\cite{attardi1994embeddable}, where Lisp code is compiled to C and easily linked into a C runtime, is conceivable and presents far fewer challenges for building applications for the NX than with an image-based implementation.
+However, due to the high performance requirements of video games and the relatively low-power platform of the NX, we decided to try and port SBCL, since SBCL produces much faster code than ECL.
+In addition, some of the obstacles presented by the NX apply in general to how most Common Lisp implementations function.
 
-The NX OS further does not provide user signal handlers.
+For example, the NX OS does not provide user signal handlers.
 This is a problem for the Garbage Collector (GC), as SBCL relies on inter-thread signalling to park threads during garbage collection.
-While SBCL does provide a safepoints mechanism for GC that is used on Windows, this mechanism had not been tested on any other platform, and as-is still did not meet the requirements of this locked-down platform.
+While SBCL does provide a safepoints mechanism for GC that is used on Windows, this mechanism is not well tested on other platforms, and as-is still did not meet the requirements of this locked-down platform.
 We discuss the GC in detail in \autoref{gc}.
@@
 
-A pure Common Lisp bootstrapping process as Durand et al.\cite{durand2019bootstrapping} outline would however not notably improve our process, as all our challenges arise from not being able to bootstrap on the desired target platform in the first place, and needing to handle the low-level system construction processes to be amenable for the NX' restrictions.
+A Common Lisp bootstrapping process as Durand et al.\cite{durand2019bootstrapping} outline where the whole target image is created on the host would work to avoid requiring compiling and executing code at runtime on the NX, however, there is not a complete implementation of this technique yet.
+User libraries which require calling foreign functions at compile time dependent on the target architecture also cannot be handled with this technique.
 
@@
 This does allow for limited cross-compilation, wherein the ``host system'' steps can be run on an operating system and platform that is not that of the target we are trying to compile for.
-However, the ``target system'' steps cannot currently be performed on a host system, and full incremental code execution is required on the target.
-As mentioned in the Introduction, we cannot perform incremental code execution on the NX, as the NX OS does not allow us to map new executable pages.
+However, the ``target system'' steps are supposed to run on the target architecture, and involve compiling, loading, and executing new system code in an existing image.
+As mentioned in the Introduction, we cannot load and execute code in the same process that compiled the code on the NX, as the NX OS does not allow us to map new executable pages.
 
@@
 \section{Relocation}\label{relocation}
+Lisp objects contain absolute pointers, so that the dumped image, which contains the memory state of the heap and other spaces, is filled with absolute pointers.
+Traditionally, image-based Lisp implementations such as SBCL map its address spaces to fixed addresses for two reasons: reloading an image into a process is easier this way, since objects can be mapped back in the new image directly to where they were in the process that dumped the image.
+In addition, some architecture backends optimize code generation by hardwiring the address of certain Lisp objects directly into the machine code.
 
-\colorbox{red}{
-  \textbf{TODO CHARLES}: Add some details here.
-}
+Because the NX operating system enforces address space layout randomization, the Lisp runtime cannot rely on being able to map its address spaces to fixed address spaces.
+Hence, the runtime must relocate all objects present in the dumped image to the address spaces placed randomly by the operating space.
+However, code objects present a special problem on the NX.
+Code objects in SBCL contain both executable machine code and absolute pointers to Lisp objects residing in the heap.
+However, executable pages are not writable and must be mapped in by the operating system from the ELF core.
+Therefore, the code and data need to be separated, at first so that the runtime can relocate the absolute pointers to the Lisp objects, and later so that the moving garbage collector is able to fix up those references.
+This requires the shrinkwrapping procedure to separate the data from the code and to rewrite the code in the image offline so that the references to the pointers are into a dislocated read/write space instead of in the code object header, which is on an executable page and hence not writable on the NX.
+The code objects are put into code sections in the executable so that the operating system can load the code and then the runtime can later fix up references to these code objects in non-code objects residing in read/write space.
 

### 2025-02-11 els2025-sbcl-nx/paper.tex
Random edits and add garbage collection section.

@@
   The Nintendo Switch (NX) is a 64-bit ARM-based platform for video games with a proprietary micro-kernel operating system.
-Notably this system does not give programs the ability to mark pages as executable at load time or give access to thread signal handlers, both of which present a significant hurdle to SBCL's intended bootstrap process and runtime operation.
+Notably this system does not give programs the ability to mark pages as executable at run time or expose access to thread signal handlers, both of which present a significant hurdle to SBCL's intended bootstrap process and runtime operation.
 We present our efforts to adapt the SBCL runtime and compiler to deploy applications onto the NX platform.
@@
 
-A Common Lisp bootstrapping process as Durand et al.\cite{durand2019bootstrapping} outline where the whole target image is created on the host would work to avoid requiring compiling and executing code at runtime on the NX, however, there is not a complete implementation of this technique yet.
-User libraries which require calling foreign functions at compile time dependent on the target architecture also cannot be handled with this technique.
+A Common Lisp bootstrapping process as Durand et al.\cite{durand2019bootstrapping} outline where the whole target image is created on the host would work to avoid requiring compiling and executing code at runtime on the NX; however, there is not a complete implementation of this technique yet.
+User libraries which require querying foreign functions at compile time where the details of target architecture matter also cannot be handled with this technique.
 
-Citing information on the NX' operating system is difficult as it is a closed-source platform with all usual information placed under NDA.
+Citing information on the NX's operating system is difficult as it is a closed-source platform with all usual information placed under NDA.
 All publicly available information is from security research such as by Roussel-Tarbouriech et al.\cite{roussel2019methodically} and reverse-engineering\cite{switchbrew}.
@@
 Lisp objects contain absolute pointers, so that the dumped image, which contains the memory state of the heap and other spaces, is filled with absolute pointers.
-Traditionally, image-based Lisp implementations such as SBCL map its address spaces to fixed addresses for two reasons: reloading an image into a process is easier this way, since objects can be mapped back in the new image directly to where they were in the process that dumped the image.
-In addition, some architecture backends optimize code generation by hardwiring the address of certain Lisp objects directly into the machine code.
+Traditionally, image-based Lisp implementations such as SBCL map their address spaces to fixed addresses for two reasons: First, reloading an image into a process is easier this way, since objects can be mapped back in the new image directly to where they were in the process that dumped the image.
+Second, some architecture backends optimize code generation by hardwiring the address of certain Lisp objects directly into the machine code.
 
@@
 Therefore, the code and data need to be separated, at first so that the runtime can relocate the absolute pointers to the Lisp objects, and later so that the moving garbage collector is able to fix up those references.
-This requires the shrinkwrapping procedure to separate the data from the code and to rewrite the code in the image offline so that the references to the pointers are into a dislocated read/write space instead of in the code object header, which is on an executable page and hence not writable on the NX.
+This requires the shrinkwrapping procedure to separate the data from the code and to rewrite the code in the image offline so that any references to the pointers are into a dislocated read/write space instead of into the code object header, which is on an executable page and hence not writable on the NX.
 The code objects are put into code sections in the executable so that the operating system can load the code and then the runtime can later fix up references to these code objects in non-code objects residing in read/write space.
@@
 \section{Garbage Collection}\label{gc}
-SBCL' default garbage collector, ``gencgc'', is a stop-the-world collector.
-In order to start collecting, it needs to ``park'' all threads in a safe location where they won't access any memory that the collector might move or change during collection.
+SBCL's default garbage collector, ``gencgc'', is a stop-the-world collector, meaning it must stop or ``park'' other threads in the process.
+This is necessary so they don't access any memory that the collector might move or change during collection.
+On RISC architectures like the ARM the Nintendo Switch runs on, the collector is also precise and must be able to identify which storage locations contain Lisp objects, so it is important that the threads are parked at code locations where the collector can safely do so.
 
-On Unix systems signals are used to park the threads.
+In the discussion that follows, a \textbf{Lisp thread} is a native thread registered to the Lisp runtime that can either be executing Lisp or foreign code.
+A native threads created by Lisp are immediately registered to the runtime.
+A native thread created by foreign code that calls into Lisp via e.g.
+a callback gets a Lisp thread associated with it by the runtime when it starts executing Lisp code for the first time.
+
+On Unix systems, the POSIX signal mechanism is used by default to park all Lisp threads besides the thread executing GC.
 By sending a signal to every other thread, each thread will enter a signal handler, which can then park the thread.
 This is advantageous, since it leverages an operating system mechanism to interrupt thread execution and frees the thread from checking whether it should park.
+The garbage collector can also use the signal context mechanism to read and process the values of all registers and the Lisp stack at the point where the thread stopped.
+On the flip side, Lisp code must be generated in such a way that the GC almost always knows how to parse Lisp objects from registers or the stack at any code location.
+However, certain instruction sequences in the generated code then need to behave atomically in the sense that it is not safe to stop for GC at those locations, so extra bookkeeping is done to defer interrupts in those sequences.
+Lisp threads in foreign code are yielded and restarted just as Lisp threads are in Lisp code, but foreign code is assumed to not access Lisp data and hence those threads ignored during collection.
 
-On Windows no equivalent mechanism is available, so a technique called ``safepoints'' is employed instead.
-The compiler injects code at strategic locations, such as at function call boundaries and loop returns, which checks whether the thread should park explicitly.
-This strategy is notably slower than using signals, since the safepoints are not free, and is finicky, as the compiler could inject too many safepoints (slowing execution of user code significantly) or too few safepoints (making threads take too long to park).
+On Windows, no equivalent signal mechanism is available, so a strategy using safepoints must be employed instead.
+A \textbf{safepoint} is a location in code which is known to be GC-safe, so that collector has enough information at that location to correctly function.
+The Lisp compiler then injects code (here termed \textbf{yieldpoint} code) at these safepoints, such as at function call boundaries and loop returns, which causes the thread to check whether it should yield for GC.
+Yieldpoint code should be inserted strategically so that it doesn't take too long for threads to park after a given thread decides to collect garbage, but also so that the overhead of checking whether to park the thread is as low as possible.
+Lisp threads in foreign code do not encounter yieldpoint code generated by the GC, but also do not need to be parked anyway, as foreign code is assumed to not access Lisp data and hence the thread is ignored during collection.
+However, since the thread is not stopped, it is possible that the thread may re-enter Lisp during a collection, in which case code for re-entry into Lisp must check if GC is running and park the thread and wait for GC to complete.
+A similar consideration applies for Lisp threads in Lisp which enter foreign code while a GCing thread is waiting for all Lisp threads in Lisp code to stop; the Lisp thread must communicate to the GC that it is entering foreign code and hence the GCing thread need not wait for it to yield anymore, since the Lisp thread is about to start executing foreign code.
+For the purposes of this paper, we will call these points of communication at foreign function call boundaries \textbf{foreign yieldpoints}.
 
-In order to make the safepoint test as small and cheap as possible SBCL does employ another trick on Windows: each safepoint simply writes to a memory region.
-When GC is triggered, this region is marked as protected, causing any thread that tries to write to it to trap, which SBCL can intercept to park the thread instead.
+To summarize, the safepoint and signal mechanisms can be seen as inverses of each other: whereas code generation for the signal mechanism must ensure that the majority of code locations are safe for GC, the safepoint mechanism only requires that strategically chosen locations are safe for GC.
+The safepoint mechanism also allows foreign code to run without stopping at the cost of extra communication at foreign call boundaries, while the signal mechanism stops all Lisp threads regardless of whether they are executing foreign code or not.
 
-On the NX we cannot install trap handlers, so the current safepoint strategy fails.
+In the existing safepoint ports, SBCL exploits hardware memory protection facilities provided by the operating system in order to make generated yieldpoint code as small as possible: GC toggles the read and write permission bits of pages in virtual memory in order to communicate to the thread various GC state transitions.
+Non-foreign yieldpoint code then consists of a single read instruction from a global page that a GC marks as non-readable when it wants all Lisp threads to park.
+The resulting page fault traps into trap handler code that yields the thread and progresses the state of the GC.
+Foreign yieldpoint code consists of a write instruction onto a thread-local page informing the GC whether the Lisp thread is entering or leaving foreign code; the GC is then able to selectively cause the relevant threads to trap and wait for the state of the GC to progress or allow the relevant threads to continue toggling the write permissions of the thread-local page accordingly.
+One other advantage of using page permissions is that it allows race-free examine + wait + use in the latter scenario where the `entry into/return from foreign code' flag word in memory needs to be read and appropriately acted on.
 
-\colorbox{red}{
-  \textbf{TODO CHARLES}: Add some details here.
-}
+\subsection{Challenges for the NX}
+As the NX also does not expose a POSIX signal mechanism, we chose to use the safepoint strategy as a starting point.
+However, as mentioned previously, even the safepoint strategy relies on hardware memory protection facilities to aid intra-thread communication between the GCing thread and other threads.
+The NX does not provide ways to install trap handlers for memory faults in release mode, so a different mechanism had to be devised.
+
+For non-foreign yieldpoint code, it sufficed to replace the trapping read instruction with a polling instruction sequence which simply checks if a global `stop now' word is true and if so, branches into a trampoline which calls into the trap handling code.
+Since there is no hardware trap and associated register context under this scheme, the trampoline also serves to spill all registers onto the Lisp stack so that the garbage collector correctly scavenges and fixes up the roots contained in the registers at the safepoint.
+
+For foreign yieldpoint code, we replace the page permission scheme by designating an additional `permission' word on the thread local page in addition to the `entry into/return from foreign code' flag word to emulate the effect of the page permission scheme: the additional word flags whether write access to the other word is allowed.
+Since the thread containing the page and the collector thread may both race to access the two words in memory, the code sequence examining the permission word and dispatching to the correct GC state handler based on the value of the flag word requires explicit synchronization primitives.
+For now, we use locks to ensure race-free communication between the collector thread and any threads which are entering or exiting foreign code.
+
+In addition to the existing safepoint mechanism requiring hardware page protection facilities, the garbage collector also used certain virtual memory tricks to quickly zero out and/or return large chunks of memory back to the operating system.
+As the NX does not expose such sophisticated virtual memory management to the application developer, we had to rely on slower and more portable means of achieving the same effect.
 

### 2025-02-11 els2025-sbcl-nx/paper.tex
Add subsections, edits, and a section on callbacks.
Since we recently managed to implement callbacks.

@@
 
+\subsection{Address Space Layout Randomization}
 Because the NX operating system enforces address space layout randomization, the Lisp runtime cannot rely on being able to map its address spaces to fixed address spaces.
@@
 The code objects are put into code sections in the executable so that the operating system can load the code and then the runtime can later fix up references to these code objects in non-code objects residing in read/write space.
+
+\subsection{Linkage between Lisp and foreign code}
+Another part of the system impacted by the need for relocation is linkage between Lisp and foreign code.
+SBCL uses a linkage table that indirects calls to foreign code from Lisp so that foreign code can be loaded by the operating system at any address on startup and resolved in the linkage table.
+However, for callbacks into Lisp from foreign code, Lisp must create function pointers to snippets of assembly for callback entry allocated in static (i.e.
+never moved by GC) Lisp space.
+These pointers can then be passed to and called from foreign code.
+To ensure that such function pointers in a saved image are valid even after the Lisp spaces containing the callback entry points are relocated, the function pointers must somehow be fixed up so they are valid after startup.
+We managed this by having the offline shrinkwrapping process move the callback entry code into a text ELF section while recording fixup information, so that the operating system can load the callback entry code into executable pages, whereby the Lisp runtime at startup can rewrite the static function pointers to point to where the code was loaded using the recorded fixup information.
 
@@
 However, since the thread is not stopped, it is possible that the thread may re-enter Lisp during a collection, in which case code for re-entry into Lisp must check if GC is running and park the thread and wait for GC to complete.
-A similar consideration applies for Lisp threads in Lisp which enter foreign code while a GCing thread is waiting for all Lisp threads in Lisp code to stop; the Lisp thread must communicate to the GC that it is entering foreign code and hence the GCing thread need not wait for it to yield anymore, since the Lisp thread is about to start executing foreign code.
+A similar consideration applies for Lisp threads in Lisp which enter foreign code while the collecting thread is waiting for all Lisp threads in Lisp code to stop; the Lisp thread must communicate to the GC that it is entering foreign code and hence the collecting thread need not wait for it to yield anymore, since the Lisp thread is about to start executing foreign code.
 For the purposes of this paper, we will call these points of communication at foreign function call boundaries \textbf{foreign yieldpoints}.
@@
 As the NX also does not expose a POSIX signal mechanism, we chose to use the safepoint strategy as a starting point.
-However, as mentioned previously, even the safepoint strategy relies on hardware memory protection facilities to aid intra-thread communication between the GCing thread and other threads.
+However, as mentioned previously, even the safepoint strategy relies on hardware memory protection facilities to aid intra-thread communication between the collecting thread and other threads.
 The NX does not provide ways to install trap handlers for memory faults in release mode, so a different mechanism had to be devised.
@@
 
-Callbacks from C to Lisp are currently left unimplemented.
-Callbacks in SBCL are implemented in part as hand-written assembly stubs, and are very hard to maintain.
-The current stubs for ARM64 do not work correctly in the presence of the shrinkwrapping steps that we make use of.
-
 Finally there are a number of improvements we've made that we would like to upstream to lessen the maintenance burden.

### 2025-02-11 els2025-sbcl-nx/paper.tex
Fix inaccuracy.
I'm still not entirely sure this is what happens but it makes sense
this way.

@@
 However, certain instruction sequences in the generated code then need to behave atomically in the sense that it is not safe to stop for GC at those locations, so extra bookkeeping is done to defer interrupts in those sequences.
-Lisp threads in foreign code are yielded and restarted just as Lisp threads are in Lisp code, but foreign code is assumed to not access Lisp data and hence those threads ignored during collection.
+Lisp threads in foreign code are stopped and restarted just as Lisp threads are in Lisp code, though only the contents of its stack are scavenged.
 
@@
 Yieldpoint code should be inserted strategically so that it doesn't take too long for threads to park after a given thread decides to collect garbage, but also so that the overhead of checking whether to park the thread is as low as possible.
-Lisp threads in foreign code do not encounter yieldpoint code generated by the GC, but also do not need to be parked anyway, as foreign code is assumed to not access Lisp data and hence the thread is ignored during collection.
-However, since the thread is not stopped, it is possible that the thread may re-enter Lisp during a collection, in which case code for re-entry into Lisp must check if GC is running and park the thread and wait for GC to complete.
+Lisp threads in foreign code do not encounter yieldpoint code generated by the GC, but also do not need to be parked anyway, as foreign code is assumed to not access Lisp data.
+Its Lisp stack can still be safely scavenged while the thread is executing foreign code.
+However, since the thread is not stopped, it is possible that the thread may re-enter Lisp during a collection, in which case code for re-entry into Lisp must check if GC is running and park the thread and wait for GC to complete before the thread can execute Lisp code again.
 A similar consideration applies for Lisp threads in Lisp which enter foreign code while the collecting thread is waiting for all Lisp threads in Lisp code to stop; the Lisp thread must communicate to the GC that it is entering foreign code and hence the collecting thread need not wait for it to yield anymore, since the Lisp thread is about to start executing foreign code.

### 2025-02-24 els2025-sbcl-nx/paper.tex
Typo/readability fixes

@@
 
-\keywords{Common Lisp, SBCL, porting, ARM, aarch64, NX, Experience Report}
+\keywords{Common Lisp, SBCL, porting, ARM, aarch64, Nintendo Switch, NX, Experience Report}
 
@@
 We cannot compile and load new code at runtime after an executable is loaded by the operating system on the NX due to platform security restrictions.
-Because large portions of the system such as the compiler of such image-based Common Lisp implementations are also written in Common Lisp, this means that the implementation must be bootstrapped and all user code compiled ahead of time off of the NX, before the compiled Lisp code in the image can then somehow be linked into a runtime built for the NX by the operating system, rather than relying on the runtime to load the image.
+Because large portions of the system such as the compiler of such image-based Common Lisp implementations are also written in Common Lisp, this restriction means that the implementation must be bootstrapped and all user code compiled ahead of time off of the NX, before the compiled Lisp code in the image can then somehow be linked into a runtime built for the NX by the operating system, rather than relying on the implementation runtime to load the image.
 We discuss the intricacies of this for an image-based implementation such as SBCL in \autoref{build} and \autoref{relocation}.
@@
 A port of a Common Lisp implementation such as ECL\cite{attardi1994embeddable}, where Lisp code is compiled to C and easily linked into a C runtime, is conceivable and presents far fewer challenges for building applications for the NX than with an image-based implementation.
-However, due to the high performance requirements of video games and the relatively low-power platform of the NX, we decided to try and port SBCL, since SBCL produces much faster code than ECL.
+However, due to the high performance requirements of video games and the relatively low-power platform of the NX, we decided to try and port SBCL, since SBCL produces much faster code than ECL especially with regards to CLOS.
 In addition, some of the obstacles presented by the NX apply in general to how most Common Lisp implementations function.
@@
 For example, the NX OS does not provide user signal handlers.
-This is a problem for the Garbage Collector (GC), as SBCL relies on inter-thread signalling to park threads during garbage collection.
-While SBCL does provide a safepoints mechanism for GC that is used on Windows, this mechanism is not well tested on other platforms, and as-is still did not meet the requirements of this locked-down platform.
+This lack of signal handlers is a problem for the Garbage Collector (GC), as SBCL relies on inter-thread signalling to park threads during garbage collection.
+While SBCL does provide a safepoints mechanism for GC that is used on Windows which similarly lacks signal handlers, this mechanism is not well tested on other platforms, and as-is still did not meet the requirements of this locked-down platform.
 We discuss the GC in detail in \autoref{gc}.
@@
 \section{Related Work}\label{relatedwork}
-Rhodes\cite{rhodes2008sbcl} outlines the methodology behind the SBCL bootstrapping process.
+Rhodes\cite{rhodes2008sbcl} outlines the methodology behind the general SBCL bootstrapping process, which we extend for our locked down target platform.
 
@@
 
-Citing information on the NX's operating system is difficult as it is a closed-source platform with all usual information placed under NDA.
+Citing information on the NX' operating system is difficult as it is a closed-source platform with all usual information placed under NDA.
 All publicly available information is from security research such as by Roussel-Tarbouriech et al.\cite{roussel2019methodically} and reverse-engineering\cite{switchbrew}.
@@
 The usual SBCL build process proceeds in several distinct phases, some of which need to be run on a ``host system'' and others on the ``target system''.
-This does allow for limited cross-compilation, wherein the ``host system'' steps can be run on an operating system and platform that is not that of the target we are trying to compile for.
+This process does allow for limited cross-compilation, wherein the ``host system'' steps can be run on an operating system and platform that is not that of the target we are trying to compile for.
 However, the ``target system'' steps are supposed to run on the target architecture, and involve compiling, loading, and executing new system code in an existing image.
@@
   We now create an almost-normal ARM64 Linux system with the same feature set as for the NX.
-This involves the usual steps as before, though with a special flag to inform some parts of the Lisp process that we're going to ultimately target the NX.
+This process involves the usual steps as before, though with a special flag to inform some parts of the Lisp process that we're going to ultimately target the NX.
 \item \code{make-host-1} (Linux$\sim$NX)
@@
 \item \code{make-host-2} (Linux$\sim$NX)
-  With the target runtime built, we build the target Lisp system (compiler and the standard library) using the Lisp cross-compiler built by the Lisp host compiler in \code{make-host-1} (Linux).
+  With the target runtime built, we build the target Lisp system (compiler and the standard library) using the Lisp cross-compiler built by the Lisp host compiler in \code{make-host-1} (Linux$\sim$NX).
 This step produces a "cold core" that the runtime can jump into, and can be done purely on the host machine.
@@
 \subsection{Address Space Layout Randomization}
-Because the NX operating system enforces address space layout randomization, the Lisp runtime cannot rely on being able to map its address spaces to fixed address spaces.
-Hence, the runtime must relocate all objects present in the dumped image to the address spaces placed randomly by the operating space.
+Because the NX operating system enforces address space layout randomization, the Lisp runtime cannot rely on being able to map its address spaces to fixed addresses.
+Hence, the runtime must relocate all objects present in the dumped image to the address spaces placed randomly by the operating system.
 However, code objects present a special problem on the NX.
@@
 Therefore, the code and data need to be separated, at first so that the runtime can relocate the absolute pointers to the Lisp objects, and later so that the moving garbage collector is able to fix up those references.
-This requires the shrinkwrapping procedure to separate the data from the code and to rewrite the code in the image offline so that any references to the pointers are into a dislocated read/write space instead of into the code object header, which is on an executable page and hence not writable on the NX.
+This restriction requires the shrinkwrapping procedure to separate the data from the code and to rewrite the code in the image offline so that any references to the pointers are into a dislocated read/write space instead of into the code object header, which is on an executable page and hence not writable on the NX.
 The code objects are put into code sections in the executable so that the operating system can load the code and then the runtime can later fix up references to these code objects in non-code objects residing in read/write space.
@@
 To ensure that such function pointers in a saved image are valid even after the Lisp spaces containing the callback entry points are relocated, the function pointers must somehow be fixed up so they are valid after startup.
-We managed this by having the offline shrinkwrapping process move the callback entry code into a text ELF section while recording fixup information, so that the operating system can load the callback entry code into executable pages, whereby the Lisp runtime at startup can rewrite the static function pointers to point to where the code was loaded using the recorded fixup information.
+We managed to solve this problem by having the offline shrinkwrapping process move the callback entry code into a text ELF section while recording fixup information, so that the operating system can load the callback entry code into executable pages, whereby the Lisp runtime at startup can rewrite the static function pointers to point to where the code was loaded using the recorded fixup information.
 
@@
 SBCL's default garbage collector, ``gencgc'', is a stop-the-world collector, meaning it must stop or ``park'' other threads in the process.
-This is necessary so they don't access any memory that the collector might move or change during collection.
+Doing so is necessary so that the threads don't access any memory that the collector might move or change during collection.
 On RISC architectures like the ARM the Nintendo Switch runs on, the collector is also precise and must be able to identify which storage locations contain Lisp objects, so it is important that the threads are parked at code locations where the collector can safely do so.
 
-In the discussion that follows, a \textbf{Lisp thread} is a native thread registered to the Lisp runtime that can either be executing Lisp or foreign code.
-A native threads created by Lisp are immediately registered to the runtime.
+In the discussion that follows, a \textit{Lisp thread} is a native thread registered to the Lisp runtime that can either be executing Lisp or foreign code.
+Native threads created by Lisp are immediately registered to the runtime.
 A native thread created by foreign code that calls into Lisp via e.g.
@@
 By sending a signal to every other thread, each thread will enter a signal handler, which can then park the thread.
-This is advantageous, since it leverages an operating system mechanism to interrupt thread execution and frees the thread from checking whether it should park.
+This method is advantageous, since it leverages an operating system mechanism to interrupt thread execution and frees the thread from checking whether it should park.
 The garbage collector can also use the signal context mechanism to read and process the values of all registers and the Lisp stack at the point where the thread stopped.
@@
 On Windows, no equivalent signal mechanism is available, so a strategy using safepoints must be employed instead.
-A \textbf{safepoint} is a location in code which is known to be GC-safe, so that collector has enough information at that location to correctly function.
-The Lisp compiler then injects code (here termed \textbf{yieldpoint} code) at these safepoints, such as at function call boundaries and loop returns, which causes the thread to check whether it should yield for GC.
+A \textit{safepoint} is a location in code which is known to be GC-safe, so that collector has enough information at that location to correctly function.
+The Lisp compiler then injects code (here termed \textit{yieldpoint} code) at these safepoints, such as at function call boundaries and loop returns, which causes the thread to check whether it should yield for GC.
 Yieldpoint code should be inserted strategically so that it doesn't take too long for threads to park after a given thread decides to collect garbage, but also so that the overhead of checking whether to park the thread is as low as possible.
@@
 A similar consideration applies for Lisp threads in Lisp which enter foreign code while the collecting thread is waiting for all Lisp threads in Lisp code to stop; the Lisp thread must communicate to the GC that it is entering foreign code and hence the collecting thread need not wait for it to yield anymore, since the Lisp thread is about to start executing foreign code.
-For the purposes of this paper, we will call these points of communication at foreign function call boundaries \textbf{foreign yieldpoints}.
+For the purposes of this paper, we will call these points of communication at foreign function call boundaries \textit{foreign yieldpoints}.
 

### 2025-02-25 els2025-sbcl-nx/paper.tex
Image
Incorporate Robert's feedback

@@
 \begin{abstract}
+  We present our efforts to adapt the SBCL runtime and compiler to deploy applications onto the NX platform.
 The Nintendo Switch (NX) is a 64-bit ARM-based platform for video games with a proprietary micro-kernel operating system.
 Notably this system does not give programs the ability to mark pages as executable at run time or expose access to thread signal handlers, both of which present a significant hurdle to SBCL's intended bootstrap process and runtime operation.
-We present our efforts to adapt the SBCL runtime and compiler to deploy applications onto the NX platform.
+To work around these hurdles we modify SBCL's build system to bootstrap on a Linux ARM system, which is similar enough to the NX to be able to compile user code.
+We then use a technique called ``shrinkwrapping'' to combine the Lisp code and data with the C runtime compiled for the NX to produce a final, fully static ELF executable that can be run on the NX, without the need for runtime code generation.
 \end{abstract}
@@
 
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{switch.jpg}
+  \caption{The Nintendo Switch (NX) handheld game console}
+\end{figure}
+
 \section{Introduction}\label{introduction}
 The Nintendo Switch (codename NX) is a handheld video game console based on the ARM 4 Cortex-A57 64-bit architecture\cite{morgan2020}.
-It features a proprietary micro-kernel operating system called the ``Nintendo Switch system software''\cite{roussel2019methodically} (NX OS), and normally only runs software that has been licensed, approved, and digitally signed and encrypted by Nintendo.
+It features a proprietary micro-kernel operating system called the ``Nintendo Switch system software''\cite{roussel2019methodically} (NX OS), and normally runs only software that has been licensed, approved, and digitally signed and encrypted by Nintendo.
 
-Developing licensed software for the NX has to be done via Nintendo's own proprietary SDK, which they only distribute under a non-disclosure agreement (NDA).
+Developing licensed software for the NX has to be done via Nintendo's own proprietary Software Development Kit (SDK), which they distribute only under a non-disclosure agreement (NDA).
 An open-source third-party alternative to the SDK is available\cite{switchbrew}, but this cannot be used for licensed software.
 
-The OS-provided runtime environment is only directly suitable for C and C++ software.
+The OS-provided runtime environment is directly suitable only for C and C++ software.
 Other game engines such as Unity, Unreal, and Godot do provide exporting functionality to the NX, meaning that ports for the runtime environments they rely on such as C\#, Lua, and GDScript have been developed, but remain closed-source.
@@
 These implementations typically provide a way to dump an image in a way where the runtime of a new process can load it.
-This contrasts with other languages such as C where code is compiled and linked `offline' into an executable whose entire code and runtime gets loaded into its own process by the operating system.
+This implementation technique contrasts with other languages such as C where code is compiled and linked `offline' into an executable whose entire code and runtime gets loaded into its own process by the operating system.
 We cannot compile and load new code at runtime after an executable is loaded by the operating system on the NX due to platform security restrictions.
 Because large portions of the system such as the compiler of such image-based Common Lisp implementations are also written in Common Lisp, this restriction means that the implementation must be bootstrapped and all user code compiled ahead of time off of the NX, before the compiled Lisp code in the image can then somehow be linked into a runtime built for the NX by the operating system, rather than relying on the implementation runtime to load the image.
-We discuss the intricacies of this for an image-based implementation such as SBCL in \autoref{build} and \autoref{relocation}.
+We discuss the intricacies of this problem for an image-based implementation such as SBCL in \autoref{build} and \autoref{relocation}.
 
@@
 
-Citing information on the NX' operating system is difficult as it is a closed-source platform with all usual information placed under NDA.
+Citing information on the operating system of the NX is difficult as it is a closed-source platform with all usual information placed under NDA.
 All publicly available information is from security research such as by Roussel-Tarbouriech et al.\cite{roussel2019methodically} and reverse-engineering\cite{switchbrew}.
@@
 However, the ``target system'' steps are supposed to run on the target architecture, and involve compiling, loading, and executing new system code in an existing image.
-As mentioned in the Introduction, we cannot load and execute code in the same process that compiled the code on the NX, as the NX OS does not allow us to map new executable pages.
+As mentioned in the introduction, we cannot load and execute code in the same process that compiled the code on the NX, as the NX OS does not allow us to map new executable pages.
 
@@
 We run this on some host system (which may not be our ARM64 Linux intermediary), using a special flag to indicate that we're building for the NX.
-We also enable the fasteval contrib, which we need to step in for any place where we would usually invoke the compiler at runtime.
+We also enable the ``fasteval'' contribution, which we need to step in for any place where we would usually invoke the compiler at runtime.
 \item \code{make-host-1} (NX) \\
@@
   Once all our code has been loaded and a final core has been produced, we ``shrinkwrap'' the image to produce assembly files.
-The details of this are outlined in \autoref{relocation}.
+The details of this technique are outlined in \autoref{relocation}.
 These assembly files can then be linked together with the SBCL C runtime compiled for the NX in step (3) with the Nintendo SDK toolchain, producing a final ELF executable.
@@
 
-A notable extra wrinkle in this procedure is that the SDK is only available for Windows.
-To accommodate this, the custom build system we developed can be run either directly from Windows using another ARM machine remotely to run the Linux bits, or it can be run from a Linux system with the ARM bits being run either locally or remotely and the Windows bits being run either remotely or through Wine\cite{amstadt1994wine}.
+A notable extra wrinkle in this procedure is that the SDK is available only for Windows.
+To accommodate this platform restriction, the custom build system we developed can be run either directly from Windows using another ARM machine remotely to run the Linux bits, or it can be run from a Linux system with the ARM bits being run either locally or remotely and the Windows bits being run either remotely or through Wine\cite{amstadt1994wine}.
 
@@
 
-To summarize, the safepoint and signal mechanisms can be seen as inverses of each other: whereas code generation for the signal mechanism must ensure that the majority of code locations are safe for GC, the safepoint mechanism only requires that strategically chosen locations are safe for GC.
+To summarize, the safepoint and signal mechanisms can be seen as inverses of each other: whereas code generation for the signal mechanism must ensure that the majority of code locations are safe for GC, the safepoint mechanism requires only that strategically chosen locations are safe for GC.
 The safepoint mechanism also allows foreign code to run without stopping at the cost of extra communication at foreign call boundaries, while the signal mechanism stops all Lisp threads regardless of whether they are executing foreign code or not.
@@
 The resulting page fault traps into trap handler code that yields the thread and progresses the state of the GC.
-Foreign yieldpoint code consists of a write instruction onto a thread-local page informing the GC whether the Lisp thread is entering or leaving foreign code; the GC is then able to selectively cause the relevant threads to trap and wait for the state of the GC to progress or allow the relevant threads to continue toggling the write permissions of the thread-local page accordingly.
+Foreign yieldpoint code consists of a write instruction onto a thread-local page, informing the GC whether the Lisp thread is entering or leaving foreign code; the GC is then able to selectively cause the relevant threads to trap and wait for the state of the GC to progress or allow the relevant threads to continue toggling the write permissions of the thread-local page accordingly.
 One other advantage of using page permissions is that it allows race-free examine + wait + use in the latter scenario where the `entry into/return from foreign code' flag word in memory needs to be read and appropriately acted on.
@@
 As the NX also does not expose a POSIX signal mechanism, we chose to use the safepoint strategy as a starting point.
-However, as mentioned previously, even the safepoint strategy relies on hardware memory protection facilities to aid intra-thread communication between the collecting thread and other threads.
+However, as mentioned previously, even the safepoint strategy relies on hardware memory protection facilities to aid inter-thread communication between the collecting thread and other threads.
 The NX does not provide ways to install trap handlers for memory faults in release mode, so a different mechanism had to be devised.
@@
 
-In addition to the existing safepoint mechanism requiring hardware page protection facilities, the garbage collector also used certain virtual memory tricks to quickly zero out and/or return large chunks of memory back to the operating system.
+In addition to the existing safepoint mechanism requiring hardware page protection facilities, the garbage collector also used certain virtual-memory tricks to quickly zero out and/or return large chunks of memory back to the operating system.
 As the NX does not expose such sophisticated virtual memory management to the application developer, we had to rely on slower and more portable means of achieving the same effect.
@@
   \includegraphics[width=\linewidth]{photo.png}
-  \caption{A photo of a demo running in the Trial Common Lisp game engine on the Nintendo Switch hardware.}
+  \caption{A demo running in the Trial Common Lisp game engine on the Nintendo Switch development hardware}
 \end{figure}
 
-Using a substitute build host that is similar enough to the target platform we can compile all code ahead of time, then substitute the base runtime and rewrite the resulting code in order to create an executable suitable for the target platform.
+Using a substitute build host that is similar enough to the target platform, we can compile all code ahead of time, then substitute the base runtime and rewrite the resulting code in order to create an executable suitable for the target platform.
 
@@
 \section{Further Work}\label{further-work}
-Currently we rely on the fasteval system to circumvent runtime compilation restrictions.
-This is especially vital for CLOS, since the discriminating function is only compiled on first call of a generic function.
+Currently, we rely on the ``fasteval'' system to circumvent runtime compilation restrictions.
+This is especially vital for CLOS, since the discriminating function is compiled only on first call of a generic function.
 However, since we know that we don't dynamically add or remove methods, we should be able to pre-compile all of these functions as well.
-We'd like to add such a CLOS freeze step to the build pipeline.
+We'd like to add such a CLOS freeze step to the build pipeline, perhaps using a technique similar to ``satiation'' as outlined by Strandh et al.\cite{strandh2014resolving}
 
@@
 We would like to thank Douglas Katzman for his help and advice for various parts of the porting effort, as well as the rest of the SBCL maintenance team for their continuous improvements to the SBCL platform.
+We would also like to thank Robert Strandh and Jan Moringe for their feedback on the draft paper.
 

### 2025-02-25 els2025-sbcl-nx/paper.tex
Incorporate Jan's feedback.

@@
 The OS-provided runtime environment is directly suitable only for C and C++ software.
-Other game engines such as Unity, Unreal, and Godot do provide exporting functionality to the NX, meaning that ports for the runtime environments they rely on such as C\#, Lua, and GDScript have been developed, but remain closed-source.
+Other game engines such as Unity, Unreal, and Godot do provide exporting functionality to the NX, meaning that ports for the runtime environments they rely on such as .NET/Mono (C#), Lua, and GDScript have been developed, but remain closed-source.
 
@@
 These implementations typically provide a way to dump an image in a way where the runtime of a new process can load it.
-This implementation technique contrasts with other languages such as C where code is compiled and linked `offline' into an executable whose entire code and runtime gets loaded into its own process by the operating system.
-We cannot compile and load new code at runtime after an executable is loaded by the operating system on the NX due to platform security restrictions.
+This implementation technique contrasts with ahead-of-time compilation languages such as C where code is compiled and linked `offline' into an executable whose entire code and runtime gets loaded into its own process by the operating system.
+On the NX, the SBCL runtime cannot compile and load new code at runtime after an executable is loaded by the operating system on the NX due to platform security restrictions.
 Because large portions of the system such as the compiler of such image-based Common Lisp implementations are also written in Common Lisp, this restriction means that the implementation must be bootstrapped and all user code compiled ahead of time off of the NX, before the compiled Lisp code in the image can then somehow be linked into a runtime built for the NX by the operating system, rather than relying on the implementation runtime to load the image.
@@
 This lack of signal handlers is a problem for the Garbage Collector (GC), as SBCL relies on inter-thread signalling to park threads during garbage collection.
-While SBCL does provide a safepoints mechanism for GC that is used on Windows which similarly lacks signal handlers, this mechanism is not well tested on other platforms, and as-is still did not meet the requirements of this locked-down platform.
+While SBCL does provide a safepoints mechanism for GC that is used on Windows which similarly lacks signal handlers, this mechanism is not well tested on other platforms, and as-is still did not meet the requirements of the locked-down NX platform.
 We discuss the GC in detail in \autoref{gc}.
@@
 \section{Related Work}\label{relatedwork}
-Rhodes\cite{rhodes2008sbcl} outlines the methodology behind the general SBCL bootstrapping process, which we extend for our locked down target platform.
+Rhodes\cite{rhodes2008sbcl} outlines the methodology behind the general SBCL bootstrapping process, which we extend for our locked-down target platform.
 
@@
 \section{Build System}\label{build}
-The usual SBCL build process proceeds in several distinct phases, some of which need to be run on a ``host system'' and others on the ``target system''.
+The usual process to build the SBCL compiler and runtime proceeds in several distinct phases, some of which need to be run on a ``host system'' and others on the ``target system''.
 This process does allow for limited cross-compilation, wherein the ``host system'' steps can be run on an operating system and platform that is not that of the target we are trying to compile for.
@@
 The SBCL C runtime contains the GC and OS glue bits that the Lisp code needs.
+The artefacts from this stage will later be used in step (10) to attach the C runtime to the ``shrinkwrapped'' core.
 \item \code{build-config} (Linux$\sim$NX) \\
   We now create an almost-normal ARM64 Linux system with the same feature set as for the NX.
-This process involves the usual steps as before, though with a special flag to inform some parts of the Lisp process that we're going to ultimately target the NX.
-\item \code{make-host-1} (Linux$\sim$NX)
-\item \code{make-target-1} (Linux$\sim$NX)
-\item \code{make-host-2} (Linux$\sim$NX)
+This process involves the usual steps as for a ``normal'' Linux ARM64 build, though with a special flag to inform some parts of the Lisp process that we're going to ultimately target the NX.
+\item \code{make-host-1} (Linux$\sim$NX) \\
+  This step proceeds as usual.
+\item \code{make-target-1} (Linux$\sim$NX) \\
+  This step proceeds as usual.
+\item \code{make-host-2} (Linux$\sim$NX) \\
   With the target runtime built, we build the target Lisp system (compiler and the standard library) using the Lisp cross-compiler built by the Lisp host compiler in \code{make-host-1} (Linux$\sim$NX).
@@
 In particular we modify \code{*features*} to include \code{:nx} and not \code{:linux}, \code{:unix}, or \code{:posix}.
-Once that is set up and ASDF has been neutered, we can compile our program "as usual" and at the end dump out a new core.
+Once that is set up and ASDF has been sufficiently tricked into believing we are on the NX, we can compile our program "as usual" and at the end dump out a new core.
 \item \code{shrinkwrap} \\
@@
 
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{build.png}
+  \caption{A diagram of the build steps and their dependencies}
+\end{figure}
+
 A notable extra wrinkle in this procedure is that the SDK is available only for Windows.
-To accommodate this platform restriction, the custom build system we developed can be run either directly from Windows using another ARM machine remotely to run the Linux bits, or it can be run from a Linux system with the ARM bits being run either locally or remotely and the Windows bits being run either remotely or through Wine\cite{amstadt1994wine}.
+To accommodate this platform restriction, the custom build system we developed to automate these steps can be run either directly from Windows using another ARM machine remotely to run the Linux bits, or it can be run from a Linux system with the ARM bits being run either locally or remotely and the Windows bits being run either remotely or through Wine\cite{amstadt1994wine}.
 
@@
 On Windows, no equivalent signal mechanism is available, so a strategy using safepoints must be employed instead.
-A \textit{safepoint} is a location in code which is known to be GC-safe, so that collector has enough information at that location to correctly function.
+A \textit{safepoint} is a location in code which is known to be GC-safe, so the collector has enough information at that location to correctly function.
 The Lisp compiler then injects code (here termed \textit{yieldpoint} code) at these safepoints, such as at function call boundaries and loop returns, which causes the thread to check whether it should yield for GC.

### 2025-02-25 els2025-sbcl-nx/paper.tex
Mention the switch 2

@@
 
+The follow-up console to the Nintendo Switch, called ``Switch 2'' has also recently been announced.
+The device is backwards compatible with the NX and we expect that the operating environment will be very similar if not identical to that of the NX, just with more RAM and a better CPU and GPU.
+We hope to receive developer access to the Switch 2 and ensure that the port works for that, too.
+
 Finally there are a number of improvements we've made that we would like to upstream to lessen the maintenance burden.

### 2025-04-09 els2025-sbcl-nx/paper.tex
Minor
Add diagrams and add more details to shrinkwrapping

@@
 \usepackage{graphicx}
+\usepackage{svg}
 \usepackage{xcolor}
@@
 We then use a technique called ``shrinkwrapping'' to combine the Lisp code and data with the C runtime compiled for the NX to produce a final, fully static ELF executable that can be run on the NX, without the need for runtime code generation.
+We further use a restricted version of ``safepoints'' to synchronise threads during garbage collection.
 \end{abstract}
@@
 The OS-provided runtime environment is directly suitable only for C and C++ software.
-Other game engines such as Unity, Unreal, and Godot do provide exporting functionality to the NX, meaning that ports for the runtime environments they rely on such as .NET/Mono (C#), Lua, and GDScript have been developed, but remain closed-source.
+Other game engines such as Unity, Unreal, and Godot do provide exporting functionality to the NX, meaning that ports for the runtime environments they rely on such as .NET/Mono (C\#), Lua, and GDScript have been developed, but remain closed-source.
 
@@
 We run this on some host system (which may not be our ARM64 Linux intermediary), using a special flag to indicate that we're building for the NX.
-We also enable the ``fasteval'' contribution, which we need to step in for any place where we would usually invoke the compiler at runtime.
+We also enable the \textit{fasteval} contribution, which we need to step in for any place where we would usually invoke the compiler at runtime.
 \item \code{make-host-1} (NX) \\
@@
 The SBCL C runtime contains the GC and OS glue bits that the Lisp code needs.
-The artefacts from this stage will later be used in step (10) to attach the C runtime to the ``shrinkwrapped'' core.
+The artefacts from this stage will later be used in step (10) to attach the C runtime to the shrinkwrapped core.
 \item \code{build-config} (Linux$\sim$NX) \\
@@
 \item \code{shrinkwrap} \\
-  Once all our code has been loaded and a final core has been produced, we ``shrinkwrap'' the image to produce assembly files.
+  Once all our code has been loaded and a final core has been produced, we \textit{shrinkwrap} the image to produce assembly files.
 The details of this technique are outlined in \autoref{relocation}.
@@
 \section{Relocation}\label{relocation}
-Lisp objects contain absolute pointers, so that the dumped image, which contains the memory state of the heap and other spaces, is filled with absolute pointers.
+Lisp objects contain absolute pointers, so that when the in-process state of all Lisp memory spaces are dumped to disk, the resulting \textit{core image} is filled with absolute pointers.
 Traditionally, image-based Lisp implementations such as SBCL map their address spaces to fixed addresses for two reasons: First, reloading an image into a process is easier this way, since objects can be mapped back in the new image directly to where they were in the process that dumped the image.
@@
 Hence, the runtime must relocate all objects present in the dumped image to the address spaces placed randomly by the operating system.
-However, code objects present a special problem on the NX.
-Code objects in SBCL contain both executable machine code and absolute pointers to Lisp objects residing in the heap.
-However, executable pages are not writable and must be mapped in by the operating system from the ELF core.
+As part of our work, we have extended the existing support for heap relocation to allow all other Lisp spaces used by SBCL to be relocated as well, with the result that the runtime can fix up all absolute pointers mapped in from the image to point to the random addresses given to the runtime by the operating system.
+In addition, we modified the Lisp code generation to produce strictly position-independent code.
+
+However, code objects still present a special problem on the NX for this relocation process.
+A code object in SBCL is represented like other Lisp objects in-memory, in that it has a header word describing the type and size of code object.
+The code object also contains slots pointing to the debug information associated with the code object as well as other boxed objects which represent constants referenced by the code instructions.
+This is so that the garbage collector can scan the code object and fix up these boxed pointers as it does with other Lisp objects.
+Finally, these slots are followed by the raw machine instructions and function entry points representing the rest compiled Lisp code.
+The code object associated with the function defined in listing~\ref{lst:function} is illustrated in figure~\ref{fig:before-shrinkwrap}.
+
+\begin{listing}[h]
+\begin{minted}[fontsize=\small]{lisp}
+(defun f ()
+  (cons '(42 . 1958) 'els2025))
+
+\end{minted}
+\caption{A Lisp function referencing the constants \code{'(42 . 1958)} and \code{'ELS2025}}
+\label{lst:function}
+\end{listing}
+
+Notably, the code object contains both executable machine code and absolute pointers to Lisp objects.
+Because the machine instructions are to be executed by the CPU, the pages the code object is allocated on must be marked executable by the runtime.
+
+\begin{figure}[h]
+  \label{fig:before-shrinkwrap}
+  \centering
+  \includesvg[width=\linewidth]{Code object before shrinkwrapping.drawio.svg}
+  \caption{Representation of the compiled code object for the function
+    \protect\texttt{\#'f}.}
+\end{figure}
+
+However, executable pages are not writable on the NX, so that the pointers to boxed Lisp data for the debug information and code constants cannot be fixed up as part of the relocation process on start-up.
+Furthermore, only the system loader is able to allocate executable pages by loading code from on-disk ELF .text sections, so that it isn't possible to first fix up boxed object pointers in the code object \emph{before} marking the code pages executable.
+In either case, the garbage collector would not be able to fix up these pointers either.
 Therefore, the code and data need to be separated, at first so that the runtime can relocate the absolute pointers to the Lisp objects, and later so that the moving garbage collector is able to fix up those references.
-This restriction requires the shrinkwrapping procedure to separate the data from the code and to rewrite the code in the image offline so that any references to the pointers are into a dislocated read/write space instead of into the code object header, which is on an executable page and hence not writable on the NX.
-The code objects are put into code sections in the executable so that the operating system can load the code and then the runtime can later fix up references to these code objects in non-code objects residing in read/write space.
+This restriction requires the shrinkwrapping procedure to separate the data from the code and to rewrite the code in the image offline so that any references in code to code constants\footnote{As for the per-code-object slot containing debug information, we modified the Lisp system so that debug information is not accessed through the code object at all but rather through a weak hash table keyed by code object.} are into a dislocated read/write space instead of into the code object itself, which is on an executable page and hence not writable on the NX.
+To demonstrate this process in more detail, our modified offline shrinkwrapping process produces two artifacts from a normal Lisp image:
+
+\begin{itemize}
+\item \code{pie-shrinkwrap-sbcl.s}: A textual assembly file is generated with all code objects in the original core image listed under \code{.text} sections.
+The pointers to debug information and code constants inside the code object are dumped out as zeroes, and instead, the code constants referenced via the ARMv8 \code{LDR} instruction are displaced into a r/w \code{.data} section shared by all code objects.
+The \code{LDR} instructions are rewritten to access the constants from this section instead.
+\item \code{pie-shrinkwrap-sbcl.o}: A binary object file with the rest of the Lisp data from the core image section is generated.
+The majority of the Lisp image is thus `shrinkwrapped' in large r/w \code{.data} section.
+\end{itemize}
+
+The reason for producing both a textual and binary artifact is that it is easier to rewrite and produce textual assembly for code, and it is faster to process and link a binary artifact for the rest of the data.
+Figure~\ref{fig:after-shrinkwrap} illustrates how the code object for the function \code{#'f} gets split up.
+The shrinkwrapping process then links both artifacts with the rest of the SBCL runtime to produce the final executable.
+The upshot of this process is that both the relocation process and the garbage collector can now scan and fix up all boxed pointers in the image at runtime without running into the read-only constraint for executable pages.
+
+\begin{figure}[h]
+  \label{fig:after-shrinkwrap}
+  \centering
+  \includesvg[width=\linewidth]{Code objects in memory after shrinkwrapping.drawio.svg}
+  \caption{Representation of the same code object after code/data
+    segregation.}
+\end{figure}
 
@@
 To ensure that such function pointers in a saved image are valid even after the Lisp spaces containing the callback entry points are relocated, the function pointers must somehow be fixed up so they are valid after startup.
-We managed to solve this problem by having the offline shrinkwrapping process move the callback entry code into a text ELF section while recording fixup information, so that the operating system can load the callback entry code into executable pages, whereby the Lisp runtime at startup can rewrite the static function pointers to point to where the code was loaded using the recorded fixup information.
+We managed to solve this problem by having the offline shrinkwrapping process move the callback entry code into another \code{.text} ELF section while recording fixup information, so that the operating system can load the callback entry code into executable pages, whereby the Lisp runtime at startup can rewrite the static function pointers to point to where the code was loaded using the recorded fixup information.
 
 \section{Garbage Collection}\label{gc}
-SBCL's default garbage collector, ``gencgc'', is a stop-the-world collector, meaning it must stop or ``park'' other threads in the process.
+SBCL's default garbage collector, \textit{gencgc}, is a stop-the-world collector, meaning it must stop or \textit{park} other threads in the process.
 Doing so is necessary so that the threads don't access any memory that the collector might move or change during collection.
@@
 \section{Further Work}\label{further-work}
-Currently, we rely on the ``fasteval'' system to circumvent runtime compilation restrictions.
+Currently, we rely on the \textit{fasteval} system to circumvent runtime compilation restrictions.
 This is especially vital for CLOS, since the discriminating function is compiled only on first call of a generic function.
 However, since we know that we don't dynamically add or remove methods, we should be able to pre-compile all of these functions as well.
-We'd like to add such a CLOS freeze step to the build pipeline, perhaps using a technique similar to ``satiation'' as outlined by Strandh et al.\cite{strandh2014resolving}
+We'd like to add such a CLOS freeze step to the build pipeline, perhaps using a technique similar to \textit{satiation} as outlined by Strandh et al.\cite{strandh2014resolving}
 

### 2025-04-09 els2025-sbcl-nx/paper.tex
Punctuation.
More detailed captions.

@@
 \end{minted}
-\caption{A Lisp function referencing the constants \code{'(42 . 1958)} and \code{'ELS2025}}
+\caption{A Lisp function referencing the constants \code{'(42 . 1958)} and \code{'ELS2025}.}
 \label{lst:function}
@@
   \includesvg[width=\linewidth]{Code object before shrinkwrapping.drawio.svg}
-  \caption{Representation of the compiled code object for the function
-    \protect\texttt{\#'f}.}
+  \caption{Representation of the compiled code object (in boldface)
+    for the function \protect\texttt{\#'f} in memory, along with the
+    cons and symbol objects it references.}
 \end{figure}
@@
   \includesvg[width=\linewidth]{Code objects in memory after shrinkwrapping.drawio.svg}
-  \caption{Representation of the same code object after code/data
-    segregation.}
+  \caption{Representation of the same code object and the cons and
+    symbol objects it references after code/data segregation and the
+    rest of the shrinkwrapping process.}
 \end{figure}

### 2025-04-10 els2025-sbcl-nx/paper.tex
Minor fixups

@@
 \section{Relocation}\label{relocation}
-Lisp objects contain absolute pointers, so that when the in-process state of all Lisp memory spaces are dumped to disk, the resulting \textit{core image} is filled with absolute pointers.
+Lisp objects contain absolute pointers, meaning when the in-process state of all Lisp memory spaces is dumped to disk, the resulting \textit{core image} is filled with absolute pointers.
 Traditionally, image-based Lisp implementations such as SBCL map their address spaces to fixed addresses for two reasons: First, reloading an image into a process is easier this way, since objects can be mapped back in the new image directly to where they were in the process that dumped the image.
@@
 Hence, the runtime must relocate all objects present in the dumped image to the address spaces placed randomly by the operating system.
-As part of our work, we have extended the existing support for heap relocation to allow all other Lisp spaces used by SBCL to be relocated as well, with the result that the runtime can fix up all absolute pointers mapped in from the image to point to the random addresses given to the runtime by the operating system.
+As part of our work, we have extended the existing support for heap relocation to allow all other Lisp spaces used by SBCL to be relocated as well, with the result being that the runtime can fix up all absolute pointers mapped in from the image to point to the random addresses given to the runtime by the operating system.
 In addition, we modified the Lisp code generation to produce strictly position-independent code.
@@
 This is so that the garbage collector can scan the code object and fix up these boxed pointers as it does with other Lisp objects.
-Finally, these slots are followed by the raw machine instructions and function entry points representing the rest compiled Lisp code.
-The code object associated with the function defined in listing~\ref{lst:function} is illustrated in figure~\ref{fig:before-shrinkwrap}.
+Finally, these slots are followed by the raw machine instructions and function entry points representing the compiled Lisp code.
+The code object associated with the function defined in \autoref{lst:function} is illustrated in \autoref{fig:before-shrinkwrap}.
 
@@
 \begin{figure}[h]
-  \label{fig:before-shrinkwrap}
   \centering
@@
     cons and symbol objects it references.}
+  \label{fig:before-shrinkwrap}
 \end{figure}
 
-However, executable pages are not writable on the NX, so that the pointers to boxed Lisp data for the debug information and code constants cannot be fixed up as part of the relocation process on start-up.
-Furthermore, only the system loader is able to allocate executable pages by loading code from on-disk ELF .text sections, so that it isn't possible to first fix up boxed object pointers in the code object \emph{before} marking the code pages executable.
+However, executable pages are not writable on the NX, meaning the pointers to boxed Lisp data for the debug information and code constants cannot be fixed up as part of the relocation process on start-up.
+Furthermore, only the system loader is able to allocate executable pages by loading code from on-disk ELF .text sections, meaning it isn't possible to first fix up boxed object pointers in the code object \emph{before} marking the code pages executable.
 In either case, the garbage collector would not be able to fix up these pointers either.
@@
 This restriction requires the shrinkwrapping procedure to separate the data from the code and to rewrite the code in the image offline so that any references in code to code constants\footnote{As for the per-code-object slot containing debug information, we modified the Lisp system so that debug information is not accessed through the code object at all but rather through a weak hash table keyed by code object.} are into a dislocated read/write space instead of into the code object itself, which is on an executable page and hence not writable on the NX.
-To demonstrate this process in more detail, our modified offline shrinkwrapping process produces two artifacts from a normal Lisp image:
-
-\begin{itemize}
-\item \code{pie-shrinkwrap-sbcl.s}: A textual assembly file is generated with all code objects in the original core image listed under \code{.text} sections.
-The pointers to debug information and code constants inside the code object are dumped out as zeroes, and instead, the code constants referenced via the ARMv8 \code{LDR} instruction are displaced into a r/w \code{.data} section shared by all code objects.
-The \code{LDR} instructions are rewritten to access the constants from this section instead.
-\item \code{pie-shrinkwrap-sbcl.o}: A binary object file with the rest of the Lisp data from the core image section is generated.
-The majority of the Lisp image is thus `shrinkwrapped' in large r/w \code{.data} section.
-\end{itemize}
-
-The reason for producing both a textual and binary artifact is that it is easier to rewrite and produce textual assembly for code, and it is faster to process and link a binary artifact for the rest of the data.
-Figure~\ref{fig:after-shrinkwrap} illustrates how the code object for the function \code{#'f} gets split up.
-The shrinkwrapping process then links both artifacts with the rest of the SBCL runtime to produce the final executable.
-The upshot of this process is that both the relocation process and the garbage collector can now scan and fix up all boxed pointers in the image at runtime without running into the read-only constraint for executable pages.
+To demonstrate this process in more detail, our modified offline shrinkwrapping process produces two artefacts from a normal Lisp image:
 
 \begin{figure}[h]
-  \label{fig:after-shrinkwrap}
   \centering
@@
     rest of the shrinkwrapping process.}
+  \label{fig:after-shrinkwrap}
 \end{figure}
+
+\begin{itemize}
+\item \code{pie-shrinkwrap-sbcl.s}: A textual assembly file is generated with all code objects in the original core image listed under \code{.text} sections.
+The pointers to debug information and code constants inside the code object are zeroed out, and the code constants referenced via the ARMv8 \code{LDR} instruction are displaced into a r/w \code{.data} section shared by all code objects.
+The \code{LDR} instructions are rewritten to access the constants from this section instead.
+\item \code{pie-shrinkwrap-sbcl.o}: A binary object file with the rest of the Lisp data from the core image section is generated.
+The majority of the Lisp image is thus `shrinkwrapped' into a large r/w \code{.data} section.
+\end{itemize}
+
+The reason for producing both a textual and a binary artefact is that it is easier to rewrite and produce textual assembly for code, and it is faster to process and link a binary artefact for the rest of the data.
+\autoref{fig:after-shrinkwrap} illustrates how the code object for the function \code{#'f} gets split up.
+The shrinkwrapping process then links both artefacts with the rest of the SBCL runtime to produce the final executable.
+The upshot of this process is that both the relocation process and the garbage collector can now scan and fix up all boxed pointers in the image at runtime without running into the read-only constraint for executable pages.
 
@@
 SBCL uses a linkage table that indirects calls to foreign code from Lisp so that foreign code can be loaded by the operating system at any address on startup and resolved in the linkage table.
-However, for callbacks into Lisp from foreign code, Lisp must create function pointers to snippets of assembly for callback entry allocated in static (i.e.
+However, for callbacks into Lisp from foreign code, Lisp must create function pointers to snippets of assembly allocated in static (i.e.
 never moved by GC) Lisp space.
 These pointers can then be passed to and called from foreign code.
-To ensure that such function pointers in a saved image are valid even after the Lisp spaces containing the callback entry points are relocated, the function pointers must somehow be fixed up so they are valid after startup.
-We managed to solve this problem by having the offline shrinkwrapping process move the callback entry code into another \code{.text} ELF section while recording fixup information, so that the operating system can load the callback entry code into executable pages, whereby the Lisp runtime at startup can rewrite the static function pointers to point to where the code was loaded using the recorded fixup information.
+To ensure that such function pointers in a saved image are valid even after the Lisp spaces containing the callback entry points are relocated, the function pointers must somehow be fixed up.
+We managed to solve this problem by having the offline shrinkwrapping process move the callback entry code into another \code{.text} ELF section while recording fixup information.
+The operating system can then load the callback entry code into executable pages, and the Lisp runtime can rewrite the static function pointers to point to where the code was loaded using the recorded fixup information.
 

### 2025-04-10 els2025-sbcl-nx/paper.tex
Fix another sentence.

@@
 \section{Relocation}\label{relocation}
-Lisp objects contain absolute pointers, meaning when the in-process state of all Lisp memory spaces is dumped to disk, the resulting \textit{core image} is filled with absolute pointers.
+Lisp objects in memory contain absolute pointers, with the result that dumping the in-process state of all Lisp memory spaces to disk results in a \textit{core image} filled with absolute pointers.
 Traditionally, image-based Lisp implementations such as SBCL map their address spaces to fixed addresses for two reasons: First, reloading an image into a process is easier this way, since objects can be mapped back in the new image directly to where they were in the process that dumped the image.

### 2025-04-25 els2025-sbcl-nx/presentation.tex
Start working on the presentation

\documentclass[14pt,t,aspectratio=169]{beamer}
\usepackage{fontspec}
\usepackage{color}
\usepackage{minted}
\usepackage{array}
\usepackage{emoji}

%% These fonts are non-free.
%% Comment out the lines if you don't have them.
\setmainfont{Equity Text A}
\setsansfont{Concourse T3}
\setmonofont{Triplicate T4}

\definecolor{bgcolor}{RGB}{255,240,240}
\definecolor{red}{RGB}{224,35,35}
\setbeamercolor{background canvas}{bg=bgcolor}
\setbeamercolor{normal text}{fg=black}
\setbeamercolor{titlelike}{fg=black}
\setbeamercolor{itemize item}{fg=red}
\setbeamercolor{enumerate item}{fg=red}
\setbeamertemplate{itemize items}[circle]
\setbeamertemplate{navigation symbols}{}
\usemintedstyle{monokai}
\newminted[lispcode]{common-lisp}{fontsize=\footnotesize}
\newminted[smalllispcode]{common-lisp}{fontsize=\scriptsize}
\def\code#1{{\color{codecolor}\texttt{#1}}}
\renewcommand{\theFancyVerbLine}{\color{darkgray}\large \oldstylenums{\arabic{FancyVerbLine}}}

\usebackgroundtemplate{\includegraphics[width=\paperwidth,height=\paperheight]{background}}
\begin{document}
{\usebackgroundtemplate{\includegraphics[width=\paperwidth,height=\paperheight]{firstpage}}
  \begin{frame}
    \color{white}
    \vspace{3cm}
    {\hspace{1.4cm} \LARGE Porting SBCL to the Nintendo Switch} \\
    \vspace{1cm}
    {\hspace{1.6cm} Yukari Hafner, Charles Zhang} \\
    \vspace{0.2cm}
    {\hspace{2.1cm}\texttt https://shirakumo.org}
  \end{frame}}

\begin{frame}
  \frametitle{The Device}
  \includegraphics[height=5cm]{switch}
  \begin{itemize}
  \item CPU: ARM 4 Cortex-A57 64-bit
  \item OS: ``Horizon OS'', proprietary micro-kernel
  \item SDK: C++, proprietary version of Clang
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Immediate Challenges}
  \begin{itemize}
  \item Everything is proprietary and under NDA \\
    \Rightarrow Scarce public information
  \item The OS is not BSD or even fully POSIX \\
    \Rightarrow Need new OS abstractions
  \item There are no inter-thread signals \\
    \Rightarrow Can't use usual GC tricks
  \item We are not allowed to create executable pages \\
    \Rightarrow No compilation at runtime
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Basic Ideas}
  \begin{itemize}
    \color{lightgray}
  \item Everything is proprietary and under NDA \\
    \textcolor{red}{\Rightarrow} \textcolor{black}{Only publicise our own interfaces}
  \item The OS is not BSD or even fully POSIX \\
    \textcolor{red}{\Rightarrow} \textcolor{black}{Write C(++) shim libraries for access}
  \item There are no inter-thread signals \\
    \textcolor{red}{\Rightarrow} \textcolor{black}{Use safepoints}
  \item We are not allowed to create executable pages \\
    \textcolor{red}{\Rightarrow} \textcolor{black}{Compile everything on linux and shrinkwrap}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Build Procedure}
  \begin{itemize}
  \item 
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Relocation}
  \begin{itemize}
  \item 
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Garbage Collection}
  \begin{itemize}
  \item 
  \end{itemize}
\end{frame}

\begin{frame}[c]{ }
  \centering
  {\Huge Live Demo}
\end{frame}

\begin{frame}
  \frametitle{Further Work}
  \begin{itemize}
  \item Optimising CLOS dispatch ahead of time \\
    \Rightarrow Christophe? \emoji{pleading-face}
  \item Optimising Trial and Kandria \\
    \Rightarrow Lots of profiling work that can be done on PC
  \item Porting to the Nintendo Switch 2 \\
    \Rightarrow As soon as plebians like us get access from almighty \\\quad Nintendo
  \end{itemize}
\end{frame}

{\usebackgroundtemplate{\includegraphics[width=\paperwidth,height=\paperheight]{firstpage}}
  \begin{frame}[c]{ }
    \centering\color{white}
    \vspace{2cm}
    {\LARGE Thank you!} \\
    \vspace{0.2cm}
    Consider supporting our work: \\
    \includegraphics[width=3cm]{patreon} \\
    {\texttt https://patreon.com/shinmera}
  \end{frame}}
\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% TeX-engine: luatex
%%% TeX-command-extra-options: "-shell-escape"
%%% End:

### 2025-04-27 els2025-sbcl-nx/presentation.tex
Finish presentation bits for myself

@@
   \item Everything is proprietary and under NDA \\
-    \Rightarrow Scarce public information
+    \Rightarrow{} Scarce public information
   \item The OS is not BSD or even fully POSIX \\
-    \Rightarrow Need new OS abstractions
+    \Rightarrow{} Need new OS abstractions
   \item There are no inter-thread signals \\
-    \Rightarrow Can't use usual GC tricks
+    \Rightarrow{} Can't use usual GC tricks
   \item We are not allowed to create executable pages \\
-    \Rightarrow No compilation at runtime
+    \Rightarrow{} No compilation at runtime
   \end{itemize}
@@
 \begin{frame}
-  \frametitle{Build Procedure}
+  \frametitle{A Standard SBCL Build}
   \begin{itemize}
-  \item 
+  \item build-config \\
+    \Rightarrow{} Gather system info
+  \item make-host-1 \\
+    \Rightarrow{} Emit C headers and support files
+  \item make-target-1 \\
+    \Rightarrow{} Compile the C runtime on the target
+  \item make-host-2 \\
+    \Rightarrow{} Cross-compile the compiler on the host
+  \item make-target-2 \\
+    \Rightarrow{} Use the compiler from the host to\\ incrementally compile the rest on the target
   \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \includegraphics[height=8.5cm]{build.png}
 \end{frame}
@@
   \item Optimising CLOS dispatch ahead of time \\
-    \Rightarrow Christophe? \emoji{pleading-face}
+    \Rightarrow{} Christophe? \emoji{pleading-face}
   \item Optimising Trial and Kandria \\
-    \Rightarrow Lots of profiling work that can be done on PC
+    \Rightarrow{} Lots of profiling work that can be done on PC
   \item Porting to the Nintendo Switch 2 \\
-    \Rightarrow As soon as plebians like us get access from almighty \\\quad Nintendo
+    \Rightarrow{} As soon as plebians like us get access from almighty \\\quad Nintendo
   \end{itemize}

### 2025-05-10 els2025-sbcl-nx/presentation.tex
Write my presentation sections
But probably horribly formatted and the code listing doesn't work
yet. Also the prose sucks.

@@
 \usepackage{emoji}
+\usepackage{svg}
 
@@
   \item We are not allowed to create executable pages \\
-    \textcolor{red}{\Rightarrow} \textcolor{black}{Compile everything on linux and shrinkwrap}
+    \textcolor{red}{\Rightarrow} \textcolor{black}{Compile everything on linux
+      and shrinkwrap}
+  \item The OS enforces strict Address Space Layout Randomization (ASLR) \\
+    \textcolor{red}{\Rightarrow} \textcolor{black}{Make all code and data fully
+      relocatable and position independent}
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{History of the Project}
+  \begin{itemize}
+  \item Initial discussion on \#sbcl and at ELS 2022 about porting SBCL
+    to Nintendo Switch \\
+  \item Decided to seriously pursue project at ELS 2023 \\
+    \Rightarrow{} Sketched out project plan in the airport afterwards
+  \item Most of the initial work and design problems solved in the
+    following months \\
+  \item Debugging GC issues and improving debugging experience in
+    2024 \\
+  \item Game engine examples stabilized, Kandria running in early 2025 \\
   \end{itemize}
@@
   \begin{itemize}
-  \item 
+  \item Lisp objects are full of absolute pointers to other objects
+  \item Lisp objects live in Lisp spaces mapped at fixed addresses
+  \item Saving an image dumps Lisp spaces to disk verbatim
+  \item Reloading an image is fast; just \texttt{mmap} the on-disk
+    spaces into the process.
+  \item \Rightarrow{} Problem for NX: enforced ASLR means no means to map spaces
+    to desired addresses
   \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Relocation (cont.)}
+  \begin{itemize}
+    \item Solution: Relocate all Lisp spaces on start up
+    \item \Rightarrow{} Problem for NX: Code generation hardwires some primitive
+      object addresses.
+    \item Solution: Change codegen to be position independent.
+    \item Code objects in SBCL also contain Lisp pointers (constants)
+    \item \Rightarrow{} Problem for NX: Relocation and moving GC needs to fix up
+      these constants, but code objects are not writable
+    \item Solution: Segregate code instructions and data in code
+      objects into different ELF sections offline, rewrite code
+      instructions to reference r/w section
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Relocation example}
+  XXX why doesn't this work?
+%   \begin{listing}
+%     \begin{minted}[fontsize=\small]{lisp}
+% (defun f ()
+%   (cons '(42 . 1958) 'els2025))
+
+%      \end{minted}
+%   \caption{A Lisp function referencing the constants \code{'(42 . 1958)} and \code{'ELS2025}.}
+%   \label{lst:function}
+%   \end{listing}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Relocation example (cont.)}
+  \begin{figure}[h]
+    \centering
+    \includesvg[width=\linewidth]{Code object before shrinkwrapping.drawio.svg}
+    \caption{Representation of the compiled code object (in boldface)
+      for the function \protect\texttt{\#'f} in memory, along with the
+      cons and symbol objects it references.}
+    \label{fig:before-shrinkwrap}
+  \end{figure}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Relocation example (cont.)}
+  \begin{figure}[h]
+    \centering
+    \includesvg[width=\linewidth]{Code objects in memory after shrinkwrapping.drawio.svg}
+    \caption{Representation of the same code object and the cons and
+      symbol objects it references after code/data segregation and the
+      rest of the shrinkwrapping process.}
+    \label{fig:after-shrinkwrap}
+  \end{figure}
 \end{frame}
@@
   \begin{itemize}
-  \item 
+  \item Garbage collection on SBCL on non-Windows platforms uses POSIX
+    signals to stop-the-world.
+  \item \Rightarrow{} Problem for NX: No POSIX signals.
+  \item Solution: Use safepoints (implemented already on Windows)
+  \item BUT: The safepoint mechanism still uses virtual memory
+    hardware exceptions to stop the current thread
+  \item \Rightarrow{} Problem for NX: No ability to handle operating system
+    exceptions
+  \item Solution: Change safepoint codegen to poll state flags
+    explicitly
+  \end{itemize}
+\end{frame}
+
+\begin{frame}
+  \frametitle{Misc Issues}
+  \begin{itemize}
+  \item SBCL uses hardware exceptions to signal some common conditions
+  \item \Rightarrow{} Solution: Change to full calls to \texttt{ERROR} and \texttt{SIGNAL}
+  \item CLOS requires runtime compilation for dispatch JIT
+  \item \Rightarrow{} Solution: Use an interpreter where absolutely necessary
+  \item The existing shrinkwrap tool is not GC-safe for precise GC
+    platforms like ARM64, random crashes with large cores
+  \item  \Rightarrow{} Solution: Complicate the build process even more by using
+    a separate SBCL with memory spaces mapped away from the offline
+    core's spaces to shrinkwrap
   \end{itemize}

### 2025-05-19 els2025-sbcl-nx/presentation.tex
Fixup presentation code sample
Make figures fit.

@@
 \setbeamertemplate{navigation symbols}{}
-\usemintedstyle{monokai}
 \newminted[lispcode]{common-lisp}{fontsize=\footnotesize}
@@
 
-\begin{frame}
+\begin{frame}[fragile]
   \frametitle{Relocation example}
-  XXX why doesn't this work?
-%   \begin{listing}
-%     \begin{minted}[fontsize=\small]{lisp}
-% (defun f ()
-%   (cons '(42 . 1958) 'els2025))
-
-%      \end{minted}
-%   \caption{A Lisp function referencing the constants \code{'(42 . 1958)} and \code{'ELS2025}.}
-%   \label{lst:function}
-%   \end{listing}
+\begin{minted}[fontsize=\small]{common-lisp}
+ (defun f ()
+   (cons '(42 . 1958) 'els2025))
+\end{minted}
+  A Lisp function referencing the constants \code{'(42 . 1958)} and \code{'ELS2025}
 \end{frame}
@@
 \begin{frame}
-  \frametitle{Relocation example (cont.)}
-  \begin{figure}[h]
-    \centering
-    \includesvg[width=\linewidth]{Code object before shrinkwrapping.drawio.svg}
-    \caption{Representation of the compiled code object (in boldface)
+  \begin{figure}
+    \begin{columns}
+      \column{0.6\linewidth}
+      \vspace{-2em}
+      \phantom{123}
+      \includesvg[scale=0.5]{Code object before
+        shrinkwrapping.drawio.svg}
+      \column{0.4\linewidth}
+      Representation of the compiled code object (in boldface)
       for the function \protect\texttt{\#'f} in memory, along with the
-      cons and symbol objects it references.}
-    \label{fig:before-shrinkwrap}
+      cons and symbol objects it references.
+    \end{columns}
   \end{figure}
@@
 \begin{frame}
-  \frametitle{Relocation example (cont.)}
-  \begin{figure}[h]
-    \centering
-    \includesvg[width=\linewidth]{Code objects in memory after shrinkwrapping.drawio.svg}
-    \caption{Representation of the same code object and the cons and
+  \begin{figure}
+    \begin{columns}
+      \column{0.7\linewidth}
+      \vspace{-1.7em}
+      \phantom{123}
+      \includesvg[scale=0.5]{Code objects in memory after
+        shrinkwrapping.drawio.svg}
+      \column{0.4\linewidth}
+      Representation of the same code object and the cons and
       symbol objects it references after code/data segregation and the
-      rest of the shrinkwrapping process.}
-    \label{fig:after-shrinkwrap}
+      rest of the shrinkwrapping process.
+    \end{columns}
   \end{figure}

### 2025-05-19 els2025-sbcl-nx/presentation.tex
Add some more issues we had to face.

@@
   \begin{itemize}
-    \item Solution: Relocate all Lisp spaces on start up
+    \item Solution: Relocate all Lisp spaces on start up (support
+    existed for most Lisp spaces already, except for one)
     \item \Rightarrow{} Problem for NX: Code generation hardwires some primitive
@@
   \item Solution: Change safepoint codegen to poll state flags
-    explicitly
+    explicitly (challenge: do this in a race-free way!)
   \end{itemize}

### 2025-05-20 els2025-sbcl-nx/presentation.tex
Minor fix
Update presentation

--- 2025-05-19 els2025-sbcl-nx/presentation.tex
@@
 \usepackage{svg}
+\usepackage[absolute,overlay]{textpos}
 
@@
 
+\definecolor{codecolor}{RGB}{50,50,50}
 \definecolor{bgcolor}{RGB}{255,240,240}
@@
     \Rightarrow{} No compilation at runtime
+  \item The OS enforces strict ASLR \\
+    \Rightarrow{} Can't hard-wire object addresses
   \end{itemize}
@@
       and shrinkwrap}
-  \item The OS enforces strict Address Space Layout Randomization (ASLR) \\
+  \item The OS enforces strict ASLR \\
     \textcolor{red}{\Rightarrow} \textcolor{black}{Make all code and data fully
-      relocatable and position independent}
+      relocatable}
   \end{itemize}
@@
   \begin{itemize}
-  \item Initial discussion on \#sbcl and at ELS 2022 about porting SBCL
-    to Nintendo Switch \\
+  \item Initial discussion and runtime port at ELS 2022 \\
   \item Decided to seriously pursue project at ELS 2023 \\
@@
     following months \\
-  \item Debugging GC issues and improving debugging experience in
-    2024 \\
-  \item Game engine examples stabilized, Kandria running in early 2025 \\
+  \item Debugging GC issues and improving debugging \\
+    experience in 2024 \\
+  \item Game engine examples stabilized, Kandria running \\
+    in early 2025 \\
   \end{itemize}
@@
   \item make-target-2 \\
-    \Rightarrow{} Use the compiler from the host to\\ incrementally compile the rest on the target
+    \Rightarrow{} Use the compiler from the host to\\
+    \quad incrementally compile the rest on the target
   \end{itemize}
@@
   \item Saving an image dumps Lisp spaces to disk verbatim
-  \item Reloading an image is fast; just \texttt{mmap} the on-disk
-    spaces into the process.
-  \item \Rightarrow{} Problem for NX: enforced ASLR means no means to map spaces
-    to desired addresses
+  \item Reloading an image is fast; just \texttt{mmap} the on-disk \\
+    spaces into the process. \\
+    \textcolor{red}{\Rightarrow} Problem: enforced ASLR means no means \\
+    \quad to map spaces to desired addresses
   \end{itemize}
@@
   \begin{itemize}
-    \item Solution: Relocate all Lisp spaces on start up
-    \item \Rightarrow{} Problem for NX: Code generation hardwires some primitive
-      object addresses.
-    \item Solution: Change codegen to be position independent.
-    \item Code objects in SBCL also contain Lisp pointers (constants)
-    \item \Rightarrow{} Problem for NX: Relocation and moving GC needs to fix up
-      these constants, but code objects are not writable
-    \item Solution: Segregate code instructions and data in code
-      objects into different ELF sections offline, rewrite code
-      instructions to reference r/w section
+    \item Solution: Relocate all Lisp spaces on start up \\
+      \textcolor{red}{\Rightarrow} Problem: Code generation hardwires some primitive \\
+      \quad object addresses \\
+      \Rightarrow{} Solution: Change codegen to be position independent
+    \item Code objects in SBCL also contain Lisp pointers \\
+      \textcolor{red}{\Rightarrow} Problem: Relocation and GC needs to fix up \\
+      \quad these constants, but code objects are not writable \\
+      \Rightarrow{} Solution: Segregate code instructions and \\
+      \quad data in code objects into different ELF sections, \\
+      \quad rewrite code instructions to reference r/w section
   \end{itemize}
@@
         shrinkwrapping.drawio.svg}
-      \column{0.4\linewidth}
+      \column{0.3\linewidth}
       Representation of the compiled code object (in boldface)
@@
       cons and symbol objects it references.
+      \column{0.1\linewidth}
     \end{columns}
@@
 \begin{frame}
-  \begin{figure}
-    \begin{columns}
-      \column{0.7\linewidth}
-      \vspace{-1.7em}
-      \phantom{123}
     \includesvg[scale=0.5]{Code objects in memory after
       shrinkwrapping.drawio.svg}
-      \column{0.4\linewidth}
+
+    \setlength{\TPHorizModule}{\textwidth}
+    \setlength{\TPVertModule}{\textwidth}
+    \begin{textblock}{0.5}(0.41,0.25)
       Representation of the same code object and the cons and
@@
       rest of the shrinkwrapping process.
-    \end{columns}
-  \end{figure}
+    \end{textblock}
 \end{frame}
@@
   \item Garbage collection on SBCL on non-Windows platforms uses POSIX
-    signals to stop-the-world.
-  \item \Rightarrow{} Problem for NX: No POSIX signals.
-  \item Solution: Use safepoints (implemented already on Windows)
-  \item BUT: The safepoint mechanism still uses virtual memory
-    hardware exceptions to stop the current thread
-  \item \Rightarrow{} Problem for NX: No ability to handle operating system
-    exceptions
-  \item Solution: Change safepoint codegen to poll state flags
-    explicitly
+    signals to stop-the-world. \\
+    \textcolor{red}{\Rightarrow} Problem: No POSIX signals. \\
+    \Rightarrow{} Solution: Use safepoints
+  \item BUT: The safepoint mechanism still uses virtual memory \\
+    hardware exceptions to stop the current thread \\
+  \textcolor{red}{\Rightarrow} Problem: No ability to handle OS exceptions \\
+  \Rightarrow{} Solution: Change safepoints to poll state flags
   \end{itemize}
@@
   \begin{itemize}
-  \item SBCL uses hardware exceptions to signal some common conditions
-  \item \Rightarrow{} Solution: Change to full calls to \texttt{ERROR} and \texttt{SIGNAL}
-  \item CLOS requires runtime compilation for dispatch JIT
-  \item \Rightarrow{} Solution: Use an interpreter where absolutely necessary
-  \item The existing shrinkwrap tool is not GC-safe for precise GC
-    platforms like ARM64, random crashes with large cores
-  \item  \Rightarrow{} Solution: Complicate the build process even more by using
-    a separate SBCL with memory spaces mapped away from the offline
-    core's spaces to shrinkwrap
+  \item SBCL uses hardware exceptions to signal some conditions \\
+    \Rightarrow{} Solution: Change to full calls to \texttt{ERROR} and \texttt{SIGNAL}
+  \item CLOS requires runtime compilation for dispatch JIT \\
+    \Rightarrow{} Solution: Use an interpreter where absolutely necessary
+  \item The existing shrinkwrap tool is not GC-safe for precise \\
+    platforms like ARM64, random crashes with large cores \\
+    \Rightarrow{} Solution: Complicate the build process even more \\
+    \quad by using a separate SBCL with memory spaces \\
+    \quad mapped away from the offline core's spaces
   \end{itemize}
@@
   \begin{itemize}
-  \item Optimising CLOS dispatch ahead of time \\
-    \Rightarrow{} Christophe? \emoji{pleading-face}
+  \item Startup time is long, build times insufferably long \\
+    \Rightarrow{} Figure out how to cache more stuff
   \item Optimising Trial and Kandria \\
